### 2023-08-07 5_manuscript/manuscript.qmd
Updated title and authors

---
title: "Assessing methods to adjust for assay differences across COVID-19 serological surveillance studies in Canada"
execute: 
  echo: false
format:
  docx: 
    reference-doc: word-style-reference-manuscript.docx
  html: default
bibliography: bib.bib
csl: american-medical-association-brackets.csl
editor: visual
crossref: 
  thm-title: "Table"
  thm-prefix: "Table"
  cor-title: "Table S"
  cor-prefix: "Table S"
  lem-title: "Figure S"
  lem-prefix: "Figure S"
  fig-title: "**Figure**"
---

```{=html}
<!---

NOTES:
-   To render both word and HTML, can type this into terminal

quarto render 5_manuscript/manuscript.qmd

-   There are limits in cross-referencing in Quarto. The table
   Caption doesn't work well with flextable package, and 
  There's no option for secondary (supplemental) series of 
  tables or figures. As workaround, we repurpose other
  series.I'm using
        @thm- for tables (Table X)
        @cor- for supplemental tables (Table SX)
        @lem- for supplemental figures (Figure SX)
  The space between "S" and the number for supplemental figures seems unavoidable
   for now. Can manually correct before submitting to journal.

Flextable package isn't playing nicely with quarto for crossreferencing at the 
moment but it's being actively worked on. To install latest package version, use 
  devtools::install_github("davidgohel/flextable")

--->
```
Jiacheng Chen^1,2^, ..., W. Alton Russell^1,2^

<br>

^1^School of Population and Global Health, McGill University, Montreal, Canada

^2^COVID-19 Immunity Task Force

<br>

**Corresponding author:**

**Key words:**

**Running title:**

{{< pagebreak >}}

```{r setup, include=FALSE}
library(ggplot2) #plots
library(data.table) #for using datatables instead of frames
library(scales) #formatting plot legends and text
library(readxl) #read.excel
library(flextable) #generating tables
# library(ftExtra)
# library(officedown) #formatting for word
# library(officer)
# library(stringr)
theme_set(theme_bw())

```

```{r}
df_samp <- read_excel("../1_data/tables.xlsx", sheet = "sample")
```

# Abstract

**Background:** A

**Methods:** A

**Results:** A

**Conclusions:** A

{{< pagebreak >}}

# Introduction

Citation [@Langham2018a]. See @fig-sample, @cor-sample, @thm-sample, and @lem-sample.

{{< pagebreak >}}

# Methods

## Subsection

A

<br>

# Results

A

<br>

# Discussion

A

<br>

{{< pagebreak >}}

# Declarations

**Funding:** A

**Conflicts:** A

**Ethics/Consent:** A

**Data and materials:** A

**Code availability:** A

**Authors' contributions: A**

{{< pagebreak >}}

# References

::: {#refs}
:::

{{< pagebreak >}}

# Tables

::: {#thm-sample}
Table caption here.

```{r}
t_samp <- as_flextable(as_grouped_data(df_samp,
                                              groups = "subhead"))
t_samp <- compose(t_samp, i = ~ !is.na(subhead), j = "col1",
              value = as_paragraph(as_chunk(subhead)))
t_samp <- fontsize(t_samp, size = 10, part = "all")
t_samp <- font(t_samp, fontname = "Times", part = "all")
t_samp <- theme_box(t_samp)
t_samp <- bg(t_samp, bg = "#EAEAEA", part = "header")
t_samp <- width(t_samp, 1, 1.7)
t_samp <- bg(t_samp, i = c(1, 4), bg = "#DDDDDD", part = "body")
t_samp <- bold(t_samp, i = c(1, 4), part = "body")
t_samp <- align(t_samp, align = "left", part = "all")

t_samp
```
:::

{{< pagebreak >}}

# Figures

```{r}
#| label: fig-sample
#| fig-cap: "Figure caption here."

knitr::include_graphics("../4_output/figs/figure.png")

```

{{< pagebreak >}}

# Supplemental materials

<br>

# A. Supplement section

{{< pagebreak >}}

# Supplemental tables

::: {#cor-sample}
Table caption here.

```{r}
t_samp <- as_flextable(as_grouped_data(df_samp,
                                              groups = "subhead"))
t_samp <- compose(t_samp, i = ~ !is.na(subhead), j = "col1",
              value = as_paragraph(as_chunk(subhead)))
t_samp <- fontsize(t_samp, size = 10, part = "all")
t_samp <- font(t_samp, fontname = "Times", part = "all")
t_samp <- theme_box(t_samp)
t_samp <- bg(t_samp, bg = "#EAEAEA", part = "header")
t_samp <- width(t_samp, 1, 1.7)
t_samp <- bg(t_samp, i = c(1, 4), bg = "#DDDDDD", part = "body")
t_samp <- bold(t_samp, i = c(1, 4), part = "body")
t_samp <- align(t_samp, align = "left", part = "all")

t_samp
```
:::

{{< pagebreak >}}

# Supplemental figures

::: {#lem-sample}
Figure caption here.

```{r}
knitr::include_graphics("../4_output/figs/figure.png")
```
:::

### 2023-08-30 2_scripts/2.0 AB Outpatient Lab Data.qmd
Upload APL data cleaning code

---
title: "2.0 AB Outpatient Lab Data Encryption (SHA-2)"
author: J. Chen
format: html
editor: visual
---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
# Load Paks
library(dplyr)
library(duawranglr)    # use this to encrypt the ids
```

```{r}
## Import Dataset
library(readxl)
RFD4682_COVID<-read_excel("~/APL/RFD4682_COVID_Serology_Data_for release June 2023.xlsx", 
    sheet = "RFD4682_Full_Data", col_types = c("text", 
        "text", "text", "text", "text", 
        "date", "text", "text", "date", "date", 
        "text", "text", "numeric", "text", 
        "text", "text", "text", "text", "date", 
        "numeric", "text", "text", 
        "date", "numeric", "numeric"))

table(RFD4682_COVID$NAAT_STATUS, useNA = 'ifany')
save(RFD4682_COVID, file = "~/APL/RFD4682.RData")

## The IDs are not encrypted, need to do it by ourselves

## Add missing value indicator for PERSON_ID and ULI_ID
RFD4682_COVID$id_per_miss<-is.na(RFD4682_COVID$PERSON_ID)
RFD4682_COVID$id_uli_miss<-is.na(RFD4682_COVID$ULI_ID)

# Use deid_dua() from {duawranglr} package to do the encryption
RFD4682_e <- deid_dua(RFD4682_COVID, 
                      id_col = 'PERSON_ID',
                      new_id_name = 'PERSON_IDe',
                      write_crosswalk = TRUE, id_length = 12,
                      crosswalk_filename='crosswalk_id_per.csv')

RFD4682_e <- deid_dua(RFD4682_e, 
                      id_col = 'ULI_ID',
                      new_id_name = 'ULI_IDe',
                      write_crosswalk = TRUE, id_length = 15)

# SHA-2 algorithm also process missing values in those two ids
# NAs in each id will have the same encrypted value, need to replace them with NA
RFD4682_e$PERSON_IDe[RFD4682_e$id_per_miss==TRUE]<-NA
RFD4682_e$ULI_IDe[RFD4682_e$id_uli_miss==TRUE]<-NA
RFD4682_e<-select(RFD4682_e, !c(id_per_miss, id_uli_miss))

## Take a look at the crosswalk file
corsswalk_uli<-read.csv(file = '~/APL/id_crosswalk_id_uli.csv')
rm(corsswalk_uli)
# Alton is the only person who keeps the crosswalks

## Save the Dataset
save(RFD4682_e, file = "~/APL/RFD4682e.RData")
```

## Check the unique individual in the original dataset

```{r}
## ULI_ID: Unique Lifetime Identifier (ULI), numerical values of Alberta PHN.
length(unique(RFD4682_e$ULI_IDe))     # Unique = 180821-1 (1 is NA)
table(is.na(RFD4682_e$ULI_IDe))       # NA = 11298

## PERSON_ID: assigned from the originating LIS
length(unique(RFD4682_e$PERSON_IDe))     # Unique = 189864-1
table(is.na(RFD4682_e$PERSON_IDe))       # NA = 2973
```

## Matt's Data
```{r}

```

### 2023-08-30 2_scripts/2.1 APL Data Quality Check.qmd
Upload APL data cleaning code

---
title: "2.1 APL Data Quality Check with additonal cleaning"
format: html
editor: visual
---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

## Load the APL data with encrypted ID variables
```{r}
load(file = "~/APL/RFD4682e.RData")
```

## Load paks
```{r}
library(dplyr)
```

```{r}
## Create subset for those NAAT Positive Cases
table(RFD4682_e$NAAT_STATUS, useNA = 'ifany')
RFD4682_NAAT<-RFD4682_e[RFD4682_e$NAAT_STATUS %in% c('Positive'),]

# Save the subset

# Load Data file

```

## The distribution of time from NAAT to Sero-assay

```{r}
hist(RFD4682_e$DAYS_SINCE_NAAT[RFD4682_e$NAAT_STATUS=="Positive"], breaks = 18,
     main="Time form NAAT to Sero-assay (Days)",
     xlab = "Days")

## The time variable is good to go
```

## Records are not patients, how many individual has been included?
```{r}
## ULI_ID: Unique Lifetime Identifier (ULI), numerical values of Alberta PHN.
length(unique(RFD4682_e$ULI_IDe))     # Unique = 180821
table(is.na(RFD4682_e$ULI_IDe))       # NA = 11298

## PERSON_ID: assigned from the originating LIS
length(unique(RFD4682_e$PERSON_IDe))     # Unique = 189864
table(is.na(RFD4682_e$PERSON_IDe))       # NA = 2973

## SERO_ID: Unique Specimen ID + Test month, unique APL dataset
length(unique(RFD4682_e$SERO_ID)) # N=214,772
table(is.na(RFD4682_e$SERO_ID))   # No missing

## Checking duplicated records by SERO_ID
tab_SERO_ID<-table(RFD4682_e$SERO_ID)
View(tab_SERO_ID)

str(tab_SERO_ID)
tab_SERO_ID.dat<-as.data.frame(tab_SERO_ID)
tab_SERO_ID.dat[tab_SERO_ID.dat$Freq==2,]
```
## Why unique PERSON_ID and ULI_ID are not matching with each other?
```{r}
## Check the discrepancies between those two

## Freq tab for ULI_ID
tab_ULI_IDe<-table(RFD4682_e$ULI_IDe)
tab_ULI_IDe.dat<-as.data.frame(tab_ULI_IDe)
colnames(tab_ULI_IDe.dat)<-c('ULI_IDe','Freq_ULI')

# tab_ULI_ID.dat.2<-tab_ULI_ID.dat[tab_ULI_ID.dat$Freq>1,]

## Freq tab for RERSON_ID
tab_PERSON_IDe<-table(RFD4682_e$PERSON_IDe)
tab_PERSON_IDe.dat<-as.data.frame(tab_PERSON_IDe)
colnames(tab_PERSON_IDe.dat)<-c('PERSON_IDe','Freq_PER')

## Add freq var to the subset
RFD4682_e$order_ID<-1:214780
RFD4682_3<-select(RFD4682_e, order_ID, SERO_ID, ULI_IDe, PERSON_IDe)
RFD4682_3<-merge(RFD4682_3, tab_ULI_IDe.dat, by='ULI_IDe',
      all.x = T, all.y = F)
RFD4682_3<-merge(RFD4682_3, tab_PERSON_IDe.dat, by='PERSON_IDe',
      all.x = T, all.y = F)

## Find Discrepancies by compare the frequencies

## NA is messy with logic, replace it with 0
RFD4682_3$Freq_ULI[is.na(RFD4682_3$Freq_ULI)==T]<-0
RFD4682_3$Freq_PER[is.na(RFD4682_3$Freq_PER)==T]<-0
tab.idfreq<-table(RFD4682_3$Freq_ULI, RFD4682_3$Freq_PER, useNA = 'ifany')
tab.idfreq
dat.idfreq<-as.data.frame(tab.idfreq)

# Scenario 0: Frequencies of ULI and Person_ID are both missing (equal too)
RFD4682_3$ID_clean<-rep(NA, 214780)
RFD4682_3$ID_clean[RFD4682_3$Freq_ULI==0 &
                   RFD4682_3$Freq_PER==0  ]<-0

# Scenario 1: Frequencies of ULI and Person_ID are equal
length(unique(RFD4682_3$ULI_IDe[(RFD4682_3$Freq_ULI==RFD4682_3$Freq_PER) &
                                 RFD4682_3$Freq_ULI>0                    &
                                 RFD4682_3$Freq_PER>0])
) # N=179,460 This is the majority of cases in the dataset, which is good
# This number matches with the numbers from the confusion tab

RFD4682_3$ID_clean[(RFD4682_3$Freq_ULI==RFD4682_3$Freq_PER) &
                                 RFD4682_3$Freq_ULI>0       &
                                 RFD4682_3$Freq_PER>0]<-1

# Scenario 2: only ULI, Person_ID are missing
# Number estimated using confusion tab: 107, 108 records
RFD4682_3$ID_clean[RFD4682_3$Freq_ULI>0 &
                   RFD4682_3$Freq_PER==0  ]<-2

# Scenario 3, ULI missing but they have PERSON_ID
# Number estimated using confusion tab: 7904, 8433 records
RFD4682_3$ID_clean[RFD4682_3$Freq_ULI==0 &
                   RFD4682_3$Freq_PER>0  ]<-3

# Scenario 4, one PERSON_ID matches with multiple ULI_IDs
# Number estimated using confusion tab: 6, 10 records
RFD4682_3$ID_clean[RFD4682_3$Freq_ULI==1 &
                   RFD4682_3$Freq_PER==2  ]<-4 # this would only fix 6 records

# Fix the rest 4 manually
RFD4682_3$ID_clean[RFD4682_3$PERSON_IDe=="9a4299b85800" &
                   is.na(RFD4682_3$ULI_IDe)==TRUE]<-4
RFD4682_3$ID_clean[RFD4682_3$PERSON_IDe=="aa6526b42bfe" &
                   is.na(RFD4682_3$ULI_IDe)==TRUE]<-4
RFD4682_3$ID_clean[RFD4682_3$PERSON_IDe=="e054534f4352" &
                   is.na(RFD4682_3$ULI_IDe)==TRUE]<-4
RFD4682_3$ID_clean[RFD4682_3$PERSON_IDe=="fcb0ca36fbb4" &
                   is.na(RFD4682_3$ULI_IDe)==TRUE]<-4

RFD4682_3$ULI_IDe[RFD4682_3$PERSON_IDe=="9a4299b85800" &
                   is.na(RFD4682_3$ULI_IDe)==TRUE]<-"3c32a58470ee228"
RFD4682_3$ULI_IDe[RFD4682_3$PERSON_IDe=="aa6526b42bfe" &
                   is.na(RFD4682_3$ULI_IDe)==TRUE]<-"f21a83130a09b98"
RFD4682_3$ULI_IDe[RFD4682_3$PERSON_IDe=="e054534f4352" &
                   is.na(RFD4682_3$ULI_IDe)==TRUE]<-"c80b7fbde166521"
RFD4682_3$ULI_IDe[RFD4682_3$PERSON_IDe=="fcb0ca36fbb4" &
                   is.na(RFD4682_3$ULI_IDe)==TRUE]<-"08bc00a9a5e8720"

# The rest are scenario 5
# N records = 2775, number matches those on confusion cross-tab
RFD4682_3$ID_clean[is.na(RFD4682_3$ID_clean)==TRUE]<-5

# Rename the variable as "scenario"
RFD4682_3<-rename(RFD4682_3, Scenario = ID_clean)
table(RFD4682_3$Scenario, useNA = 'ifany')

# Apply approperiate individual ids to each records, according to its scenario
RFD4682_3$clean_IDe<-rep(NA, 214780)
RFD4682_3$clean_IDe<-if_else(RFD4682_3$Scenario %in% c(3), RFD4682_3$PERSON_IDe, RFD4682_3$ULI_IDe)
```

## Scenario Examples Showcase
```{r}
# Example with each Scenarios
RFD4682_3[RFD4682_3$Freq_ULI==6 & RFD4682_3$Freq_PER==5 ,]     # Scenario 5
RFD4682_3[RFD4682_3$Freq_ULI==15 & RFD4682_3$Freq_PER==15 ,]   # Scenario 1
RFD4682_3[RFD4682_3$Freq_ULI==2 & RFD4682_3$Freq_PER==0 ,]     # Scenario 2
RFD4682_3[RFD4682_3$Freq_ULI==0 & RFD4682_3$Freq_PER==5 ,]     # Scenario 3
RFD4682_3[RFD4682_3$Freq_ULI==1 & RFD4682_3$Freq_PER==2 ,]     # Scenario 4
RFD4682_3[RFD4682_3$Freq_ULI==0 & RFD4682_3$Freq_PER==0 ,]     # Scenario 0
```

## Add clean_IDe and Scenario back to dataset (encrypted)
```{r}
RFD4682_e<-merge(RFD4682_e, select(RFD4682_3, order_ID, Scenario, clean_IDe),
                      by = 'order_ID', all = TRUE)
```

#### ULI ID and PERSON_ID, in NAAT+ subset
```{r}
## ULI_ID: Unique Lifetime Identifier (ULI), numerical values of Alberta PHN.
length(unique(RFD4682_NAAT$ULI_ID))     # N = 5933
table(is.na(RFD4682_NAAT$ULI_ID))       # NA = 0

## PERSON_ID: assigned from the originating LIS
length(unique(RFD4682_NAAT$PERSON_ID)) # N = 5940
table(is.na(RFD4682_NAAT$PERSON_ID))   # NA = 0

## Check the discrepancies between those two
tab_ULI_ID<-table(RFD4682_NAAT$ULI_ID)
tab_ULI_ID.dat<-as.data.frame(tab_ULI_ID)
colnames(tab_ULI_ID.dat)<-c('ULI_ID','Freq_ULI')
tab_ULI_ID.dat.2<-tab_ULI_ID.dat[tab_ULI_ID.dat$Freq>1,]

tab_PERSON_ID<-table(RFD4682_NAAT$PERSON_ID)
tab_PERSON_ID.dat<-as.data.frame(tab_PERSON_ID)
colnames(tab_PERSON_ID.dat)<-c('PERSON_ID','Freq_PER')
tab_PERSON_ID.dat.2<-tab_PERSON_ID.dat[tab_PERSON_ID.dat$Freq>1,]

## Add freq var to 
RFD4682_NAAT_3<-select(RFD4682_NAAT, SERO_ID, ULI_ID, PERSON_ID)
RFD4682_NAAT_3<-merge(RFD4682_NAAT_3, tab_ULI_ID.dat.2, by='ULI_ID',
      all.x = T, all.y = F)
RFD4682_NAAT_3<-merge(RFD4682_NAAT_3, tab_PERSON_ID.dat.2, by='PERSON_ID',
      all.x = T, all.y = F)

RFD4682_NAAT_4<-RFD4682_NAAT_3[is.na(RFD4682_NAAT_3$Freq_ULI)==F |
               is.na(RFD4682_NAAT_3$Freq_PER)==F,]

RFD4682_NAAT_4$freq_minus<-RFD4682_NAAT_4$Freq_ULI-RFD4682_NAAT_4$Freq_PER

print(RFD4682_NAAT_4[RFD4682_NAAT_4$freq_minus>0 |
               is.na(RFD4682_NAAT_4$freq_minus)==TRUE,])

## Save the dataset
save(RFD4682_e, file = '~/APL/RFD4682e.RData')

## Conclusion:
## Use ULI_ID, in NAAT+ subset, seven people have 2 diff PERSON_IDs
## Unique ULI_ID: N=5940
```

```{r}
length(unique(RFD4682_e$clean_IDe))
table(is.na(unique(RFD4682_e$clean_IDe)), useNA = 'ifany')

# Unique Individual in the dataset: 188,722 (beware of missing values)
```

## Check N=2865 cases with missing in both ULI and Person_ID
```{r}
RFD4682_e[RFD4682_e$Scenario==0,]
## Surprisingly, most of the records in this group has some useful variables,
## such as: age, gender and Anti_RBD test results

# Age
table(RFD4682_e$AGE_AT_COLLECTION[RFD4682_e$Scenario==0], useNA = 'ifany') # 0.03%
# Gender
table(RFD4682_e$GENDER[RFD4682_e$Scenario==0], useNA = 'ifany') # 0.03%
# N-Ig
table(RFD4682_e[RFD4682_e$Scenario==0,]$`N-IgG_RESULT`, useNA = 'ifany') # Can't show all
table(is.na(RFD4682_e[RFD4682_e$Scenario==0,]$`N-IgG_RESULT`), useNA = 'ifany') # 92.95%
table(is.na(RFD4682_e$`N-IgG_RESULT`), useNA = 'ifany') # 92.95%

# Ig Anti-RBD
table(is.na(RFD4682_e[RFD4682_e$Scenario==0,]$`RBD-IgGII_RESULT`), useNA = 'ifany')
# no missing, good

# FSA
table(is.na(RFD4682_e[RFD4682_e$Scenario==0,]$PAT_FSA), useNA = 'ifany') # 9.49%

# Vaccination
table(is.na(RFD4682_e[RFD4682_e$Scenario==0,]$VACCINATION_STATUS), useNA = 'ifany')
# No miss, good
table(RFD4682_e[RFD4682_e$Scenario==0,]$VACCINATION_STATUS, useNA = 'ifany')
table(RFD4682_e$VACCINATION_STATUS, useNA = 'ifany')

# NAAT Status
table(is.na(RFD4682_e[RFD4682_e$Scenario==0,]$NAAT_STATUS), useNA = 'ifany') # 92.98%
table(is.na(RFD4682_e$NAAT_STATUS), useNA = 'ifany')

id_temp<-RFD4682_e$SERO_ID[RFD4682_e$VACCINATION_STATUS %in% c("Pre-Immunization")]
View(id_temp)
max(id_temp, na.rm = T) # 2020 Nov
min(id_temp, na.rm = T)

# Conclusion: The data quality of this part is good, but we cannot use those in our project
```

## The Whole APL dataset
```{r}
# The NAAT status variable
table(RFD4682_e$NAAT_STATUS, useNA = 'ifany')

# The Vaccination Status Variable
table(RFD4682_e$VACCINATION_STATUS, useNA = 'ifany')

# Cross-tab of those two
table(RFD4682_e$NAAT_STATUS, RFD4682_e$VACCINATION_STATUS, useNA = 'ifany')
```
## Categorize the time since last NAAT test
```{r}
hist(RFD4682_e$DAYS_SINCE_NAAT, breaks=20)

# Create an indicator
RFD4682_e$DAYS_NAAT_cat<-rep(NA, 214780)
RFD4682_e$DAYS_NAAT_cat[RFD4682_e$DAYS_SINCE_NAAT>=0 &
                        RFD4682_e$DAYS_SINCE_NAAT<=90 ]<-1  # <3 months
RFD4682_e$DAYS_NAAT_cat[RFD4682_e$DAYS_SINCE_NAAT>90 &
                        RFD4682_e$DAYS_SINCE_NAAT<=180 ]<-2  # 4-6 months
RFD4682_e$DAYS_NAAT_cat[RFD4682_e$DAYS_SINCE_NAAT>180 &
                        RFD4682_e$DAYS_SINCE_NAAT<=270 ]<-3  # 4-6 months
RFD4682_e$DAYS_NAAT_cat[RFD4682_e$DAYS_SINCE_NAAT>270 &
                        is.na(RFD4682_e$DAYS_SINCE_NAAT)==F ]<-4  # 4-6 months
table(RFD4682_e$DAYS_NAAT_cat, useNA = 'ifany')

# Cross-tabulation between NAAT and NAAT_DAYS categories
# Anti-N groups
table(RFD4682_e$DAYS_NAAT_cat[is.na(RFD4682_e$`N-IgG_RESULT`)==F], 
      RFD4682_e$NAAT_STATUS[is.na(RFD4682_e$`N-IgG_RESULT`)==F], useNA = 'ifany')

# Anti-RBD groups
table(RFD4682_e$DAYS_NAAT_cat[is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F &
                              RFD4682_e$VACCINATION_STATUS %in% c('Not Immunized')], 
      RFD4682_e$NAAT_STATUS[is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F &
                              RFD4682_e$VACCINATION_STATUS %in% c('Not Immunized')], useNA = 'ifany')
```

## Vaccination Status
```{r}
table(RFD4682_NAAT$VACCINATION_STATUS, useNA = 'ifany')
hist(RFD4682_NAAT$DAYS_SINCE_IMM)

# Create day gap

## NAs in NAAT Days var
table(is.na(RFD4682_NAAT$DAYS_SINCE_NAAT)) # Great, no NAs

## NAs in Vaccination Days var
table(is.na(RFD4682_NAAT$DAYS_SINCE_IMM)) # Only 911 missing, not match with Vacc Status

## Calculate the Gap
RFD4682_NAAT$DAYS_Gap<-rep(NA,6165)
RFD4682_NAAT$DAYS_Gap<-RFD4682_NAAT$DAYS_SINCE_NAAT - RFD4682_NAAT$DAYS_SINCE_IMM
table(is.na(RFD4682_NAAT$DAYS_Gap)) #911 Missing, 0 vaccinated at the same day of sample collection
hist(RFD4682_NAAT$DAYS_Gap, 
     main = 'Gap between NAAT and Vaccination',
     xlab = 'Days')

## Create the indicator
RFD4682_NAAT$Vacc_status<-rep(NA, 6165)
RFD4682_NAAT$Vacc_status[RFD4682_NAAT$VACCINATION_STATUS %in% c("Not Immunized")]<-1
RFD4682_NAAT$Vacc_status[RFD4682_NAAT$DAYS_Gap>=0 & is.na(RFD4682_NAAT$DAYS_Gap)==F]<-2
RFD4682_NAAT$Vacc_status[RFD4682_NAAT$DAYS_Gap<0 & is.na(RFD4682_NAAT$DAYS_Gap)==F]<-3
table(RFD4682_NAAT$Vacc_status, useNA = 'ifany')
```

### 2023-10-23 5_manuscript/manuscript.qmd
prepare str of second paper
Updated methods section (still a plan)

@@
 
+Two types of antibodies to be detected by assays: Anti-S and Anti-N:
+
+anti-S (humoral response to infection or vaccination) and anti-N (marker of natural infection)
+
+Waning of immunity against SARS-COV-2 after infection and immunization, indicated by decreasing blood Ab level overtime.
+
+\[insert the figure of declining Ab with re-activation of immunization\]
+
 # Methods
 
-## Subsection
+**Notes**: Unlike the DBS vs VBS manuscript, the methods section of this paper is mostly [unfinished]{.underline}.
+This section only shows the plan and the flow chart of the project in general, while the details of it needs constantly updating and polishing overtime, as the project goes.
+The potential obstacles were listed below:
 
-A
+-   AB Antenatal Data has not arrived yet.
+
+-   In AB Outpatient Lab Data, they changed the assay (Diasorin S --\> Abbott S) in the midway, this would affect the validity of analysis based on quantitative test results, as the results from different assay are not directly comparable.
+
+## Dataset
+
+To estimate the seroprevalence and attack rate in Alberta, we used data from the Alberta Outpatient Lab (APL), Alberta Antenatal Testing, and Canadian Blood Services datasets.
+CBS has tested \>450,000 blood donations for both SARS-CoV-2 anti-S and anti-N levels since March 2020.
+Data include donor's sex, self-reported race/ethnicity, and age.
+From their address, neighborhood material/social deprivation and urban/rural indicator have been derived.
+This analysis uses donations tested in Alberta (N=XX) only.
+
+Alberta Health Services (AHS) tested \>160,000 Albertan outpatient labs (APL) and \>6,000 antenatal blood draws for anti-N and anti-S antibodies from June 2020 until December 2021.
+These data contain all the same variables as the blood donors except for race/ethnicity.
+Besides of serology test results and demographic data, CBS has linked the administrative data for the Alberta cohort (**is the link finished?**), which provided additional on their hospitalizations, clinic visits, diagnoses, COVID vaccinations, and long-term disability claims.
+Similarly, AHS has completed a similar linkage in the Albertan outpatient lab (APL) data and Alberta Antenatal Blood Test data.
+See Table.X below for more details.
+
+## **Seroprevalence and attack rate**
+
+To bridge the gap between test positivity and population-level seroprevalence, we will employ the Rogen-Gladen (RG) equation and Bayesian models.
+Following seroprevalence estimation, we will further estimate the attack rate while adjusting for waning immunity.
+These estimations will help to answer what proportion of the Alberta outpatient population that has been infected by SARS-CoV-2.
+
+We presented both seroprevalence and attack rate in a unified graph (**Fig**. X), displaying changes over time on a four-month scale.
+
+## **Proportion sufficiently protected:**
+
+In evaluating whether participants possess immunity levels sufficient to shield them from future infections, from either vaccination or infection, we will apply an age- and sex-adjusted threshold of protection to the quantitative Anti-S assay results.
+We will also present the fluctuation of the proportion sufficiently protected over time in another figure, and we will conduct sensitivity analyses in which we compare alternative threshold levels.
+The details of the project flowchart is presented in the **Fig**.
+Y below.
 

### 2023-10-26 5_manuscript/manuscript.qmd
Updated methods section, still a plan

@@
 
-{{< pagebreak >}}
+The Covid-19 pandemic overview, global and what it looks like in Canada.
 
-Two types of antibodies to be detected by assays: Anti-S and Anti-N:
+Serosurveillance programs in Canada, and its importance
+
+Two types of antibodies to be detected by assays: Anti-S and Anti-N. Why we treated them differently in data analysis. Quantitative and qualitative test results.
 
@@
 
+Seropravalence and attack rate, why we need to employ RG equation or Bayesian model to estimated seropravalence, and why we also need models to adjust for waning immunity.
+
+{{< pagebreak >}}
+
 # Methods
 
-**Notes**: Unlike the DBS vs VBS manuscript, the methods section of this paper is mostly [unfinished]{.underline}.
+**Notes**: Unlike the DBS vs VBS manuscript, the methods section of this project is mostly [unfinished]{.underline}.
 This section only shows the plan and the flow chart of the project in general, while the details of it needs constantly updating and polishing overtime, as the project goes.
@@
 
-To estimate the seroprevalence and attack rate in Alberta, we used data from the Alberta Outpatient Lab (APL), Alberta Antenatal Testing, and Canadian Blood Services datasets.
+To estimate the SARS-CoV-2 seroprevalence and attack rate in Alberta, we used data from the Alberta Outpatient Lab (APL), Alberta Antenatal Testing, and Canadian Blood Services (CBS) datasets.
 CBS has tested \>450,000 blood donations for both SARS-CoV-2 anti-S and anti-N levels since March 2020.
@@
 These data contain all the same variables as the blood donors except for race/ethnicity.
-Besides of serology test results and demographic data, CBS has linked the administrative data for the Alberta cohort (**is the link finished?**), which provided additional on their hospitalizations, clinic visits, diagnoses, COVID vaccinations, and long-term disability claims.
-Similarly, AHS has completed a similar linkage in the Albertan outpatient lab (APL) data and Alberta Antenatal Blood Test data.
-See Table.X below for more details.
+Besides of serology test results and demographic data, CBS has linked the administrative data for the Alberta cohort (**is the linkage finished?**), which provided additional information on their hospitalizations, clinic visits, diagnoses, COVID vaccinations, and long-term disability claims.
+Similarly, AHS has completed a linkage in the Albertan outpatient lab (APL) data and Alberta Antenatal Blood Test data.
+We did not conduct analysis using a merged data, but instead, we conducted analysis separately for each data and presented the results in a unified graph (**Fig**.
+Y).
+**Table.X** below sumarrized the differences among three data utilized in the study.
+For more detail, see **Table.
+Z** in the result section.
+
+**Table.X** Summary of difference among serosurveillance datasets from CBS and AHS
+
+\[insert a table here\]
 
@@
 
-To bridge the gap between test positivity and population-level seroprevalence, we will employ the Rogen-Gladen (RG) equation and Bayesian models.
-Following seroprevalence estimation, we will further estimate the attack rate while adjusting for waning immunity.
-These estimations will help to answer what proportion of the Alberta outpatient population that has been infected by SARS-CoV-2.
-
-We presented both seroprevalence and attack rate in a unified graph (**Fig**. X), displaying changes over time on a four-month scale.
+To evaluate the severity of Covid-19 pandemic overtime in Alberta province, first we calculated the anti-N assay test positivity.
+Then, to bridge the gap between test positivity and population-level seroprevalence, we employed the Rogen-Gladen (RG) equation and Bayesian models.
+Following seroprevalence estimation, we further estimated the attack rate while adjusting for waning immunity.
+These estimations are dessignated to answer what proportion of the Alberta population that has been infected by SARS-CoV-2.
+We presented both seroprevalence and attack rate in a unified graph (**Fig**.
+X), displaying changes over time on a four-month scale.
 
@@
 
-In evaluating whether participants possess immunity levels sufficient to shield them from future infections, from either vaccination or infection, we will apply an age- and sex-adjusted threshold of protection to the quantitative Anti-S assay results.
-We will also present the fluctuation of the proportion sufficiently protected over time in another figure, and we will conduct sensitivity analyses in which we compare alternative threshold levels.
+In evaluating whether participants possess immunity levels sufficient to shield them from future infections, from either vaccination or infection, we applied an age- and sex-adjusted threshold of protection to the quantitative Anti-S assay results.
+The threshold is cited from a valid source \[source of threshold\], indicating sufficient immunity against future SARS-CoV-2 infection.
+We also presented the fluctuation of the proportion sufficiently protected over time in another figure (**Fig.** Z), and we conducted sensitivity analyses in which we compare alternative threshold levels.
 The details of the project flowchart is presented in the **Fig**.
 Y below.
+
+Fig.Y Project Flowchart
+
+\[Insert the flowchart here\]
 
@@
 
-A
+**Table.Z** Difference among serosurveillance datasets from CBS and AHS
+
+ANOVA test, p-values...
 

### 2023-12-04 2_scripts/2.1 APL Data Quality Check.qmd
APL Figures

@@
 ## Load the APL data with encrypted ID variables
+
 ```{r}
-load(file = "~/APL/RFD4682e.RData")
+load(file = "../1_data/private/RFD4682e.RData")
 ```
@@
 ## Load paks
+
 ```{r}
-library(dplyr)
+library(dplyr)         # Data management
+library(lubridate)     # Manage variables in Date formats
+library(zoo)           # for yearmonth Date format
+library(reshape2)      # for melt() function
 ```
@@
 ## Records are not patients, how many individual has been included?
+
 ```{r}
@@
 ```
+
 ## Why unique PERSON_ID and ULI_ID are not matching with each other?
+
 ```{r}
@@
 ## Scenario Examples Showcase
+
 ```{r}
@@
 ## Add clean_IDe and Scenario back to dataset (encrypted)
+
 ```{r}
@@
 #### ULI ID and PERSON_ID, in NAAT+ subset
+
 ```{r}
@@
 ## Check N=2865 cases with missing in both ULI and Person_ID
+
 ```{r}
@@
 ## The Whole APL dataset
+
 ```{r}
@@
 ```
+
 ## Categorize the time since last NAAT test
+
 ```{r}
@@
 ## Vaccination Status
+
 ```{r}
@@
 ```
+
+Estimation on number of individuals, using cleaned version of IDe
+
+```{r}
+# Total num of individual in the data:
+table(is.na(unique(RFD4682_e$clean_IDe)), useNA = 'ifany')    # N=188,722, plus 1 NA
+
+# Number of participants with available Anti-N results:
+antin_na<-is.na(unique(RFD4682_e[is.na(RFD4682_e$clean_IDe)==F &
+                             is.na(RFD4682_e$`N-IgG_RESULT`)==F,
+                             ]$clean_IDe))
+table(antin_na, useNA = 'ifany')  # N=39450
+
+# Number of participants with available Anti-S results:
+antis_na<-is.na(unique(RFD4682_e[is.na(RFD4682_e$clean_IDe)==F &
+                             is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F,
+                             ]$clean_IDe))
+table(antin_na, useNA = 'ifany') # N=188,688
+
+table(is.na(RFD4682_e$`N-IgG_RESULT`)==F,
+      is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F, useNA = 'ifany')
+```
+Some additional cleaning/editing, making anti-S quantitative
+```{r}
+# Missing values
+summary(is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F) 
+# N=36, beware of "not available" values, better recode those as NAs
+
+# Remove all the numbers, see what left
+S_chr<-RFD4682_e$`RBD-IgGII_RESULT` %>%
+  gsub("[[:digit:]]+", "",.)
+table(S_chr, useNA = 'ifany')
+## As we can see, the use of units in this data is quite messy, needs additonal cleaning
+
+# How many of them use unit "AU/ml" (arbitury unit)?
+AU<-RFD4682_e$`RBD-IgGII_RESULT` %>%
+  grepl("ml", .,fixed = F, ignore.case = T)
+table(AU) # N=109798
+
+# index
+index<-RFD4682_e$`RBD-IgGII_RESULT` %>%
+  grepl("index", .,fixed = F, ignore.case = T)
+table(index) # N=104946
+```
+#### Yuan found this: anti-N is only available for the Omicron stage
+```{r}
+# The APL Data:
+RFD4682_e$COLLECTION_DATE %>%
+  summary()
+
+# Anti-N assay results is only available to N=39,450 participants
+RFD4682_e[is.na(RFD4682_e$clean_IDe)==F &
+          is.na(RFD4682_e$`N-IgG_RESULT`)==F,]$COLLECTION_DATE %>%
+  summary()
+# Anti-N: 2021/12/14 -- 2022/10/14
+
+# Anti-S assay results is available to most of the participants
+RFD4682_e[is.na(RFD4682_e$clean_IDe)==F &
+          is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F,]$COLLECTION_DATE %>%
+  summary()
+```
+#### A quick clean of the Anti-N data
+```{r}
+# Remove unecessary 'index'
+table(RFD4682_e$`N-IgG_RESULT`)
+
+RFD4682_e$N<-RFD4682_e$`N-IgG_RESULT` %>%
+  gsub("Index", "",.) %>%
+  as.numeric()
+
+hist(RFD4682_e$N, breaks=25)
+
+# Results
+table(RFD4682_e$`N-IgG_INTERP`, useNA = 'ifany')
+```
+#### Calculate positivity directly
+
+```{r}
+# Create month var
+RFD4682_e$Col_Date<-RFD4682_e$COLLECTION_DATE %>%
+  as.Date(., tz="UTC") # a short time var, date only
+
+library(lubridate)
+library(zoo)
+
+# Only keep the month, remove the data
+RFD4682_e$mois <- month(RFD4682_e$Col_Date)
+RFD4682_e$annee <- year(RFD4682_e$Col_Date)
+RFD4682_e$Col_Month<-paste(RFD4682_e$annee, RFD4682_e$mois, 
+                           sep = "-")
+table(RFD4682_e$Col_Month, useNA='ifany')
+# as char variable if needed for calculation, use as.yearmon() from {zoo} or just use the Col_Date var
+```
+
+```{r}
+# Calculate positivity for each month
+tb1<-RFD4682_e[is.na(RFD4682_e$`N-IgG_RESULT`)==F,] %>%
+  group_by(Col_Month) %>%
+  summarise(n = n())
+
+tb2<-RFD4682_e[RFD4682_e$`N-IgG_INTERP`=="Positive",] %>%
+  group_by(Col_Month) %>%
+  summarise(n.pos = n())
+
+tb1<-merge(tb1, tb2, by = "Col_Month", all.x = T, all.y = F)
+tb1$prop.n<-tb1$n.pos/tb1$n
+tb1$Col_Month<-as.yearmon(tb1$Col_Month)
+tb1<-tb1[order(tb1$Col_Month),]
+
+# Add R-G adjusted prevalances
+# Sen=0.881, Spec = 0.947
+# Source:
+tb1$prop.n_RG<- (tb1$prop.n+0.881-1)/(0.881+0.947-1)
+
+# remove NAs
+tb1<-tb1[!is.na(tb1$Col_Month),]
+
+tb1 %>%
+  with(plot(y=.$prop.n, x = .$Col_Month, type="b",
+            ylab="positivity", xlab="Month",col="blue")
+       )
+```
+
+#### Apply ggplot, plot two together
+```{r}
+tb1_melt<-melt(select(tb1, Col_Month, prop.n, prop.n_RG), id.vars = "Col_Month")
+
+tb1_melt %>%
+  ggplot(aes(x=Col_Month, y=value, colour=variable)) +
+  geom_point(aes(colour=variable)) +
+  geom_line(aes(colour=variable))  +
+  theme_bw()
+```
+
+```{r}
+
+plot(x=temp$Time ,y=temp$Positivity, type = 'b', 
+     col="blue", xlab="Time: 2022", ylab="Test Positivity",
+     main="Anti-N Assay Positivity in Omicron Wave, using APL Data")
+plot(x=temp$Time ,y=temp$Positivity_RG, 
+     type = 'b', col="red")
+```

### 2023-12-04 2_scripts/CITF_CBS_desc_stats.Rmd
APL Figures

---
title: "CITF copy of CBS serology project data: descriptive stats"
author: "Ahmed Moustafa Ahmed and Tanya Murphy"
date: "Last updated: Oct 14, 2022, Created: May 5, 2022"
output:
  html_document:
    toc: true
    number_sections: true
    code_folding: hide
    df_print: paged
---

|---------------------------------------------------------------------------------|
| Generating CBS descriptive statistics:                                          |                
|   1. Make sure to have the up-to-date tables ready from CBS_ETL_function script |
|---------------------------------------------------------------------------------|

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(haven)
library(lubridate)
library(tidyverse) # loads readr
```

## CITF database

Canadian Blood Services (CBS) seroprevalence project data: CBS shares its seroprevalence results with basic donor demographics with the CITF for internal use, which includes seroprevalence estimation by the CITF Data Management and Analysis team and collaborations with students and researchers on analyses deemed of high importance to CITF's objectives (rapid dissemination of COVID-19 immunity surveillance results to decision makers).

``` {r postgresql-tables}
# to read the prepared data tables
library(DBI)
library(RPostgres)

### Test connection arguments
con <- dbConnect(
    RPostgres::Postgres(),
    dbname = 'cbs0', 
    host = '132.216.183.71', 
    port = 5432, # or any other port specified by your DBAuser = "rstudio", 
    user = "jchen", # CITF-provided postgres name and password
    rstudioapi::askForPassword("Database password:") 
)
# Set search path (does not include 'public', which is for internal CITF use)
dbSendQuery(con, 'set search_path = "$user", "students"') 
dbListTables(con)

# you may keep intermediate tables in your schema: yyu.[tablename]

```

## Total Donations, One-time and Repeat Donors
```{r stats1}
data_all   <- dbGetQuery(con, "SELECT * FROM students.copy_cbs_combined") # all data 
data_demog <- dbGetQuery(con, "SELECT * FROM students.copy_cbs_demog") # socio-demog table
data_sero <- dbGetQuery(con, "SELECT * FROM students.copy_cbs_sero") # serology table
repeat_donors <- data_demog %>% filter(sero_num > 1) # repeat donors

#class(data_all$sampledate)

# Basic stats------------------------------------------------------
n_donations <- length(data_all$pid) # total number of donations N=743,999
# or
n_donation2 <- sum(data_demog$sero_num) # sum of total donations for each donor
# N=778,662
#cat("Total number of donations:", n_donations2)

n_donors <- length(data_demog$pid) # number of unique donors
# length(unique(data_demog$pid)) # check

n_onetime_donors <- sum(data_demog$sero_num == 1) # number of one-time donors

n_repeat_donors  <- sum(data_demog$sero_num > 1) # number of repeat donors
#------------------------------------------------------------------

# All Donors-------------------------------------------------------
ever_n_positive  <- sum(data_demog$npos_ever == TRUE, na.rm = T) # All donors ever n positive on either Abbott or Roche (mostly Roche from Jan 2021)

ever_n_positive_abbott  <- sum(data_demog$npos_ever_abbott == TRUE, na.rm = T) # All donors ever n positive on Abbott

ever_n_positive_roche   <- sum(data_demog$npos_ever_roche == TRUE, na.rm = T)  # All donors ever n positive on Roche
#------------------------------------------------------------------

# Repeat Donors----------------------------------------------------
r_ever_n_positive <- sum(repeat_donors$npos_ever, na.rm = T) # Repeat donors ever n positive on either Abbott or Roche (mostly Roche from Jan 2021)

r_ever_n_positive_abbott <- sum(repeat_donors$npos_ever_abbott, na.rm = T) # Repeat donors ever n positive on Abbott

r_ever_n_positive_roche  <- sum(repeat_donors$npos_ever_roche, na.rm = T)   # Repeat donors ever n positive on Roche

# Create the mode function.
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

r_mode_donations <- getmode(repeat_donors$sero_num) # Most repeated number of donations

#------------------------------------------------------------------
```

### Stats from pos to neg /// neg to pos for anti-N
```{r stats2}

# Serology---------------------------------------------------------
# 74 donors went from p to n, the timepassed is from the first positive to the first negative presented in the histogram
repeat_sero_p_to_n <- data_sero %>% group_by(pid) %>% arrange(pid, sampledate) %>% mutate(sero_num = n()) %>% filter(sero_num > 1 & grepl('Roche', assay_type)) %>% filter(cummax(cur_result_n == "Roche Positive") > 0) %>% mutate(timepassed = sampledate - min(sampledate)) %>%  filter(cummax(cur_result_n == "Roche Negative") > 0) %>% filter(row_number()==1)

n_repeat_sero_p_to_n <- length(unique(repeat_sero_p_to_n$pid))

# Donors going from negative to positive = 2017, timepassed is from first negative to first positive
repeat_sero_n_to_p <- data_sero %>% group_by(pid) %>% arrange(pid, sampledate) %>% mutate(sero_num = n()) %>% filter(sero_num > 1 & grepl('Roche', assay_type)) %>% filter(cummax(cur_result_n == "Roche Negative") > 0) %>% mutate(timepassed = sampledate - min(sampledate)) %>%  filter(cummax(cur_result_n == "Roche Positive") > 0) %>% filter(row_number()==1)

# Very few = 7 observations, went from positive to negative then to positive again
repeat_sero_p_n_p <- data_sero %>% group_by(pid) %>% arrange(pid, sampledate) %>% mutate(sero_num = n()) %>% filter(sero_num > 1 & grepl('Roche', assay_type)) %>% filter(cummax(cur_result_n == "Roche Positive") > 0) %>% mutate(timepassed = sampledate - min(sampledate)) %>%  filter(cummax(cur_result_n == "Roche Negative") > 0) %>% filter(cur_result_n == "Roche Positive") %>% filter(row_number()==1)

n_repeat_sero_p_n_p <- length(unique(repeat_sero_p_n_p$pid))

# 7 Donors going from negative to positive to negative
# repeat_sero_n_p_n <- data_sero %>% group_by(pid) %>% arrange(pid, sampledate) %>% mutate(sero_num = n()) %>% filter(sero_num > 1 & grepl('Roche', assay_type)) %>% filter(cummax(cur_result_n == "Roche Negative") > 0) %>% mutate(timepassed = sampledate - min(sampledate)) %>%  filter(cummax(cur_result_n == "Roche Positive") > 0) %>% filter(cur_result_n == "Roche Negative") %>% filter(row_number()==1)

#------------------------------------------------------------------

```

### Chunk for presenting the results of the above (used in Rmarkdown report)
```{r results}
library("formattable")

a <- c("Total number of donations:", "Total number of donors:", "Number of repeat donors:", "Mode Donations:", "Repeat donors ever N-positive Roche:", "Number of Repeat donor anti-N positive to negative (Roche):", "Number of Repeat donor anti-N positive to negative to positive (Roche):")

b <- c(n_donations, n_donors, n_repeat_donors, r_mode_donations, r_ever_n_positive_roche, n_repeat_sero_p_to_n, n_repeat_sero_p_n_p)

table <- paste(a, b)
table <- as.data.frame(table)
table

hist(main = "Histogram timepassed between donation from first positive to negative results",
     xlab = "timepassed between donations",
     as.numeric(repeat_sero_p_to_n$timepassed))

```

### A chunk to perform some quality checks
```{r checks}
# data checks----------------------------------------------------
#length(unique(data_all$pid)) # check that the result from demogs table and full table match

#check <- data_demog
#length(unique(data_demog$pid)) # to make sure there is one line per donor
#table(data_demog$discrep) # to make sure there is one line per donor

#check <- data_demog %>% filter(sero_num==1) # checking that all rows are uniqe for one-time donors in demog table
#length(unique(check$pid)) # checking that all rows are uniqe for one-time donors in demog table

#check <- data_demog %>% filter(sero_num>1) # checking that all rows are unique for repeat donors in demog table
#length(unique(check$pid)) # checking that all rows are unique for repeat donors in demog table

#check <- data_demog %>% filter(npos_ever==TRUE) # checking that both this value and sum function value are the same
#length(unique(check$pid)) # checking that all rows are unique for npos_ever in demog table

#check <- data_demog %>% filter(sero_num >1 & npos_ever==TRUE) # checking that both this value and sum function value are the same

# check to see if all donations are tested
#sum(data_demog$sero_num)
#length(unique(data_other$pid))
#----------------------------------------------------------

# `r format(round(n_donations, 1), nsmall = 1)`
# `r format(round(n_donors, 1), nsmall = 1)`
# `r format(round(n_onetime_donors, 1), nsmall = 1)`
# `r format(round(n_repeat_donors, 1), nsmall = 1)`

#data_all %>% filter(sampledate >= "2022-03-01" & sampledate < "2022-04-01") # checking number of donations to match CBS reports
#data_jan <- read_sas("/home/amoustafa/data/mcgilldata_jan2_22mar2022.sas7bdat") # discrepancy in cbs jan reporting between v1 and v2

```

```{r aggregate-request}

data <- dbGetQuery(con, "SELECT * FROM students.copy_cbs_combined")

data2 <- data %>% filter(!is.na(roche_n) | !is.na(roche_s)) # when we started having roche result is 2020-12-14

age_groups_fun <- function(variable) {
  age_group = cut(variable, 
      breaks = c(-Inf,10 ,20,30,40,50,60 ,70,80,90, Inf) ,
      labels = c("0-9 years"
                  ,"10-19 years","20-29 years","30-39 years"
                  ,"40-49 years","50-59 years","60-69 years"
                  ,"70-79 years","80-89 years","90+ years"),
                  right = FALSE)
                  return(age_group)}

province_fun <- function(var) {
  fsa_f = as.character(substr(var,1,1))
  prov =  ifelse(fsa_f == "A", "NL",
          ifelse(fsa_f == "B", "NS",
          ifelse(fsa_f == "C", "PE",
          ifelse(fsa_f == "E", "NB",
          ifelse(fsa_f == "G" | fsa_f == "H" | fsa_f == "J", "QC",
          ifelse(fsa_f == "K" | fsa_f == "L" | fsa_f == "M" | fsa_f == "N" | fsa_f == "P", "ON",
          ifelse(fsa_f == "R", "MB",
          ifelse(fsa_f == "S", "SK",
          ifelse(fsa_f == "T", "AB",
          ifelse(fsa_f == "V", "BC",
          ifelse(fsa_f == "X", "NU/NT",
          ifelse(fsa_f == "Y", "YT", NA
            ))))))))))))
          return(prov)}

data3 <- data2 %>% mutate(age = 2022 - dob, age_group = age_groups_fun(age)) %>% mutate(month_year = format(as.Date(data2$sampledate), "%Y-%m")) %>% mutate(province = province_fun(fsa)) %>%  mutate(log_roche_s = log(roche_s)) # for natural log do we take the absolute value or change negative values to zero???

table(data3$province, useNA = "ifany") # check

data3 %>% filter(province == "ON", age_group == "10-19 years", month_year == "2020-12") # check

data4 <- data3 %>% 
  select(age_group, month_year, province, roche_s, log_roche_s) %>%
  group_by(age_group, month_year, province) %>% 
  summarise(mean_s = mean(roche_s), mean_log_s=mean(log_roche_s),
            sd_s=sd(roche_s), sd_log_s=sd(log_roche_s)) %>% 
  arrange(age_group, province, month_year)

```
## Save the data to server
```{r}
CBS<-data3                  # The cleaned CBS data, at least one Roche_N and Roche_s
CBS_desc<-data4             # Data descriptive table
CBS_Ab<-CBS[CBS$province=="AB",]  # AB subset

save(CBS, file = '../1_data/private/CBS.RData')
save(CBS_Ab, file = '../1_data/private/CBS_Ab.RData')
save(CBS_desc, file = '../1_data/private/CBS_desc.RData')
```

```{r close-dbi-connection}
# Close connection to postgres (not needed once data loaded into R)
dbDisconnect(con)

```

### 2023-12-05 2_scripts/2.1 APL Data Quality Check.qmd
Add RG Adj Scripts

@@
 ```
-#### Calculate positivity directly
 
+#### Add a year-month var
 ```{r}
-# Create month var
-RFD4682_e$Col_Date<-RFD4682_e$COLLECTION_DATE %>%
-  as.Date(., tz="UTC") # a short time var, date only
+# only keep the month, remove the data
+RFD4682_e$Col_Month<-format(as.Date(RFD4682_e$COLLECTION_DATE), "%Y-%m")
+table(RFD4682_e$Col_Month, useNA='ifany')
 
-library(lubridate)
-library(zoo)
-
-# Only keep the month, remove the data
-RFD4682_e$mois <- month(RFD4682_e$Col_Date)
-RFD4682_e$annee <- year(RFD4682_e$Col_Date)
-RFD4682_e$Col_Month<-paste(RFD4682_e$annee, RFD4682_e$mois, 
-                           sep = "-")
-table(RFD4682_e$Col_Month, useNA='ifany')
 # as char variable if needed for calculation, use as.yearmon() from {zoo} or just use the Col_Date var
@@
 
+#### Save a cleaned ver of RFD4682e
 ```{r}
-# Calculate positivity for each month
-tb1<-RFD4682_e[is.na(RFD4682_e$`N-IgG_RESULT`)==F,] %>%
-  group_by(Col_Month) %>%
-  summarise(n = n())
-
-tb2<-RFD4682_e[RFD4682_e$`N-IgG_INTERP`=="Positive",] %>%
-  group_by(Col_Month) %>%
-  summarise(n.pos = n())
-
-tb1<-merge(tb1, tb2, by = "Col_Month", all.x = T, all.y = F)
-tb1$prop.n<-tb1$n.pos/tb1$n
-tb1$Col_Month<-as.yearmon(tb1$Col_Month)
-tb1<-tb1[order(tb1$Col_Month),]
-
-# Add R-G adjusted prevalances
-# Sen=0.881, Spec = 0.947
-# Source:
-tb1$prop.n_RG<- (tb1$prop.n+0.881-1)/(0.881+0.947-1)
-
-# remove NAs
-tb1<-tb1[!is.na(tb1$Col_Month),]
-
-tb1 %>%
-  with(plot(y=.$prop.n, x = .$Col_Month, type="b",
-            ylab="positivity", xlab="Month",col="blue")
-       )
+save(RFD4682_e, file = "../1_data/private/RFD4682e.RData")
 ```
-
-#### Apply ggplot, plot two together
-```{r}
-tb1_melt<-melt(select(tb1, Col_Month, prop.n, prop.n_RG), id.vars = "Col_Month")
-
-tb1_melt %>%
-  ggplot(aes(x=Col_Month, y=value, colour=variable)) +
-  geom_point(aes(colour=variable)) +
-  geom_line(aes(colour=variable))  +
-  theme_bw()
-```
-
-```{r}
-
-plot(x=temp$Time ,y=temp$Positivity, type = 'b', 
-     col="blue", xlab="Time: 2022", ylab="Test Positivity",
-     main="Anti-N Assay Positivity in Omicron Wave, using APL Data")
-plot(x=temp$Time ,y=temp$Positivity_RG, 
-     type = 'b', col="red")
-```

### 2023-12-05 2_scripts/2.2 Rogen-Gladen Adj.qmd
Add RG Adj Scripts

---
title: "2.2 Rogen-Gladen Adjustment"
format: html
editor: visual
---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

## Running Code

## Load dataset
```{r}
# The CBS Ab subset data
load(file = "../1_data/private/CBS_Ab.RData")
load(file = "../1_data/private/RFD4682e.RData")
```

## Load necc paks
```{r}
library(dplyr)         # Data management
library(reshape2)      # for melt() function
library(ggplot)        # ggplot
library(zoo)           # year-month format
```
# ============= Part I. The Changing of Anti-N Pos APL ===============#

#### Calculate positivity directly

```{r}
# Calculate positivity for each month
tb1<-RFD4682_e[is.na(RFD4682_e$`N-IgG_RESULT`)==F,] %>%
  group_by(Col_Month) %>%
  summarise(n = n())

tb2<-RFD4682_e[RFD4682_e$`N-IgG_INTERP`=="Positive",] %>%
  group_by(Col_Month) %>%
  summarise(n.pos = n())

tb1<-merge(tb1, tb2, by = "Col_Month", all.x = T, all.y = F)
tb1$prop.n<-tb1$n.pos/tb1$n
tb1$Col_Month<-as.yearmon(tb1$Col_Month)
tb1<-tb1[order(tb1$Col_Month),]

# Add R-G adjusted prevalances
# Sen=0.844, Spec = 1.0
# Source: 
tb1$prop.n_RG<- (tb1$prop.n+1-1)/(1+0.844-1)

# remove NAs
tb1<-tb1[!is.na(tb1$Col_Month),]

tb1 %>%
  with(plot(y=.$prop.n, x = .$Col_Month, type="b",
            ylab="positivity", xlab="Month",col="blue")
       )
```

#### Apply ggplot, plot two together
```{r}
APL_melt<-melt(select(tb1, Col_Month, prop.n, prop.n_RG), id.vars = "Col_Month")

APL_melt %>%
  ggplot(aes(x=Col_Month, y=value, colour=variable)) +
  geom_point(aes(colour=variable)) +
  geom_line(aes(colour=variable))  +
  theme_bw()
```

# ============= Part II. The Changing of Anti-N Pos Overtime ===============#

#### Anti-N in CBS data (Roche)
```{r}
CBS_Ab %>%
  with(summary(roche_n))

hist(log10(CBS_Ab$roche_n))

table(CBS_Ab$interp_roche_n, useNA = 'ifany')
# 54713/(124008-1) = 44.12% much higher than APL

# Negative Cases: roche_n<1, clearly 1.0 is the COI, as it published by Roche
CBS_Ab %>%
  .[.$interp_roche_n==0, ] %>%
  with(summary(roche_n))

# Positive Cases: roche_n>=1
CBS_Ab %>%
  .[.$interp_roche_n==1, ] %>%
  with(summary(roche_n))
```

```{r}
# Check the data collection time

# Abbott only - not exist
CBS_Ab %>%
  .[!is.na(.$abbott_n) & is.na(.$roche_n), ] %>%
  with(summary(sampledate))

# Abbott & Roche
CBS_Ab %>%
  .[!is.na(.$abbott_n) & !is.na(.$roche_n), ] %>%
  with(summary(sampledate))

# Roche only
CBS_Ab %>%
  .[is.na(.$abbott_n) & !is.na(.$roche_n), ] %>%
  with(summary(sampledate))
```
## Making result table for ggplot
```{r}
# prepare melt data
tb1<-CBS_Ab%>%
  group_by(month_year) %>%
  summarise(n = n(), )

tb2<-CBS_Ab[CBS_Ab$interp_roche_n==1,] %>%
  group_by(month_year) %>%
  summarise(n.pos = n(), )

tb1<-merge(tb1, tb2, by = "month_year", all.x = T, all.y = F)
tb1$prop.n<-tb1$n.pos/tb1$n
tb1$month_year<-as.yearmon(tb1$month_year)
tb1<-tb1[order(tb1$month_year),]

# Add R-G adjusted prevalances
# Sen=99.8%, Spec = 99.5%
# Source:https://www.rochecanada.com/media/covid-19--an-additional-contribution-to-the-testing-landscape-in
tb1$prop.n_RG_h<- (tb1$prop.n+0.995-1)/(0.995+0.998-1)

# Third-party sens & spec
# Sen=90.6, spec = 100 
# Source: https://meridian.allenpress.com/aplm/article/145/1/32/443501/Comparative-Clinical-Evaluation-of-the-Roche
tb1$prop.n_RG_l<- (tb1$prop.n+1-1)/(1+0.906-1)
```

#### Apply ggplot, plot two together
```{r}
CBS_melt<-melt(select(tb1, month_year, prop.n, prop.n_RG_h, prop.n_RG_l), id.vars = "month_year")

CBS_melt %>%
  ggplot(aes(x=month_year, y=value, colour=variable)) +
  geom_point(aes(colour=variable)) +
  geom_line(aes(colour=variable))  +
  geom_hline(yintercept=0, colour='red', linetype=2) +
  theme_bw() + 
  scale_fill_discrete(name = "", labels = c("Unadjusted", "RG_adjusted: Roche", "RG_adjusted: Other study")) +
  theme(legend.title = element_blank())
```

#### Plot CBS and APL all together
```{r}
# Create a merged data
APL_melt$data<-rep("APL", length(APL_melt$Col_Month))
colnames(APL_melt)<-c("month_year", "variable", "value" , "data")

CBS_melt$data<-rep("CBS", length(CBS_melt$month_year))
# keep only RG_I (sens=0.9)
CBS_melt1<-CBS_melt %>% .[.$variable!="prop.n_RG_h",]
CBS_melt1$variable<-as.character(CBS_melt1$variable)
CBS_melt1$variable[CBS_melt1$variable!="prop.n"]<-'prop.n_RG'

APLBS<-rbind.data.frame(APL_melt, CBS_melt1)

# Do ggplot
p01<-APLBS %>%
  ggplot(aes(x=month_year, y=value, colour=variable)) +
  geom_point(aes(colour=variable, shape=data)) +
  geom_line(aes(colour=variable, linetype=data))  +
  geom_hline(yintercept=0, colour='red', linetype=2) +
  xlab("Time") + ylab("Estimated Prevalence") +
  theme_bw()

p01$labels$colour<-"Adjustment"

p01
```

## What's the reason

### 2023-12-14 2_scripts/2.2 Rogen-Gladen Adj.qmd
Wrap up the RG adj for anti-N, update manuscript

@@
 
-## What's the reason
+## What's the reason of the gap? 
+```{r}
+# The gap between adjusted and unadjused - Rogen-Gladen equation is the cause
+# In our case, the sensitivity is where most variance come from.
+Specificity is good over time, and we can used the number published by the manufacturer (Roche & Abbott).
+Since sens is imperfect (0.8--1.0) and specificity is mostly 100% trustworthy (~1.0), the R-G equation and be simplified as:
+
+#     True Prevalence = Observed Test Positivity/sens
+# According to the equation above, the worse the sens (smaller in value), the larger the gap between the adjusted (blue) and unadjusted (red). The gap is larger among unadjusted lines (red).
+
+# Between data (APL, CBS, CANPATH) diff is not the major focus of this study, Yuan's paper will focus on those. For this paper, the major focus should be the difference between adjusting methods
+```

### 2023-12-14 5_manuscript/manuscript.qmd
Wrap up the RG adj for anti-N, update manuscript

@@
 
+The gap between adjusted and unadjused - Rogen-Gladen equation is the cause
+
+In our case, the sensitivity is where most variance come from.
+Specificity is good over time, and we can used the number published by the manufacturer (Roche & Abbott).
+Since sens is imperfect (0.8--1.0) and specificity is mostly 100% trustworthy (\~1.0), the R-G equation and be simplified as:
+
+#### **True Prevalence = Observed Test Positivity/sens**
+
+According to the equation above, the worse the sens (smaller in value), the larger the gap between the adjusted (blue) and unadjusted (red). The gap is larger among unadjusted lines (red).
+
+Between data (APL, CBS, CANPATH) diff is not the major focus of this study, Yuan's paper will focus on those. For this paper, the major focus should be the difference between adjusting methods
+
 <br>

### 2023-12-18 2_scripts/2.2 Rogen-Gladen Adj.qmd
add anti-S figures

@@
 library(reshape2)      # for melt() function
-library(ggplot)        # ggplot
+library(ggplot2)        # ggplot
 library(zoo)           # year-month format
@@
 ```
+
+##==================== Part III. Anti-S Positivity Overtime ================##
+
+#### This time we will use anti-S only from the CBS data, since among-data diff is not the major focus of this project
+
+```{r}
+# Roche_s: like other Roche assay, COI = 0.4, those < 0.4 = negative and recorded as 0.4
+
+# Histogram 
+hist(CBS_Ab$roche_s[CBS_Ab$roche_s>0.4]) # extreme values
+summary(CBS_Ab$roche_s) # This time they have 10,0000 as max, much larger than CLSA's 2500
+
+# Histogram log-transformed
+hist(CBS_Ab$log_roche_s[CBS_Ab$roche_s>0.4]) # Aman did log_e transform
+summary(CBS_Ab$log_roche_s)
+
+# Binary
+table(CBS_Ab$interp_roche_s, useNA = 'ifany')
+# Majority (87.50%) were test positive
+```
+#### Calculate Anti-S Positivity
+```{r}
+# prepare melt data
+tb1<-CBS_Ab%>%
+  group_by(month_year) %>%
+  summarise(n = n(), )
+
+tb2<-CBS_Ab[CBS_Ab$interp_roche_s==1,] %>%
+  group_by(month_year) %>%
+  summarise(s.pos = n(), )
+
+tb1<-merge(tb1, tb2, by = "month_year", all.x = T, all.y = F)
+tb1$prop.s<-tb1$s.pos/tb1$n
+tb1$month_year<-as.yearmon(tb1$month_year)
+tb1<-tb1[order(tb1$month_year),]
+
+# Third-party sens & spec
+# Sen=97.95, spec = 99.95 
+# Source:https://www.sciencedirect.com/science/article/pii/S016609342100210X
+tb1$prop.s_RG_l<- (tb1$prop.s+1-0.9995)/(1+0.9795-0.9995)
+
+CBS_melt2<-melt(select(tb1, month_year, prop.s, prop.s_RG_l), id.vars = "month_year")
+
+p02<-CBS_melt2 %>%
+  ggplot(aes(x=month_year, y=value, colour=variable)) +
+  geom_point(aes(colour=variable)) +
+  geom_line(aes(colour=variable))  +
+  geom_hline(yintercept=0, colour='red', linetype=2) +
+  xlab("Time") + ylab("Estimated Prevalence") +
+  theme_bw()
+
+p02$labels$colour<-"Adjustment"
+p02
+```
+#### Add additional COI 0.8 and 15
+
+```{r}
+CBS_Ab$interp_roche_s0.8<-if_else(CBS_Ab$roche_s>0.8, 1, 0)
+
+# elevating COI from 0.4 to 0.8 only increase N(-) by 2
+
+CBS_Ab$interp_roche_s15<-if_else(CBS_Ab$roche_s>15, 1, 0)
+table(CBS_Ab$interp_roche_s15, useNA = 'ifany')
+
+# elevating COI from 0.4 to 15 increase N(-) by 2000
+```
+#### Plot 03:  0.4 and 15 COI together in one single plot
+```{r}
+# Creating the N_prop data
+tb1<-CBS_Ab%>%
+  group_by(month_year) %>%
+  summarise(n = n(), )
+
+tb2<-CBS_Ab[CBS_Ab$interp_roche_s==1,] %>%
+  group_by(month_year) %>%
+  summarise(s.pos = n(), )
+
+tb3<-CBS_Ab[CBS_Ab$interp_roche_s15==1,] %>%
+  group_by(month_year) %>%
+  summarise(s.pos15 = n(), )
+
+tb1<-merge(tb1, tb2, by = "month_year", all.x = T, all.y = F)
+tb1<-merge(tb1, tb3, by = "month_year", all.x = T, all.y = F)
+
+tb1$prop.s<-tb1$s.pos/tb1$n
+tb1$prop.s15<-tb1$s.pos15/tb1$n
+tb1$month_year<-as.yearmon(tb1$month_year)
+tb1<-tb1[order(tb1$month_year),]
+```
+
+```{r}
+# Melt the data
+CBS_melt3<-melt(select(tb1, month_year, prop.s, prop.s15), id.vars = "month_year")
+
+p03<-CBS_melt3 %>%
+  ggplot(aes(x=month_year, y=value, colour=variable)) +
+  geom_point(aes(colour=variable)) +
+  geom_line(aes(colour=variable))  +
+  geom_hline(yintercept=0, colour='red', linetype=2) +
+  xlab("Time") + ylab("Estimated Prevalence") +
+  theme_bw()
+
+p03$labels$colour<-"COI"
+p03
+```
+#### Add RG-adjustment to the plot, using the sens and spec related to detection of virus neutrolization, rather than detection of presence of Antibodies.
+```{r}
+# N and proportions
+tb1<-CBS_Ab%>%
+  group_by(month_year) %>%
+  summarise(n = n(), )
+
+tb2<-CBS_Ab[CBS_Ab$interp_roche_s0.8==1,] %>%
+  group_by(month_year) %>%
+  summarise(s.pos0.8 = n(), )
+
+tb3<-CBS_Ab[CBS_Ab$interp_roche_s15==1,] %>%
+  group_by(month_year) %>%
+  summarise(s.pos15 = n(), )
+
+tb1<-merge(tb1, tb2, by = "month_year", all.x = T, all.y = F)
+tb1<-merge(tb1, tb3, by = "month_year", all.x = T, all.y = F)
+
+tb1$prop.s0.8<-tb1$s.pos0.8/tb1$n
+tb1$prop.s15<-tb1$s.pos15/tb1$n
+tb1$month_year<-as.yearmon(tb1$month_year)
+tb1<-tb1[order(tb1$month_year),]
+```
+
+```{r}
+# Add R-G adjusted props
+# Source: https://diagnostics.roche.com/global/en/products/params/elecsys-anti-sars-cov-2-s.html
+# Sample size is small, find a larger one?
+
+tb1$prop.s0.8_RG<- (tb1$prop.s0.8+0.875-1)/(0.875+0.994-1)
+tb1$prop.s15_RG<- (tb1$prop.s15+1-1)/(1+0.925-1)
+
+# melt the data
+CBS_melt4<-melt(select(tb1, month_year, prop.s0.8, prop.s15,
+                       prop.s0.8_RG, prop.s15_RG), id.vars = "month_year")
+CBS_melt4$adj[CBS_melt4$variable %in% c("prop.s0.8", "prop.s15")]<-"no adjustment"
+CBS_melt4$adj[CBS_melt4$variable %in% c("prop.s0.8_RG", "prop.s15_RG")]<-"RG adjusted"
+table(CBS_melt4$variable, CBS_melt4$adj)
+
+CBS_melt4$COI[CBS_melt4$variable %in% c("prop.s0.8", "prop.s0.8_RG")]<-"0.8"
+CBS_melt4$COI[CBS_melt4$variable %in% c("prop.s15", "prop.s15_RG")]<-"15"
+table(CBS_melt4$variable, CBS_melt4$COI)
+```
+```{r}
+## ggplot using melted data
+p04<-CBS_melt4 %>%
+  ggplot(aes(x=month_year, y=value, colour=adj)) +
+  geom_point(aes(colour=adj, shape=COI)) +
+  geom_line(aes(colour=adj, linetype=COI))  +
+  geom_hline(yintercept=0, colour='red', linetype=2) +
+  xlab("Time") + ylab("Estimated Prevalence") +
+  theme_bw()
+
+p04$labels$colour<-"Adjustment"
+
+p04
+
+# plot COI = 0.8/15 seperately
+```

### 2023-12-19 5_manuscript/manuscript.qmd
Update the methods section

@@
 ```{r}
+# Load a sample dataset
 df_samp <- read_excel("../1_data/tables.xlsx", sheet = "sample")
@@
 
--   AB Antenatal Data has not arrived yet.
+-   APL assay data before the Omicron wave was mostly missing, Carmen said they have already approved the data access but the data is still on its way.
 
 -   In AB Outpatient Lab Data, they changed the assay (Diasorin S --\> Abbott S) in the midway, this would affect the validity of analysis based on quantitative test results, as the results from different assay are not directly comparable.
+Hopefully this can be clarified in the updated version of the dataset.
 
@@
 
-To estimate the SARS-CoV-2 seroprevalence and attack rate in Alberta, we used data from the Alberta Outpatient Lab (APL), Alberta Antenatal Testing, and Canadian Blood Services (CBS) datasets.
-CBS has tested \>450,000 blood donations for both SARS-CoV-2 anti-S and anti-N levels since March 2020.
-Data include donor's sex, self-reported race/ethnicity, and age.
+To estimate the SARS-CoV-2 seroprevalence and attack rate in Alberta, we used data from the Alberta Outpatient Lab (APL), Canadian Blood Services (CBS) and Canadian Partnership for Tomorrow's Health (CANPATH).
+Alberta Health Services (AHS) tested \>160,000 APL blood draws for SARS-CoV-2 anti-N and anti-S antibodies from June 2020 until December 2021.
+CBS has tested \>450,000 blood donations for both antibodies levels since March 2020.
+CANPATH has provided data with sample size \> XX.
+Besides of assay results, all three datasets include demographic variables such as donor's sex, self-reported race/ethnicity, and age.
 From their address, neighborhood material/social deprivation and urban/rural indicator have been derived.
-This analysis uses donations tested in Alberta (N=XX) only.
 
-Alberta Health Services (AHS) tested \>160,000 Albertan outpatient labs (APL) and \>6,000 antenatal blood draws for anti-N and anti-S antibodies from June 2020 until December 2021.
-These data contain all the same variables as the blood donors except for race/ethnicity.
-Besides of serology test results and demographic data, CBS has linked the administrative data for the Alberta cohort (**is the linkage finished?**), which provided additional information on their hospitalizations, clinic visits, diagnoses, COVID vaccinations, and long-term disability claims.
-Similarly, AHS has completed a linkage in the Albertan outpatient lab (APL) data and Alberta Antenatal Blood Test data.
+Besides of serology test results and demographic data, CBS has linked the administrative data for the Alberta cohort, which provided additional information on their hospitalizations, clinic visits, diagnoses, COVID vaccinations, and long-term disability claims.
+Similarly, AHS and CANPATH have completed linkages and included additional vaccination history.
+
 We did not conduct analysis using a merged data, but instead, we conducted analysis separately for each data and presented the results in a unified graph (**Fig**.
 Y).
-**Table.X** below sumarrized the differences among three data utilized in the study.
+**Table.X** below summarized the differences among three data utilized in the study.
 For more detail, see **Table.
@@
 We presented both seroprevalence and attack rate in a unified graph (**Fig**.
-X), displaying changes over time on a four-month scale.
+X), displaying changes over time on a monthly scale.
 
@@
 
-In evaluating whether participants possess immunity levels sufficient to shield them from future infections, from either vaccination or infection, we applied an age- and sex-adjusted threshold of protection to the quantitative Anti-S assay results.
-The threshold is cited from a valid source \[source of threshold\], indicating sufficient immunity against future SARS-CoV-2 infection.
-We also presented the fluctuation of the proportion sufficiently protected over time in another figure (**Fig.** Z), and we conducted sensitivity analyses in which we compare alternative threshold levels.
+In evaluating whether participants possess humoral immunity levels sufficient to shield them from future infections, from either vaccination or infection, we applied a threshold of protection to the quantitative Anti-S assay results.
+The threshold is cited from the package insert of each assays \[Roche and Abbott\], indicating sufficient antibody level to neutralize over 50% virus, with certain times of dilution (1:20).
+We also presented the fluctuation of the proportion sufficiently protected over time in another figure (**Fig.** Z), and we conducted sensitivity analyses in which we compared alternative threshold levels.
 The details of the project flowchart is presented in the **Fig**.
@@
 \[Insert the flowchart here\]
-
-The gap between adjusted and unadjused - Rogen-Gladen equation is the cause
-
-In our case, the sensitivity is where most variance come from.
-Specificity is good over time, and we can used the number published by the manufacturer (Roche & Abbott).
-Since sens is imperfect (0.8--1.0) and specificity is mostly 100% trustworthy (\~1.0), the R-G equation and be simplified as:
-
-#### **True Prevalence = Observed Test Positivity/sens**
-
-According to the equation above, the worse the sens (smaller in value), the larger the gap between the adjusted (blue) and unadjusted (red). The gap is larger among unadjusted lines (red).
-
-Between data (APL, CBS, CANPATH) diff is not the major focus of this study, Yuan's paper will focus on those. For this paper, the major focus should be the difference between adjusting methods
 
@@
 
-A
+The gap between adjusted and unadjused - Rogen-Gladen equation is the cause
+
+In our case, the sensitivity is where most variance come from.
+Specificity is good over time, and we can used the number published by the manufacturer (Roche & Abbott).
+Since sens is imperfect (0.8--1.0) and specificity is mostly 100% trustworthy (\~1.0), the R-G equation and be simplified as:
+
+#### **True Prevalence = Observed Test Positivity/sens**
+
+According to the equation above, the worse the sens (smaller in value), the larger the gap between the adjusted (blue) and unadjusted (red). The gap is larger among unadjusted lines (red).
+
+Between data (APL, CBS, CANPATH) diff is not the major focus of this study, Yuan's paper will focus on those. For this paper, the major focus should be the difference between adjusting methods
 

### 2024-01-08 5_manuscript/manuscript.qmd
add more details in methods

@@
 df_samp <- read_excel("../1_data/tables.xlsx", sheet = "sample")
+df_dataset <- read_excel("../1_data/tables.xlsx", sheet = "dataset")
 ```
@@
 
+(beware of the period of data collection for each dataset)
+
 To estimate the SARS-CoV-2 seroprevalence and attack rate in Alberta, we used data from the Alberta Outpatient Lab (APL), Canadian Blood Services (CBS) and Canadian Partnership for Tomorrow's Health (CANPATH).
@@
 Then, to bridge the gap between test positivity and population-level seroprevalence, we employed the Rogen-Gladen (RG) equation and Bayesian models.
-Following seroprevalence estimation, we further estimated the attack rate while adjusting for waning immunity.
-These estimations are dessignated to answer what proportion of the Alberta population that has been infected by SARS-CoV-2.
+Following seroprevalence estimation, we further estimated the attack rate while adjusting for waning immunity using Bayesian models.
+These estimations are dessignated to answer what proportion of the target population that has been infected by SARS-CoV-2.
 We presented both seroprevalence and attack rate in a unified graph (**Fig**.
 X), displaying changes over time on a monthly scale.
+
+Rogan Gladen equation:
+
+$$
+Prevalence = \frac{TestPositivity + (Spec-1)}{Spec + (Sens-1))}
+$$
 
@@
 ```
+
+Table 2 Caption Here
+```{r}
+t_dataset<-as_flextable(df_dataset)
+t_dataset <- fontsize(t_dataset, size = 10, part = "all")
+t_dataset <- font(t_dataset, fontname = "Times", part = "all")
+t_dataset <- theme_box(t_dataset)
+t_dataset <- bg(t_dataset, bg = "#EAEAEA", part = "header")
+t_dataset <- width(t_dataset, 1, 1.7)
+t_dataset <- bg(t_dataset, i = c(1, 4), bg = "#DDDDDD", part = "body")
+t_dataset <- bold(t_dataset, i = c(1, 4), part = "body")
+t_dataset <- align(t_dataset, align = "left", part = "all")
+
+t_dataset
+```
+
 :::

### 2024-01-11 5_manuscript/manuscript.qmd
updated methods section, tabs & figs

--- 2023-12-19 5_manuscript/manuscript.qmd
@@
 ```
-Jiacheng Chen^1,2^, ..., W. Alton Russell^1,2^
+Jiacheng Chen^1,2^, Yuan Yu^1^,..., W. Alton Russell^1,2^
 
@@
 theme_set(theme_bw())
-
 ```
 
-```{r}
+```{r, include=FALSE}
 # Load a sample dataset
 df_samp <- read_excel("../1_data/tables.xlsx", sheet = "sample")
+
+# This table to summarize three dataset we used
+df_dataset <- read_excel("../1_data/tables.xlsx", sheet = "dataset")
 ```
@@
 
-Citation [@Langham2018a]. See @fig-sample, @cor-sample, @thm-sample, and @lem-sample.
+Citation [@Langham2018a]. See @fig-sample, @cor-sample, @thm-sample, @thm-data, and @lem-sample.
 
@@
 To estimate the SARS-CoV-2 seroprevalence and attack rate in Alberta, we used data from the Alberta Outpatient Lab (APL), Canadian Blood Services (CBS) and Canadian Partnership for Tomorrow's Health (CANPATH).
-Alberta Health Services (AHS) tested \>160,000 APL blood draws for SARS-CoV-2 anti-N and anti-S antibodies from June 2020 until December 2021.
+Alberta Health Services (AHS) tested \>160,000 APL blood draws for SARS-CoV-2 anti-N and anti-S antibodies from June 2020 until October 2022.
 CBS has tested \>450,000 blood donations for both antibodies levels since March 2020.
@@
 
-Besides of serology test results and demographic data, CBS has linked the administrative data for the Alberta cohort, which provided additional information on their hospitalizations, clinic visits, diagnoses, COVID vaccinations, and long-term disability claims.
+Besides of serology test results and demographic data, CBS has linked the administrative data for the cohort, which provided additional information on their hospitalizations, clinic visits, diagnoses, COVID vaccinations, and long-term disability claims.
 Similarly, AHS and CANPATH have completed linkages and included additional vaccination history.
 
-We did not conduct analysis using a merged data, but instead, we conducted analysis separately for each data and presented the results in a unified graph (**Fig**.
-Y).
-**Table.X** below summarized the differences among three data utilized in the study.
-For more detail, see **Table.
-Z** in the result section.
+The three datasets feature different data collection time frames.
+The CBS dataset has the most extended period, ranging from December 2020 to February 2023.
+APL's data collection spans from June 2020 to October 2022, but it's important to note that anti-N assay results are only available starting from December 2021.
+Meanwhile, the CANPATH dataset is accessible for a shorter duration, from September 2020 to May 2021.
 
-**Table.X** Summary of difference among serosurveillance datasets from CBS and AHS
+We did not conduct analysis using a merged data, but instead, we conducted analysis separately for each data and presented the results in a graph panel (@fig-seroprevalence).
+@thm-data below summarized the differences among three data utilized in the study.
 
-\[insert a table here\]
+@thm-data CBS, CANPATH and APL
+
+\[insert @thm-data here if necessary\]
 
@@
 Then, to bridge the gap between test positivity and population-level seroprevalence, we employed the Rogen-Gladen (RG) equation and Bayesian models.
-Following seroprevalence estimation, we further estimated the attack rate while adjusting for waning immunity.
-These estimations are dessignated to answer what proportion of the Alberta population that has been infected by SARS-CoV-2.
-We presented both seroprevalence and attack rate in a unified graph (**Fig**.
-X), displaying changes over time on a monthly scale.
+Following seroprevalence estimation, we further estimated the attack rate while adjusting for waning immunity using Bayesian models.
+These estimations are dessignated to answer what proportion of the target population that has been infected by SARS-CoV-2.
+We presented both seroprevalence and attack rate in a unified graph (@fig-seroprevalence), displaying changes over time on a monthly scale.
+
+Rogan Gladen equation:
+
+$$
+Prevalence = \frac{TestPositivity + (Spec-1)}{Spec + (Sens-1))}
+$$
+Bayesian models adjusting for sensitivity:
+
+$$
+
+$$
+
+Bayesian models adjusting for waning immunity:
+$$
+
+$$
 
@@
 The threshold is cited from the package insert of each assays \[Roche and Abbott\], indicating sufficient antibody level to neutralize over 50% virus, with certain times of dilution (1:20).
-We also presented the fluctuation of the proportion sufficiently protected over time in another figure (**Fig.** Z), and we conducted sensitivity analyses in which we compared alternative threshold levels.
-The details of the project flowchart is presented in the **Fig**.
-Y below.
+We also presented the fluctuation of the proportion sufficiently protected over time in another figure (@fig-immunity), and we conducted sensitivity analyses in which we compared alternative threshold levels.
+The details of the sensitivity analysis is presented in the **Fig**.
+Y (haven't scratched this one yet).
 
-Fig.Y Project Flowchart
-
-\[Insert the flowchart here\]
+\[Insert @fig-immunity here if necessary\]
 
@@
 # Results
-
-**Table.Z** Difference among serosurveillance datasets from CBS and AHS
 
@@
 
-The gap between adjusted and unadjused - Rogen-Gladen equation is the cause
+Notes: The gap between Rogen-Gladen adjusted sero-prevalence and unadjused sero-positivity.
 
@@
 
-#### **True Prevalence = Observed Test Positivity/sens**
+$$
+Prevalence = \frac{TestPositivity}{Sens}
+$$
 
@@
 ::: {#thm-sample}
-Table caption here.
+This is a sample table
 
@@
 
+::: {#thm-data}
+The table below displays the distribution of age groups and gender across all three datasets.
+It also details the provinces where the data was collected, along with the corresponding data collection periods.
+
+```{r}
+t_dataset<-as_flextable(df_dataset)
+t_dataset <- fontsize(t_dataset, size = 10, part = "all")
+t_dataset <- font(t_dataset, fontname = "Times", part = "all")
+
+t_dataset <- bold(t_dataset, i = 1, part = "header")
+t_dataset <- bold(t_dataset, i = c(2, 6, 9), part = "body")
+
+#t_dataset <- theme_box(t_dataset)
+#t_dataset <- bg(t_dataset, bg = "#EAEAEA", part = "header")
+#t_dataset <- width(t_dataset, 1, 1.7)
+#t_dataset <- bg(t_dataset, i = c(1, 4), bg = "#DDDDDD", part = "body")
+#t_dataset <- bold(t_dataset, i = c(1, 4), part = "body")
+#t_dataset <- align(t_dataset, align = "left", part = "all")
+
+t_dataset
+```
+:::
+
 {{< pagebreak >}}
@@
 knitr::include_graphics("../4_output/figs/figure.png")
+```
 
+```{r}
+#| label: fig-seroprevalence
+#| fig-cap: "The figure panel above presents the temporal changes in seroprevalence and attack rate, grouped by province and data source.
+Each figure displays the unadjusted seropositivity, the seroprevalence adjusted using the Rogan-Gladen equation, the seroprevalence adjusted by Bayesian models, and the attack rates.
+The calculations of seropositivity, seroprevalence, and attack rate are based on the qualitative results of the anti-N assay.
+#| (Just a scratch, not the real fig)"
+
+knitr::include_graphics("../4_output/figs/Figure.1 Seroprelance and attack rate by province and data source.png")
+```
+
+```{r}
+#| label: fig-immunity
+#| fig-cap: "The figure panel above depicts the temporal fluctuations in quantitative anti-S assay results, grouped by province and data source.
+Two cutoff indices (COIs) are utilized for different public health interpretations.
+The COI of 0.8, recommended by the manufacturer (Roche), is optimal for identifying the presence of antibodies in the blood sample.
+Meanwhile, a COI of 15 is indicative of a threshold for sufficient humoral immunity against the SARS-CoV-2 virus.
+We also applied Rogen-Gladen equation and Bayesian models to adjust for test sensitivity.
+#| (Just one fig, will add more to build a panel)"
+
+knitr::include_graphics("../4_output/figs/Figure.2 Immunity.png")
 ```

### 2024-01-19 5_manuscript/manuscript.qmd
Minor updates on methods, add topics to discussion

@@
 -   APL assay data before the Omicron wave was mostly missing, Carmen said they have already approved the data access but the data is still on its way.
+Updates Jan-16-2024: Carmen said more data will be available soon.
 
@@
 Hopefully this can be clarified in the updated version of the dataset.
+Our currently analysis assume the Anti-S used in omicron wave are all Abbott's, based on the paper published by Carmen in the post-vax stage: https://doi.org/10.1080/23744235.2022.2080250
 
@@
 
-Besides of serology test results and demographic data, CBS has linked the administrative data for the cohort, which provided additional information on their hospitalizations, clinic visits, diagnoses, COVID vaccinations, and long-term disability claims.
-Similarly, AHS and CANPATH have completed linkages and included additional vaccination history.
+**Notes**: *The paragraph below describes briefly about the time frame of data collection in those datasets.
+CBS is still actively updating (last ver downloaded on Nov 30th) and APL is also actively updating, we are currently waiting for additional linked variables from Carmen.*
 
@@
 
-\[insert @thm-data here if necessary\]
+[insert @thm-data here if necessary]
 
@@
 Prevalence = \frac{TestPositivity + (Spec-1)}{Spec + (Sens-1))}
-$$
-Bayesian models adjusting for sensitivity:
+$$ Bayesian models adjusting for sensitivity:
 
-$$
+\$\$
 
-$$
+\$\$
 
-Bayesian models adjusting for waning immunity:
-$$
+Bayesian models adjusting for waning immunity: \$\$
 
-$$
+\$\$
 
-## **Proportion sufficiently protected:**
+## **NT50 and corresponding antibody level:**
 
@@
 
-\[Insert @fig-immunity here if necessary\]
+[Insert @fig-immunity here if necessary]
 
@@
 In our case, the sensitivity is where most variance come from.
-Specificity is good over time, and we can used the number published by the manufacturer (Roche & Abbott).
-Since sens is imperfect (0.8--1.0) and specificity is mostly 100% trustworthy (\~1.0), the R-G equation and be simplified as:
+Specificity is good over time, and we shall not only take the numbers published by the manufacturer (Roche & Abbott) into consideration.
+Since sens is imperfect (0.8--1.0) and specificity is mostly 100% trustworthy (\~1.0), the R-G equation can be simplified as:
 
@@
 
-Between data (APL, CBS, CANPATH) diff is not the major focus of this study, Yuan's paper will focus on those. For this paper, the major focus should be the difference between adjusting methods
+**Notes**: *Variances among data (APL, CBS, CANPATH) is not the major focus of this study, Yuan's paper will focus on those by evaluating the representativeness of the serosurveillance data.
+For this paper, the major focus should be the difference between adjusting methods.
+The potential readers of this papers includes those who are using serosurveillance data to monitor the progress of Covid pandemic, or other pathogens of interests which possess similar features as SARS-CoV-2.*
+
+The take-home messages would be:
+
+1.
+Without any adjustment, the public health inference on test positivity is very limited.
+It only estimates the proportion of test positive results among blood samples collected in the serosurveillance study.
+2.
+Adjusting for certain factors will change the public health inference, meanwhile it will also improve its generalizability.
+When the imperfect nature of the seroassays is taken into account, the adjusted proportion of positive test results (seroprevalence) will be able to represent the target population in our target population, from which our sample has been collected.
+Upon the seroprevalence, after adjusting for seroreversion using Bayesian models, we get attack rates of SARS-CoV-2, which estimates the proportion of population infected by the virus since the beginning of pandemic.
+3.
+There's gap among all those different measurements, and also they have different public health inference.
+Future researchers should be cautious when making references to those different measurements.
 
@@
 
+## Supplement Methods
+
+Besides of serology test results and demographic data, CBS has linked the administrative data for the cohort, which provided additional information on their hospitalizations, clinic visits, diagnoses, COVID vaccinations, and long-term disability claims.
+Similarly, AHS and CANPATH have completed linkages and included additional vaccination history.
+
+**Notes**: *Current ver of APL has completed very limited data linkage, which only includes PCR test results/time and vaccination time.
+The paragraph briefly describe the data linkage for three datasets, however for now we do not need those variables from linked administrative data.*
+
 {{< pagebreak >}}

### 2024-01-22 2_scripts/2.3 Prob Model for PCR pos.qmd
Add a new script about using anti-N assay results to predic PCR+ in the past

---
title: "2.3 Prob Model for PCR Positivity Estimation"
format: html
editor: visual
---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

The aim of this qmd file is providing an alternative way to monitor the severity of the outbreak. Besides of cumulative incidence in a certain time interval (like 3 months) we build a probabilistic model to estimate the prob of having a positive PCR test results in last three months. Those two are similar but not exactly the same.

By calculating monthly average of the prob(PCR+ in past 3 months) we got an estimation in each month. By linking all those dots together we got a curve, suggesting the changing of severity of the pandemic overtime.

## Running Code

## Load packages
```{r}
library(dplyr)         # Data management
library(ggplot2)       # ggplot
library(zoo)           # year-month format
library(lmtest)        # for LR test, comparing models
#library(reshape2)      # for melt() function
```

## Load dataset
```{r}
# The CBS Ab subset data
load(file = "../1_data/private/CBS_Ab.RData")
load(file = "../1_data/private/RFD4682e.RData")
```

##============================ Part I. Probablistic Model ====================##

## Create a small subset for PCR+ data
```{r}
APL_PCR<-RFD4682_e %>%
  .[is.na(.$`N-IgG_RESULT`)==F & .$NAAT_STATUS %in% c("Negative","Positive"),]
APL_PCR$NAAT[APL_PCR$NAAT_STATUS=="Positive"]<-1
APL_PCR$NAAT[APL_PCR$NAAT_STATUS=="Negative"]<-0
APL_PCR$NAAT<-as.numeric(APL_PCR$NAAT)
table(APL_PCR$NAAT_STATUS, APL_PCR$NAAT, useNA = 'ifany')

# A subset for those who got results within 90 days
APL_PCR90<-APL_PCR %>%
  .[.$DAYS_SINCE_NAAT>=0 & .$DAYS_SINCE_NAAT<=90,]
```

## Build regression model using the data
```{r}
# Model 1: add age and sex as covariates, 90 days only
plm01<-glm(NAAT ~ AGE_AT_COLLECTION + GENDER  + N, data = APL_PCR90,
           family = binomial(link = "logit"))
summary(plm01)

# Model 2: add age and sex as covariates, everyone
plm02<-glm(NAAT ~ AGE_AT_COLLECTION + GENDER  + N, data = APL_PCR,
           family = binomial(link = "logit"))
summary(plm02)

# Model 3, add NAAT time into the model and compare it with model 2
plm03<-glm(NAAT ~ AGE_AT_COLLECTION + GENDER  + N + DAYS_SINCE_NAAT, data = APL_PCR,
           family = binomial(link = "logit"))
summary(plm03)

lrtest(plm02, plm03) # Sig difference, much lower AIC
```
Based on model 2 we can also add time since last PCR test into analysis, and it will improve AIC greatly for model fitting. However, this model is not applicable for other datasets as the PCR test info is missing.

```{r}
# Add predicted prob back to the data
APL_PCR90$NAAT_p<-predict(plm01, type = 'response')
hist(APL_PCR90$NAAT)

sum(APL_PCR90$NAAT)    # N=1008
sum(APL_PCR90$NAAT_p)  # N=1008
```

##====================== Part II. Applying the model to APL data =============##

Apply the APL_90 model to everyone in the dataset, then plot the difference btw those who has PCR test results (observed) and those who do not (predicted).

```{r}

```

### 2024-01-24 2_scripts/2.3 Prob Model for PCR pos.qmd
Finish the model to predict 3-m PCR+

@@
 
+Compared to model 1, model 2 is predicting  "ever predicted" rather than "infected in past three months" as it does in model 1. The only one we can apply to the APL data is the model 1.
+
+#### 2.1 Prepare for the dataset
 ```{r}
+#those tested within last 90 days
+summary(APL_PCR90$DAYS_SINCE_NAAT) # Model 1 is built upon this dataset
 
+# Add an indicator for those who had PCR test in past 90 days
+RFD4682_e$PCR90<-rep(0, length(RFD4682_e$order_ID))
+RFD4682_e$PCR90[RFD4682_e$order_ID %in% APL_PCR90$order_ID]<-1
+table(RFD4682_e$PCR90)
+
+# Clean the gender variable (level unknown)
+table(RFD4682_e$GENDER)
+APL_Clean<-RFD4682_e
+APL_Clean$GENDER[APL_Clean$GENDER=="Unknown"]<-NA
+
+# Recode NAAT as 0/1/9
+APL_Clean$NAAT[APL_Clean$NAAT_STATUS=="Positive"]<-1
+APL_Clean$NAAT[APL_Clean$NAAT_STATUS=="Negative"]<-0
+APL_Clean$NAAT[APL_Clean$NAAT_STATUS=="No Prior NAT"]<-9
+APL_Clean$NAAT<-as.numeric(APL_Clean$NAAT)
+table(APL_Clean$NAAT_STATUS, APL_Clean$NAAT, useNA = 'ifany')
+
+# Apply model 1 to APL data (full)
+APL_Clean$NAAT_p<-predict(plm01, newdata = APL_Clean, type = 'response')
+
+# Create NAAT_m, using real NAAT results for those tested <=90 days
+APL_Clean$NAAT_m <- APL_Clean$NAAT_p
+APL_Clean$NAAT_m[APL_Clean$PCR90==1 & APL_Clean$NAAT_STATUS=="Positive"]<-1
+APL_Clean$NAAT_m[APL_Clean$PCR90==1 & APL_Clean$NAAT_STATUS=="Negative"]<-0
+
+summary(APL_Clean$NAAT_m)
+table(APL_Clean$NAAT_m==1) # number matches
+table(APL_Clean$NAAT_m==0)
 ```
+
+#### 2.2 Based on new data, making plot seperately
+
+```{r}
+# Calculate positivity for each month
+
+# tb1-the data (PCR<90) used to build model
+tb1<-APL_Clean[APL_Clean$PCR90==1,] %>%
+  group_by(Col_Month) %>%
+  summarise(n = n() , pos= mean(NAAT_m))
+tb1$PCR90<-rep(1, length(tb1$Col_Month))
+
+# tb2-we applied the model to the rest of APL (PCR>90) 
+tb2<-APL_Clean[APL_Clean$PCR90==0,] %>%
+  group_by(Col_Month) %>%
+  summarise(n = n() , pos= mean(NAAT_m, na.rm=TRUE))
+tb2$PCR90<-rep(0, length(tb2$Col_Month))
+
+# Merge the data to take a look
+tb3<-merge(tb1, tb2, by = "Col_Month", all.x = T, all.y = T)
+tb3$Col_Month<-as.yearmon(tb3$Col_Month)
+tb3<-tb3[order(tb3$Col_Month),]
+
+# Compare the difference
+# Paired t-test
+t.test(tb3$pos.x[18:27], tb3$pos.y[18:27], paired=1)
+# p-value=0.2761, no sig diff between observed (<90D) and the rest
+```
+#### The plots
+```{r}
+APL_Pmelt<-rbind.data.frame(select(tb1, Col_Month, pos, PCR90), 
+                            select(tb2[18:27,], Col_Month, pos, PCR90))
+APL_Pmelt$PCR90[APL_Pmelt$PCR90==0]<-"Predicted"
+APL_Pmelt$PCR90[APL_Pmelt$PCR90==1]<-"Observed"
+
+APL_Pmelt %>%
+  ggplot(aes(x=Col_Month, y=pos, group=PCR90, colour=PCR90)) +
+  geom_point() +
+  geom_line()  +
+  geom_hline(yintercept=0, colour='blue', linetype=2) +
+  theme_bw() 
+```

### 2024-02-07 2_scripts/2.3 Prob Model for PCR pos.qmd
Data cleaning and import from previous seroassay studies, mutinomial log reg model building

@@
 Besides of cumulative incidence in a certain time interval (like 3 months) we build a probabilistic model to estimate the prob of having a positive PCR test results in last three months.
-Those two are similar but not exactly the same.
+Those two are similar and we can use prob(inf.
+in past 3-month) to estimate the 3-month rolling incidence.
 
@@
 #library(reshape2)     # for melt() function
+library(readxl)        # for reading .xlsx data
+library(nnet)          # for multinomial log regression
 ```
@@
 load(file = "../1_data/private/RFD4682e.RData")
+```
+
+## Dataset from previous studies for model building
+```{r}
+## Data-01:France Leon, for negative ref, use non-reactive samples before week 69 (Oct 2020)
+library(readxl)
+X1_pone <- read_excel("../1_data/private/Prob Model/1-pone.xlsx", 
+    sheet = "ST1_Dataset", col_types = c("numeric", 
+        "text", "numeric", "numeric", "numeric", 
+        "numeric", "text", "text", "text", 
+        "text", "text", "text", "text", "text", 
+        "text", "text", "text", "text", "text", 
+        "text", "text", "text", "text", "text", 
+        "text", "text", "text"))
+dta_01<-select(X1_pone, 1:7, 27)
+dta_01s<-dta_01[dta_01$Result=="NonReac" &
+                dta_01$`Donation type`!="Convalescent plasma" &
+                dta_01$`Donation Week`<69,]
+
+# keep only assay, age, sex
+dta_01s<-select(dta_01s, 1,5,6,7)
+colnames(dta_01s)<-c("roche_n", "pid", "age", "gender")
+tb1<-as.data.frame(table(dta_01s$pid))
+colnames(tb1)<-c("pid","freq")
+tb1$sample_wt<-1/tb1$freq
+dta_01s<-merge(dta_01s, select(tb1, 1,3), by="pid",all.x = T, all.y = F)
+
+## add interp var and study ID
+dta_01s$study<-rep(1, length(dta_01s$pid))
+dta_01s$time_pso<-rep("never_inf", length(dta_01s$pid))
+rm(tb1, tb2)
+```
+
+```{r}
+## Data-03, pre-Covid negative ref only, from Jamaica (Carribean)
+X3_mmc2 <- read_excel("../1_data/private/Prob Model/3-mmc2.xlsx", 
+    range = "A2:M124", na = "NT")
+dta_03<-select(X3_mmc2, 2,6)
+dta_03$pid<-c(1:length(dta_03$`Elecsys Index Value`))
+dta_03$study<-rep(3, length(dta_03$pid))
+dta_03$time_pso<-rep("never_inf", length(dta_03$pid))
+
+dta_03s<-dta_03[!(is.na(dta_03$`Elecsys Index Value`) &
+                  is.na(dta_03$`Architect IgG Index Value`)),]
+
+# add weights
+dta_03s$sample_wt<-rep(1, length(dta_03s$pid))
+
+# correct var names
+colnames(dta_03s)<-c("roche_n", "abbott_n",
+                     colnames(dta_03s[,3:6]))
+```
+
+```{r}
+## Data-04, 2/8 months data PSO, extracted from Y. Nagakama et al.
+X4_Nakagama_all <- read_excel("../1_data/private/Prob Model/4-Nakagama_all.xlsx")
+dta_04<-X4_Nakagama_all 
+dta_04$time_pso[dta_04$time_pso=="2 months"]<-"less_3_month"
+dta_04$time_pso[dta_04$time_pso=="8 months"]<-"more_3_month"
+
+# All participants been sampled twice, wt=0.5
+dta_04$sample_wt<-rep(0.5, length(dta_04$pid))
+
+# correct col names
+colnames(dta_04)<-c("study", colnames(dta_04[,2:6]))
+```
+
+```{r}
+## Data-06, shared by Dr Paul Bieniasz
+X6_Paul <- read_excel("../1_data/private/Prob Model/6-2024_02_06_SR1407_McGill.xlsx", 
+    sheet = "data_import", skip = 1)
+dta_06<-select(X6_Paul, 1:6, 9)
+colnames(dta_06)<-c("age", "gender", "pid", "time",
+                    "abbott_n", "roche_n", "study")
+dta_06s<-dta_06[!(is.na(dta_06$abbott_n) &
+                  is.na(dta_06$roche_n)),]
+
+# add time_pso based on time
+dta_06s$time_pso[dta_06s$time<=90]<-"less_3_month"
+dta_06s$time_pso[dta_06s$time> 90]<-"more_3_month"
+table(dta_06s$time_pso, useNA = 'ifany')
+
+# add weights
+tb1<-as.data.frame(table(dta_06s$pid))
+colnames(tb1)<-c("pid","freq")
+tb1$sample_wt<-1/tb1$freq
+dta_06s<-merge(dta_06s, select(tb1, 1,3), by="pid",all.x = T, all.y = F)
+rm(tb1, tb2)
+
+dta_06s<-select(dta_06s, -c("time"))
+```
+
+#### bind all those four data by rows
+```{r}
+dta_all<-dplyr::bind_rows(dta_01s, dta_03s, dta_04, dta_06s)
 ```
@@
 sum(APL_PCR90$NAAT_p)  # N=1008
+```
+
+## Multinomial Regression Model:
+```{r}
+mnlm01<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ roche_n, data = dta_all)
+summary(mnlm01)
+
+(summary(mnlm01))$coeffcients/summary(mnlm01)$standard.errors
+
+# Model 02 takes sample weights into account
+mnlm02<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ roche_n, 
+                 data = dta_all, weights = sample_wt)
+summary(mnlm02)
+
+(summary(mnlm01))$coeffcients/summary(mnlm01)$standard.errors
 ```

### 2024-02-08 2_scripts/2.3 Prob Model for PCR pos.qmd
Updated codes for figures

@@
 library(ggplot2)       # ggplot
+library(ggpubr)     #ggplot2 QQ plot for normality check
 library(zoo)           # year-month format
@@
 ```{r}
-# The CBS Ab subset data
-load(file = "../1_data/private/CBS_Ab.RData")
+# The CBS full data
+load(file = "../1_data/private/CBS.RData")
+table(CBS$province, useNA = 'ifany')
+
 load(file = "../1_data/private/RFD4682e.RData")
@@
 dta_all<-dplyr::bind_rows(dta_01s, dta_03s, dta_04, dta_06s)
+
+# add log quant assay
+dta_all$log_roche_n<-log(dta_all$roche_n)    # COI for log_n = 0
+dta_all$log_abbott_n<-log(dta_all$abbott_n)  # COI for log)n = 0.3365 (COI = 1.4)
+
+# save the rbinded harmonized data
+save(dta_all, file="../1_data/private/Prob Model/dta_all.RData")
 ```
@@
 summary(mnlm02)
+```
+Notes on sample weights: About the target population estimates we have in the study, it is "the proportion of being infected in past 3 months" at a certain time point (each month).
+So, we can regard it as a point prevalence of the target population.
+In this case, samples sampled at different months should be unweighted as they contributed independently to each month's point prevalence estimates.
 
-(summary(mnlm01))$coeffcients/summary(mnlm01)$standard.errors
+As for the models we have been working on recently, it's a little bit different.
+The data we used are longitudinal but we used it in a cross-sectional manner: we pooled the data from all time point together then used quant assay data predicting the probability of being infected in the past three months.
+In this case, we should apply weights.
+Otherwise, individuals who were sampled multiple times would disproportionately influence the model parameter estimates, so they need to be downweighted.”
+
+```{r}
+# model 03: using log-transformed assay
+
+# check the distribution of roche_n (<COI)
+ggdensity(dta_all$roche_n[dta_all$roche_n<1])     # mostly normal, but skewed
+ggqqplot(dta_all$roche_n[dta_all$roche_n<1])      # same above
+shapiro.test(dta_all$roche_n[dta_all$roche_n<1])  # not normal
+
+# check the distribution of log_e roche_n (<COI)
+ggdensity(dta_all$log_roche_n[dta_all$log_roche_n<0])     # mostly normal, but skewed
+ggqqplot(dta_all$log_roche_n[dta_all$log_roche_n<0])      # same above
+shapiro.test(dta_all$log_roche_n[dta_all$log_roche_n<0])  # not normal
+
+# what if log-10 transformed
+shapiro.test(log10(dta_all$roche_n[dta_all$roche_n<1]))
+
+# check the positive cases roche-n
+ggdensity(dta_all$roche_n[dta_all$roche_n>1])     # mostly normal, but skewed
+ggqqplot(dta_all$roche_n[dta_all$roche_n>1])      # same above
+shapiro.test(dta_all$roche_n[dta_all$roche_n>1])  # not normal
+
+# log-e transformed, >COI
+ggdensity(dta_all$log_roche_n[dta_all$log_roche_n>0])     # mostly normal, but skewed
+ggqqplot(dta_all$log_roche_n[dta_all$log_roche_n>0])      # same above
+shapiro.test(dta_all$log_roche_n[dta_all$log_roche_n>0])  # not normal
+```
+```{r}
+# Model 02 takes sample weights into account
+mnlm03<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ log_roche_n, 
+                 data = dta_all, weights = sample_wt)
+summary(mnlm03)
 ```
@@
 Compared to model 1, model 2 is predicting  "ever predicted" rather than "infected in past three months" as it does in model 1. The only one we can apply to the APL data is the model 1.
+
+need to rewrite this section
 
@@
 ```
+
+##============ Part III. Apply Model to CBS Data ================
+
+```{r}
+# Add log roche_n to the CBS data
+CBS$log_roche_n<-log(CBS$roche_n)
+
+CBS_prob_r<-as.data.frame(predict(mnlm03, newdata = CBS, "probs")) # num of >3 months = 0
+mean(CBS_prob_r[,1], na.rm = T)  # 59.97%
+mean(CBS_prob_r[,2], na.rm = T)  # 27.70%
+mean(CBS_prob_r[,3], na.rm = T)  # 12.33%
+
+CBSs<-cbind.data.frame(select(CBS, pid, sampledate, sex, age, fsa,
+                        province, roche_n, abbott_n, roche_s, log_roche_n),
+                 CBS_prob_r)
+```
+
+#### Make by Prov Plots
+```{r}
+# Recategorize by regions
+table(CBSs$province)
+
+# 12 provinces, recatgorized into five regions
+CBSs$region<-rep(NA, length(CBSs$pid))
+CBSs$region[CBSs$province %in% c("NL", "PE", "NS", "NB")]<-"Atlantic"
+CBSs$region[CBSs$province %in% c("QC", "ON")]<-"Central"
+CBSs$region[CBSs$province %in% c("AB", "MB", "SK")]<-"Prairie"
+CBSs$region[CBSs$province %in% c("BC")]<-"BC"
+CBSs$region[CBSs$province %in% c("NU/NT", "YT")]<-"North"
+
+table(CBSs$province, CBSs$region, useNA = 'ifany')
+
+# Creat sampling months
+CBSs$samplemonth<-as.yearmon(CBSs$sampledate)
+table(CBSs$samplemonth, useNA = 'ifany')
+
+# Creat a ever inf indicator
+CBSs$ever_inf <- 1-CBSs$never_inf
+```
+
+#### Creating melt data
+```{r}
+# create by-month est for each region
+tb_bc<-CBSs[CBSs$region=="BC",] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
+            less_3_month=mean(less_3_month, na.rm = T))
+tb_bc$region<-rep("BC", length(tb_bc$n))
+
+tb_at<-CBSs[CBSs$region=="Atlantic",] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
+            less_3_month=mean(less_3_month, na.rm = T))
+tb_at$region<-rep("Atlantic", length(tb_at$n))
+
+tb_ce<-CBSs[CBSs$region=="Central",] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
+            less_3_month=mean(less_3_month, na.rm = T))
+tb_ce$region<-rep("Central", length(tb_ce$n))
+
+tb_pr<-CBSs[CBSs$region=="Prairie",] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
+            less_3_month=mean(less_3_month, na.rm = T))
+tb_pr$region<-rep("Prairie", length(tb_pr$n))
+
+tb_no<-CBSs[CBSs$region=="North",] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
+            less_3_month=mean(less_3_month, na.rm = T))
+tb_no$region<-rep("North", length(tb_no$n))
+
+tb_all<-rbind.data.frame(tb_bc, tb_pr, tb_ce, tb_at, tb_no)
+```
+
+#### Make plots by region
+```{r}
+# Reshape meltted data
+tb_melt<-reshape2::melt(select(tb_all, 1,3:5), 
+                        id.vars = c("samplemonth", "region"),
+                        variable.name = "inf",
+                        value.name = "prop")
+
+p01<-tb_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=inf, colour=inf)) +
+  geom_point(aes(colour=inf)) +
+#  geom_line(aes(colour=inf))  +
+  geom_smooth(method = "loess", aes(colour=inf)) +
+  theme_bw()
+
+p01 + facet_wrap(~region)
+```
+## Part IV Apply model to APL data
+```{r}
+summary(RFD4682_e$N)
+# The codebook does not specify what types of anti-N assay is applied, given the number we assume they are using Abbott anti-N assay (small numbers)
+```
+
+## Part V Apply model to CANPATH data
+```{r}
+Canpath <- read.csv("~/COVID-assay-adjust-methods/1_data/private/CANPATH/DAO-543759_ResearcherDataset_Serology_Results_74503par.csv")
+Canpath_all <- read.csv("~/COVID-assay-adjust-methods/1_data/private/CANPATH/DAO-543759_ResearcherDataset_Qx_96014par_1125var.csv")
+
+table(Canpath$C1_CITF_ASSAY_ID)
+
+```

### 2024-02-13 2_scripts/2.1 APL Data Quality Check.qmd
Updated APL data cleaning code, match for quintmat & quintsoc info

@@
 ```
+
 Some additional cleaning/editing, making anti-S quantitative
+
 ```{r}
@@
 ```
+
 #### Yuan found this: anti-N is only available for the Omicron stage
+
 ```{r}
@@
 ```
+
+#### As required by Matthew, add quintmat & qunitsoc variable
+```{r}
+# load a matching scale, by 6-digit postal codes
+pcode_match <- read.delim("~/COVID-assay-adjust-methods/1_data/private/ot_ccp_postalcodecodeequivalence.txt")
+
+# load updated version of APL
+library(readxl)
+RFD4682_v2 <- read_excel("1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
+    sheet = "RFD4682_Full_Data", col_types = c("text", 
+        "numeric", "text", "numeric", "numeric", 
+        "date", "text", "text", "date", "date", 
+        "text", "numeric", "numeric", "text", 
+        "text", "text", "text", "text", "numeric", 
+        "numeric", "numeric", "numeric", 
+        "numeric", "numeric", "numeric"))
+View(RFD4682_v2)
+
+# match 6-digit postal codes to the first ver
+RFD4682_e$PAT_POSTALCODE <- RFD4682_v2$PAT_POSTALCODE
+
+# cleanup NA
+length(RFD4682_e$PAT_POSTALCODE[RFD4682_e$PAT_POSTALCODE=="N/A"]) # needs to cleanup "N/A"
+RFD4682_e$PAT_POSTALCODE[RFD4682_e$PAT_POSTALCODE=="N/A"]<-NA
+table(is.na(RFD4682_e$PAT_POSTALCODE))
+
+# does the postal code match?
+# create a truncated ver of 6-digit postal code
+RFD4682_e$PAT3<-substr(RFD4682_e$PAT_POSTALCODE, 1, 3)
+
+RFD4682_e$PAT_match<-if_else(RFD4682_e$PAT_FSA==RFD4682_e$PAT3, 1, 0)
+table(RFD4682_e$PAT_match, useNA = "ifany") # 1065 unmatch, 3276 missing
+
+RFD4682_e %>%
+  .[.$PAT_match == 0 &  !is.na(.$PAT3),] %>%
+  select(PAT_POSTALCODE, PAT3, PAT_FSA)
+# N=1756 unmatched cases btw APL ver 1 & 2. Talked to M. Knight and he would like to use the updated ver
+
+RFD4682_e<-merge(RFD4682_e, select(pcode_match, PCODE, QUINTMAT, QUINTSOC),
+      by.x = "PAT_POSTALCODE", by.y="PCODE", all.x = T, all.y = F)
+
+# a clean version of if for M.Knight
+RFD4682_e1<-select(RFD4682_e, -c("PAT3", "PAT_match", "N", "Col_Month"))
+
+```
+
 #### A quick clean of the Anti-N data
+
 ```{r}
@@
 #### Add a year-month var
+
 ```{r}
@@
 #### Save a cleaned ver of RFD4682e
+
 ```{r}
 save(RFD4682_e, file = "../1_data/private/RFD4682e.RData")
+
+save(RFD4682_e1, file = "../1_data/private/RFD4682e1.RData")
 ```

### 2024-02-21 2_scripts/2.2.a CBS Preprocessing and Fig.qmd
CBS processing and figs, APL/Canpath will be added later

---
title: "2.2.a CBS Preprocessing and Monthly Positivity Figures"
format: html
editor: visual
author: J.Chen (let me know if you run into any troubles with those codes)
---

## Quarto

This qmd file is to share the code of CBS data cleaning and monthly test positivity figures with Y.Yu, who will apply Bayesian models to adjust for waning sensitivity and specificity.

## Running Code

## Load dataset
```{r}
# The CBS Data, extracted using "CITF_CBS_desc_stats.Rmd"
load(file = "../1_data/private/CBS.RData")
# Notes: Access to CBS data got denied, ask Amen for access. Y.Yuan has confirmed on 21-Feb-2024 the ver 2023-11-30 of CBS is still the most up-to-date ver.

# APL Data - will be updated soon

# Canpath data - will be updated soon
```

## Load necessary paks
```{r}
library(dplyr)         # Data management
library(reshape2)      # for melt() function
library(ggplot2)       # ggplot
library(zoo)           # for year-month format
```

## Load dataset
```{r}
# CBS
load(file = "../1_data/private/CBS.RData")
```

##==================== Part I. Preprocessing Data ============================##

#### 1.1 CBS data
```{r}
# CBS data is clean, no need for preprocessing

# Add region variable for CBS
table(CBS$province)

# 12 provinces, re-catgorized into six regions
# region divided according to Yuan's study. In CBS data since the number of samples from QC and North prov are too small (N=13, 85+83). We decided to exclude samples from those three prov from analysis.
# Also, since APL provides AB data only, we will divide prairie prov into AB and SK+MB. This should apply to Canpath data as well.

CBS$region<-rep(NA, length(CBS$pid))
CBS$region[CBS$province %in% c("NL", "PE", "NS", "NB")]<-"Atlantic"
CBS$region[CBS$province %in% c("ON")]<-"Ontario"
CBS$region[CBS$province %in% c("AB")]<-"Alberta"
CBS$region[CBS$province %in% c("MB", "SK")]<-"SK & MB"
CBS$region[CBS$province %in% c("BC")]<-"BC"

# In this way, QC, NU/NT, and YT will be coded as 'NA'
table(CBS$province, CBS$region, useNA = 'ifany')

# Creat sampling months (need {zoo} pak)
CBS$samplemonth<-as.yearmon(CBS$sampledate)
table(CBS$samplemonth, useNA = 'ifany')
```

#### 1.2 APL data
```{r}
# APL data needs additional cleaning, will update this part soon!
```

#### 1.3 Canpath data
```{r}
# Canpath data needs additional cleaning, will update this part soon!
```

##==================== Part II. Test Positivity & RG =========================##

#### 2.1 CBS data
```{r}
# prepare melt data
tb1<-CBS%>%
  group_by(samplemonth, region, .add=TRUE) %>%
  summarise(n = n(), )

tb2<-CBS[CBS$interp_roche_n==1,] %>%
  group_by(samplemonth, region, .add=TRUE) %>%
  summarise(n.pos = n(), )

tb3<-merge(tb1, tb2, by = c("samplemonth", "region"), all.x = T, all.y = F)
tb3$prop.n<-tb3$n.pos/tb3$n
CBS_pr<-tb3[!is.na(tb3$region),]

# Add R-G adjusted prevalances
# Sen=99.8%, Spec = 99.5%
# Source:https://www.rochecanada.com/media/covid-19--an-additional-contribution-to-the-testing-landscape-in
# tb1$prop.n_RG_h<- (tb1$prop.n+0.995-1)/(0.995+0.998-1)

# Third-party sens & spec
# Sen=97.2%, spec = 99.8% 
# Source: https://assets.publishing.service.gov.uk/media/5f05887ad3bf7f2be6e02161/Evaluation__of_sensitivity_and_specificity_of_4_commercially_available_SARS-CoV-2_antibody_immunoassays.pdf
CBS_pr$prop.n_RG<- (CBS_pr$prop.n+0.998-1)/(0.998+0.972-1)

## According to Yuan's request, add sens & spec to all the obs
CBS_pr$sens<-rep(0.972, length(CBS_pr$samplemonth))
CBS_pr$spec<-rep(0.998, length(CBS_pr$samplemonth))
CBS_pr$data<-rep("CBS", length(CBS_pr$samplemonth))

## drop useless intermediate tables
rm(tb1, tb2, tb3)
```
We only applied the sens & spec from third-party studies, instead of the numbers published online by Roche.

#### 2.2 APL data
```{r}

```

#### 2.3 Canpath data
```{r}

```

#### Merge all three into one
```{r}

# save all those data
save(CBS_pr, file="../3_intermediate/CBS_pr.RData" )
```

##==================== Part III.  Figures =========================##

#### 3.1 CBS: Make plots by region
```{r}
# Reshape meltted data
CBS_melt<-reshape2::melt(select(CBS_pr, samplemonth, region,
                                       prop.n, prop.n_RG), 
                        id.vars = c("samplemonth", "region"),
                        variable.name = "Adj",
                        value.name = "prop")

p01<-CBS_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=Adj, colour=Adj)) +
  geom_point(aes(colour=Adj)) +
  geom_smooth(method = "loess", aes(colour=Adj), se=FALSE) +
#  geom_line(aes(colour=Adj))  + #non-smooth curve
  theme_bw()

p01 + facet_wrap(~region)
```

### 2024-02-21 2_scripts/2.3 Prob Model for PCR pos.qmd
CBS processing and figs, APL/Canpath will be added later

@@
 library(dplyr)         # Data management
-library(ggplot2)       # ggplot
-library(ggpubr)     #ggplot2 QQ plot for normality check
+library(ggplot2)       # ggplot2
+library(ggpubr)        # ggplot2 QQ plot for normality check
 library(zoo)           # year-month format
@@
 
-# 12 provinces, recatgorized into five regions
+# 12 provinces, recatgorized into six regions
+# region divided according to Yuan's study. In CBS data since the number of samples from QC and North prov are too small (N=13, 85+83). We decided to exclude samples from those two prov from analysis.
+# Also, since APL provides AB data only, we will divide prairie prov into AB and SKS+MN. This should apply to Canpath data as well.
+
 CBSs$region<-rep(NA, length(CBSs$pid))
 CBSs$region[CBSs$province %in% c("NL", "PE", "NS", "NB")]<-"Atlantic"
-CBSs$region[CBSs$province %in% c("QC", "ON")]<-"Central"
-CBSs$region[CBSs$province %in% c("AB", "MB", "SK")]<-"Prairie"
+CBSs$region[CBSs$province %in% c("ON")]<-"Ontario"
+CBSs$region[CBSs$province %in% c("AB")]<-"Alberta"
+CBSs$region[CBSs$province %in% c("MB", "SK")]<-"SK & MB"
 CBSs$region[CBSs$province %in% c("BC")]<-"BC"
-CBSs$region[CBSs$province %in% c("NU/NT", "YT")]<-"North"
 
+# In this way, QC, NU/NT, and YT will be coded as 'NA'
 table(CBSs$province, CBSs$region, useNA = 'ifany')
@@
 CBSs$ever_inf <- 1-CBSs$never_inf
+```
+#### Save the dataset
+```{r}
+save(CBSs, file='../1_data/private/CBSs.RData')
 ```
@@
 
-tb_ce<-CBSs[CBSs$region=="Central",] %>%
+tb_on<-CBSs[CBSs$region=="Ontario",] %>%
   group_by(samplemonth) %>%
@@
             less_3_month=mean(less_3_month, na.rm = T))
-tb_ce$region<-rep("Central", length(tb_ce$n))
+tb_on$region<-rep("Ontario", length(tb_on$n))
 
-tb_pr<-CBSs[CBSs$region=="Prairie",] %>%
+tb_ab<-CBSs[CBSs$region=="Alberta",] %>%
   group_by(samplemonth) %>%
@@
             less_3_month=mean(less_3_month, na.rm = T))
-tb_pr$region<-rep("Prairie", length(tb_pr$n))
+tb_ab$region<-rep("Alberta", length(tb_ab$n))
 
-tb_no<-CBSs[CBSs$region=="North",] %>%
+tb_skm<-CBSs[CBSs$region=="SK & MB",] %>%
   group_by(samplemonth) %>%
@@
             less_3_month=mean(less_3_month, na.rm = T))
-tb_no$region<-rep("North", length(tb_no$n))
+tb_skm$region<-rep("SK & MB", length(tb_skm$n))
 
-tb_all<-rbind.data.frame(tb_bc, tb_pr, tb_ce, tb_at, tb_no)
+tb_all<-rbind.data.frame(tb_bc, tb_ab, tb_skm, tb_on, tb_at)
+# clean NA's
+tb_all<-tb_all[!is.na(tb_all$samplemonth),]
 ```

### 2024-02-26 5_manuscript/manuscript.qmd
Update the methods section for prob models, also add bullitins for each paragraphs to be added

@@
 ```
-Jiacheng Chen^1,2^, Yuan Yu^1^,..., W. Alton Russell^1,2^
+Jiacheng Chen^1,2^, Yuan Yu^1^, Sheila O'Brien^3^, Yu Nakagama^4^ ..., W. Alton Russell^1,2^
 
@@
 ^2^COVID-19 Immunity Task Force
+
+^3^Canadian Blood Services
+
+^4^Graduate School of Medicine, Osaka City University
 
@@
 
-Citation [@Langham2018a]. See @fig-sample, @cor-sample, @thm-sample, @thm-data, and @lem-sample.
+Test of citation: [@Langham2018a]. See @fig-sample, @cor-sample, @thm-sample, @thm-data, and @lem-sample.
 
-The Covid-19 pandemic overview, global and what it looks like in Canada.
+The Covid-19 pandemic overview, global/national (in Canada) trend and how it envolved overtime in each regions of Canada. Serosurveillance programs in Canada, and its importance.
 
-Serosurveillance programs in Canada, and its importance
+Two types of antibodies to be detected by assays: anti-spike and anti-nucleocapsid. Why we treated them differently in data analysis, one for quantitative and anothor one for qualitative test results.
 
-Two types of antibodies to be detected by assays: Anti-S and Anti-N. Why we treated them differently in data analysis. Quantitative and qualitative test results.
+-   Anti-S (humoral response to infection or vaccination) and anti-N (marker of natural infection)
 
-anti-S (humoral response to infection or vaccination) and anti-N (marker of natural infection)
+-   Anti-S level elevates after infection and can be boosted by vax, while anti-N level elevates with infection (even drops) but didn't change so much with vax.
 
-Waning of immunity against SARS-COV-2 after infection and immunization, indicated by decreasing blood Ab level overtime.
+-   Anti-S level wanes slower than anti-N level after infection
 
-\[insert the figure of declining Ab with re-activation of immunization\]
+Previous studies have found an association between seropositivity/ab levels and intensity of immunity (humoral + cellular).
+Declining ab level suggesting waning of immunity against SARS-COV-2 after infection and immunization.
 
-Seropravalence and attack rate, why we need to employ RG equation or Bayesian model to estimated seropravalence, and why we also need models to adjust for waning immunity.
+Seropravalence and rolling incidence, why we need to employ RG equation or Bayesian model to estimated seropravalence, and why we also need probabilistic models to predict past infection using early 2020 and pre-Covid neg ref data.
 
@@
 
-**Notes**: Unlike the DBS vs VBS manuscript, the methods section of this project is mostly [unfinished]{.underline}.
-This section only shows the plan and the flow chart of the project in general, while the details of it needs constantly updating and polishing overtime, as the project goes.
-The potential obstacles were listed below:
+```{=html}
+<!---
 
--   APL assay data before the Omicron wave was mostly missing, Carmen said they have already approved the data access but the data is still on its way.
-Updates Jan-16-2024: Carmen said more data will be available soon.
+NOTES:
+- The methods section of this project is partly unfinished, and the details of it needs constantly updating overtime, as the project goes. The potential obstacles were listed below:
 
--   In AB Outpatient Lab Data, they changed the assay (Diasorin S --\> Abbott S) in the midway, this would affect the validity of analysis based on quantitative test results, as the results from different assay are not directly comparable.
-Hopefully this can be clarified in the updated version of the dataset.
-Our currently analysis assume the Anti-S used in omicron wave are all Abbott's, based on the paper published by Carmen in the post-vax stage: https://doi.org/10.1080/23744235.2022.2080250
+1. APL assay data before the Omicron wave was mostly missing, only Omicron wave data is available. Also we do not know what assay they used for anti-N (probably Abbott anti-N).
 
+2.
+In AB Outpatient Lab Data, they changed the assay (Diasorin S -- Abbott S) in the midway, this would affect the validity of analysis based on quantitative test results, as the results from different assay are not directly comparable.
+Hopefully this can be clarified by the APL data team.
+Our currently analysis assume the Anti-N/S assays used in omicron wave are all Abbott's, based on the paper published by Carmen in the post-vax stage: https://doi.org/10.1080/23744235.2022.2080250
+
+3.
+Canpath: At least two types of anti-N assays are utilized: Mount Sinai Hosp anti-N (a in-lab developed assay) and Abbott anti-N (a commercial assay).
+Need to figure out which prov at what time used what assay, so we can apply sens for adjustment accordingly.
+
+4.
+CBS: The data we have least problem with.
+CBS is still actively updating (last ver downloaded on Nov 30th) and Yuan has the access to the most updated ver of CBS.
+(last checked on 21-Feb-2024, CBS 2023-11-30 is still the last version)
+--->
+```
 ## Dataset
 
-To estimate the SARS-CoV-2 seroprevalence and attack rate in Alberta, we used data from the Alberta Outpatient Lab (APL), Canadian Blood Services (CBS) and Canadian Partnership for Tomorrow's Health (CANPATH).
+To estimate the SARS-CoV-2 seroprevalence and rolling incidence in Alberta, we used data from the Alberta Outpatient Lab (APL), Canadian Blood Services (CBS) and Canadian Partnership for Tomorrow's Health (CANPATH).
 Alberta Health Services (AHS) tested \>160,000 APL blood draws for SARS-CoV-2 anti-N and anti-S antibodies from June 2020 until October 2022.
@@
 From their address, neighborhood material/social deprivation and urban/rural indicator have been derived.
-
-**Notes**: *The paragraph below describes briefly about the time frame of data collection in those datasets.
-CBS is still actively updating (last ver downloaded on Nov 30th) and APL is also actively updating, we are currently waiting for additional linked variables from Carmen.*
 
@@
 
-## **Seroprevalence and attack rate**
+## **Seroprevalence and rolling incidence**
 
@@
 Then, to bridge the gap between test positivity and population-level seroprevalence, we employed the Rogen-Gladen (RG) equation and Bayesian models.
-Following seroprevalence estimation, we further estimated the attack rate while adjusting for waning immunity using Bayesian models.
-These estimations are dessignated to answer what proportion of the target population that has been infected by SARS-CoV-2.
-We presented both seroprevalence and attack rate in a unified graph (@fig-seroprevalence), displaying changes over time on a monthly scale.
+
+Following seroprevalence estimation, we further estimated monthly rolling incidence proportion of SARS-CoV-2 infection.
+We developed a multinomial logistic regression model predicting the possibility of past infection based on the anti-N assay results, adjusting for age and gender.
+Then we estimated the probability of past infection in a monthly scale using population averages.
+
+We presented both seroprevalence and rolling incidence proportion in a unified graph (@fig-seroprevalence), displaying changes over time on a monthly scale.
 
@@
 
-Bayesian models adjusting for waning immunity: \$\$
+Bayesian models adjusting for waning immunity:
+
+\$\$
 
@@
 
+## Attack rates - TBD
+
+The cumulative attack rates is no longer of interest, as the majority of the Canadian population has been infected at least once since the Omicron wave.
+As the comparison rolling incidence proportion is a measurement with more meaningful public health implications.
+
+## Mixture  Models - TBD
+
+In mixture models, we assume the dist of certain variables (in our case it is S/Co of Anti-N/S assay results) has two (or maybe three) distinct (though partially overlapping) distributions.
+Mixture models were utilized by previous studies to estimate cumulative incidence.
+
+<https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1008728>
+
+<https://www.sciencedirect.com/science/article/pii/S175543652200024X>
+
 <br>
@@
 
-ANOVA test, p-values...
+A paragraph about test sample positivity, Rogen-Gladen adjusted seroprevalence and Bayesian model adjusted seroprevalence.
+Describe the trend of changing overtime, difference and consistency among different adjusting methos, regions and datasets.
+Cite figures accordingly if necessary.
+
+A paragraph about estimated proportion of ever infected & infected in past 3 month.
+
+A paragraph about estimated proportion with sufficient humoral immunity to protect them from future possible SARS-CoV-2 infection.
+This proportion is also adjusted by Rogen-Gladen equation and Bayesian models.
 
@@
 
-Notes: The gap between Rogen-Gladen adjusted sero-prevalence and unadjused sero-positivity.
+The gap between Rogen-Gladen adjusted sero-prevalence and unadjused sero-positivity: In our case, the sensitivity is where most variance come from.
+As the proof of excellent assay performance, assay manufacturer published the sensitivity and specificity of their assay.
+However, their test performance figures derived from hospitalized patients with more severe symptoms, who were samples at the time when sero antibody level reachd its highest level.
+Due to the representativeness issue, it is not recommended only take the test performance figures published by the assays manufacturer into consideration.
 
-In our case, the sensitivity is where most variance come from.
-Specificity is good over time, and we shall not only take the numbers published by the manufacturer (Roche & Abbott) into consideration.
-Since sens is imperfect (0.8--1.0) and specificity is mostly 100% trustworthy (\~1.0), the R-G equation can be simplified as:
+Since sens is imperfect (0.85--1.0) and specificity is mostly 100% trustworthy (\~1.0), the R-G equation can be simplified as:
 
@@
 
-According to the equation above, the worse the sens (smaller in value), the larger the gap between the adjusted (blue) and unadjusted (red). The gap is larger among unadjusted lines (red).
+According to the equation above, the worse the sensitivity (smaller in value), the larger the gap between the adjusted and unadjusted.
+We found larger gaps among the Abbott assay results, as the sensitivity we used for Abbott assays are smaller, when compared to sensitivity we used for Roche assays.
 
-**Notes**: *Variances among data (APL, CBS, CANPATH) is not the major focus of this study, Yuan's paper will focus on those by evaluating the representativeness of the serosurveillance data.
+```{=html}
+<!---
+Notes: 
+Variances among data (APL, CBS, CANPATH) is not the major focus of this study, Yuan's paper will focus on those by evaluating the representativeness of the serosurveillance data.
 For this paper, the major focus should be the difference between adjusting methods.
-The potential readers of this papers includes those who are using serosurveillance data to monitor the progress of Covid pandemic, or other pathogens of interests which possess similar features as SARS-CoV-2.*
+The potential readers of this papers includes those who are using serosurveillance data to monitor the progress of Covid pandemic, or other pathogens of interests which possess similar features as SARS-CoV-2.
 
-The take-home messages would be:
+So we won't discuss the among-dataset differences on measurements in the discussion section
+--->
+```
 
+Difference/Consistence among adjusting methods, regions in Canada. Is our finding consistent with previous studies?
+
+Strength and Limitation
+
+Public health implications:
 1.
@@
 3.
-There's gap among all those different measurements, and also they have different public health inference.
+There's some differences among all those different quantitative measurements, and also they have different public health inference.
 Future researchers should be cautious when making references to those different measurements.
@@
 
-**Notes**: *Current ver of APL has completed very limited data linkage, which only includes PCR test results/time and vaccination time.
-The paragraph briefly describe the data linkage for three datasets, however for now we do not need those variables from linked administrative data.*
+```{=html}
+<!---
+Notes: Current ver of APL has completed very limited data linkage, which only includes PCR test results/time and vaccination time.
+The paragraph briefly describe the data linkage for three datasets, however for now we do not need those variables from linked administrative data.
+--->
+```
+
+A short paragraph about weighting: Our analysis is based on individual level.
+As we are calculating measurements at a monthly scale, samples sampled at different months were unweighted as they contributed independently to each month's point prevalence estimates.
+While in all of those three dataset, it's possible to have one individual being sampled multiple times within the same month, in this case we applied weights to each individual according to the number of samples they have in the same month.
 

### 2024-03-05 5_manuscript/manuscript.qmd
Incoporated all the comments received by 27-Feb-2024

@@
 
-The Covid-19 pandemic overview, global/national (in Canada) trend and how it envolved overtime in each regions of Canada. Serosurveillance programs in Canada, and its importance.
+\[Be concise in this section\]
 
-Two types of antibodies to be detected by assays: anti-spike and anti-nucleocapsid. Why we treated them differently in data analysis, one for quantitative and anothor one for qualitative test results.
+\[The Covid-19 pandemic overview\]: Brief, general intro on Covid-19. Serosurveillance programs in Canada, and its importance. Use previous serosurveillence studies as examples.
+
+\[Two types of antibodies to be detected by assays\]: Anti-spike and anti-nucleocapsid.
+Why we treated them differently in data analysis, one for quantitative and anothor one for qualitative test results.
 
@@
 
-Previous studies have found an association between seropositivity/ab levels and intensity of immunity (humoral + cellular).
+\[Ab level and immunity\]: Previous studies have found an association between seropositivity/ab levels and intensity of immunity (humoral + cellular).
 Declining ab level suggesting waning of immunity against SARS-COV-2 after infection and immunization.
 
-Seropravalence and rolling incidence, why we need to employ RG equation or Bayesian model to estimated seropravalence, and why we also need probabilistic models to predict past infection using early 2020 and pre-Covid neg ref data.
+\[Our measurement and adjustment\]: Seropravalence and rolling incidence, why we need to employ RG equation or Bayesian model to estimated seropravalence, and why we also need probabilistic models to predict past infection using early 2020 and pre-Covid neg ref data.
 
@@
 Prevalence = \frac{TestPositivity + (Spec-1)}{Spec + (Sens-1))}
-$$ Bayesian models adjusting for sensitivity:
+$$
+
+Bayesian models adjusting for sensitivity:
 
@@
 
-Bayesian models adjusting for waning immunity:
-
-\$\$
-
-\$\$
-
-## **NT50 and corresponding antibody level:**
-
-In evaluating whether participants possess humoral immunity levels sufficient to shield them from future infections, from either vaccination or infection, we applied a threshold of protection to the quantitative Anti-S assay results.
-The threshold is cited from the package insert of each assays \[Roche and Abbott\], indicating sufficient antibody level to neutralize over 50% virus, with certain times of dilution (1:20).
-We also presented the fluctuation of the proportion sufficiently protected over time in another figure (@fig-immunity), and we conducted sensitivity analyses in which we compared alternative threshold levels.
-The details of the sensitivity analysis is presented in the **Fig**.
-Y (haven't scratched this one yet).
-
-[Insert @fig-immunity here if necessary]
+\[Bayesian models adjusting for waning sensitivity\]: Equation (above), model input and output. What covariates we've adjusted for.
 
@@
 
+## **NT50 and corresponding antibody level:**
+
+In evaluating whether participants possess humoral immunity levels sufficient to shield them from future infections, from either vaccination or infection, we applied a threshold of protection to the quantitative Anti-S assay results.
+The threshold is cited from the package insert of each assays \[Roche and Abbott\], indicating sufficient antibody level to neutralize over 50% virus, with certain times of dilution (1:20).
+We also presented the fluctuation of the proportion sufficiently protected over time in another figure (@fig-immunity), and we conducted sensitivity analyses in which we compared alternative threshold levels.
+The details of the sensitivity analysis is presented in the **Fig**.
+Y (haven't scratched this one yet).
+
+[Insert @fig-immunity here if necessary]
+
 <br>
@@
 
-A paragraph about test sample positivity, Rogen-Gladen adjusted seroprevalence and Bayesian model adjusted seroprevalence.
+\[Describe fig: seroprevalence\] A paragraph about test sample positivity, Rogen-Gladen adjusted seroprevalence and Bayesian model adjusted seroprevalence.
 Describe the trend of changing overtime, difference and consistency among different adjusting methos, regions and datasets.
@@
 
-A paragraph about estimated proportion of ever infected & infected in past 3 month.
+\[Describe fig: rolling inc prop\] A paragraph about estimated proportion of ever infected & infected in past 3 month.
 
-A paragraph about estimated proportion with sufficient humoral immunity to protect them from future possible SARS-CoV-2 infection.
+\[Describe fig: NT50 and corresponding Ab level\] A paragraph about estimated proportion with sufficient humoral immunity to protect them from future possible SARS-CoV-2 infection.
 This proportion is also adjusted by Rogen-Gladen equation and Bayesian models.
@@
 As the proof of excellent assay performance, assay manufacturer published the sensitivity and specificity of their assay.
-However, their test performance figures derived from hospitalized patients with more severe symptoms, who were samples at the time when sero antibody level reachd its highest level.
+However, their test performance figures derived from hospitalized patients with more severe symptoms, who were samples at the time when sero antibody level reached its highest level.
 Due to the representativeness issue, it is not recommended only take the test performance figures published by the assays manufacturer into consideration.
@@
 ```
+\[Difference/Consistence among adjusting methods, regions in Canada\] Is our finding consistent with previous studies?\]
 
-Difference/Consistence among adjusting methods, regions in Canada. Is our finding consistent with previous studies?
+\[Strength and Limitation\]
 
-Strength and Limitation
+\[Public health implications\]
 
-Public health implications:
-1.
-Without any adjustment, the public health inference on test positivity is very limited.
+1.      Without any adjustment, the public health inference on test positivity is very limited.
 It only estimates the proportion of test positive results among blood samples collected in the serosurveillance study.
-2.
-Adjusting for certain factors will change the public health inference, meanwhile it will also improve its generalizability.
+
+2.      Adjusting for certain factors will change the public health inference, meanwhile it will also improve its generalizability.
 When the imperfect nature of the seroassays is taken into account, the adjusted proportion of positive test results (seroprevalence) will be able to represent the target population in our target population, from which our sample has been collected.
 Upon the seroprevalence, after adjusting for seroreversion using Bayesian models, we get attack rates of SARS-CoV-2, which estimates the proportion of population infected by the virus since the beginning of pandemic.
-3.
-There's some differences among all those different quantitative measurements, and also they have different public health inference.
+
+3.      There's some differences among all those different quantitative measurements, and also they have different public health inference.
 Future researchers should be cautious when making references to those different measurements.
@@
 ```
-
 A short paragraph about weighting: Our analysis is based on individual level.

### 2024-03-05 2_scripts/2.1.1 Canpath Data.qmd
Updated codes for model w age + sex

---
title: "2.1.1 Canpath Data Check"
format: html
editor: visual
---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

The aim of this qmd file is cleaning the Canpath data. Those code is unnecessary to publish with the project.

## Running Code

## Load packages
```{r}
library(dplyr)         # Data management
library(zoo)           # year-month format
```

## Load the dataset
```{r}
# The Questionnaire data
Canpath_qx <- read.csv("../1_data/private/CANPATH/DAO-543759_ResearcherDataset_Qx_96014par_1125var.csv")

# The linked administrative data
Canpath_admin <- read.csv("../1_data/private/CANPATH/DAO-543759_ResearcherDataset_Serology_Admin_25727par.csv")

# The serology data
Canpath_ser <- read.csv("../1_data/private/CANPATH/DAO-543759_ResearcherDataset_Serology_Results_74503par.csv")
```

### 2024-03-05 2_scripts/2.3 Prob Model for PCR pos.qmd
Updated codes for model w age + sex

@@
 ## Load packages
+
 ```{r}
@@
 ## Load dataset
+
 ```{r}
@@
 
+# The CBSs data, for prob model only
+load(file = "../1_data/private/CBSs.RData")
+
 load(file = "../1_data/private/RFD4682e.RData")
@@
 ## Dataset from previous studies for model building
+
 ```{r}
@@
 #### bind all those four data by rows
+
 ```{r}
@@
 ## Create a small subset for PCR+ data
+
 ```{r}
@@
 ## Build regression model using the data
+
 ```{r}
@@
 ```
+
 Based on model 2 we can also add time since last PCR test into analysis, and it will improve AIC greatly for model fitting.
@@
 ## Multinomial Regression Model:
+
 ```{r}
@@
 ```
+
 Notes on sample weights: About the target population estimates we have in the study, it is "the proportion of being infected in past 3 months" at a certain time point (each month).
@@
 In this case, we should apply weights.
-Otherwise, individuals who were sampled multiple times would disproportionately influence the model parameter estimates, so they need to be downweighted.”
+Otherwise, individuals who were sampled multiple times would disproportionately influence the model parameter estimates, so they need to be downweighted."
 
@@
 ```
+
 ```{r}
@@
 #### 2.1 Prepare for the dataset
+
 ```{r}
@@
 ```
+
 #### The plots
+
 ```{r}
@@
 ```
+
+#### 2.3 Model adjusting for age and sex
+
+The multinominal model we applied to APL before (mnlm01, 02, 03), none of those three model has adjusted for covariates.
+However with the data we have we could adjust for those two variables in Roches assays.
+
+This subsection is to compare two models, one with age & sex adjusted and another one adjusted for nothing (mnlm03).
+To make two models comparable we will only use the subset with only age&sex data available
+
+```{r}
+mnlm03.a<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ log_roche_n, 
+                 data = dta_all[!is.na(dta_all$age) & 
+                                !is.na(dta_all$gender),], 
+                 weights = sample_wt)
+summary(mnlm03.a)
+
+mnlm03.b<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ log_roche_n + 
+                     age + gender, 
+                 data = dta_all[!is.na(dta_all$age) & 
+                                !is.na(dta_all$gender),], 
+                 weights = sample_wt)
+summary(mnlm03.b)
+
+lrtest(mnlm03.a, mnlm03.b)
+```
+
+```{r}
+# what about adding age & sex seperately?
+mnlm03.b_age<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ log_roche_n + 
+                     age, 
+                 data = dta_all[!is.na(dta_all$age) & 
+                                !is.na(dta_all$gender),], 
+                 weights = sample_wt)
+summary(mnlm03.b_age)
+lrtest(mnlm03.b_age, mnlm03.b)
+
+mnlm03.b_sex<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ log_roche_n + 
+                     gender, 
+                 data = dta_all[!is.na(dta_all$age) & 
+                                !is.na(dta_all$gender),], 
+                 weights = sample_wt)
+summary(mnlm03.b_sex)
+lrtest(mnlm03.b_sex, mnlm03.b)
+```
+
+**Conclusion**: 1.Adjusting for age and gender does not improve the model fit significantly.
+2.Adjusting for gender only will have limited improvement on model fit.
+Not adjusting for age & sex is okay, it will allow us to utilize more neg ref data, as age & sex info is usually missing in those data.
 
@@
 
-#### Make by Prov Plots
+```{r}
+## Applying the model with age & sex
+CBS$gender<-CBS$sex
+CBS_prob_r2<-as.data.frame(predict(mnlm03.b, newdata = CBS, "probs")) # num of >3 months = 0
+mean(CBS_prob_r2[,1], na.rm = T)  # 59.49%
+mean(CBS_prob_r2[,2], na.rm = T)  # 32.45%
+mean(CBS_prob_r2[,3], na.rm = T)  # 8.06%
+
+CBSs2<-cbind.data.frame(select(CBS, pid, sampledate, sex, age, fsa,
+                        province, roche_n, abbott_n, roche_s, log_roche_n),
+                 CBS_prob_r2)
+```
+
+#### Make by Region Plots
+
 ```{r}
@@
 CBSs$ever_inf <- 1-CBSs$never_inf
+
+# Copy new vars to CBSs2
+CBSs2$region<-CBSs$region
+CBSs2$samplemonth<-CBSs$samplemonth
+CBSs2$ever_inf <- 1-CBSs2$never_inf
 ```
+
 #### Save the dataset
+
 ```{r}
@@
 #### Creating melt data
+
 ```{r}
@@
 
+```{r}
+# create by-month est for each region, using CBSs2 data (model w age & sex)
+tb_bc<-CBSs2[CBSs2$region=="BC",] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
+            less_3_month=mean(less_3_month, na.rm = T))
+tb_bc$region<-rep("BC", length(tb_bc$n))
+
+tb_at<-CBSs2[CBSs2$region=="Atlantic",] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
+            less_3_month=mean(less_3_month, na.rm = T))
+tb_at$region<-rep("Atlantic", length(tb_at$n))
+
+tb_on<-CBSs2[CBSs2$region=="Ontario",] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
+            less_3_month=mean(less_3_month, na.rm = T))
+tb_on$region<-rep("Ontario", length(tb_on$n))
+
+tb_ab<-CBSs2[CBSs2$region=="Alberta",] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
+            less_3_month=mean(less_3_month, na.rm = T))
+tb_ab$region<-rep("Alberta", length(tb_ab$n))
+
+tb_skm<-CBSs2[CBSs2$region=="SK & MB",] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , ever_inf= mean(ever_inf, na.rm = T), 
+            less_3_month=mean(less_3_month, na.rm = T))
+tb_skm$region<-rep("SK & MB", length(tb_skm$n))
+
+tb_all2<-rbind.data.frame(tb_bc, tb_ab, tb_skm, tb_on, tb_at)
+# clean NA's
+tb_all2<-tb_all2[!is.na(tb_all2$samplemonth),]
+```
+
 #### Make plots by region
+
 ```{r}
@@
 
+tb_melt2<-reshape2::melt(select(tb_all2, 1,3:5), 
+                        id.vars = c("samplemonth", "region"),
+                        variable.name = "inf",
+                        value.name = "prop")
+
+# 1. The figure using simple model
 p01<-tb_melt %>%
@@
   geom_point(aes(colour=inf)) +
-#  geom_line(aes(colour=inf))  +
-  geom_smooth(method = "loess", aes(colour=inf)) +
+  geom_line(aes(colour=inf))  +
   theme_bw()
@@
 p01 + facet_wrap(~region)
+
+# 2. The model using simple model + age & sex
+p02<-tb_melt2 %>%
+  ggplot(aes(x=samplemonth, y=prop, group=inf, colour=inf)) +
+  geom_point(aes(colour=inf)) +
+  geom_line(aes(colour=inf))  +
+  theme_bw()
+
+p02 + facet_wrap(~region)
 ```
+
 ## Part IV Apply model to APL data
+
 ```{r}
@@
 ## Part V Apply model to CANPATH data
+
 ```{r}

### 2024-03-07 2_scripts/2.1.1 Canpath Data.qmd
Data quality check for Canpath data

@@
 ## Load the dataset
+
 ```{r}
@@
 ```
+
+##=============== Part.I Clean, merge and create a small subset===============##
+
+#### 1.1 We need those variables from each dataset:
+
+-   The questionnaire (qx) data:
+
+```{r}
+# age, pid, gender, location (FSA)
+
+# age var:
+summary(Canpath_qx$C_SDC_AGE) # N=3559 missing, our population is old median=64
+hist(Canpath_qx$C_SDC_AGE)    # skewed normal dist, lack of young adults (18-35)
+
+# sex/gender: use sex here
+summary(Canpath_qx$C_SDC_SEX) # N=3559 missing
+table(Canpath_qx$C_SDC_SEX, useNA = 'ifany')
+# 0=male, 1=female. Much more f than m.
+
+# location - 3-digit FSA
+summary(Canpath_qx$C_ADM_FSA) # N=4569 missing, ~5%
+table(is.na(Canpath_qx$C_ADM_FSA), useNA = 'ifany')
+
+# province - not available, we can create one with FSA
+```
+
+-   The admin data: info about sampling
+
+```{r}
+# sample ID, 
+summary(Canpath_admin$C1_SAMPLE_ID)
+
+# cohort
+summary(Canpath_admin$C1_ADM_COHORT_ID)
+
+# date of collection
+summary(Canpath_admin$C1_ADM_COLLECT_DATE)
+
+# sample type
+summary(Canpath_admin$C1_SAMPLE_TYPE)
+
+# date of procession
+summary(Canpath_admin$C1_SAMPLE_PROCESS_DATE)
+```
+
+-   The serology data: sero assay results
+```{r}
+# assay ID (three Sinai Hosp in-lab, 2 Abbott)
+table(Canpath_ser$C1_CITF_ASSAY_ID, useNA = 'ifany')
+# 10010-00: Gingras anti-N
+# 10020-00: Gingras anti-S
+# 10030-00: Gingras anti-S-RBD
+# 20010-03: Abbott anti-N
+# 20012-03: Abbott anti-S-RBD
+
+# assay COI
+table(Canpath_ser$C1_SAMPLE_ANTIGEN_CUTOFF, useNA = 'ifany')
+table(Canpath_ser$C1_SAMPLE_ANTIGEN_CUTOFF, Canpath_ser$C1_CITF_ASSAY_ID)
+# one single COI was applied to each assay, excellent!
+
+# what is max value (upper dynamic range) for each assay?
+table(Canpath_ser$C1_SAMPLE_ANTIGEN_MAXIMUM, useNA = 'ifany')
+# not quite useful
+
+# what is the assay results?
+summary(Canpath_ser$C1_SAMPLE_RESULT, useNA = 'ifany')
+hist(Canpath_ser$C1_SAMPLE_RESULT[Canpath_ser$C1_SAMPLE_RESULT<5])
+
+# Interpretation of the results 1/0
+table(Canpath_ser$C1_SAMPLE_RESULTS_DESCRIPTION, useNA = 'ifany')
+```
+
+-   All for three: pid (participants ID) and sid (sample ID)
+```{r}
+# sample_ID
+
+# Sample ID
+# admin data
+summary(Canpath_admin$C1_SAMPLE_ID)
+table(is.na(Canpath_admin$C1_SAMPLE_ID), useNA = 'ifany') # N=8349 missing, be careful
+length(unique(Canpath_admin$C1_SAMPLE_ID))                # all non-missing has unique sample ID
+
+# ser data
+summary(Canpath_ser$C1_ADM_SPECIMEN_ID)
+table(is.na(Canpath_ser$C1_ADM_SPECIMEN_ID), useNA = 'ifany')  # N=13632 missing, be careful
+length(unique(Canpath_ser$C1_ADM_SPECIMEN_ID))                 # not all non-missing has unique sample ID
+
+# Research ID
+# qx data:
+summary(Canpath_qx$ResearcherID)                        # N=96014, no missing
+table(is.na(Canpath_qx$ResearcherID), useNA = 'ifany')
+length(unique(Canpath_qx$ResearcherID))                 # all 96014 rid are unique
+
+# admin data:
+summary(Canpath_admin$ResearcherID)                     # N=25727, no missing
+table(is.na(Canpath_admin$ResearcherID), useNA = 'ifany')
+length(unique(Canpath_admin$ResearcherID))              # N=25724, duplicates
+
+  ## find duplicates
+  n_admin <- data.frame(table(Canpath_admin$ResearcherID))
+  n_admin[n_admin$Freq > 1,]          # 3 duplicates, sample pp with two samples (diff sample ID)
+
+# ser data
+summary(Canpath_ser$ResearcherID)                      # N=74503, no missing
+table(is.na(Canpath_ser$ResearcherID), useNA = 'ifany')
+length(unique(Canpath_ser$ResearcherID))               # N=25975, duplicates
+```
+#### 1.2 Merge three data
+```{r}
+# select the variables we need from each subset:
+# qx subset:
+qx_s<-select(Canpath_qx, ResearcherID, C_ADM_STUDY_DATASET, C_SDC_AGE, C_SDC_SEX,
+             C_ADM_FSA)
+# apply shorter var names:
+colnames(qx_s)<-c("rid", "study_id", "age", "sex", "fsa")
+
+# admin subset
+admin_s<-select(Canpath_admin, ResearcherID, C1_SAMPLE_ID,
+                C1_ADM_COHORT_ID, C1_ADM_COLLECT_DATE, C1_SAMPLE_TYPE, 
+                C1_SAMPLE_PROCESS_DATE)
+colnames(admin_s)<-c("rid", "sid", "cohort_id", "date_col", "stype", "date_pro")
+
+# ser data, use anti-N only
+ser_s<-select(Canpath_ser[Canpath_ser$C1_CITF_ASSAY_ID %in% c("10010-00", "20010-03"),],
+              ResearcherID, C1_ADM_SPECIMEN_ID, C1_CITF_ASSAY_ID, C1_SAMPLE_DATE_ASSAY_RUN,
+              C1_SAMPLE_RESULT, C1_SAMPLE_RESULTS_DESCRIPTION)
+colnames(ser_s)<-c("rid", "sid", "assay_id", "date_asy", "result", "pos")
+
+# reshape the subset
+# ser_sw<-reshape(ser_s, idvar = "sid", timevar = "assay_id", direction = "wide")
+# reshape is unnecessary
+```
+```{r}
+# merge those small subsets
+m1<-merge(admin_s, ser_s[!is.na(ser_s$rid) & !is.na(ser_s$sid),],
+          by = c("rid", "sid"), all.x = F, all.y = T)
+m2<-merge(m1, qx_s, by='rid', all.x = T, all.y = F)
+
+ser_s %>%
+  .[!is.na(.$rid) & is.na(.$sid),]
+
+# is cohort_id same as study_id?
+table(m2$cohort_id, m2$study_id, useNA = 'ifany')
+# looks messy, use FSA to determine prov of residency then
+
+# unique participants (rid)
+summary(m2$rid)
+table(is.na(m2$rid), useNA = 'ifany')          # N=21536, no missing
+
+# unique samples (sid)
+summary(m2$sid)
+table(is.na(m2$sid), useNA = 'ifany')          # N=21536, no missing
+length(unique(m2$sid))                         # N=21349, some pp has multiple samples
+```
+#### Apply FSA-province conversion
+```{r}
+# this chunk is provided by Y.Yu
+province_fun <- function(var) {
+  fsa_f = as.character(substr(var,1,1))
+  prov =  ifelse(fsa_f == "A", "NL",
+                 ifelse(fsa_f == "B", "NS",
+                        ifelse(fsa_f == "C", "PE",
+                               ifelse(fsa_f == "E", "NB",
+                                      ifelse(fsa_f == "G" | fsa_f == "H" | fsa_f == "J", "QC",
+                                             ifelse(fsa_f == "K" | fsa_f == "L" | fsa_f == "M" | fsa_f == "N" | fsa_f == "P", "ON",
+                                                    ifelse(fsa_f == "R", "MB",
+                                                           ifelse(fsa_f == "S", "SK",
+                                                                  ifelse(fsa_f == "T", "AB",
+                                                                         ifelse(fsa_f == "V", "BC",
+                                                                                ifelse(fsa_f == "X", "NU/NT",
+                                                                                       ifelse(fsa_f == "Y", "YT", NA
+                                                                                       ))))))))))))
+
+  return(prov)}
+
+# apply the function
+m2$prov<-province_fun(m2$fsa)
+table(m2$prov, m2$study_id, useNA = 'ifany')
+
+# need to clean NA's
+table(m2$cohort_id, m2$study_id, useNA = 'ifany')
+table(m2$prov, m2$cohort_id, useNA = 'ifany')
+
+## 1) group by prov (FSA converted)
+  m2$region[m2$prov %in% c("NB", "NL", "NS", "PE")]<-"Atlantic"
+  m2$region[m2$prov %in% c("AB")]<-"Alberta"
+  m2$region[m2$prov %in% c("BC")]<-"BC"
+  m2$region[m2$prov %in% c("ON")]<-"Ontario"
+  m2$region[m2$prov %in% c("QC")]<- "QC"  # recode it as NA later
+  m2$region[m2$prov %in% c("SK")]<- "SK & MB"
+table(m2$prov, m2$region, useNA = 'ifany')
+
+## 2) clean NA, use study_id
+table(m2[is.na(m2$region),]$study_id, 
+      m2[is.na(m2$region),]$cohort_id, useNA = 'ifany')
+  m2$region[is.na(m2$region) &
+            m2$cohort_id == 1] <- "Atlantic"
+  m2$region[is.na(m2$region) &
+            m2$cohort_id == 3] <- "BC"
+  m2$region[is.na(m2$region) &
+            m2$cohort_id == 5] <- "SK & MB"
+  m2$region[is.na(m2$region) &
+            m2$cohort_id == 6] <- "Ontario"
+  m2$region[is.na(m2$region) & is.na(m2$cohort_id) &
+            m2$study_id == "ATP"] <- "Alberta"
+  m2$region[is.na(m2$region) & is.na(m2$cohort_id) &
+            m2$study_id == "OHS"] <- "Ontario"
+table(m2$region, useNA = 'ifany')
+
+## recode QC as missing
+  m2$region[m2$region=="QC"]<-NA # N=4 coded as missing (NA)
+```
+## see what is the assay used in each region
+```{r}
+m2 %>%
+  select(region, assay_id) %>%
+  group_by(region, assay_id) %>%
+  summarise(Freq=n())
+```
+## save the data
+```{r}
+Canpaths<-m2
+rm(m1, m2, n_admin, ser_sw)
+save(Canpaths, file = "../1_data/private/CANPATH/Canpaths.RData")
+```

### 2024-03-11 2_scripts/2.1.1 Canpath Data.qmd
Exp: model based on samples from traning data

@@
   summarise(Freq=n())
+
+# clean the data var
+m2$date_col<-as.Date(m2$date_col, format = "%Y-%m-%d")
+m2$date_asy<-as.Date(m2$date_asy, format = "%Y-%m-%d")
+
+# when those assay was applied?
+m2 %>%
+  group_by(region) %>%
+  summarise(Freq=n(),
+            tcol_min = min(date_col, na.rm=T),
+            tcol_max = max(date_col, na.rm=T))
 ```

### 2024-03-11 2_scripts/2.3 Prob Model for PCR pos.qmd
Exp: model based on samples from traning data

@@
 
-##============ Part III. Apply Model to CBS Data ================
+##============ Part III-1 Apply Model to CBS Data ================
 
@@
 
-## Part IV Apply model to APL data
+## Part III-2 Apply model to APL data
 
@@
 
-## Part V Apply model to CANPATH data
+## Part III-3 Apply model to CANPATH data
 
@@
 table(Canpath$C1_CITF_ASSAY_ID)
+```
 
+## Part IV Misc Analysis
+
+1.
+About the age dist in training data and CBS: The difference on age distribution might be a potential problem.
+Generally speaking, blood donors (CBS, FR-Leon data) is relatively older, while Covid-19 convalescent patients are younger.
+```{r}
+# The FR-Leon data
+summary(dta_all[dta_all$study==1,]$age) # mean=46, median = 48
+hist(dta_all[dta_all$study==1,]$age)
+
+# UK - Convalscent Pts
+summary(dta_all[dta_all$study==6,]$age) # mean=44.54, median = 47
+hist(dta_all[dta_all$study==6,]$age, breaks = 25) # two waves?
+
+# overall dist (FR-Leon + UK)
+summary(dta_all[dta_all$study %in% c(1,6),]$age) # mean=46, median = 48
+hist(dta_all[dta_all$study %in% c(1,6),]$age, breaks = 25)
+# overall it looks good
+
+# CBS
+summary(CBSs$age) # one ourlier = 168, who is this??
+hist(CBSs$age[CBSs$age<100])
+
+# CBS by anti-N pos
+hist(CBSs$age[CBSs$age<100 & 
+              CBSs$roche_n>1])
+
+# KS test: UK+FR-Leon vs CBS
+ks.test(dta_all[dta_all$study %in% c(1,6),]$age, 
+        CBSs$age[CBSs$age<100])
+# as KS test suggests, the distribution of age btw two groups is different
 ```
+```{r}
+# create a sampling prob variable, based on the age dist in CBS
+CBSs$age_group<-CBS$age_group
+
+tb1<-as.data.frame(prop.table(table(CBS$age_group)))
+colnames(tb1)<-c("age_group","age_w")
+tb1$n_sp<-round(tb1$age_w*650, digits = 0)
+
+dta_all$age_group[dta_all$age<20 & dta_all$age>0]<-"10-19 years"
+dta_all$age_group[dta_all$age<30 & dta_all$age>20]<-"20-29 years"
+dta_all$age_group[dta_all$age<40 & dta_all$age>30]<-"30-39 years"
+dta_all$age_group[dta_all$age<50 & dta_all$age>40]<-"40-49 years"
+dta_all$age_group[dta_all$age<60 & dta_all$age>50]<-"50-59 years"
+dta_all$age_group[dta_all$age<70 & dta_all$age>60]<-"60-69 years"
+dta_all$age_group[dta_all$age<80 & dta_all$age>70]<-"70-79 years"
+dta_all$age_group[dta_all$age>80]<-"80+ years"
+table(dta_all$age_group, useNA = 'ifany')
+
+dta_all<-merge(dta_all, tb1, by="age_group", all.x=T, all.y=F)
+```
+
+#### Create a sample based on the sampling prob:
+```{r}
+# a simple prob sampling:
+set.seed(2048)
+dta_sp<-slice_sample(dta_all[!is.na(dta_all$age_w),], n=650,
+             weight_by = dta_all[!is.na(dta_all$age_w),]$age_w,
+             replace = FALSE)
+
+# how does age looks like in the sample?
+hist(dta_sp$age)
+
+# Stratified sampling (Alternative):
+dta1<-dta_all[!is.na(dta_all$age_group == "10-19 years"),] %>%
+  slice_sample(n=tb1$n_sp[2], replace = FALSE)
+dta2<-dta_all[!is.na(dta_all$age_group == "20-29 years"),] %>%
+  slice_sample(n=tb1$n_sp[3], replace = FALSE)
+dta3<-dta_all[!is.na(dta_all$age_group == "30-39 years"),] %>%
+  slice_sample(n=tb1$n_sp[4], replace = FALSE)
+dta4<-dta_all[!is.na(dta_all$age_group == "40-49 years"),] %>%
+  slice_sample(n=tb1$n_sp[5], replace = FALSE)
+dta5<-dta_all[!is.na(dta_all$age_group == "50-59 years"),] %>%
+  slice_sample(n=tb1$n_sp[6], replace = FALSE)
+dta6<-dta_all[!is.na(dta_all$age_group == "60-69 years"),] %>%
+  slice_sample(n=tb1$n_sp[7], replace = FALSE)
+dta7<-dta_all[!is.na(dta_all$age_group == "70-79 years"),] %>%
+  slice_sample(n=tb1$n_sp[8], replace = FALSE)
+dta8<-dta_all[!is.na(dta_all$age_group == "80+ years"),] %>%
+  slice_sample(n=sum(tb1$n_sp[9:10]), replace = FALSE)
+dta_sp<-rbind.data.frame(dta1, dta2, dta3, dta4, dta5,
+                         dta6, dta7, dta8)
+rm(dta1, dta2, dta3, dta4, dta5, dta6, dta7, dta8)
+
+# KS test
+ks.test(dta_sp$age, CBSs$age[CBSs$age<100])
+```
+
+## Regression on sampled data:
+```{r}
+mnlm03.s<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ log_roche_n + 
+                     age + gender, 
+                 data = dta_sp,
+                 weights = sample_wt)
+summary(mnlm03.s)
+```

### 2024-03-12 2_scripts/2.1.1 Canpath Data.qmd
update Canpath cleaning code

@@
 Canpath_ser <- read.csv("../1_data/private/CANPATH/DAO-543759_ResearcherDataset_Serology_Results_74503par.csv")
+
+# The new MTP Qx subset
+Canpath_qx_MTP<-read.csv("../1_data/private/CANPATH/DAO-543759_ResearcherDataset_Qx_1114par_1127varMTP.csv")
 ```
@@
 ```{r}
-# sample_ID
+# sample_ID: only available in admin and ser data
 
@@
 
-# Research ID
+# Research ID: invididual ID
 # qx data:
@@
   n_admin <- data.frame(table(Canpath_admin$ResearcherID))
-  n_admin[n_admin$Freq > 1,]          # 3 duplicates, sample pp with two samples (diff sample ID)
+  n_admin[n_admin$Freq > 1,]          # 3 duplicates, one pp with two samples (diff sample ID), collected on diff day. Drop the first one
 
@@
 table(is.na(Canpath_ser$ResearcherID), useNA = 'ifany')
-length(unique(Canpath_ser$ResearcherID))               # N=25975, duplicates
+length(unique(Canpath_ser$ResearcherID))               # N=25975, duplicates due to same sample tested for multiple assays
 ```
@@
              C_ADM_FSA)
+
+# don't forget MTP subset
+qx_s_mtp<-select(Canpath_qx_MTP, ResearcherID, C_ADM_STUDY_ID, 
+             C_SDC_AGE, C_SDC_SEX, C_ADM_FSA,
+             C1_SDC_AGE, C1_SDC_SEX, C1_ADM_FSA)
 # apply shorter var names:
 colnames(qx_s)<-c("rid", "study_id", "age", "sex", "fsa")
+colnames(qx_s_mtp)<-c("rid", "study_id", "age", "sex", "fsa", 
+                      "age1", "sex1", "fsa1")
+
+# identify those missed from original data:
+qx_s_mtp[qx_s_mtp$rid %in% qx_s$rid,] # not missed, N=699
+qx_mtpmiss<-qx_s_mtp[!(qx_s_mtp$rid %in% qx_s$rid),] # missed, N=445
+
+# rbind those two into one:
+qx_s<-rbind.data.frame(qx_s, 
+                       select(qx_mtpmiss, -c("age1", "sex1", "fsa1")))
+
+# clean study_id
+qx_s$study_id[qx_s$study_id=='5']<-"MTP"
+table(qx_s$study_id, useNA = 'ifany')
 
@@
                 C1_SAMPLE_PROCESS_DATE)
-colnames(admin_s)<-c("rid", "sid", "cohort_id", "date_col", "stype", "date_pro")
+colnames(admin_s)<-c("rid", "sid_admin", "cohort_id", "date_col", "stype", "date_pro")
 
@@
               C1_SAMPLE_RESULT, C1_SAMPLE_RESULTS_DESCRIPTION)
-colnames(ser_s)<-c("rid", "sid", "assay_id", "date_asy", "result", "pos")
+colnames(ser_s)<-c("rid", "sid_ser", "assay_id", "date_asy", "result", "pos")
+
+table(ser_s$rid %in% qx_mtpmiss$rid, useNA = 'ifany')
+table(admin_s$rid %in% qx_mtpmiss$rid, useNA = 'ifany')
 
@@
 ```
+#### Data quality check of those small dataset
+```{r}
+# admin_s: missing
+table(is.na(admin_s$rid), useNA = 'ifany')     # rid no missing, not all unique (3 duplicates)
+# remove duplicates: drop the earliest sample
+admin_s<-admin_s[!admin_s$sid %in% c("DB0006218", "DB0009025", "DB0009310"),]
+
+table(is.na(admin_s$sid), useNA = 'ifany')     # sid 8349 missing
+length(unique(admin_s$sid[!is.na(admin_s$sid)])) # non-missing sids all unique
+
+# ser_s (only 2 anti-N assays): missing
+table(is.na(ser_s$rid), useNA = 'ifany')      # rid no missing
+length(unique(ser_s$rid))                     # duplicates exists
+
+  n_ser <- data.frame(table(ser_s$rid))
+  n_ser[n_ser$Freq > 1,]
+# some sample has been tested multiple times by the same assay
+# none of those testing has been result in inconsistency of anti-N positivity
+
+# in ser_s, see where rid is duplicated: all used Mount Sinai Hosp Assay (anti-N)
+ser_rid_dup<-ser_s[ser_s$rid %in% n_ser[n_ser$Freq > 1,]$Var1, ]
+table(ser_s[ser_s$rid %in% n_ser[n_ser$Freq > 1,]$Var1, ]$assay_id, useNA='ifany')
+
+# need to deduplicate ser_s, only keep one of duplicated records
+ser_rid_dup$id<-c(1:404)
+ser_rid_dup %>%
+  distinct(rid, .keep_all=TRUE) # N=163 （404-163 = 241 de-duplicated)
+
+ser_s<-ser_s %>%
+  distinct(rid, .keep_all=TRUE) # N=25839, number matches
+```
+
 ```{r}
 # merge those small subsets
-m1<-merge(admin_s, ser_s[!is.na(ser_s$rid) & !is.na(ser_s$sid),],
-          by = c("rid", "sid"), all.x = F, all.y = T)
+m1<-merge(admin_s, ser_s,
+          by = c("rid"), all.x = F, all.y = T)
 m2<-merge(m1, qx_s, by='rid', all.x = T, all.y = F)
 
-ser_s %>%
-  .[!is.na(.$rid) & is.na(.$sid),]
+# sid match
+m2$sid_match[!is.na(m2$sid_admin) & !is.na(m2$sid_ser) &
+             m2$sid_admin == m2$sid_ser]<-"match"
+m2$sid_match[!is.na(m2$sid_admin) & !is.na(m2$sid_ser) &
+             m2$sid_admin != m2$sid_ser]<-"mismatch"
+m2$sid_match[!is.na(m2$sid_admin) & is.na(m2$sid_ser)]<-"ser miss"
+m2$sid_match[is.na(m2$sid_admin) & !is.na(m2$sid_ser)]<-"admin miss"
+m2$sid_match[is.na(m2$sid_admin) & is.na(m2$sid_ser)]<-"all miss"
+table(m2$sid_match, useNA = 'ifany')
+
+# work on this part later #
 
@@
 ```
+
+##===================== Part.II Estimate Seroprevalence =================
+
+**Notes:** Seroprevalence estimation requires individual analysis and a little bit weighting.
+Those who has been sampled multiple times should be down-weighted properly.
+Those who are sampled multiple times in different months will not be down-weighted.
+
+M.Knight's data cleaning strategy is much more straightforward, as he only need to locate the participants who provided sero data, so matching on researcherID makes senses.
+In our case we need to treat sero data more carefully, as in Canpath sero data, one individual could provide multiple samples, and those samples were tested by multiple assays (Abbott N/S, Mount Sinai Hosp N/S/S-RBD).
+
+```{r}
+
+```
+
+##

### 2024-03-14 2_scripts/2.1.1 Canpath Data.qmd
Finished Canpath data cleaning

@@
 ---
-title: "2.1.1 Canpath Data Check"
+title: "2.1.1 Canpath Data Cleaning"
 format: html
@@
 
-Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.
+The aim of this qmd file is cleaning the Canpath data.
 
-The aim of this qmd file is cleaning the Canpath data. Those code is unnecessary to publish with the project.
+In Canpath data we've removed three individuals who provided multiple samples, and several samples tested for multiple times.
+Those samples tested for multiple times did not result in inconsistent interpretations on assay results.
+See codes below for more details on data cleaning.
+
+In cleaned ver of Canpath data all rid (researcherID, participant ID) and sid (sample ID) are unique and 1-to-1, though some sid is missing.
 
@@
 library(dplyr)         # Data management
-library(zoo)           # year-month format
+# library(zoo)           # year-month format
 ```
@@
 ```
+
 #### 1.2 Merge three data
@@
 table(admin_s$rid %in% qx_mtpmiss$rid, useNA = 'ifany')
-
-# reshape the subset
-# ser_sw<-reshape(ser_s, idvar = "sid", timevar = "assay_id", direction = "wide")
-# reshape is unnecessary
 ```
@@
 length(unique(admin_s$sid[!is.na(admin_s$sid)])) # non-missing sids all unique
+# remove duplicates: drop the earliest sample
+ser_s<-ser_s[!ser_s$sid %in% c("DB0009310"),]
 
@@
 # need to deduplicate ser_s, only keep one of duplicated records
-ser_rid_dup$id<-c(1:404)
+ser_rid_dup$id<-c(1:402)
 ser_rid_dup %>%
-  distinct(rid, .keep_all=TRUE) # N=163 （404-163 = 241 de-duplicated)
+  distinct(rid, .keep_all=TRUE) # N=162 （402-162 = 240 de-duplicated)
 
@@
 m2$sid_match[is.na(m2$sid_admin) & is.na(m2$sid_ser)]<-"all miss"
-table(m2$sid_match, useNA = 'ifany')
+table(m2$sid_match, useNA = 'ifany') # no mismtach, good
 
-# work on this part later #
+# drop unnecessary sids
+m2$sid<-m2$sid_ser
+m2<-select(m2, -c("sid_admin", "sid_ser", "sid_match"))
 
@@
 summary(m2$rid)
-table(is.na(m2$rid), useNA = 'ifany')          # N=21536, no missing
+table(is.na(m2$rid), useNA = 'ifany')          # N=25839, no missing
+length(unique(m2$rid))                         # all unique
 
@@
 summary(m2$sid)
-table(is.na(m2$sid), useNA = 'ifany')          # N=21536, no missing
-length(unique(m2$sid))                         # N=21349, some pp has multiple samples
+table(is.na(m2$sid), useNA = 'ifany')          # N=21346, 4493 missing
+length(unique(m2$sid[!is.na(m2$sid)]))         # N=21346, no missing all unique
 ```
@@
   m2$region[m2$prov %in% c("QC")]<- "QC"  # recode it as NA later
-  m2$region[m2$prov %in% c("SK")]<- "SK & MB"
+  m2$region[m2$prov %in% c("SK", "MB")]<- "SK & MB"
 table(m2$prov, m2$region, useNA = 'ifany')
@@
 ```
-## see what is the assay used in each region
+
+#### see what is the assay used in each region
 ```{r}
@@
 
-# clean the data var
-m2$date_col<-as.Date(m2$date_col, format = "%Y-%m-%d")
-m2$date_asy<-as.Date(m2$date_asy, format = "%Y-%m-%d")
+# clean the date var (codes provided by M.Knight)
 
-# when those assay was applied?
+## The collection date var
+m2$date_col1<-gsub("/","-", m2$date_col) # replace "/" with "-"
+m2$date_col1<-as.Date(lubridate::parse_date_time(m2$date_col1,
+                                                 c("%y-%m-%d","%m-%d-%y")),tz = "UTC")
+
+## The assay running date var
+m2$date_asy[m2$date_asy=="21-07-16"]<-"2021-07-16"
+m2$date_asy1<-as.Date(m2$date_asy, format = "%Y-%m-%d")
+
+## cleaning temp var
+m2$date_asy<-m2$date_asy1
+m2$date_col<-m2$date_col1
+m2<-select(m2, -c("date_col1", "date_asy1"))
+
+## Beware some discrepancies: sample collection time is later than assay date
+## N=4
+m2[m2$date_asy < m2$date_col &
+             !is.na(m2$date_asy) & !is.na(m2$date_col),]
+
+# when those assay was collected?
 m2 %>%
@@
 ```
-## save the data
+
+#### 1.3 save the data
 ```{r}
 Canpaths<-m2
-rm(m1, m2, n_admin, ser_sw)
+rm(m1, m2, n_admin, ser_sw, n_ser, qx_mtpmiss, ser_rid_dup)
+rm(admin_s, qx_s, qx_s_mtp, ser_s)
 save(Canpaths, file = "../1_data/private/CANPATH/Canpaths.RData")
+
+# Save to the shared folder 
+save(Canpaths, file = "~/../../jmy/CANPATH/Canpaths.RData")
 ```
-
-##===================== Part.II Estimate Seroprevalence =================
-
-**Notes:** Seroprevalence estimation requires individual analysis and a little bit weighting.
-Those who has been sampled multiple times should be down-weighted properly.
-Those who are sampled multiple times in different months will not be down-weighted.
-
-M.Knight's data cleaning strategy is much more straightforward, as he only need to locate the participants who provided sero data, so matching on researcherID makes senses.
-In our case we need to treat sero data more carefully, as in Canpath sero data, one individual could provide multiple samples, and those samples were tested by multiple assays (Abbott N/S, Mount Sinai Hosp N/S/S-RBD).
-
-```{r}
-
-```
-
-##

### 2024-03-14 2_scripts/2.2 Rogen-Gladen Adj.qmd
Finished Canpath data cleaning

@@
 ## Load necc paks
+
 ```{r}
@@
 ```
-# ============= Part I. The Changing of Anti-N Pos APL ===============#
+
+# ============= Part I. The Changing of Anti-N Pos APL ===============
 
@@
 #### Apply ggplot, plot two together
+
 ```{r}
@@
 
-# ============= Part II. The Changing of Anti-N Pos Overtime ===============#
+# ============= Part II. The Changing of Anti-N Pos Overtime ===============
 
 #### Anti-N in CBS data (Roche)
+
 ```{r}
@@
 ```
+
 ## Making result table for ggplot
+
 ```{r}
@@
 #### Plot CBS and APL all together
+
 ```{r}
@@
 ## What's the reason of the gap?
+
 ```{r}
@@
 ```
+
 #### Calculate Anti-S Positivity
+
 ```{r}
@@
 ```
+
 #### Add additional COI 0.8 and 15
@@
 ```
+
 #### Plot 03: 0.4 and 15 COI together in one single plot
+
 ```{r}
@@
 ```
+
 #### Add RG-adjustment to the plot, using the sens and spec related to detection of virus neutrolization, rather than detection of presence of Antibodies.
+
 ```{r}

### 2024-03-14 2_scripts/2.2.a CBS Preprocessing and Fig.qmd
Finished Canpath data cleaning

@@
 ---
-title: "2.2.a CBS Preprocessing and Monthly Positivity Figures"
+title: "2.2.a Preprocessing and Monthly Positivity Figures"
 format: html
@@
 
+For APL
+For Canpath cleaning & pre-processing, please refer to "2.1.1 Canpath Data.qmd"
+
 ## Running Code
@@
 ## Load necessary paks
+
 ```{r}
@@
 load(file = "../1_data/private/CBS.RData")
+
+# Canpath - clean ver
+load(file = "../1_data/private/CANPATH/Canpaths.RData")
+# load from jmy foler
+# load(file = "~/../../jmy/CANPATH/Canpaths.RData")
 ```
@@
 #### 1.1 CBS data
+
 ```{r}
@@
 #### 1.2 APL data
+
 ```{r}
@@
 #### 1.3 Canpath data
-```{r}
-# Canpath data needs additional cleaning, will update this part soon!
-```
+
+**Notes:** In Canpath data we've removed three individuals who provided multiple samples, and several samples tested for multiple times.
+Those samples tested for multiple times did not result in inconsistent interpretation on assay results.
+See "2.1.1 Canpath Date.Rmd" for more details on data cleaning.
+
+In Canpath data all rid (researcherID, participant ID) and sid (sample ID) are unique and 1-to-1, though some sid is missing.
 
@@
 #### 2.1 CBS data
+
 ```{r}
@@
 ```
+
 We only applied the sens & spec from third-party studies, instead of the numbers published online by Roche.
@@
 #### 2.2 APL data
+
+**Notes:** Seroprevalence estimation requires individual analysis and a little bit weighting.
+Those who has been sampled multiple times should be down-weighted properly.
+Those who are sampled multiple times in different months will not be down-weighted.
+
 ```{r}
@@
 #### 2.3 Canpath data
+
 ```{r}
@@
 #### Merge all three into one
+
 ```{r}
@@
 #### 3.1 CBS: Make plots by region
+
 ```{r}

### 2024-03-18 2_scripts/2.1.1 Canpath Data.qmd
Update fig code for Canpath

@@
 ```
+
 #### Data quality check of those small dataset
@@
 ```
+#### 1.3 Add additional variables
+```{r}
+# add race variable: the coding on race is a bit messy in Canpath
+# 14 variables were used to indicate diff racial minorities
+# The cleaning code is provided by M.Knight
+m2<-merge(m2, select(Canpath_qx, "ResearcherID",
+          "C1_SDC_EB_ARAB","C1_SDC_EB_BLACK","C1_SDC_EB_CHINESE",
+         "C1_SDC_EB_FILIPINO","C1_SDC_EB_JAPANESE",
+         "C1_SDC_EB_KOREAN","C1_SDC_EB_LATIN","C1_SDC_EB_S_ASIAN",
+         "C1_SDC_EB_SE_ASIAN","C1_SDC_EB_W_ASIAN","C1_SDC_EB_WHITE",
+         "C1_SDC_EB_OTHER", "C1_SDC_EB_OTHER_OTSP","C1_SDC_EB_CA"),
+         by.x = 'rid', by.y ="ResearcherID", 
+         all.x = T, all.y = F
 
-#### 1.3 save the data
+)
+
+# Reorder the variables to fit Matthew's code:
+m3<-cbind.data.frame(m2[,1:8], m2[,17:30], m2[,9:16])
+m2<-m3
+rm(m3)
+
+# If possible, add a short para in the discussion section explaining why we can not categorize the indigenous people. Consult Yuan's email for more detail.
+```
+
+```{r}
+# define whites and minorities:
+
+#Racialized minorities - no indigenous responses included (applied later in for loop)
+rm_indn<-c("Afro Latino Caribbean decent","East Indian descent, born in Kenya","Indo-Caribbean","white/asian",
+           "SE Asian (Taiwanese)","i am South African, I carry in my veins European, African, Asian and indigenous Southern Afircan",
+           "West Indian from Guyana","Guyanese - West Indian","South American","caucasian - european, indian, iraqi",
+           "Ugandian of South Asian heritage","Trinidadian","White, Chinese, West Asian", "Japanese/Irish","Chinese, Indian, English",
+           "East Indian born in Guyana","West-Indian","South African of mixed races","half Japanese half White",
+           "Armenian","South Asian from East Africa","Indo Caribbean","Japanese and European","Black/white",
+           "Mixed race Chinese-Guyanese-White","East Asian","west indian",
+           "White, Indigeneous, Latin-American","Indian born in the Caribbean.",
+           "White and Khoisan","European Hispanic","Armenian mother, Irish father, but born in England",
+           "Persian","Guyanese,south asian descent","Mixed heritage (white and black)","Caribbean, Mixed Race",
+           "Assyrian","Caucasian African","WEST INDIAN/CHINESE/","Latino and European","Of Indian decent - Born in Caribbean",
+           "White/Southeast Asian","mixed european/chinese","Indo-caribbean","Mixed Race (Black, White, Hispanic,Soth Asian)",
+           "South Asian/ Mixed South african","Canadian, German & Caribbean","Black Canadian",
+           "East Indian","Mixed European & Southeast Asian","Je suis Afro Latino-AmÃ\u0083Â\u0083Ã\u0082Â©ricain","latino-amÃ\u0083Â\u0083Ã\u0082Â©ricain occidental",
+           "Nord africain","Berbere","Berbere (Afrique du nord)","Berbere Nord Africain","amazigh","maghrebine","Nord-Africain",
+           "Juif marocain","Nord africaine","90% white 10%Jewish-Arab","Canadian and Afrikaans",
+           "mixed with Jamaican", "guianese of east indian descent",
+           "caribbean chinese", "chinese and french background and other",
+           "arabe-africaine", "berbère","persian",
+           "indo-european","colombienne","japanese & swedish",
+           "east indian","middle east",
+           "armenian",
+           "north african","south american, african, chinese","middle eastern",
+           "middle east, and central europe","cherokee",
+           "irish and guyanese","african","zimbawayan",
+           "dutch indonesian",
+           "trinidadian","south america - biracial",
+           "punjabi","west indian of east indian descent",
+           "mixed race - mother part black/father white caucasian","jamaican",
+           "canadian chinese","cambodgienne","africain","asian","mauricien",
+           "taiwanese","middle east/ persion",
+           "1/4 middle eastern (syria which is now lebanon) from paternal grandfather",
+           "mixed afghan and english","tunisien",
+           "jamaican.","west  indian","zoroastrian (ancient persian ancestry)",
+           "white/west asian","kabyle (berbère) d'afrique du nord","white mexican jewish")
+
+#Racialized minorities - indigenous responses included
+rm_indy<-c("Metis/European descent","White and aboriginal","Metis","Indigenous / European","Metis French Canadian",
+           "Metis Canadian","Blanche ascendance micmak","Metis ancestry","white a bit of indigenous","métis",
+           "french metis", "30% indigenous metis/70% white","metis",
+           "culturally white, 3/16 aboriginal","mik'maw","inuit / european", "autochtone",
+           rm_indn)
+
+#White or Caucasian
+w<-c("white newfoundlander","White British",
+     "White passing","Caucasian","White - British Decent",
+     "White, British descent","Caucasian white", "White - Canadian Irish descent",
+     "white french canadian", "Caucasian female","White - Canadian","caucasian","White Canadian Scottish parents","Canadian Caucasian","White British Descent",
+     "Canadian - White - 4 generations",
+     "white, British descent","CAUCASIAN",
+     "Caucasian - European descent","Canadian white Caucasian","white Canadian" ,
+     "Canadian White","White Franco Ontarian","Caucasian Canadian",
+     "white  Canadian       North American", "White Canadian born","White Canadian, Parents of Ukrainian and French background",
+     "White Anglo-Saxon","White/Cdn born/European grandparents",
+     "Canadian, white","White caucasian","white not Europe.",
+     "Hebrew (Jewish) Canadian white","White (Canadian)","White","White from Canada",
+     "White  English decent=","White bilingual Canadian - Family from Quebec","white canadian","White  Canadian",
+     "White Canadian, no known ethnicities","White (Causcasian)","white","blanc Canadienne","Canadienne blanche",
+     "BLANCHE","Blanc.", "Blanc (Canada)","Canadienne blanc","Caucasien","Blanc - Canadien",
+     "Blanc canadienne","Blanc du canada","Blanc canadien","Blanc canadian",
+     "Blanche (Canadienne)","une femme de race blanche","blanc","Blanche","Blanc canada","Blanc","Blanche, Canadienne",
+     "blanc quebecois","caucasien",
+     "White & Canadian","White Canadian","blanc du canada","White- true blue CANADIAN","White Scottish Descent",
+     "blanche canadienne","Blanc Canadien","canadien francais blanc",
+     "white, scottish and other",
+     "ccaucasian","caucasienne",
+     "white european","white, mainly scottish/english/irish",
+     "white, of european decent with mi'kmaq ancestors several generations removed.",
+     "canadian caucasian", "white, norwegian and scottish",
+     "canadien francais, caucasien","european caucasian",
+     "white celtic")
+```
+
+```{r}
+# a for loop for creating race var:
+m2$race<-NULL
+
+for(i in 1:nrow(m2)){
+  m2$race[i]<-case_when(
+    #select pnts and nothing else
+    !is.na(m2[i,22]) & 
+      sum(is.na(m2[i,9:22])) == 13 ~ "pnts",
+
+    #select only 1 white ethnicity
+    !is.na(m2[i,19]) & 
+      sum(is.na(m2[i,9:22])) == 13 ~ "White",
+
+    #select only 1 racialized minority ethnicity
+    sum(!is.na(m2[i,9:18])) == 1 &
+      sum(is.na(m2[i,9:22])) == 13 ~ "Racialized minority",
+
+    #select 1 white ethnicity and at least 1 racialized minority ethnicity
+    sum(!is.na(m2[i,9:18])) > 0 & 
+      !is.na(m2[i,19]) &
+      sum(is.na(m2[i,22])) ~ "Racialized minority",
+
+    #select 2+ racialized minority ethnicities
+    sum(!is.na(m2[i, 9:18])) > 1 &
+      sum(is.na(m2[i, c(19, 22)])) == 2 ~ "Racialized minority",
+
+    #select at least 1 racialized minority ethnicity and at least 1 "other" ethnicity
+    sum(!is.na(m2[i, 9:18])) > 0 &
+      sum(is.na(m2[i, c(19, 22)])) == 2 &
+      sum(!is.na(m2[i, 20,21])) > 0 ~ "Racialized minority",
+
+    #select white, but also identify as a racialized minority using text box
+    sum(is.na(m2[i, c(9:18, 22)])) == 11 &
+      !is.na(m2[i,19]) &
+      sum(!is.na(m2[i, c(20:21)])) >= 1 &
+      m2[i, c(21)] %in% (rm_indn) ~ "Racialized minority",
+
+    #select white, but also identify as white using text box
+    sum(is.na(m2[i, c(9:18, 22)])) == 11 &
+      !is.na(m2[i,19]) &
+      sum(!is.na(m2[i, c(20:21)])) >= 1 &
+      m2[i, c(21)] %in% (w) ~ "White",
+
+    #select other and identify as a racialized minority using text box
+    sum(is.na(m2[i, c(9:19, 22)])) == 12 &
+      m2[i, 21] %in% (rm_indn) ~ "Racialized minority",
+
+    #select other and identify as white using text box
+    sum(is.na(m2[i, c(9:19, 22)])) == 12 &
+      m2[i, 21] %in% (w) ~ "White",
+
+    TRUE ~ NA)
+}
+
+table(m2$race, useNA = 'ifany')
+
+# pnt = prefer not to answer
+table(m2[,22], m2$race, useNA = 'ifany')
+
+# remove loose race vars
+m2<-select(m2, -c(9:22))
+```
+
+#### 1.4 save the data
 ```{r}
@@
 save(Canpaths, file = "~/../../jmy/CANPATH/Canpaths.RData")
+# Save to the shared folder as .csv format
+write.csv(Canpaths, file = "~/../../jmy/CANPATH/Canpaths.csv", row.names=FALSE)
 ```

### 2024-03-18 2_scripts/2.2.a CBS Preprocessing and Fig.qmd
Update fig code for Canpath

@@
 
+**Notes:** In Canpath data we don't have to worry about weighting, as one people has been sampled only once, then tested by multiple asays (either Gingras x3 or Abbott x2)
+
+##### 2.3.1 Pre-process Canpath data
 ```{r}
+Canpaths %>%
+  select(region, assay_id) %>%
+  group_by(region, assay_id) %>%
+  summarise(Freq=n())
 
+# remove 9 AB cases with Gingras abti-N (Mount Sinai Hosp)
+Canpaths<-Canpaths %>%
+  .[!(.$region=="Alberta" & .$assay_id=="10010-00"),]
+
+# remove 2 BC cases with Abbott anti-N
+Canpaths<-Canpaths %>%
+  .[!(.$region=="BC" & .$assay_id=="20010-03"),]
+
+# create samplemonth var
+Canpaths$samplemonth<-as.yearmon(Canpaths$date_col)
+table(Canpaths$samplemonth, useNA = 'ifany')
+
+# see samplemonth by assay
+Canpaths %>%
+  .[!is.na(.$region),] %>%
+  select(samplemonth, assay_id) %>%
+  group_by(samplemonth, assay_id) %>%
+  summarise(Freq=n())
+```
+
+```{r}
+# create sero-prevalance tab
+tb1<-Canpaths%>%
+  group_by(samplemonth, region, .add=TRUE) %>%
+  summarise(n = n(), )
+
+tb2<-Canpaths[Canpaths$pos==1,] %>%
+  group_by(samplemonth, region, .add=TRUE) %>%
+  summarise(n.pos = n(), )
+
+tb3<-merge(tb1, tb2, by = c("samplemonth", "region"), all.x = T, all.y = F)
+tb3$prop.n<-tb3$n.pos/tb3$n
+tb3$prop.n[is.na(tb3$prop.n)]<-0
+
+Canpaths_pr<-tb3[!is.na(tb3$region) &
+                 !is.na(tb3$samplemonth),]
+
+# Third-party sens & spec: Abbott-N
+# Sen=92.7%, spec = 99.9% 
+# Source: https://assets.publishing.service.gov.uk/media/5f05887ad3bf7f2be6e02161/Evaluation__of_sensitivity_and_specificity_of_4_commercially_available_SARS-CoV-2_antibody_immunoassays.pdf
+Canpaths_pr$prop.n_RG<-rep(NA, 38)
+Canpaths_pr[Canpaths_pr$region=="Alberta",]$prop.n_RG<- 
+  (Canpaths_pr[Canpaths_pr$region=="Alberta",]$prop.n+0.999-1)/(0.999+0.927-1)
+
+# Third-party sens & spec: Mount Sinai Hosp (Gingras), from T.Murphy
+# Sen=83.5%, spec = 96.4%
+Canpaths_pr[!Canpaths_pr$region=="Alberta",]$prop.n_RG<- 
+  (Canpaths_pr[!Canpaths_pr$region=="Alberta",]$prop.n+0.999-1)/(0.999+0.927-1)
+
+## According to Yuan's request, add sens & spec to all the obs
+Canpaths_pr$sens[Canpaths_pr$region=="Alberta"]<-0.927
+Canpaths_pr$sens[!Canpaths_pr$region=="Alberta"]<-0.835
+Canpaths_pr$spec[Canpaths_pr$region=="Alberta"]<-0.999
+Canpaths_pr$spec[!Canpaths_pr$region=="Alberta"]<-0.964
+
+Canpaths_pr$data<-rep("Canpaths", length(Canpaths_pr$samplemonth))
+
+## drop useless intermediate tables
+rm(tb1, tb2, tb3)
 ```
@@
   geom_point(aes(colour=Adj)) +
-  geom_smooth(method = "loess", aes(colour=Adj), se=FALSE) +
-#  geom_line(aes(colour=Adj))  + #non-smooth curve
+#  geom_smooth(method = "loess", aes(colour=Adj), se=FALSE) +
+  geom_line(aes(colour=Adj))  + #non-smooth curve
   theme_bw()
@@
 ```
+
+#### 3.2 Canpath: make plots by region
+```{r}
+# Reshape meltted data
+Canpaths_melt<-reshape2::melt(select(Canpaths_pr, samplemonth, region,
+                                       prop.n, prop.n_RG), 
+                        id.vars = c("samplemonth", "region"),
+                        variable.name = "Adj",
+                        value.name = "prop")
+
+p02<-Canpaths_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=Adj, colour=Adj)) +
+  geom_point(aes(colour=Adj)) +
+#  geom_smooth(method = "loess", aes(colour=Adj), se=FALSE) +
+  geom_line(aes(colour=Adj))  + 
+  theme_bw()+
+  scale_y_continuous(limits = c(0,1))
+
+p02 + facet_wrap(~region)
+
+```

### 2024-04-04 2_scripts/2.2.a CBS Preprocessing and Fig.qmd
clean the code, do the abbott~roche model

@@
 
-For APL
-For Canpath cleaning & pre-processing, please refer to "2.1.1 Canpath Data.qmd"
+For APL For Canpath cleaning & pre-processing, please refer to "2.1.1 Canpath Data.qmd"
 
@@
 ##### 2.3.1 Pre-process Canpath data
+
 ```{r}
@@
 #### 3.2 Canpath: make plots by region
+
 ```{r}

### 2024-04-04 2_scripts/2.3 Prob Model for PCR pos.qmd
clean the code, do the abbott~roche model

@@
 
+# The APL data
 load(file = "../1_data/private/RFD4682e.RData")
@@
 dta_01s$time_pso<-rep("never_inf", length(dta_01s$pid))
-rm(tb1, tb2)
+rm(tb1, tb2, X1_pone)
 ```
@@
                      colnames(dta_03s[,3:6]))
+rm(X3_mmc2)
 ```
@@
 ## Data-04, 2/8 months data PSO, extracted from Y. Nagakama et al.
-X4_Nakagama_all <- read_excel("../1_data/private/Prob Model/4-Nakagama_all.xlsx")
+X4_Nakagama_all <- read_excel("../1_data/private/Prob Model/4-anonymized_rocheabbott.xlsx",
+                              col_types = c("numeric", "text", "numeric", 
+                                            "numeric", "numeric", "numeric"))
+
 dta_04<-X4_Nakagama_all
-dta_04$time_pso[dta_04$time_pso=="2 months"]<-"less_3_month"
-dta_04$time_pso[dta_04$time_pso=="8 months"]<-"more_3_month"
-
-# All participants been sampled twice, wt=0.5
-dta_04$sample_wt<-rep(0.5, length(dta_04$pid))
 
 # correct col names
-colnames(dta_04)<-c("study", colnames(dta_04[,2:6]))
+colnames(dta_04)<-c("age","gender","month_pso","abbott_n","roche_n","pid")
+dta_04$pso <- dta_04$month_pso*30
+hist(dta_04$pso, breaks = seq(0,540, by=30)) # not so much samples within 3-9 months
+
+dta_04$time_pso[dta_04$pso<=90]<-"less_3_month"
+dta_04$time_pso[dta_04$pso<=180 & dta_04$pso>90]<-"3-6_month"
+dta_04$time_pso[dta_04$pso<=270 & dta_04$pso>180]<-"6-9_month"
+dta_04$time_pso[dta_04$pso>270]<-"more_9_month"
+table(dta_04$time_pso, useNA = 'ifany')
+
+# All participants been sampled only once, wt=1
+dta_04$sample_wt<-rep(1, length(dta_04$pid))
+dta_04$study<-rep(4, length(dta_04$pid))
+
+# clean the unused data
+rm(X4_Nakagama_all)
 ```
@@
 dta_06<-select(X6_Paul, 1:6, 9)
-colnames(dta_06)<-c("age", "gender", "pid", "time",
+colnames(dta_06)<-c("age", "gender", "pid", "pso",
                     "abbott_n", "roche_n", "study")
@@
 # add time_pso based on time
-dta_06s$time_pso[dta_06s$time<=90]<-"less_3_month"
-dta_06s$time_pso[dta_06s$time> 90]<-"more_3_month"
+dta_06s$time_pso[dta_06s$pso<=90]<-"less_3_month"
+dta_06s$time_pso[dta_06s$pso<=180 & dta_06s$pso>90]<-"3-6_month"
+dta_06s$time_pso[dta_06s$pso<=270 & dta_06s$pso>180]<-"6-9_month"
+dta_06s$time_pso[dta_06s$pso>270]<-"more_9_month"
 table(dta_06s$time_pso, useNA = 'ifany')
+hist(dta_06s$pso, breaks = seq(0,540, by=30)) # some data from month 3-9
 
@@
 dta_06s<-merge(dta_06s, select(tb1, 1,3), by="pid",all.x = T, all.y = F)
-rm(tb1, tb2)
-
-dta_06s<-select(dta_06s, -c("time"))
+rm(tb1, tb2, X6_Paul)
 ```
@@
 ```{r}
-dta_all<-dplyr::bind_rows(dta_01s, dta_03s, dta_04, dta_06s)
+dta_all<-dplyr::bind_rows(dta_01s, dta_03s, 
+                          select(dta_04, -c("month_pso")), dta_06s)
 
-# add log quant assay
+# add log quant assay (log_e transformation)
 dta_all$log_roche_n<-log(dta_all$roche_n)    # COI for log_n = 0
-dta_all$log_abbott_n<-log(dta_all$abbott_n)  # COI for log)n = 0.3365 (COI = 1.4)
+dta_all$log_abbott_n<-log(dta_all$abbott_n)  # COI for log_n = 0.3365 (COI = 1.4)
+
+# clean the gender var
+table(dta_all$gender)
+dta_all$gender[dta_all$gender %in% c("m")]<-"M"
+dta_all$gender[dta_all$gender %in% c("f")]<-"F"
 
@@
 save(dta_all, file="../1_data/private/Prob Model/dta_all.RData")
+
+# summary analysis
+hist(dta_all$pso, breaks = seq(0,540, by=30))
+table(dta_all$time_pso, useNA = 'ifany')
+
+# remove all intermediate data
+rm(dta_01, dta_01s, dta_03, dta_03s, dta_04, dta_06, dta_06s)
 ```
@@
 
-## Create a small subset for PCR+ data
+#### 1.1 Multinomial Regression Model:
 
 ```{r}
-APL_PCR<-RFD4682_e %>%
-  .[is.na(.$`N-IgG_RESULT`)==F & .$NAAT_STATUS %in% c("Negative","Positive"),]
-APL_PCR$NAAT[APL_PCR$NAAT_STATUS=="Positive"]<-1
-APL_PCR$NAAT[APL_PCR$NAAT_STATUS=="Negative"]<-0
-APL_PCR$NAAT<-as.numeric(APL_PCR$NAAT)
-table(APL_PCR$NAAT_STATUS, APL_PCR$NAAT, useNA = 'ifany')
-
-# A subset for those who got results within 90 days
-APL_PCR90<-APL_PCR %>%
-  .[.$DAYS_SINCE_NAAT>=0 & .$DAYS_SINCE_NAAT<=90,]
-```
-
-## Build regression model using the data
-
-```{r}
-# Model 1: add age and sex as covariates, 90 days only
-plm01<-glm(NAAT ~ AGE_AT_COLLECTION + GENDER  + N, data = APL_PCR90,
-           family = binomial(link = "logit"))
-summary(plm01)
-
-# Model 2: add age and sex as covariates, everyone
-plm02<-glm(NAAT ~ AGE_AT_COLLECTION + GENDER  + N, data = APL_PCR,
-           family = binomial(link = "logit"))
-summary(plm02)
-
-# Model 3, add NAAT time into the model and compare it with model 2
-plm03<-glm(NAAT ~ AGE_AT_COLLECTION + GENDER  + N + DAYS_SINCE_NAAT, data = APL_PCR,
-           family = binomial(link = "logit"))
-summary(plm03)
-
-lrtest(plm02, plm03) # Sig difference, much lower AIC
-```
-
-Based on model 2 we can also add time since last PCR test into analysis, and it will improve AIC greatly for model fitting.
-However, this model is not applicable for other datasets as the PCR test info is missing.
-
-```{r}
-# Add predicted prob back to the data
-APL_PCR90$NAAT_p<-predict(plm01, type = 'response')
-hist(APL_PCR90$NAAT)
-
-sum(APL_PCR90$NAAT)    # N=1008
-sum(APL_PCR90$NAAT_p)  # N=1008
-```
-
-## Multinomial Regression Model:
-
-```{r}
+# Model 01: simple model, unweighted
 mnlm01<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ roche_n, data = dta_all)
@@
 
-# Model 02 takes sample weights into account
+# Model 02: takes sample weights into account
 mnlm02<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ roche_n, 
@@
 ```{r}
-# Model 02 takes sample weights into account
+# Model 03  takes sample weights into account
 mnlm03<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ log_roche_n, 
@@
 summary(mnlm03)
+## use estimates/SE and 1.96 to calculate p-value manually, should be <0.001
 ```
 
-##====================== Part II. Applying the model to APL data =============##
+##====================== Part II. Applying the model to Sero data =============##
 
-Apply the APL_90 model to everyone in the dataset, then plot the difference btw those who has PCR test results (observed) and those who do not (predicted).
-
-Compared to model 1, model 2 is predicting "ever predicted" rather than "infected in past three months" as it does in model 1. The only one we can apply to the APL data is the model 1.
-
-need to rewrite this section
-
-#### 2.1 Prepare for the dataset
-
-```{r}
-#those tested within last 90 days
-summary(APL_PCR90$DAYS_SINCE_NAAT) # Model 1 is built upon this dataset
-
-# Add an indicator for those who had PCR test in past 90 days
-RFD4682_e$PCR90<-rep(0, length(RFD4682_e$order_ID))
-RFD4682_e$PCR90[RFD4682_e$order_ID %in% APL_PCR90$order_ID]<-1
-table(RFD4682_e$PCR90)
-
-# Clean the gender variable (level unknown)
-table(RFD4682_e$GENDER)
-APL_Clean<-RFD4682_e
-APL_Clean$GENDER[APL_Clean$GENDER=="Unknown"]<-NA
-
-# Recode NAAT as 0/1/9
-APL_Clean$NAAT[APL_Clean$NAAT_STATUS=="Positive"]<-1
-APL_Clean$NAAT[APL_Clean$NAAT_STATUS=="Negative"]<-0
-APL_Clean$NAAT[APL_Clean$NAAT_STATUS=="No Prior NAT"]<-9
-APL_Clean$NAAT<-as.numeric(APL_Clean$NAAT)
-table(APL_Clean$NAAT_STATUS, APL_Clean$NAAT, useNA = 'ifany')
-
-# Apply model 1 to APL data (full)
-APL_Clean$NAAT_p<-predict(plm01, newdata = APL_Clean, type = 'response')
-
-# Create NAAT_m, using real NAAT results for those tested <=90 days
-APL_Clean$NAAT_m <- APL_Clean$NAAT_p
-APL_Clean$NAAT_m[APL_Clean$PCR90==1 & APL_Clean$NAAT_STATUS=="Positive"]<-1
-APL_Clean$NAAT_m[APL_Clean$PCR90==1 & APL_Clean$NAAT_STATUS=="Negative"]<-0
-
-summary(APL_Clean$NAAT_m)
-table(APL_Clean$NAAT_m==1) # number matches
-table(APL_Clean$NAAT_m==0)
-```
-
-#### 2.2 Based on new data, making plot seperately
-
-```{r}
-# Calculate positivity for each month
-
-# tb1-the data (PCR<90) used to build model
-tb1<-APL_Clean[APL_Clean$PCR90==1,] %>%
-  group_by(Col_Month) %>%
-  summarise(n = n() , pos= mean(NAAT_m))
-tb1$PCR90<-rep(1, length(tb1$Col_Month))
-
-# tb2-we applied the model to the rest of APL (PCR>90) 
-tb2<-APL_Clean[APL_Clean$PCR90==0,] %>%
-  group_by(Col_Month) %>%
-  summarise(n = n() , pos= mean(NAAT_m, na.rm=TRUE))
-tb2$PCR90<-rep(0, length(tb2$Col_Month))
-
-# Merge the data to take a look
-tb3<-merge(tb1, tb2, by = "Col_Month", all.x = T, all.y = T)
-tb3$Col_Month<-as.yearmon(tb3$Col_Month)
-tb3<-tb3[order(tb3$Col_Month),]
-
-# Compare the difference
-# Paired t-test
-t.test(tb3$pos.x[18:27], tb3$pos.y[18:27], paired=1)
-# p-value=0.2761, no sig diff between observed (<90D) and the rest
-```
-
-#### The plots
-
-```{r}
-APL_Pmelt<-rbind.data.frame(select(tb1, Col_Month, pos, PCR90), 
-                            select(tb2[18:27,], Col_Month, pos, PCR90))
-APL_Pmelt$PCR90[APL_Pmelt$PCR90==0]<-"Predicted"
-APL_Pmelt$PCR90[APL_Pmelt$PCR90==1]<-"Observed"
-
-APL_Pmelt %>%
-  ggplot(aes(x=Col_Month, y=pos, group=PCR90, colour=PCR90)) +
-  geom_point() +
-  geom_line()  +
-  geom_hline(yintercept=0, colour='blue', linetype=2) +
-  theme_bw() 
-```
-
-#### 2.3 Model adjusting for age and sex
+#### 2.1 Model adjusting for age and sex
 
@@
 
-##============ Part III-1 Apply Model to CBS Data ================
+#### 2.2 Apply Model to CBS Data
 
@@
 CBS_prob_r<-as.data.frame(predict(mnlm03, newdata = CBS, "probs")) # num of >3 months = 0
-mean(CBS_prob_r[,1], na.rm = T)  # 59.97%
-mean(CBS_prob_r[,2], na.rm = T)  # 27.70%
-mean(CBS_prob_r[,3], na.rm = T)  # 12.33%
+mean(CBS_prob_r[,1], na.rm = T)  # never: 59.82%
+mean(CBS_prob_r[,4], na.rm = T)  # <3 months: 25.31%
+mean(CBS_prob_r[,2], na.rm = T)  # 3-6 month: 3.87%
+mean(CBS_prob_r[,3], na.rm = T)  # 6-9 months: 2.98%
+mean(CBS_prob_r[,5], na.rm = T)  # >9 months: 8.02%
+
+CBS_prob_r$ever_inf <- 1-CBS_prob_r$never_inf
+CBS_prob_r$less_6_month <- CBS_prob_r$less_3_month + CBS_prob_r$`3-6_month`
+CBS_prob_r$less_9_month <- CBS_prob_r$less_6_month + CBS_prob_r$`6-9_month`
 
@@
                         province, roche_n, abbott_n, roche_s, log_roche_n),
-                 CBS_prob_r)
+                 CBS_prob_r[,c(4,6,7,8)])
 ```
@@
 CBS$gender<-CBS$sex
+
 CBS_prob_r2<-as.data.frame(predict(mnlm03.b, newdata = CBS, "probs")) # num of >3 months = 0
@@
 
-# Creat a ever inf indicator
-CBSs$ever_inf <- 1-CBSs$never_inf
-
 # Copy new vars to CBSs2
@@
 CBSs2$samplemonth<-CBSs$samplemonth
-CBSs2$ever_inf <- 1-CBSs2$never_inf
 ```
@@
 
+```{r}
+# create by-month estimates, by region
+# create by-month est for each region
+tb_all<-CBSs %>%
+  group_by(samplemonth, region) %>%
+  summarise(n = n() , 
+            ever_inf= mean(ever_inf, na.rm = T), 
+            less_3_month=mean(less_3_month, na.rm = T),
+            less_6_month=mean(less_6_month, na.rm=T),
+            less_9_month=mean(less_9_month, na.rm=T),
+            )
+# clean NAs
+tb_all<-tb_all[!is.na(tb_all$region), ]
+```
+
+##===================== Part III. Exp: Reg Model Abbott N~Roche N =============##
+
+This idea came from Alton, we can use the data we have for prob model training to generate another model: Abbott-N~Roche-N.
+Then apply the model to the CBS data and generate the sero-prevalence curve, see if it gets closer to the APL curve
+
+```{r}
+# The linear model using log_e transformed abbott_n and roche_n
+
+# lm01, simple model
+lm01<-glm(log_abbott_n ~ log_roche_n, 
+          family = "gaussian",
+          data = dta_all[!is.na(dta_all$age) &
+                         !is.na(dta_all$gender),], weights = sample_wt)
+summary(lm01) # AIC=1398
+
+# lm02, add age and sex
+lm02<-glm(log_abbott_n ~ log_roche_n + age + gender, 
+          family = "gaussian",
+          data = dta_all[!is.na(dta_all$age) &
+                         !is.na(dta_all$gender),], weights = sample_wt)
+summary(lm02) # AIC=1401.1, increased a bit
+
+# lm02.a, add age only
+lm02.a<-glm(log_abbott_n ~ log_roche_n + age, 
+          family = "gaussian",
+          data = dta_all[!is.na(dta_all$age) &
+                         !is.na(dta_all$gender),], weights = sample_wt)
+summary(lm02.a) # AIC=1400, increased a bit
+
+# lm02.b, add sex only
+lm02.b<-glm(log_abbott_n ~ log_roche_n + gender, 
+          family = "gaussian",
+          data = dta_all[!is.na(dta_all$age) &
+                         !is.na(dta_all$gender),], weights = sample_wt)
+summary(lm02.b) # AIC=1399.1, increased a bit
+
+lrtest(lm02, lm01) # p=0.65, not so much diff
+
+## Let's just use simple model lm01 then
+
+# refit model 01, using those with missing age/sex info
+lm01<-glm(log_abbott_n ~ log_roche_n, 
+          family = "gaussian",
+          data = dta_all, weights = sample_wt)
+summary(lm01) 
+```
+
+#### Apply the model lm01 to the CBS data
+```{r}
+log_abbott_n<-predict(lm01, newdata = CBSs, "response")
+summary(log_abbott_n)
+
+CBSs$log_abbott_n <- log_abbott_n
+CBSs$abbott_n_hat <- exp(CBSs$log_abbott_n)
+summary(CBSs$abbott_n_hat)
+
+abbott_n<-CBSs %>%
+  select(abbott_n, abbott_n_hat) %>%
+  .[is.na(.$abbott_n)==F,]
+
+plot(abbott_n$abbott_n, abbott_n$abbott_n_hat)
+# do a paired t-test
+t.test(abbott_n$abbott_n, abbott_n$abbott_n_hat,
+       paired = TRUE)
+# paired t-test, log transformed
+abbott_n$log_abbott_n<-log(abbott_n$abbott_n)
+abbott_n$log_abbott_n_hat<-log(abbott_n$abbott_n_hat)
+t.test(abbott_n$log_abbott_n, abbott_n$log_abbott_n_hat,
+       paired = TRUE)
+
+# Apply COI=1.4
+CBSs$pos_abbott<-if_else(CBSs$abbott_n_hat<=1.4, 0, 1, missing = NA)
+# Apply Roche COI=1.0
+CBSs$pos_roche<-if_else(CBSs$roche_n<=1, 0, 1, missing = NA)
+```
+
+#### plot: sero-prevalence, using predicted Abbott positivity
+```{r}
+tb_abb<-CBSs %>%
+  group_by(samplemonth, region) %>%
+  summarise(n = n() , 
+            pos_abbott= sum(pos_abbott, na.rm = T), 
+            pos_roche= sum(pos_roche, na.rm = T)
+            )
+tb_abb$prop.abbott<-tb_abb$pos_abbott/tb_abb$n
+tb_abb$prop.roche<-tb_abb$pos_roche/tb_abb$n
+
+# clean NAs
+tb_abb<-tb_abb[!is.na(tb_abb$region), ]
+
+tb_abb_melt<-reshape2::melt(select(tb_abb, 1:2,6:7), 
+                        id.vars = c("samplemonth", "region"),
+                        variable.name = "assay",
+                        value.name = "prop")
+
+p03<-tb_abb_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=assay, colour=assay)) +
+  geom_point(aes(colour=assay)) +
+  geom_line(aes(colour=assay))  +
+  theme_bw()
+
+p03 + facet_wrap(~region)
+```
+
 #### Make plots by region
@@
 # Reshape meltted data
-tb_melt<-reshape2::melt(select(tb_all, 1,3:5), 
+tb_melt<-reshape2::melt(select(tb_all, 1:2,4:7), 
                         id.vars = c("samplemonth", "region"),
@@
 
-## Part III-2 Apply model to APL data
+#### 2.3 Apply model to APL data
 
@@
 
-## Part III-3 Apply model to CANPATH data
+#### 2.4 Apply model to CANPATH data
 

### 2024-04-05 2_scripts/2.2.a CBS Preprocessing and Fig.qmd
minor changes, need to work on plot part later

@@
 # geom_smooth(method = "loess", aes(colour=Adj), se=FALSE) +
-  geom_line(aes(colour=Adj))  + #non-smooth curve
+  geom_line(aes(colour=Adj))  + # non-smooth curve
   theme_bw()

### 2024-04-05 2_scripts/2.3 Prob Model for PCR pos.qmd
minor changes, need to work on plot part later

@@
 
+#### 3.1 The linear (log) model
 ```{r}
@@
 
-#### Apply the model lm01 to the CBS data
+#### 3.2 Apply the model lm01 to the CBS data
 ```{r}
@@
 
-#### plot: sero-prevalence, using predicted Abbott positivity
+#### 3.3 plot: sero-prevalence, using predicted Abbott positivity
 ```{r}

### 2024-04-09 2_scripts/2.3 Prob Model for PCR pos.qmd
Abbott prob models, xyplot

@@
 library(nnet)          # for multinomial log regression
+#library(expss)         # for labeling variables
+library(ggExtra)       # for marginal dist on ggplot
 ```
@@
 load(file = "../1_data/private/RFD4682e.RData")
+
+# The cleaned ver of Canpath data
+
 ```
@@
 ## Data-04, 2/8 months data PSO, extracted from Y. Nagakama et al.
+## The Convalescent pts data
 X4_Nakagama_all <- read_excel("../1_data/private/Prob Model/4-anonymized_rocheabbott.xlsx",
@@
 rm(X4_Nakagama_all)
+```
+
+```{r}
+## The negative control data, updated on 7-Apr-2024
+dta_04_neg <- read_excel("../1_data/private/Prob Model/4-anonymized_rocheabbott.xlsx",
+                              sheet = "true_neg",
+                              col_types = c("text", "numeric", "numeric", 
+                                            "numeric", "numeric", "numeric"))
+
+# correct col names
+colnames(dta_04_neg)<-c("gender","age","month_pso","abbott_n","roche_n","pid")
+dta_04_neg$pso <- dta_04_neg$month_pso*30
+
+# All participants been sampled only once, wt=1
+dta_04_neg$sample_wt<-rep(1, length(dta_04_neg$pid))
+dta_04_neg$study<-rep(4.1, length(dta_04_neg$pid))
 ```
@@
                           select(dta_04, -c("month_pso")), dta_06s)
+dta_all<-dplyr::bind_rows(dta_all, 
+                          select(dta_04_neg, -c("month_pso")))
 
@@
 
+# Label the study var
+dta_all$study<-as.character(dta_all$study)
+dta_all$study[dta_all$study=='1']<-"Leon_FR"
+dta_all$study[dta_all$study=='3']<-"Jamaica"
+dta_all$study[dta_all$study=='4']<-"Osaka_JP_pos"
+dta_all$study[dta_all$study=='4.1']<-"Osaka_JP_neg"
+dta_all$study[dta_all$study=='6']<-"UK"
+
+dta_all$time_pso[is.na(dta_all$time_pso)]<-"never_inf"
+
 # save the rbinded harmonized data
@@
 # remove all intermediate data
-rm(dta_01, dta_01s, dta_03, dta_03s, dta_04, dta_06, dta_06s)
+rm(dta_01, dta_01s, dta_03, dta_03s, dta_04, dta_04_neg, dta_06, dta_06s)
 ```
@@
 
-#### 1.1 Multinomial Regression Model:
+#### 1.1 Multinomial Regression Model (Roche):
 
@@
 
+#### 1.2 Model mnlm04 same model but for abbott
+```{r}
+# Model 04  takes sample weights into account
+mnlm04<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ log_abbott_n, 
+                 data = dta_all, weights = sample_wt)
+summary(mnlm04)
+## use estimates/SE and 1.96 to calculate p-value manually, should be <0.001
+```
+
 ##====================== Part II. Applying the model to Sero data =============##
@@
 
-#### 2.2 Apply Model to CBS Data
+#### 2.2 Apply Model (Roche) to CBS Data
 
@@
 
-#### Make by Region Plots
+#### 2.2.2 Make by Region Plots
 
@@
 ```
+#### 2.2.3 Make plots by region
+
+```{r}
+# Reshape meltted data
+tb_melt<-reshape2::melt(select(tb_all, 1:2,4:7), 
+                        id.vars = c("samplemonth", "region"),
+                        variable.name = "inf",
+                        value.name = "prop")
+
+tb_melt2<-reshape2::melt(select(tb_all2, 1,3:5), 
+                        id.vars = c("samplemonth", "region"),
+                        variable.name = "inf",
+                        value.name = "prop")
+
+# 1. The figure using simple model
+p01<-tb_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=inf, colour=inf)) +
+  geom_point(aes(colour=inf)) +
+  geom_line(aes(colour=inf))  +
+  theme_bw()
+
+p01 + facet_wrap(~region)
+
+# 2. The model using simple model + age & sex
+p02<-tb_melt2 %>%
+  ggplot(aes(x=samplemonth, y=prop, group=inf, colour=inf)) +
+  geom_point(aes(colour=inf)) +
+  geom_line(aes(colour=inf))  +
+  theme_bw()
+
+p02 + facet_wrap(~region)
+```
+
+#### 2.3 Apply model to APL data
+
+#### 2.3.1 Load & Clean
+```{r}
+summary(RFD4682_e$N)
+# The codebook does not specify what types of anti-N assay is applied, given the value we assume they are using Abbott anti-N assay (small numbers)
+
+# create a subset of APL
+APLs<-RFD4682_e %>%
+  select(order_ID, AGE_AT_COLLECTION, GENDER, N, Col_Month, PAT3)
+
+# unify colnames
+colnames(APLs)<-c("order_ID","age","gender","abbott_n","samplemonth","fsa")
+APLs$samplemonth<-as.yearmon(APLs$samplemonth)
+
+# add log_abbott_n
+## find 0 values bf transformation
+APLs$abbott_n[APLs$abbott_n==0 & 
+              !is.na(APLs$abbott_n)]<-0.001
+
+APLs$log_abbott_n <- log(APLs$abbott_n)
+```
+
+#### 2.3.2 Apply model (Abbott)
+```{r}
+# Apply the model mnlm04
+APLs_prob_r<-as.data.frame(predict(mnlm04, newdata = APLs, "probs")) # num of >3 months = 0
+mean(APLs_prob_r[,1], na.rm = T)  # never: 53.28%
+mean(APLs_prob_r[,4], na.rm = T)  # <3 months: 24.21%
+mean(APLs_prob_r[,2], na.rm = T)  # 3-6 month: 5.40%
+mean(APLs_prob_r[,3], na.rm = T)  # 6-9 months: 4.66%
+mean(APLs_prob_r[,5], na.rm = T)  # >9 months: 12.46%
+
+APLs_prob_r$ever_inf <- 1-APLs_prob_r$never_inf
+APLs_prob_r$less_6_month <- APLs_prob_r$less_3_month + APLs_prob_r$`3-6_month`
+APLs_prob_r$less_9_month <- APLs_prob_r$less_6_month + APLs_prob_r$`6-9_month`
+
+APLs<-cbind.data.frame(APLs, APLs_prob_r[,c(4,6,7,8)])
+```
+
+```{r}
+# create melt data (monthly rolling inc prop estimates)
+tb_APL<-APLs %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            ever_inf= mean(ever_inf, na.rm = T), 
+            less_3_month=mean(less_3_month, na.rm = T),
+            less_6_month=mean(less_6_month, na.rm=T),
+            less_9_month=mean(less_9_month, na.rm=T),
+            )
+# clean NAs
+tb_APL<-tb_APL[!is.na(tb_APL$samplemonth) & !is.na(tb_APL$ever_inf), ]
+
+```
+
+#### 2.3.3 making Plots
+```{r}
+tb_melt_APL<-reshape2::melt(select(tb_APL, 1,3:6), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "inf",
+                        value.name = "prop")
+
+p06<-tb_melt_APL %>%
+  ggplot(aes(x=samplemonth, y=prop, group=inf, colour=inf)) +
+  geom_point(aes(colour=inf)) +
+  geom_line(aes(colour=inf))  +
+  theme_bw()
+p06
+```
+**Additional plotting** 
+```{r}
+# merge monthly summary table: CBS + APL
+tb_CBS<-tb_all[tb_all$region=="Alberta",] %>%
+  select(-2)
+tb_CBS$data<-rep("CBS", length(tb_CBS$samplemonth))
+
+tb_APL$data<-rep("APL", length(tb_APL$samplemonth))
+tb_APLCBS<-bind_rows(select(tb_CBS, -2), select(tb_APL, -2))
+
+```
+
+```{r}
+# create melt data:
+tb_melt_2<-reshape2::melt(tb_APLCBS, 
+                        id.vars = c("samplemonth", "data"),
+                        variable.name = "inf",
+                        value.name = "prop")
+
+p07<-tb_melt_2 %>%
+  ggplot(aes(x=samplemonth, y=prop, group=data, colour=data)) +
+  geom_point(aes(colour=data)) +
+  geom_line(aes(colour=data))  +
+  theme_bw()
+p07 + facet_wrap(~inf)
+```
+
+#### 2.4 Apply model to CANPATH data
+
+```{r}
+Canpath <- read.csv("~/COVID-assay-adjust-methods/1_data/private/CANPATH/DAO-543759_ResearcherDataset_Serology_Results_74503par.csv")
+Canpath_all <- read.csv("~/COVID-assay-adjust-methods/1_data/private/CANPATH/DAO-543759_ResearcherDataset_Qx_96014par_1125var.csv")
+
+table(Canpath$C1_CITF_ASSAY_ID)
+```
 
@@
 
-#### Make plots by region
+#### 3.4 Plot the correlation btw Roche-N and Abbott-N:
+```{r}
+# The clear linear relation btw log_roche and log_abbott:
+plot(CBSs$log_roche_n, CBSs$log_abbott_n) # this linear association is intriduced by lm01
+
+# How it looks like in normal scale?
+plot(CBSs$roche_n, CBSs$abbott_n_hat)     # a smooth curve
+
+dta_all$study<-as.factor(dta_all$study)
+
+# using our training data:
+p04<-dta_all %>%
+  .[!is.na(.$log_abbott_n) & !is.na(.$log_roche_n),] %>%
+  ggplot(aes(x=log_roche_n, y=log_abbott_n, 
+             study, colour=study)) +
+  geom_point(aes(colour=as.factor(study))) +
+  geom_vline(xintercept = 0, linewidth=1.0, linetype="dotted", col="red") +
+  geom_hline(yintercept = log(1.4), linewidth=1.0, linetype="dotted", col="red") +
+  xlab("Roche Anti-N S/Co ratio (log scale)") + 
+  ylab("Abbott Anti-N S/Co ratio (log scale)") +
+  labs(color='') +
+  scale_colour_manual(labels = c("Jamaica (-)", "Osaka, JP (-)",
+                                 "Osaka, JP (+)","UK (+)"), 
+                      values = c("red", "skyblue", "darkgreen", "purple")) +
+  theme_bw() + 
+  theme(legend.position = "bottom")
+
+p04
+```
 
 ```{r}
-# Reshape meltted data
-tb_melt<-reshape2::melt(select(tb_all, 1:2,4:7), 
-                        id.vars = c("samplemonth", "region"),
-                        variable.name = "inf",
-                        value.name = "prop")
+## Another figure: add CBS matched data
+dta_CBS<-CBSs[!is.na(CBSs$abbott_n) & !is.na(CBSs$roche_n),]  %>%
+  select(log_roche_n, abbott_n)
+dta_CBS$study<-rep("CBS matched", length(dta_CBS$log_roche_n))
+dta_CBS$log_abbott_n<-log(dta_CBS$abbott_n)
 
-tb_melt2<-reshape2::melt(select(tb_all2, 1,3:5), 
-                        id.vars = c("samplemonth", "region"),
-                        variable.name = "inf",
-                        value.name = "prop")
+dta_CBS<-bind_rows(select(dta_all, study, log_roche_n, log_abbott_n), 
+                   select(dta_CBS, study, log_roche_n, log_abbott_n))
 
-# 1. The figure using simple model
-p01<-tb_melt %>%
-  ggplot(aes(x=samplemonth, y=prop, group=inf, colour=inf)) +
-  geom_point(aes(colour=inf)) +
-  geom_line(aes(colour=inf))  +
-  theme_bw()
+p05<-dta_CBS %>%
+  .[!is.na(.$log_abbott_n) & !is.na(.$log_roche_n),] %>%
+  ggplot(aes(x=log_roche_n, y=log_abbott_n, 
+             study, colour=study)) +
+  geom_point(aes(colour=as.factor(study))) +
+  geom_vline(xintercept = 0, linewidth=1.0, linetype="dotted", col="red") +
+  geom_hline(yintercept = log(1.4), linewidth=1.0, linetype="dotted", col="red") +
+  xlab("Roche Anti-N S/Co ratio (log scale)") + 
+  ylab("Abbott Anti-N S/Co ratio (log scale)") +
+  labs(color='') +
+  scale_colour_manual(labels = c("CBS","Jamaica (-)", "Osaka, JP (-)",
+                                 "Osaka, JP (+)","UK (+)"), 
+                      values = c("lightblue","red", "blue", "darkgreen", "purple")) +
+  theme_bw() + 
+  theme(legend.position = "bottom")
 
-p01 + facet_wrap(~region)
-
-# 2. The model using simple model + age & sex
-p02<-tb_melt2 %>%
-  ggplot(aes(x=samplemonth, y=prop, group=inf, colour=inf)) +
-  geom_point(aes(colour=inf)) +
-  geom_line(aes(colour=inf))  +
-  theme_bw()
-
-p02 + facet_wrap(~region)
+p05
 ```
-
-#### 2.3 Apply model to APL data
-
+#### Alton suggests using marginal dist with {ggExtra}
 ```{r}
-summary(RFD4682_e$N)
-# The codebook does not specify what types of anti-N assay is applied, given the number we assume they are using Abbott anti-N assay (small numbers)
-```
-
-#### 2.4 Apply model to CANPATH data
-
-```{r}
-Canpath <- read.csv("~/COVID-assay-adjust-methods/1_data/private/CANPATH/DAO-543759_ResearcherDataset_Serology_Results_74503par.csv")
-Canpath_all <- read.csv("~/COVID-assay-adjust-methods/1_data/private/CANPATH/DAO-543759_ResearcherDataset_Qx_96014par_1125var.csv")
-
-table(Canpath$C1_CITF_ASSAY_ID)
+ggMarginal(p04, groupColour = TRUE, groupFill = TRUE) # log-scale
+ggMarginal(p05, groupColour = TRUE, groupFill = TRUE) # with CBS matched data
 ```

### 2024-04-11 2_scripts/2.3 Prob Model for PCR pos.qmd
Update Roche~Abbott models

@@
 APLs<-RFD4682_e %>%
-  select(order_ID, AGE_AT_COLLECTION, GENDER, N, Col_Month, PAT3)
+  select(order_ID, AGE_AT_COLLECTION, GENDER, N, `N-IgG_INTERP`, Col_Month, PAT3)
 
 # unify colnames
-colnames(APLs)<-c("order_ID","age","gender","abbott_n","samplemonth","fsa")
+colnames(APLs)<-c("order_ID","age","gender","abbott_n", "pos_abbott","samplemonth","fsa")
 APLs$samplemonth<-as.yearmon(APLs$samplemonth)
@@
 
-#### 3.1 The linear (log) model
+#### 3.1.1 The linear (log) model: Abbott ~ Roche
 ```{r}
@@
 
-#### 3.2 Apply the model lm01 to the CBS data
+#### 3.1.2  The linear (log) model: Roche ~ Abbott
+```{r}
+lm01r<-glm(log_roche_n ~ log_abbott_n, 
+          family = "gaussian",
+          data = dta_all, weights = sample_wt)
+summary(lm01r) 
+```
+
+#### 3.2.1 Apply the model lm01 to the CBS data
 ```{r}
@@
 
-#### 3.3 plot: sero-prevalence, using predicted Abbott positivity
+#### 3.2.2 Apply model lm01r to APL data
+```{r}
+APLs$log_roche_n<-predict(lm01r, newdata = APLs, "response")
+
+APLs$roche_n_hat <- exp(APLs$log_roche_n)
+summary(APLs$roche_n_hat)
+
+# Apply Roche COI=1.0
+APLs$pos_roche<-if_else(APLs$roche_n_hat<=1, 0, 1, missing = NA)
+# recode Abbott pos as num
+APLs$pos_abbott[APLs$pos_abbott=="Negative"]<-0
+APLs$pos_abbott[APLs$pos_abbott=="Positive"]<-1
+APLs$pos_abbott<-as.numeric(APLs$pos_abbott)
+```
+
+#### 3.3.1 plot: sero-prevalence, using predicted Abbott positivity
 ```{r}
@@
 
+#### 3.3.2 plot: sero-prevalence, using predicted Roche pos
+```{r}
+tb_ro<-APLs %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_abbott= sum(pos_abbott, na.rm = T), 
+            pos_roche= sum(pos_roche, na.rm = T)
+            )
+tb_ro$prop.abbott<-tb_ro$pos_abbott/tb_ro$n
+tb_ro$prop.roche<-tb_ro$pos_roche/tb_ro$n
+
+# clean NAs
+tb_ro<-tb_ro[!is.na(tb_ro$samplemonth) & 
+             !tb_ro$pos_abbott==0, ]
+
+tb_ro_melt<-reshape2::melt(select(tb_ro, -c(2:4)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "assay",
+                        value.name = "prop")
+
+p03r<-tb_ro_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=assay, colour=assay)) +
+  geom_point(aes(colour=assay)) +
+  geom_line(aes(colour=assay))  +
+  theme_bw() 
+p03r
+```
+
 #### 3.4 Plot the correlation btw Roche-N and Abbott-N:
@@
              study, colour=study)) +
-  geom_point(aes(colour=as.factor(study))) +
+  geom_point(aes(colour=as.factor(study)), alpha=0.5) +
   geom_vline(xintercept = 0, linewidth=1.0, linetype="dotted", col="red") +
@@
                                  "Osaka, JP (+)","UK (+)"), 
-                      values = c("red", "skyblue", "darkgreen", "purple")) +
+                      values = c("red", "blue", "darkgreen", "purple")) +
   theme_bw() + 
@@
              study, colour=study)) +
-  geom_point(aes(colour=as.factor(study))) +
+  geom_point(aes(colour=as.factor(study)), alpha=0.40) +
   geom_vline(xintercept = 0, linewidth=1.0, linetype="dotted", col="red") +

### 2024-04-19 2_scripts/2.3 Prob Model for PCR pos.qmd
Update the codes for figure about Abb~Roch transformation

@@
 ```
+#### 3.3.3 Plot both roche and abbott, raw/transformed in on figure
+```{r}
+# modify melted data
+
+## raw and transformed
+tb_ro_melt$raw[tb_ro_melt$assay=="prop.abbott"]<-"raw"
+tb_ro_melt$raw[tb_ro_melt$assay=="prop.roche"]<-"transformed"
+
+tb_abb_melt2<-tb_abb_melt %>%
+  .[.$region=="Alberta",] %>%
+  select(1,3,4)
+tb_abb_melt2$raw[tb_abb_melt2$assay=="prop.roche"]<-"raw"
+tb_abb_melt2$raw[tb_abb_melt2$assay=="prop.abbott"]<-"transformed"
+
+## Alberta only
+tb_alberta<-bind_rows(tb_ro_melt, tb_abb_melt2)
+
+## add a data source indicator
+table(tb_alberta$assay, tb_alberta$raw)
+tb_alberta$data[tb_alberta$assay=="prop.abbott"
+                & tb_alberta$raw=="raw"]<-"APL"
+tb_alberta$data[tb_alberta$assay=="prop.roche"
+                & tb_alberta$raw=="transformed"]<-"APL"
+tb_alberta$data[tb_alberta$assay=="prop.abbott"
+                & tb_alberta$raw=="transformed"]<-"CBS"
+tb_alberta$data[tb_alberta$assay=="prop.roche"
+                & tb_alberta$raw=="raw"]<-"CBS"
+table(tb_alberta$data, tb_alberta$assay)
+table(tb_alberta$data, tb_alberta$raw)
+```
+
+Alton's comments:
+- solid/dash to indicate raw/transformed
+- red/blue to indicate Abbott/Roche
+
+```{r}
+# making the plot, 
+p03.3<-tb_alberta %>%
+  ggplot(aes(x=samplemonth, y=prop, colour=data)) +
+  geom_point(aes(colour=data, shape=raw)) +
+  geom_line(aes(colour=data, linetype=raw))  +
+  theme_bw() + 
+  xlab("Time of Sample Collection") + 
+  ylab("Sero-prevalence") 
+p03.3
+```
 
@@
 p04
+```
+#### Add alternative COI
+```{r}
+## add two new COI, those two were transformed using the model lm01 and lm01r
+df.temp<-as.data.frame(c(log(1.0)))
+colnames(df.temp)<-c("log_roche_n")
+COI_abbott_t<-predict(lm01, newdata = df.temp, "response")
+
+df.temp<-as.data.frame(c(log(1.4)))
+colnames(df.temp)<-c("log_abbott_n")
+COI_roche_t<-predict(lm01r, newdata = df.temp, "response")
+
+p04 +   
+  geom_vline(xintercept = COI_roche_t, linewidth=1.0, linetype="dotted", col="orange") +
+  geom_hline(yintercept = COI_abbott_t, linewidth=1.0, linetype="dotted", col="orange")
+
+## Use model lm01 coeff to calculate the Roche COI
+COI_roche_t2<-(log(1.4)-as.numeric(lm01$coefficients[1]))/as.numeric(lm01$coefficients[2])
+COI_abbott_t2<-(log(1.0)-as.numeric(lm01r$coefficients[1]))/as.numeric(lm01r$coefficients[2])
+
+p04<-p04 +   
+  geom_vline(xintercept = COI_roche_t2, linewidth=1.0, linetype="dotted", col="darkgreen") +
+  geom_hline(yintercept = COI_abbott_t2, linewidth=1.0, linetype="dotted", col="darkgreen")
 ```

### 2024-04-22 2_scripts/2.3 Prob Model for PCR pos.qmd
update figures for AABB abstract (unfinished)

@@
 table(tb_alberta$data, tb_alberta$raw)
+
+## create a merged indicator for study + transformed
+## we could use interaction() in ggplot instead
+tb_alberta$study<-paste(tb_alberta$data, 
+                        rep(" (", length(tb_alberta$data)),
+                        tb_alberta$raw, 
+                        rep(")", length(tb_alberta$data) ))
 ```
@@
 - red/blue to indicate Abbott/Roche
-
-```{r}
-# making the plot, 
+- Merge the color & linetype indicator in the legend
+
+```{r}
+# making the plot, either those codes will work
 p03.3<-tb_alberta %>%
-  ggplot(aes(x=samplemonth, y=prop, colour=data)) +
-  geom_point(aes(colour=data, shape=raw)) +
-  geom_line(aes(colour=data, linetype=raw))  +
+  ggplot(aes(x=samplemonth, y=prop, colour=study, linetype=study)) +
+  geom_point() +
+  geom_line()  +
   theme_bw() + 
   xlab("Time of Sample Collection") + 
-  ylab("Sero-prevalence") 
+  ylab("Sero-prevalence")+ 
+  scale_colour_manual("",   values = c("red", "red", "blue", "blue"))+
+  scale_linetype_manual("", values=c(1,3,1,3))
+
+# use this chunk to skip the paste()
+p03.lab<-c("APL (raw)", "CBS (raw)", 
+           "APL (transformed)", "CBS (transformed)")
+p03.3<-tb_alberta %>%
+  ggplot(aes(x=samplemonth, y=prop, colour=interaction(data, raw),
+                                    linetype=interaction(data, raw))) +
+  geom_point() +
+  geom_line()  +
+  theme_bw() +
+  xlab("Time of Sample Collection") + 
+  ylab("Sero-prevalence")+ 
+  scale_colour_manual("", values = c("red", "blue", "red", "blue"),
+                      labels = p03.lab)+
+  scale_linetype_manual("", values=c(1,1,3,3), labels = p03.lab) +
+  theme(legend.position="top")+
+  guides(color = guide_legend(nrow = 2))
+
 p03.3
@@
 ```
+#### 3.5 Create figure panel for AABB conference abstract
+
+#### 3.5.1 Panel A: the serosurveillance figure and RG/Bayesian adjustment
+Why: they only alow on figure.
+```{r}
+## Load Yuan's RG data
+rg_all <- read.csv("../1_data/private/df_time_rg_all.csv", row.names=1)
+rg_ab  <- rg_all[rg_all$region=="Alberta" &
+                 rg_all$source %in% c("APL", "CBS") &
+                 !(rg_all$model=="RAW"),]
+rg_ab$yearmon <- as.yearmon(rg_ab$yearmon)
+
+## merge rg_ab and tb_alberta
+tb_alberta<-merge(tb_alberta, select(rg_ab, 2:6),
+      by.x = c("data", "samplemonth"), by.y = c("source", "yearmon"),
+      all = T)   # need to ask Yuan about the updated CBS data
+
+tb_ab_rg<-tb_alberta %>%
+  select(-c("assay", "study")) %>%
+  .[.$raw=="raw" & !is.na(.$raw),] %>%
+  select(-c("raw"))
+
+tb_ab_rg<-reshape2::melt(tb_ab_rg, 
+                      id.vars = c("data", "samplemonth",
+                                  "rg_lower", "rg_upper"),
+                      variable.name = "rg",
+                      value.name = "prop")
+tb_ab_rg$rg_lower[tb_ab_rg$rg=="prop"]<-NA
+tb_ab_rg$rg_upper[tb_ab_rg$rg=="prop"]<-NA
+```
+
+```{r}
+# rebuild the figure
+p03.5lab<-c("APL (raw)", "CBS (raw)", 
+            "APL (RG-Bayesian)", "CBS (RG-Bayesian)")
+
+p03.5<-tb_ab_rg %>%
+  ggplot(aes(x=samplemonth, y=prop, colour=interaction(data,rg),
+                                    linetype=interaction(data,rg))) +
+  geom_point() +
+  geom_line()  +
+  geom_errorbar(aes(ymin=rg_lower, ymax=rg_upper), width=0.05)+
+  theme_bw() + 
+  xlab("Time of Sample Collection") + 
+  ylab("Sero-prevalence") +
+  scale_colour_manual("", values = c("red", "blue", "red", "blue"),
+                      labels = p03.5lab) +
+  scale_linetype_manual("", values=c(1,1,3,3), labels = p03.5lab) +
+  theme(legend.position="top")+
+  guides(color = guide_legend(nrow = 2))
+
+p03.5
+```
+
+#### 3.5.2 Panel B: transformed assay result fig
+```{r}
+p03.3
+```
+
+#### 3.5.3 Panel C: rolling incidence fig
 

### 2024-04-24 2_scripts/2.3 Prob Model for PCR pos.qmd
Finished coding for AABB conference fig

@@
 #### 3.5.3 Panel C: rolling incidence fig
+```{r}
+## recorder the dataset, relevel inf variable
+tb_melt_2$inf.f<-as.factor(tb_melt_2$inf)
+tb_melt_2<-transform(tb_melt_2, 
+                     inf.f = factor(inf.f, levels = c("ever_inf", "less_9_month",
+                                                     "less_6_month", "less_3_month"))
+)
+# replace with friendly name for figuers
+levels(tb_melt_2$inf.f)<-c("ever infected", "infected <9 month",
+                           "infected <6 month", "infected <3 month")
+## make figures p07
+p07<-tb_melt_2 %>%
+  ggplot(aes(x=samplemonth, y=prop, group=data, colour=data)) +
+  geom_point() +
+  geom_line()  +
+  theme_bw()   +
+  xlab("Time of Sample Collection") + 
+  ylab("Rolling Incidence Proportion") +
+  scale_colour_manual("", values = c("red", "blue"))
+p07
+
+# add facet names
+p07<-p07 + facet_wrap(~inf.f, nrow=1) +
+    theme(axis.text.x = element_text(angle = 30, vjust=0.8, hjust=0.8))
+p07
+```
+
+#### 3.5.4 Make all three figures into one
+```{r}
+# use cowplot package
+p08.upper<-cowplot::plot_grid(p03.5, p03.3 + ylab(""), nrow=1,
+                              labels = c("A", "B"))
+
+png(file="../4_output/figs/Fig.X1 AABB Conference.png", 
+    width=2400, height=1800, res=300)
+cowplot::plot_grid(p08.upper, p07 + theme(legend.position = c(0.95, 0.93),
+                                          legend.background = element_blank()),
+                   labels = c("","C"), ncol=1, rel_heights = c(1, 1))
+dev.off()
+```
 

### 2024-04-25 2_scripts/2.3 Prob Model for PCR pos.qmd
Minor updates on figures for AABB conference

--- 2024-04-22 2_scripts/2.3 Prob Model for PCR pos.qmd
@@
 ```
+The gap between CBS and APL seropositivity estimates for October 2023 was X% using raw seropositivity, Y% after transforming CBS to Abbott, and Z% after transforming APL to Roche.
+
+```{r}
+# How gaps were calculated: 
+# X: Oct 2023, btw raw CBS & APL, Y: APL vs transformed CBS, Z: CBS vs transformed APL
+tb_alberta %>%
+  select(data, samplemonth, prop, raw) %>%
+  .[.$samplemonth == as.yearmon("Oct 2022"),] %>%
+  mutate(X = .$prop[4] - .$prop[1]) %>%
+  mutate(Y = .$prop[3] - .$prop[1]) %>%
+  mutate(Z = .$prop[4] - .$prop[2])
+```
 
@@
 #### 3.5.3 Panel C: rolling incidence fig
+```{r}
+## recorder the dataset, relevel inf variable
+tb_melt_2$inf.f<-as.factor(tb_melt_2$inf)
+tb_melt_2<-transform(tb_melt_2, 
+                     inf.f = factor(inf.f, levels = c("ever_inf", "less_9_month",
+                                                     "less_6_month", "less_3_month"))
+)
+# replace with friendly name for figuers
+levels(tb_melt_2$inf.f)<-c("Ever infected", "Infected <9 months ago",
+                           "Infected <6 months ago", "Infected <3 months ago")
+## make figures p07
+p07<-tb_melt_2 %>%
+  ggplot(aes(x=samplemonth, y=prop, group=data, colour=data)) +
+  geom_point() +
+  geom_line()  +
+  theme_bw()   +
+  xlab("Time of Sample Collection") + 
+  ylab("Rolling Incidence Proportion") +
+  scale_colour_manual("", values = c("red", "blue"))
+p07
+
+# add facet names
+p07<-p07 + facet_wrap(~inf.f, nrow=1) +
+    theme(axis.text.x = element_text(angle = 30, vjust=0.8, hjust=0.8))
+p07
+```
+
+#### 3.5.4 Make all three figures into one
+```{r}
+# use cowplot package
+p08.upper<-cowplot::plot_grid(p03.5 + xlab ("") + ylab("Seroprevalence"), 
+                              p03.3 + xlab ("") + ylab(""), nrow=1,
+                              labels = c("A", "B"))
+
+png(file="../4_output/figs/Fig.X1 AABB Conference.png", 
+    width=2400, height=1800, res=300)
+cowplot::plot_grid(p08.upper, p07 + theme(legend.position = c(0.95, 0.93),
+                                          legend.background = element_blank()),
+                   labels = c("","C"), ncol=1, rel_heights = c(1, 1))
+dev.off()
+```
 

### 2024-04-30 2_scripts/2.3 Prob Model for PCR pos.qmd
codes for sensiticity analysis

@@
 dev.off()
+
+# tiff format
+tiff(file="../4_output/figs/Fig.X1 AABB Conference.tiff", 
+    width=2400, height=1800, res=300, compression = "zip")
+cowplot::plot_grid(p08.upper, p07 + theme(legend.position = c(0.95, 0.93),
+                                          legend.background = element_blank()),
+                   labels = c("","C"), ncol=1, rel_heights = c(1, 1))
+dev.off()
 ```
@@
 
-1.
+#### 4.1.
 About the age dist in training data and CBS: The difference on age distribution might be a potential problem.
@@
 ```
+
+#### 4.2 Sensitivity Analysis
+
+Both Alton and Yuan have address the importance of doing sensitivity analysis for the transformation models (Roche--Abbott).
+Originally we trained those two models with external matched data (Jamaica-Osaka-UK).
+We wonder if model would fit similarity with our own matched data in CBS.
+
+#### 4.2.1 Using Jamaica-Osaka-UK data:
+```{r}
+# Roche-> Abbott
+summary(lm01)
+
+# Abbott-> Roche
+
+summary(lm01r)
+```
+#### 4.2.2 Refit using CBS matched data
+```{r}
+CBSs2<-CBSs %>%
+  .[!is.na(.$roche_n) & !is.na(.$abbott_n),] #CBS matched data
+
+CBSs2$log_abbott_n<-log(CBSs2$abbott_n)
+CBSs2$log_roche_n<-log(CBSs2$roche_n)
+
+# clean inf values with log(0.01)
+CBSs2$log_abbott_n[CBSs2$log_abbott_n==-Inf]<-log(0.01)
+
+# lm05: Roche--> Abbott, using matched CBS
+lm05<-glm(log_abbott_n ~ log_roche_n, 
+          family = "gaussian",
+          data = CBSs2)
+summary(lm05)
+
+# lm05r: Abbott--> Roche, using matched CBS
+lm05r<-glm(log_roche_n~log_abbott_n , 
+          family = "gaussian",
+          data = CBSs2)
+summary(lm05r)
+# model 5r should be focused
+```
+#### Adjusting the weights of those with negative test results
+```{r}
+# add a indicator for positivity of two assays
+CBSs2$ppnn[CBSs2$abbott_n > 1.4 & CBSs2$roche_n > 1]<-"pp"
+CBSs2$ppnn[CBSs2$abbott_n > 1.4 & CBSs2$roche_n < 1]<-"np"
+CBSs2$ppnn[CBSs2$abbott_n <=1.4 & CBSs2$roche_n < 1]<-"nn"
+CBSs2$ppnn[CBSs2$abbott_n <=1.4 & CBSs2$roche_n > 1]<-"pn"
+
+CBSs2$samplewt <- rep(1, length(CBSs2$pid))
+CBSs2$samplewt[CBSs2$ppnn %in% c("np","nn")]<-0.01
+
+# refit model 05r
+lm05r.w<-glm(log_roche_n~log_abbott_n , 
+          family = "gaussian", weights = samplewt,
+          data = CBSs2)
+summary(lm05r.w)
+# after down-weighting, the betas became closer to 
+```

### 2024-05-06 2_scripts/2.3 Prob Model for PCR pos.qmd
Update figures and manuscript

@@
 p03.5
+
+png(file="../4_output/figs/Fig.1 Seroprevalence.png", 
+    width=2400, height=1800, res=300)
+p03.5
+dev.off()
 ```
@@
 p03.3
+
+png(file="../4_output/figs/Fig.2 Transformed.png", 
+    width=2400, height=1800, res=300)
+p03.3
+dev.off()
 ```
@@
 p07
+
+png(file="../4_output/figs/Fig.3 Rolling-Incidence.png", 
+    width=2400, height=1800, res=300)
+p07 + facet_wrap(~inf.f, nrow=2)
+dev.off()
 ```

### 2024-05-06 5_manuscript/manuscript.qmd
minor updates on prob models
Update the codes for figure about Abb~Roch transformation
Update figures and manuscript

@@
 3.
-Canpath: At least two types of anti-N assays are utilized: Mount Sinai Hosp anti-N (a in-lab developed assay) and Abbott anti-N (a commercial assay).
+Canpath (removed): At least two types of anti-N assays are utilized: Mount Sinai Hosp anti-N (a in-lab developed assay) and Abbott anti-N (a commercial assay).
 Need to figure out which prov at what time used what assay, so we can apply sens for adjustment accordingly.
+As the assay difference project is mainly comparing CBS to APL, and Canpath has very few data available at later phases of the pandemic when sero-prevalence diverge between APL & CBS, we decided to have it excluded from analysis.
 
@@
 CBS is still actively updating (last ver downloaded on Nov 30th) and Yuan has the access to the most updated ver of CBS.
-(last checked on 21-Feb-2024, CBS 2023-11-30 is still the last version)
+(last checked on 30-Apr-2024, CBS 2023-11-30 is the version we are working on, lastest CBS has 2023 data updated) As the assay difference project is mainly comparing CBS to APL, we don't have to include 2023 CBS data as APL has no comparable data.
 --->
@@
 
-To estimate the SARS-CoV-2 seroprevalence and rolling incidence in Alberta, we used data from the Alberta Outpatient Lab (APL), Canadian Blood Services (CBS) and Canadian Partnership for Tomorrow's Health (CANPATH).
+To estimate the SARS-CoV-2 seroprevalence and rolling incidence in Alberta, we used data from the Alberta Outpatient Lab (APL) and Canadian Blood Services (CBS).
 Alberta Health Services (AHS) tested \>160,000 APL blood draws for SARS-CoV-2 anti-N and anti-S antibodies from June 2020 until October 2022.
 CBS has tested \>450,000 blood donations for both antibodies levels since March 2020.
-CANPATH has provided data with sample size \> XX.
-Besides of assay results, all three datasets include demographic variables such as donor's sex, self-reported race/ethnicity, and age.
-From their address, neighborhood material/social deprivation and urban/rural indicator have been derived.
+The CBS Alberta subset has tested 124,008 samples in the same data collection period as CBS.
+Besides of assay results, both datasets include demographic variables such as donor's sex, self-reported race/ethnicity, age, and Forward Sortation Area (FSA) codes.
+From their FSA codes, neighborhood material/social deprivation and urban/rural indicator have been derived.
 
-The three datasets feature different data collection time frames.
+Two datasets have different data collection time frames.
 The CBS dataset has the most extended period, ranging from December 2020 to February 2023.
-APL's data collection spans from June 2020 to October 2022, but it's important to note that anti-N assay results are only available starting from December 2021.
-Meanwhile, the CANPATH dataset is accessible for a shorter duration, from September 2020 to May 2021.
+Meanwhile, APL's data collection spans from June 2020 to October 2022.
+However, the anti-N assay results are only available starting from December 2021 in the APL data.
+@thm-data summarized the differences among two datasets.
 
-We did not conduct analysis using a merged data, but instead, we conducted analysis separately for each data and presented the results in a graph panel (@fig-seroprevalence).
-@thm-data below summarized the differences among three data utilized in the study.
-
-@thm-data CBS, CANPATH and APL
+@thm-data CBS and APL
 
@@
 
-## **Seroprevalence and rolling incidence**
+Regarding the data we used to train the regression models, we identified several studies conducted in the early phases of the Covid-19 pandemic.
+Blood samples are collected from France, Jamaica, Japan and UK \[**add citation**\], with known sample positivity at the time of collection (more details are available at the supplemental methods section).
+The samples in Jamaica, Japan and UK were tested both for Abbott anti-N and Roche anti-N assay, which makes them suitable for transformation model training.
+
+## **Seroprevalence adjusted by RG and Bayesian models**
 
 To evaluate the severity of Covid-19 pandemic overtime in Alberta province, first we calculated the anti-N assay test positivity.
-Then, to bridge the gap between test positivity and population-level seroprevalence, we employed the Rogen-Gladen (RG) equation and Bayesian models.
-
-Following seroprevalence estimation, we further estimated monthly rolling incidence proportion of SARS-CoV-2 infection.
-We developed a multinomial logistic regression model predicting the possibility of past infection based on the anti-N assay results, adjusting for age and gender.
-Then we estimated the probability of past infection in a monthly scale using population averages.
-
-We presented both seroprevalence and rolling incidence proportion in a unified graph (@fig-seroprevalence), displaying changes over time on a monthly scale.
+Then, to bridge the gap between test positivity and population-level seroprevalence, we employed the adjustment using Rogen-Gladen (RG) equation and Bayesian models.
+We conducted analysis separately for CBS and APL data then presented the results in figure X (@fig-seroprevalence).
 
@@
 
+Where as "Prevalence" indicate RG adjusted seroprevalence estimates.
+"Test Positivity" indicates the proportion of test samples with positive results.
+"Spec" and "Sens" indicate the specificity and sensitivity of the assay, respectively.
+
 Bayesian models adjusting for sensitivity:
 
-\$\$
+$$
+\frac{A(1-W_{AA})}{N_{Group1}} \times (1-W_{H})P_{inf} = C_{sensitivity} \times Index_{a} \times 1600
+$$
 
-\$\$
+\[Bayesian models adjusting for waning sensitivity\]: Equation (above) and what each factor indicates (What model input and output. What covariates we've adjusted for...).
 
-\[Bayesian models adjusting for waning sensitivity\]: Equation (above), model input and output. What covariates we've adjusted for.
+## **Seroprevalence based on raw/transformed data**
+
+Following seroprevalence estimation, we further estimated monthly rolling incidence proportion of SARS-CoV-2 infection.
+Using matched data from samples tested with both assays early in the pandemic, we developed log-linear regression models to transform CBS Roche qualitative results to Abbott and applied the Abbott cutoff index and made the opposite transformation to convert APL to Roche.
+The seroprevalence estimates based on raw and transformed data are presented in figure X (@fig-transformed).
+
+## **Rolling Incidence**
+
+Next, we developed multinomial logistic regression models using data collected in the early pandemic to estimate the probability a tested sample was tested 0-3, 4-6, 7-9, or \>9 months from infection based on the qualitative anti-N assay results.
+Then we applied this model to the CBS and APL datasets to estimate rolling incidence from 2021 to 2023.
+
+We presented the rolling incidence proportion estimates in a paneled figure (@fig-rolling-incidence), displaying regression-based estimation of proportion of population infected in the last ever or in the past 3, 6, or 9 months.
 
@@
 
-[Insert @fig-immunity here if necessary]
-
 <br>
@@
 Describe the trend of changing overtime, difference and consistency among different adjusting methos, regions and datasets.
-Cite figures accordingly if necessary.
+(Cite figures accordingly if necessary.)The datasets had similar seropositivity earlier in the pandemic but diverged after May 2022 due to greater waning sensitivity in the Abbott assay used by APL (Figure X).RG and Bayesian adjustments had minimal impact, increasing sero-prevalence estimates by 2.04% for the CBS Alberta subset and 7.76% for APL (Figure X).
 
-\[Describe fig: rolling inc prop\] A paragraph about estimated proportion of ever infected & infected in past 3 month.
+\[Describe fig: seroprevalence based on raw/transformed data\] Our regression-based approaches increased concordance between estimated seropositivity (Figure X).
+However, we still observed divergence after May 2022.The gap between CBS and APL seropositivity estimates for October 2022 was 48% using raw seropositivity, 15% after transforming CBS to Abbott, and 14% after transforming APL to Roche.
+
+\[Describe fig: rolling inc prop\] In the rolling incidence estimates, we obderved more similar estimates of rolling incidence in each panels.
+The difference on rolling incidence estimates ar October 2022 are X%, X%, X% and X%, for ever infected %, 3-month, 6-month, and 9-month rolling incidence, respectively.
 
@@
 
-\[Strength and Limitation\]
+\[Strength and Limitation\] In our sensitivity analysis, we found the regression models used for Roche--\>Abbott transformation is sensitive to the ratio of positive vs negative cases in the training data.
 
@@
 It only estimates the proportion of test positive results among blood samples collected in the serosurveillance study.
+
+2.      Serosurveillance estimates are sensitive to choice of assay, and the most common adjustment methods (RG) do little to make estimates more concordant, particularly later in the pandemic when differences in waning sensitivity become important.
+
+3.
 
@@
 Future researchers should be cautious when making references to those different measurements.
+
+Future studies should...
+When using matched data to train regression models for assay results transformation, negative samples should be down-weighted if the proportion of negative cases is unbalanced high.
 
@@
 #| label: fig-seroprevalence
-#| fig-cap: "The figure panel above presents the temporal changes in seroprevalence and attack rate, grouped by province and data source.
-Each figure displays the unadjusted seropositivity, the seroprevalence adjusted using the Rogan-Gladen equation, the seroprevalence adjusted by Bayesian models, and the attack rates.
-The calculations of seropositivity, seroprevalence, and attack rate are based on the qualitative results of the anti-N assay.
-#| (Just a scratch, not the real fig)"
+#| fig-cap: "Figure 1 presents the raw and Rogen Gladen-adjusted seroprevalence estimated from the APL and CBS data with 95% credible intervals.
+The calculations of seropositivity, seroprevalence are based on the qualitative results of the anti-N assay."
 
-knitr::include_graphics("../4_output/figs/Figure.1 Seroprelance and attack rate by province and data source.png")
+knitr::include_graphics("../4_output/figs/Fig.1 Seroprevalence.png")
+```
+
+```{r}
+#| label: fig-transformed
+#| fig-cap: "Figure 2 depicts raw seroprevalence and seroprevalence transformed from Roche to Abbott (APL) and Abbott to Roche (APL) using a log-linear regression model. "
+
+knitr::include_graphics("../4_output/figs/Fig.2 Transformed.png")
+```
+
+```{r}
+#| label: fig-rolling-incidence
+#| fig-cap: "Figure 3 presents regression-based estimate of rolling incidence (proportion of population infected in the last ever or in the past 3, 6, or 9 months). "
+
+knitr::include_graphics("../4_output/figs/Fig.3 Rolling-Incidence.png")
 ```
@@
 
-Besides of serology test results and demographic data, CBS has linked the administrative data for the cohort, which provided additional information on their hospitalizations, clinic visits, diagnoses, COVID vaccinations, and long-term disability claims.
+(to-be-deleted: we do not utilize most of the demographic variabes in our analysis) Besides of serology test results and demographic data, CBS has linked the administrative data for the cohort, which provided additional information on their hospitalizations, clinic visits, diagnoses, COVID vaccinations, and long-term disability claims.
 Similarly, AHS and CANPATH have completed linkages and included additional vaccination history.
@@
 The paragraph briefly describe the data linkage for three datasets, however for now we do not need those variables from linked administrative data.
+
+Delete this part later
 --->
 ```
+
 A short paragraph about weighting: Our analysis is based on individual level.
@@
 While in all of those three dataset, it's possible to have one individual being sampled multiple times within the same month, in this case we applied weights to each individual according to the number of samples they have in the same month.
+In regression model training, individuals who provided multiple samples were also down-weighted according to the number of samples they provided.
+
+A paragraph about the training data: Cite the table summarizing the features of training data (Table S1). Clarify which data has matched assay results.
 
@@
 # Supplemental tables
+
+Add a table here summarizing the sample size, data source of the training data we used.
 

### 2024-05-08 2_scripts/2.3 Prob Model for PCR pos.qmd
Update code for sensitivity analysis and figs.

@@
 ```
+
 #### Adjusting the weights of those with negative test results
@@
 ```
+
+#### 4.2.3 Add regression lines to the figure
+```{r}
+# Modify p04 for sensitivity analysis
+p04<-dta_all %>%
+  .[!is.na(.$log_abbott_n) & !is.na(.$log_roche_n),] %>%
+  ggplot(aes(x=log_roche_n, y=log_abbott_n, 
+             study, colour=study)) +
+  geom_point(aes(colour=as.factor(study)), alpha=0.5) +
+  geom_vline(xintercept = 0, linewidth=1.0, linetype="dotted", col="darkgrey") +
+  geom_hline(yintercept = log(1.4), linewidth=1.0, linetype="dotted", col="darkgrey") +
+  xlab("Roche Anti-N S/Co ratio (log scale)") + 
+  ylab("Abbott Anti-N S/Co ratio (log scale)") +
+  labs(color='') +
+  scale_colour_manual(labels = c("Jamaica (-)", "Osaka, JP (-)",
+                                 "Osaka, JP (+)","UK (+)"), 
+                      values = c("red", "blue", "darkgreen", "purple")) +
+  theme_bw() + 
+  theme(legend.position = "bottom")
+
+p04.2a<-p04 + geom_abline(aes(intercept = -1.60, slope = 0.68, color="Roche->Abbott",),
+                  size=0.8) +
+      geom_abline(aes(intercept = -1.68, slope = 0.78, color="Abbott->Roche",), 
+                  size=0.8) +
+scale_colour_manual(values = c("red", "blue", "darkgreen", "purple","blue","red"),
+                    guide = guide_legend(nrow = 3,
+                            override.aes = list(
+                            linetype = c(rep("blank", 4), "solid", "solid"),
+                            shape = c(rep(16, 4), NA, NA)))) +
+     ggtitle("Jamaica-Osaka-UK matched data")
+```
+```{r}
+# Calculating Beta estimates for model 01r-reverted
+(-1)*lm01r$coefficients[1]/lm01r$coefficients[2] #beta_0
+1/lm01r$coefficients[2]                          #beta_1
+
+# Calculating Beta estimates for model 05r-reverted
+(-1)*lm05r$coefficients[1]/lm05r$coefficients[2] #beta_0
+1/lm05r$coefficients[2]                          #beta_1
+
+# Calculating Beta estimates for model 05rw-reverted
+(-1)*lm05r.w$coefficients[1]/lm05r.w$coefficients[2] #beta_0
+1/lm05r.w$coefficients[2]                            #beta_1
+```
+```{r}
+# Refit models for CBS data, modify p05
+p04.2b<-dta_CBS %>%
+  .[!is.na(.$log_abbott_n) & !is.na(.$log_roche_n) &
+    .$study=="CBS matched",] %>%
+  ggplot(aes(x=log_roche_n, y=log_abbott_n, 
+             study, colour=study)) +
+  geom_point(aes(colour=as.factor(study)), alpha=0.40) +
+  geom_vline(xintercept = 0, linewidth=1.0, linetype="dotted", col="darkgrey") +
+  geom_hline(yintercept = log(1.4), linewidth=1.0, linetype="dotted", col="darkgrey") +
+  xlab("Roche Anti-N S/Co ratio (log scale)") + 
+  ylab("Abbott Anti-N S/Co ratio (log scale)") +
+  labs(color='') +
+  scale_colour_manual(labels = c("CBS matched"), 
+                      values = c("lightblue")) +
+  theme_bw() + 
+  theme(legend.position = "bottom")
+
+p04.2b<-p04.2b + geom_abline(aes(intercept = -1.87, slope = 0.64, color="Roche->Abbott",),
+                  size=0.8) +
+         geom_abline(aes(intercept = 2.02, slope = 2.26, color="Abbott->Roche",), 
+                  size=0.8) +
+         geom_abline(aes(intercept = -2.14, slope = 0.85, 
+                         color="Abbott->Roche (down-weighted)",), 
+                  size=0.8, linetype=2) +
+         scale_colour_manual(values = c("lightblue","blue","red","red"),
+                             guide = guide_legend(nrow=2,
+                             override.aes = list(
+                             linetype = c(rep("blank", 1), "solid", "solid","dashed"),
+                             shape = c(rep(16, 1), NA, NA, NA))
+                             )) +
+         ggtitle("CBS matched data")
+p04.2b
+```
+
+```{r}
+## make two into one paneled fig
+png(file="../4_output/figs/Fig.S1 Sensitivity Analysis.png", 
+    width=2400, height=1440, res=300)
+cowplot::plot_grid(p04.2a, 
+                   p04.2b  + ylab(""), nrow=1,
+                   align = 'h',
+                   labels = c("A", "B"))
+dev.off()
+```

### 2024-05-16 2_scripts/2.4 PRNT50.qmd
Update analysis for anti-S PRNT50 (unfinished)

---
title: "2.4 PRNT50"
format: html
editor: visual
author: J.Chen (let me know if you run into any troubles with those codes)
---

## Quarto

This qmd file including the codes we utilized for estimating the proportion of the population having Ab titer high enough to protect them from possible future infections. The estimation is based on quant anti-S assay results (CBS-Roche-S and APL-Abbott-S-RBD) and the corresponding Ab cutoff related to PRNT50 (1:118).

## Running Code

## Load dataset

```{r}
# The CBS Data, extracted using "CITF_CBS_desc_stats.Rmd"
load(file = "../1_data/private/CBSs.RData") # a clean subset of CBS
# Notes: Y.Yuan confirmed on 01-May-2024 the ver 2023-11-30 of CBS is no longer the last ver. The most recent CBS data has 2023 data included but it's not adding new info to our project so we are still using the 2023-11-30 ver of CBS.

# APL Data
load()
load(file = "../1_data/private/APLs.RData") # a clean subset
```

## Load necessary paks

```{r}
library(dplyr)         # Data management
#library(reshape2)      # for melt() function
library(ggplot2)       # ggplot
library(zoo)           # for year-month format
```

## 1. Clean the APL data
```{r}
APLs<-merge(APLs, select(RFD4682_e, order_ID, 
                   'RBD-IgGII_RESULT', `RBD-IgGII_INTERP`,`RBD-IgGII_RESULT_DATE`
                   ), by="order_ID",all=T)
colnames(APLs)<-c(names(APLs)[1:15], "abbott_s", "pos_abbott_s", "month_abbott_s")
APLs$abb_s_index<-grepl("index", APLs$abbott_s, ignore.case = T)
APLs$abb_s_AU<-grepl("ml", APLs$abbott_s, ignore.case = T)
table(APLs$abb_s_index, APLs$abb_s_AU)
# 36 samples missing abbott_s info
APLs$abb_s_index[APLs$abb_s_index==0 & APLs$abb_s_AU==0]<-NA
APLs$abb_s_AU[is.na(APLs$abb_s_index)]<-NA

APLs$abb_s_AU<-as.numeric(APLs$abb_s_AU)
APLs$abb_s_index<-as.numeric(APLs$abb_s_index)
table(APLs$abb_s_index, APLs$abb_s_AU, useNA = 'ifany')

# remove units
APLs$abbott_s<-gsub('[a-z]', '', APLs$abbott_s)
APLs$abbott_s<-gsub('[A-Z]', '', APLs$abbott_s)
APLs$abbott_s<-gsub('[/<>]', '', APLs$abbott_s)
APLs$abbott_s<-as.numeric(APLs$abbott_s)

# check the COI
summary(APLs[APLs$abb_s_index==1 & APLs$pos_abbott_s=="Positive",]$abbott_s)
# 1.4 - 9.6 the COI is 1.4 not 1?
summary(APLs[APLs$abb_s_index==1 & APLs$pos_abbott_s=="Negative",]$abbott_s)
# 8.1 index but negative?

summary(APLs[APLs$abb_s_AU==1 & APLs$pos_abbott_s=="Positive",]$abbott_s)
# 50 - 800000 the COI = 50 good
summary(APLs[APLs$abb_s_AU==1 & APLs$pos_abbott_s=="Negative",]$abbott_s)
# 0-49, COI = 50 good

# COI for anti_n
summary(APLs[APLs$pos_abbott==1 & !is.na(APLs$abbott_n),]$abbott_n)
# 1.4 - 14.19 the COI = 1.4
summary(APLs[APLs$pos_abbott==0 & !is.na(APLs$abbott_n),]$abbott_n)
# 0.001 - 1.39

table(APLs$pos_abbott)
```
```{r}
# abbott_n misclassified as abbott_s with index?
APLs %>%
  .[is.na(.$abbott_n) & .$abb_s_index==1 & !is.na(.$abb_s_index),] # N=104946

APLs %>%
  .[is.na(.$abbott_n) & .$abb_s_AU==1 & !is.na(.$abb_s_AU),] # N=69108

APLs %>%
  .[!is.na(.$abbott_n) & .$abb_s_index==TRUE & !is.na(.$abb_s_index),] # N=0

APLs %>%
  .[!is.na(.$abbott_n) & .$abb_s_AU==1 & !is.na(.$abb_s_AU),]  # N=40690

APLs %>%
  .[!is.na(.$abbott_n) & is.na(.$abb_s_AU),]                   # N=36

table(is.na(APLs[APLs$abb_s_AU==1,]$abbott_n))

# month of data collection for those five groups:
# Group 1: "Abbott-S" (Index) only
fivenum(APLs[APLs$abb_s_index==1,]$samplemonth, na.rm = T)

# Group 2: True Abbott-S (AU) only
fivenum(APLs[is.na(APLs$abbott_n) & APLs$abb_s_AU==1,]$samplemonth, na.rm = T)

# Group 3: True Abbott-S (AU) and Abbott-N?
fivenum(APLs[!is.na(APLs$abbott_n) & APLs$abb_s_AU==1,]$samplemonth, na.rm = T)

# Group 4: no available anti-S data, only Abbott-N
fivenum(APLs[!is.na(APLs$abbott_n) & is.na(APLs$abb_s_AU),]$samplemonth, na.rm = T)
```
**Notes**: So no matter what it really is it can't be Abbott-S-RBD and shall not be included in anti-S analysis.

## 2. Applying the COI

#### 2.1 Apply COI to APL (Abbott-S) data
```{r}
APLs$pos_abbott_s2<-rep(NA, 214780)
APLs$pos_abbott_s2[APLs$abb_s_AU==1 & !is.na(APLs$abb_s_AU) &
                   APLs$abbott_s>=402.36 & !is.na(APLs$abbott_s)]<-"Positive"
APLs$pos_abbott_s2[APLs$abb_s_AU==1 & !is.na(APLs$abb_s_AU) &
                     APLs$abbott_s<402.36 & !is.na(APLs$abbott_s)]<-"Negative"
table(APLs$pos_abbott_s2, useNA = 'ifany')

table(APLs$pos_abbott_s, APLs$pos_abbott_s2, useNA = 'ifany')
```
## 3. Making the Plot

## 3.1 Prepare the melt data
```{r}
# recode pos/neg
APLs$pos_abbott_s[APLs$pos_abbott_s=="Positive"]<-1
APLs$pos_abbott_s[APLs$pos_abbott_s=="Negative"]<-0
APLs$pos_abbott_s<-as.numeric(APLs$pos_abbott_s)
APLs$pos_abbott_s2[APLs$pos_abbott_s2=="Positive"]<-1
APLs$pos_abbott_s2[APLs$pos_abbott_s2=="Negative"]<-0
APLs$pos_abbott_s2<-as.numeric(APLs$pos_abbott_s2)

tb_abbs<-APLs %>%
  .[.$abb_s_AU==1,] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
            )
tb_abbs$prop.abbs1<-tb_abbs$pos_abbott_s/tb_abbs$n
tb_abbs$prop.abbs2<-tb_abbs$pos_abbott_s2/tb_abbs$n

# clean NAs
tb_abbs<-tb_abbs[!is.na(tb_abbs$samplemonth),]

tb_abbs_melt<-reshape2::melt(select(tb_abbs, -c(2:4)), 
                        id.vars = c("samplemonth"),
                        variable.name = "COI",
                        value.name = "prop")

p09<-tb_abbs_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
  geom_point(aes(colour=COI)) +
  geom_line(aes(colour=COI))  +
  theme_bw() 
p09
```

### 2024-05-17 2_scripts/2.4 PRNT50.qmd
Update PRNT50 analysis (unfinished)

@@
 ```
+### Apply COI to CBS (Roche-S) data
+```{r}
+
+```
+
 ## 3. Making the Plot
@@
                         value.name = "prop")
+```
 
+## 3.2 Making the Plot
+```{r}
+# APL
 p09<-tb_abbs_melt %>%

### 2024-05-21 2_scripts/2.4 PRNT50.qmd
Finished anti-S analysis

@@
 ```{r}
+CBSs$pos_roche_s<-CBS$interp_roche_s
+CBSs_AB<-CBSs[CBSs$province=="AB",]
 
+# CBS used Roche-S, one ob has missing values
+table(is.na(CBSs_AB$roche_s))
+
+# Roche COI: 0.8 & 178.81
+CBSs_AB$pos_roche_s2<-rep(NA, 124008)
+CBSs_AB$pos_roche_s2[CBSs_AB$roche_s>=178.81 & !is.na(CBSs_AB$roche_s)]<-1
+CBSs_AB$pos_roche_s2[CBSs_AB$roche_s< 178.81 & !is.na(CBSs_AB$roche_s)]<-0
+table(CBSs_AB$pos_roche_s2, useNA = 'ifany')
+
+# Comparison btw COI 0.8 & 178.81
+table(CBSs_AB$pos_roche_s, CBSs_AB$pos_roche_s2, useNA = 'ifany')
+```
+**Notes:** After increasing the COI from 0.8 to 178.81, 6732 of 108,514 (6.20%) samples were re-categorized as negative.
+The gap is smaller than what we observed in APL data, which has gaps of around 10%.
+
+#### 2.2 Roche-S <--> Abbott-S transformation
+
+#### 2.2.1 Apply Transformation
+```{r}
+# Abbott-S = 82.5 + 15.54*Roche-S
+# Source: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8552734/
+
+# APL
+APLs$roche_st<-(APLs$abbott_s-82.5)/15.54
+summary(APLs$roche_st)
+
+## add positivity, COI=0.8
+APLs$pos_roche_st<-rep(NA, 214780)
+APLs$pos_roche_st[APLs$abb_s_AU==1 & !is.na(APLs$abb_s_AU) &
+                   APLs$roche_st>=0.8 & !is.na(APLs$roche_st)]<-1
+APLs$pos_roche_st[APLs$abb_s_AU==1 & !is.na(APLs$abb_s_AU) &
+                     APLs$roche_st<0.8 & !is.na(APLs$roche_st)]<-0
+table(APLs$pos_roche_st, useNA = 'ifany')
+
+## add immunity, COI = 178.81
+APLs$pos_roche_st2<-rep(NA, 214780)
+APLs$pos_roche_st2[APLs$abb_s_AU==1 & !is.na(APLs$abb_s_AU) &
+                   APLs$roche_st>=178.81 & !is.na(APLs$roche_st)]<-1
+APLs$pos_roche_st2[APLs$abb_s_AU==1 & !is.na(APLs$abb_s_AU) &
+                     APLs$roche_st<178.81 & !is.na(APLs$roche_st)]<-0
+table(APLs$pos_roche_st, APLs$pos_roche_st2, useNA = 'ifany')
+```
+```{r}
+# CBS
+CBSs_AB$abbott_st<-82.5+CBSs_AB$roche_s*15.54
+summary(CBSs_AB$abbott_st)
+
+## add positivity, COI=50, all positive (COI lower than min after transformation)
+CBSs_AB$pos_abbott_st<-rep(NA, 124008)
+CBSs_AB$pos_abbott_st[CBSs_AB$abbott_st>=50 & !is.na(CBSs_AB$abbott_st)]<-1
+CBSs_AB$pos_abbott_st[CBSs_AB$abbott_st<50 & !is.na(CBSs_AB$abbott_st)]<-0
+table(CBSs_AB$pos_abbott_st, useNA = 'ifany')
+
+## add immunity, COI = 178.81
+CBSs_AB$pos_abbott_st2<-rep(NA, 124008)
+CBSs_AB$pos_abbott_st2[CBSs_AB$abbott_st>=402.36 & !is.na(CBSs_AB$abbott_st)]<-1
+CBSs_AB$pos_abbott_st2[CBSs_AB$abbott_st<402.36 & !is.na(CBSs_AB$abbott_st)]<-0
+table(CBSs_AB$pos_abbott_st, CBSs_AB$pos_abbott_st2, useNA = 'ifany')
+```
+#### 2.2.2 Creating merged dta
+```{r}
+tb_abbst<-APLs %>%
+  .[.$abb_s_AU==1,] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_roche_st= sum(pos_roche_st, na.rm = T), 
+            pos_roche_st2= sum(pos_roche_st2, na.rm = T)
+            )
+tb_abbst$prop.rost1<-tb_abbst$pos_roche_st/tb_abbst$n
+tb_abbst$prop.rost2<-tb_abbst$pos_roche_st2/tb_abbst$n
+
+# merge abbs & abbst table
+tb_abbs<-merge(tb_abbs, select(tb_abbst, -c(2:4)), all=T, by="samplemonth")
+
+# clean NAs
+tb_abbs<-tb_abbs[!is.na(tb_abbs$samplemonth),]
+
+tb_melt_t1<-reshape2::melt(select(tb_abbs, samplemonth, prop.abbs1, prop.rost1, study), 
+                        id.vars = c("samplemonth", "study"),
+                        variable.name = "COI",
+                        value.name = "prop")
+
+tb_melt_t2<-reshape2::melt(select(tb_abbs, samplemonth, prop.abbs2, prop.rost2, study), 
+                        id.vars = c("samplemonth", "study"),
+                        variable.name = "COI",
+                        value.name = "prop")
+```
+
+```{r}
+tb_rost<-CBSs_AB %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_abbott_st= sum(pos_abbott_st, na.rm = T), 
+            pos_abbott_st2= sum(pos_abbott_st2, na.rm = T)
+            )
+tb_rost$prop.abbst1<-tb_rost$pos_abbott_st/tb_rost$n
+tb_rost$prop.abbst2<-tb_rost$pos_abbott_st2/tb_rost$n
+
+# merge abbs & abbst table
+tb_ros<-merge(tb_ros, select(tb_rost, -c(2:4)), all=T, by="samplemonth")
+
+# clean NAs
+tb_ros<-tb_ros[!is.na(tb_ros$samplemonth),]
+
+tb_melt_t3<-reshape2::melt(select(tb_ros, samplemonth, prop.ros1, prop.abbst1, study),
+                        id.vars = c("samplemonth", "study"),
+                        variable.name = "COI",
+                        value.name = "prop")
+tb_melt_t4<-reshape2::melt(select(tb_ros, samplemonth, prop.ros2, prop.abbst2, study),
+                        id.vars = c("samplemonth", "study"),
+                        variable.name = "COI",
+                        value.name = "prop")
 ```
@@
 
-## 3.1 Prepare the melt data
+## 3.1.1 Prepare the melt data: APL
 ```{r}
@@
 
+## 3.1.2 Prepare the melt data: CBS
+```{r}
+tb_ros<-CBSs_AB %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_roche_s= sum(pos_roche_s, na.rm = T), 
+            pos_roche_s2= sum(pos_roche_s2, na.rm = T)
+            )
+tb_ros$prop.ros1<-tb_ros$pos_roche_s/tb_ros$n
+tb_ros$prop.ros2<-tb_ros$pos_roche_s2/tb_ros$n
+
+tb_ros_melt<-reshape2::melt(select(tb_ros, -c(2:4)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "COI",
+                        value.name = "prop")
+```
+
 ## 3.2 Making the Plot
+
+#### 3.2.1 Making the plot seperately for CBS & APL
 ```{r}
@@
 p09
+
+# CBS
+p09b<-tb_ros_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
+  geom_point(aes(colour=COI)) +
+  geom_line(aes(colour=COI))  +
+  theme_bw() 
+p09b
 ```
+#### 3.2.2 Making APL & CBS in one plot
+```{r}
+# creating merged summary tab
+tb_abbs_melt$study<-rep("APL", length(tb_abbs_melt$samplemonth))
+tb_ros_melt$study<-rep("CBS", length(tb_ros_melt$samplemonth))
+
+tb_APLCBS2<-rbind.data.frame(tb_abbs_melt, tb_ros_melt)
+
+tb_APLCBS2$COI<-as.character(tb_APLCBS2$COI)
+tb_APLCBS2$COI[tb_APLCBS2$COI=="prop.abbs1"]<-"50"
+tb_APLCBS2$COI[tb_APLCBS2$COI=="prop.abbs2"]<-"402.36"
+tb_APLCBS2$COI[tb_APLCBS2$COI=="prop.ros1"]<-"0.8"
+tb_APLCBS2$COI[tb_APLCBS2$COI=="prop.ros2"]<-"178.81"
+
+tb_APLCBS2$COI_ty[tb_APLCBS2$COI %in% c("50", "0.8")]<-"Positivity"
+tb_APLCBS2$COI_ty[tb_APLCBS2$COI %in% c("402.36", "178.81")]<-"Immunity"
+```
+
+```{r}
+# ggplot
+p09.lab<-c("APL-Abbott (COI=402)", "CBS-Roche (COI=179)", 
+           "APL-Abbott (COI=50)", "CBS-Roche (COI=0.8)")
+p09c<-tb_APLCBS2 %>%
+  ggplot(aes(x=samplemonth, y=prop, colour=interaction(study, COI_ty),
+                                    linetype=interaction(study, COI_ty))) +
+  geom_point() +
+  geom_line()  +
+  theme_bw() +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_colour_manual("", values = c("red", "blue", "red", "blue"),
+                      labels = p09.lab)+
+  scale_linetype_manual("", values=c(3,3,1,1), labels = p09.lab) 
+#  theme(legend.position="top")+
+#  guides(color = guide_legend(nrow = 2))
+
+p09c
+```
+#### 3.2.3 Plot for transformed curved
+```{r}
+# Plot 1
+p09d.dta<-rbind.data.frame(tb_melt_t1, tb_melt_t3)
+
+p09d.dta$COI<-as.character(p09d.dta$COI)
+p09d.dta$COI_ty[p09d.dta$COI %in% c("prop.abbs1", "prop.ros1")]<-"Raw"
+p09d.dta$COI_ty[p09d.dta$COI %in% c("prop.abbst1", "prop.rost1")]<-"Transformed"
+
+p09d.lab<-c("APL-Abbott (COI=50)", "CBS-Roche (COI=0.8)", 
+           "APL-Roche (COI=0.8)", "CBS-Abbott (COI=50)")
+p09d<-p09d.dta %>%
+  ggplot(aes(x=samplemonth, y=prop, colour=interaction(study, COI_ty),
+                                    linetype=interaction(study, COI_ty))) +
+  geom_point() +
+  geom_line()  +
+  theme_bw() +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_colour_manual("", values = c("red", "blue", "red", "blue"),
+                      labels = p09d.lab)+
+  scale_linetype_manual("", values=c(1,1,3,3), labels = p09d.lab) 
+#  theme(legend.position="top")+
+#  guides(color = guide_legend(nrow = 2))
+
+p09d
+```
+
+```{r}
+# Plot 2
+p09e.dta<-rbind.data.frame(tb_melt_t2, tb_melt_t4)
+
+p09e.dta$COI<-as.character(p09e.dta$COI)
+p09e.dta$COI_ty[p09e.dta$COI %in% c("prop.abbs2", "prop.ros2")]<-"Raw"
+p09e.dta$COI_ty[p09e.dta$COI %in% c("prop.abbst2", "prop.rost2")]<-"Transformed"
+
+p09e.lab<-c("APL-Abbott (COI=402)", "CBS-Roche (COI=179)", 
+           "APL-Roche (COI=179)", "CBS-Abbott (COI=402)")
+p09e<-p09e.dta %>%
+  ggplot(aes(x=samplemonth, y=prop, colour=interaction(study, COI_ty),
+                                    linetype=interaction(study, COI_ty))) +
+  geom_point() +
+  geom_line()  +
+  theme_bw() +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_colour_manual("", values = c("red", "blue", "red", "blue"),
+                      labels = p09e.lab)+
+  scale_linetype_manual("", values=c(1,1,3,3), labels = p09e.lab) 
+#  theme(legend.position="top")+
+#  guides(color = guide_legend(nrow = 2))
+
+p09e
+```

### 2024-05-27 2_scripts/2.4 PRNT50.qmd
Anti-S sensitivity analysis update

@@
 ```
+## 4 Misc Analysis
+
+#### 4.1 Sensitivity Analysis
+```{r}
+# Add vax variable
+APLs<-merge(APLs, select(RFD4682_e, order_ID, VACCINATION_STATUS),
+      by="order_ID", all=T)
+
+APLs$vax<-APLs$VACCINATION_STATUS
+APLs<-select(APLs, -c("VACCINATION_STATUS"))
+table(APLs$vax, useNA = 'ifany')
+
+# Add days since imm (vax)
+APLs<-merge(APLs, select(RFD4682_e, order_ID, DAYS_SINCE_IMM),
+      by="order_ID", all=T)
+```
+
+```{r}
+# re-do the melt data
+tb_abbsv<-APLs %>%
+  .[.$abb_s_AU==1 & !(.$vax %in% c("Not Immunized", "Pre-Immunization")) &
+    .$DAYS_SINCE_IMM>=28 & !is.na(.$DAYS_SINCE_IMM),] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
+            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
+            )
+tb_abbsv$prop.abbs1<-tb_abbsv$pos_abbott_s/tb_abbsv$n
+tb_abbsv$prop.abbs2<-tb_abbsv$pos_abbott_s2/tb_abbsv$n
+
+# clean NAs
+tb_abbsv<-tb_abbsv[!is.na(tb_abbsv$samplemonth),]
+
+tb_abbsv_melt<-reshape2::melt(select(tb_abbsv, -c(2:4)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "COI",
+                        value.name = "prop")
+```
+
+```{r}
+# re-do the plot
+# APL vax only
+p09v0<-tb_abbsv_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
+  geom_point(aes(colour=COI)) +
+  geom_line(aes(colour=COI))  +
+  theme_bw() + 
+  ylim(c(0,1))
+p09v0
+```
+## 4.2 The vaccination rate among APL
+The APL data Carmen shared with us has vaccination info, the 2024 Jan ver of it even has the vax brand.
+While the CBS data we have doesnot include vax info.
+However, a paper written by Sheila has the vax% of those in CBS data, which we can use for comparison on vax%.
+
+```{r}
+# create vax% cat var
+APLs$vax_cat<-rep(NA, 214780)
+APLs$vax_cat[APLs$vax %in% c("1 Dose","1-2 Doses","2 Doses","3 Doses",
+                             "4 Doses")]<-1
+APLs$vax_cat[APLs$vax %in% c("Not Immunized", "Pre-Immunization")]<-0
+table(APLs$vax, APLs$vax_cat, useNA = 'ifany')
+```
+```{r}
+# create merged data
+tb_vax<-APLs %>%
+  .[.$abb_s_AU==1,] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            n_vax= sum(vax_cat, na.rm = T)
+            )
+tb_vax$prop.vax<-tb_vax$n_vax/tb_vax$n
+
+# clean NAs
+tb_vax<-tb_vax[!is.na(tb_vax$samplemonth),]
+
+# make the plot
+p10<-tb_vax %>%
+  ggplot(aes(x=samplemonth, y=prop.vax)) +
+  geom_point() +
+  geom_line()  +
+  geom_smooth(colour="red") +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  ylab("Vaccinated%")
+p10
+```
+**Alton's comments:** "It might be worth comparing to reported vax rate for Alberta similar to what Sheila and Steve did in that figure." As they used (S+ N-)% to estimate the vax%, it will inevitably results in underestimation, as those who are vaccinated and infected was not counted.
+
+```{r}
+# create an addition indicator for S+ and N-
+table(APLs$pos_abbott, APLs$pos_abbott_s, useNA = 'ifany')
+
+APLs$vax_est<-rep(0, 214780)
+APLs$vax_est[APLs$pos_abbott==0 & APLs$pos_abbott_s==1]<-1
+APLs$vax_est[is.na(APLs$pos_abbott) | is.na(APLs$pos_abbott_s)]<-NA
+table(APLs$vax_est)
+```
+```{r}
+tb_vax_est<-APLs %>%
+  .[.$abb_s_AU==1 & !is.na(.$pos_abbott_s) & !is.na(.$pos_abbott),] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            n_vax= sum(vax_est, na.rm = T)
+            )
+tb_vax_est$prop.vax<-tb_vax_est$n_vax/tb_vax_est$n
+
+# clean NAs
+tb_vax_est<-tb_vax_est[!is.na(tb_vax_est$samplemonth),]
+
+# make the plot
+p11<-tb_vax_est %>%
+  ggplot(aes(x=samplemonth, y=prop.vax)) +
+  geom_point() +
+  geom_line()  +
+  geom_smooth(colour="red") +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  ylab("Vaccinated% Est by anti-S&N")
+p11
+```

### 2024-05-28 2_scripts/2.4 PRNT50.qmd
Vax% estimation for PRNT50 (unfinished)

--- 2024-05-21 2_scripts/2.4 PRNT50.qmd
@@
 ## 1. Clean the APL data
+
+#### 1.1 Use the APLs created
 ```{r}
@@
 **Notes**: So no matter what it really is it can't be Abbott-S-RBD and shall not be included in anti-S analysis.
+Updates on May 28th: Carmen replied and confirmed they are misplaced anti-N assay results.
+
+#### 1.2 Move the misplaced anti-N back to where they are
+```{r}
+APLs$abbott_n[APLs$abb_s_index==1 &
+              !is.na(APLs$abbott_s)]<-APLs$abbott_s[APLs$abb_s_index==1 & 
+                                                   !is.na(APLs$abbott_s)]
+APLs$abbott_s[APLs$abb_s_index==1 & !is.na(APLs$abbott_s)]<-NA
+
+# add anti-N interpretation
+APLs$pos_abbott[APLs$abbott_n<1.4 & !is.na(APLs$abbott_n)]<-0
+APLs$pos_abbott[APLs$abbott_n>=1.4 & !is.na(APLs$abbott_n)]<-1
+
+# redo the anti-S interpretation
+APLs$pos_abbott_s<-rep(NA,214780)
+APLs$pos_abbott_s[APLs$abbott_s<50 & !is.na(APLs$abbott_s)]<-0
+APLs$pos_abbott_s[APLs$abbott_s>=50 & !is.na(APLs$abbott_s)]<-1
+table(APLs$pos_abbott_s, useNA = 'ifany')
+```
 
@@
 ```
+## 4 Misc Analysis
+
+#### 4.1 Sensitivity Analysis
+```{r}
+# Add vax variable
+APLs<-merge(APLs, select(RFD4682_e, order_ID, VACCINATION_STATUS),
+      by="order_ID", all=T)
+
+APLs$vax<-APLs$VACCINATION_STATUS
+APLs<-select(APLs, -c("VACCINATION_STATUS"))
+table(APLs$vax, useNA = 'ifany')
+
+# Add days since imm (vax)
+APLs<-merge(APLs, select(RFD4682_e, order_ID, DAYS_SINCE_IMM),
+      by="order_ID", all=T)
+```
+**Notes:** In this part we also do the analysis for those got vaccinated >=28 days, however it does not change the shape of the lines so much. So we just group by doses (0/>=1).
+
+```{r}
+# re-do the melt data: no vax
+tb_abbsv0<-APLs %>%
+  .[.$abb_s_AU==1 & (.$vax %in% c("Not Immunized", "Pre-Immunization")),] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
+            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
+            )
+tb_abbsv0$prop.abbs1<-tb_abbsv0$pos_abbott_s/tb_abbsv0$n
+tb_abbsv0$prop.abbs2<-tb_abbsv0$pos_abbott_s2/tb_abbsv0$n
+
+# clean NAs
+tb_abbsv0<-tb_abbsv0[!is.na(tb_abbsv0$samplemonth),]
+
+tb_abbsv0_melt<-reshape2::melt(select(tb_abbsv0, -c(2:4)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "COI",
+                        value.name = "prop")
+
+# re-do the melt data: at least 1 dose
+tb_abbsv1<-APLs %>%
+  .[.$abb_s_AU==1 & .$vax %in% c("1 Dose","1-2 Doses","2 Doses","3 Doses",
+                             "4 Doses") ,] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
+            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
+            )
+tb_abbsv1$prop.abbs1<-tb_abbsv1$pos_abbott_s/tb_abbsv1$n
+tb_abbsv1$prop.abbs2<-tb_abbsv1$pos_abbott_s2/tb_abbsv1$n
+
+# clean NAs
+tb_abbsv1<-tb_abbsv1[!is.na(tb_abbsv1$samplemonth),]
+
+tb_abbsv1_melt<-reshape2::melt(select(tb_abbsv1, -c(2:4)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "COI",
+                        value.name = "prop")
+```
+
+```{r}
+# re-do the plot
+# APL: not vaccinated
+p09v0<-tb_abbsv0_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
+  geom_point(colour="red") +
+  geom_line(aes(linetype=COI), colour="red")  +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_linetype_manual("", values=c(1,3), labels = c("APL-Abbott (COI=50)", 
+                                                      "APL-Abbott (COI=402)"))+
+  theme(legend.position='top') + 
+  ggtitle("Stratum: not vaccinated")
+p09v0
+```
+```{r}
+# Plot: at least 1 dose
+p09v1<-tb_abbsv1_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
+  geom_point(colour="red") +
+  geom_line(aes(linetype=COI), colour="red")  +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_linetype_manual("", values=c(1,3), labels = c("APL-Abbott (COI=50)", 
+                                                      "APL-Abbott (COI=402)"))+
+  theme(legend.position='top')+
+  ggtitle("Stratum: at least 1 dose")
+p09v1
+```
+#### Figure panel
+```{r}
+# use cowplot package
+p09all.upper<-cowplot::plot_grid(p09c + ylab("Proportion") +
+                           theme(legend.position='top') +
+                           guides(color = guide_legend(nrow = 2)), 
+                           p09v0 + xlab ("") + ylab(""), nrow=1,
+                           labels = c("A", "B"),  align = "h")
+p09all.upper
+
+p09all.lower<-cowplot::plot_grid(
+                           NULL,  p09v1 + theme(legend.position="none"), nrow=1,
+                           labels = c("", "C"))
+p09all.lower
+
+png(file="../4_output/figs/Fig.4 Immunity.png", 
+    width=2400, height=2000, res=300)
+cowplot::plot_grid(p09all.upper, p09all.lower, nrow=2, rel_heights = c(1, 0.8))
+dev.off()
+```
+
+## 4.2 The vaccination rate among APL
+The APL data Carmen shared with us has vaccination info, the 2024 Jan ver of it even has the vax brand.
+While the CBS data we have doesnot include vax info.
+However, a paper written by Sheila has the vax% of those in CBS data, which we can use for comparison on vax%.
+
+#### 4.2.1 Vax rate by APL vax data
+```{r}
+# create vax% cat var
+APLs$vax_cat<-rep(NA, 214780)
+APLs$vax_cat[APLs$vax %in% c("1 Dose","1-2 Doses","2 Doses","3 Doses",
+                             "4 Doses")]<-1
+APLs$vax_cat[APLs$vax %in% c("Not Immunized", "Pre-Immunization")]<-0
+table(APLs$vax, APLs$vax_cat, useNA = 'ifany')
+```
+```{r}
+# create merged data
+tb_vax<-APLs %>%
+  .[.$abb_s_AU==1,] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            n_vax= sum(vax_cat, na.rm = T)
+            )
+tb_vax$prop.vax<-tb_vax$n_vax/tb_vax$n
+
+# clean NAs
+tb_vax<-tb_vax[!is.na(tb_vax$samplemonth),]
+
+# make the plot
+p10<-tb_vax %>%
+  ggplot(aes(x=samplemonth, y=prop.vax)) +
+  geom_point() +
+  geom_line()  +
+  geom_smooth(colour="red") +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  ylab("Vaccinated%")
+p10
+```
+**Alton's comments:** "It might be worth comparing to reported vax rate for Alberta similar to what Sheila and Steve did in that figure." As they used (S+ N-)% to estimate the vax%, it will inevitably results in underestimation, as those who are vaccinated and infected was not counted.
+
+#### 4.2.2 Vax by anti-S and anti-N
+```{r}
+# create an addition indicator for S+ and N-
+table(APLs$pos_abbott, APLs$pos_abbott_s, useNA = 'ifany')
+
+APLs$vax_est<-rep(0, 214780)
+APLs$vax_est[APLs$pos_abbott==0 & APLs$pos_abbott_s==1]<-1
+APLs$vax_est[is.na(APLs$pos_abbott) | is.na(APLs$pos_abbott_s)]<-NA
+table(APLs$vax_est)
+```
+```{r}
+tb_vax_est<-APLs %>%
+#  .[!is.na(.$pos_abbott_s) & !is.na(.$pos_abbott),] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            n_vax= sum(vax_est, na.rm = T)
+            )
+tb_vax_est$prop.vax<-tb_vax_est$n_vax/tb_vax_est$n
+
+# clean NAs
+tb_vax_est<-tb_vax_est[!is.na(tb_vax_est$samplemonth),]
+
+# make the plot
+p11<-tb_vax_est %>%
+  ggplot(aes(x=samplemonth, y=prop.vax)) +
+  geom_point() +
+  geom_line()  +
+  geom_smooth(colour="red") +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  ylab("Vaccinated% Est by anti-S&N")
+p11
+```
+**Notes:** Estimating the vax% using the anti-S (+) and anti-N (-) assay results.
+For our analysis, this way of estimation might be problematic.
+We don't have pre-Omicron wave anti-N data in APL  and the prevalence of ever infected (S+ & N+) raised sharply in 2022.
+This will result underestimated vax% in 2022, as presented in the fig above.
+
+#### 4.3.3 The Vax rate in AB general population
+```{r}
+# load the data from Canada Statistics
+vax_AB<-read.csv("https://health-infobase.canada.ca/src/data/covidLive/vaccination-coverage-map.csv")
+
+# create AB subset and clean the data
+vax_AB<-vax_AB[vax_AB$prename=="Alberta",]
+vax_ABs<-vax_AB %>%
+  select(week_end, numtotal_atleast1dose, proptotal_atleast1dose)
+colnames(vax_ABs)<-c("weekend", "num.1dose", "prop.1dose")
+vax_ABs$weekend<-as.yearmon(vax_ABs$weekend)
+
+# calculate monthly mean prop for vax rate
+tb_vax_AB<-vax_ABs %>%
+  group_by(weekend) %>%
+  summarise(ave.1dose = mean(prop.1dose))
+
+tb_vax_AB<-tb_vax_AB[!is.na(tb_vax_AB$ave.1dose),]
+tb_vax_AB$data<-rep("AB_general", length(tb_vax_AB$weekend))
+
+d01<-tb_vax_AB
+colnames(d01)<-c("samplemonth", "prop.vax", "data")
+```
+
+##### creating a merged data
+```{r}
+# Estimated vaccination rate: by anti-S + and anti-N neg
+View(tb_vax_est)
+d02<-select(tb_vax_est, c(1,4))
+d02$data<-rep("S_pos_N_Neg", length(d02$samplemonth))
+
+# Estimated vaccination rate: by anti-S only
+View(tb_abbs)
+d03<-select(tb_abbs, samplemonth, prop.abbs1, study)
+colnames(d03)<-c("samplemonth", "prop.vax", "data")
+d03$data<-rep("S_pos", length(d03$samplemonth))
+
+# Reported vax rate in APL
+View(tb_vax)
+d04<-select(tb_vax, c(1,4))
+d04$data<-rep("APL_vax", length(d04$samplemonth))
+
+# merge all
+dta.vax<-rbind.data.frame(d01, d02, d03, d04)
+rm(d01, d02, d03, d04)
+
+dta.vax$prop.vax[dta.vax$data=="AB_general"]<-dta.vax$prop.vax[dta.vax$data=="AB_general"]/100
+```
+
+##### Making the plot
+```{r}
+p12<-dta.vax %>%
+  ggplot(aes(x=samplemonth, y=prop.vax, group=data, colour=data)) +
+  geom_point(aes(colour=data)) +
+  geom_line(aes(colour=data))  +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  ylab("Vaccinated%")
+p12
+```

### 2024-06-03 2_scripts/2.4 PRNT50.qmd
Add a fig with major vax milestones in AB

@@
 ```
+#### Add time indicator to the figure
+```{r}
+# time when vax is available to Alberta health care workers and first responders
+# Jan: medical workers
+# Feb: Phase-1: age 75+
+# Mar: Phase-2: group A
+# Apr: Phase-2: group B, C
+# May: Phase-3: age 12+
+
+p09v1 + 
+  geom_vline(xintercept = as.yearmon(c("2021-01", "2021-02", "2021-03",
+                                       "2021-04", "2021-05"))
+             , linetype="dotted", size = 1.0, color="grey") +
+  geom_text(aes(x=as.yearmon("Jan 2021"), label="\nHealth Care Workers", y=0.25), 
+            colour="navy", angle=90)+
+  geom_text(aes(x=as.yearmon("Feb 2021"), label="\nPhase1: aged 75+", y=0.25), 
+            colour="navy", angle=90)+
+  geom_text(aes(x=as.yearmon("Mar 2021"), label="\nPhase2A: aged 65+", y=0.25), 
+            colour="navy", angle=90)+
+  geom_text(aes(x=as.yearmon("Apr 2021"), label="\nPhase2B, 2C: aged 18+ with risks", y=0.40), 
+            colour="navy", angle=90)+
+  geom_text(aes(x=as.yearmon("May 2021"), label="\nPhase3: aged 12+", y=0.25), 
+            colour="navy", angle=90)
+```
+
 #### Figure panel
@@
 
-#### 4.3.3 The Vax rate in AB general population
+#### 4.2.3 The Vax rate in AB general population
 ```{r}

### 2024-06-04 2_scripts/2.4 PRNT50.qmd
Update anti-S stratified analysis (unfinished)

--- 2024-05-27 2_scripts/2.4 PRNT50.qmd
@@
 ## 1. Clean the APL data
+
+#### 1.1 Use the APLs created
+
 ```{r}
@@
 ```
+
 **Notes**: So no matter what it really is it can't be Abbott-S-RBD and shall not be included in anti-S analysis.
+Updates on May 28th: Carmen replied and confirmed they are misplaced anti-N assay results.
+
+#### 1.2 Move the misplaced anti-N back to where they are
+
+```{r}
+APLs$abbott_n[APLs$abb_s_index==1 &
+              !is.na(APLs$abbott_s)]<-APLs$abbott_s[APLs$abb_s_index==1 & 
+                                                   !is.na(APLs$abbott_s)]
+APLs$abbott_s[APLs$abb_s_index==1 & !is.na(APLs$abbott_s)]<-NA
+
+# add anti-N interpretation
+APLs$pos_abbott[APLs$abbott_n<1.4 & !is.na(APLs$abbott_n)]<-0
+APLs$pos_abbott[APLs$abbott_n>=1.4 & !is.na(APLs$abbott_n)]<-1
+
+# redo the anti-S interpretation
+APLs$pos_abbott_s<-rep(NA,214780)
+APLs$pos_abbott_s[APLs$abbott_s<50 & !is.na(APLs$abbott_s)]<-0
+APLs$pos_abbott_s[APLs$abbott_s>=50 & !is.na(APLs$abbott_s)]<-1
+table(APLs$pos_abbott_s, useNA = 'ifany')
+```
 
@@
 #### 2.1 Apply COI to APL (Abbott-S) data
+
 ```{r}
@@
 ```
+
 ### Apply COI to CBS (Roche-S) data
+
 ```{r}
@@
 ```
+
 **Notes:** After increasing the COI from 0.8 to 178.81, 6732 of 108,514 (6.20%) samples were re-categorized as negative.
@@
 
-#### 2.2 Roche-S <--> Abbott-S transformation
+#### 2.2 Roche-S \<--\> Abbott-S transformation
 
 #### 2.2.1 Apply Transformation
+
 ```{r}
@@
 ```
+
 #### 2.2.2 Creating merged dta
+
 ```{r}
@@
 ## 3.1.1 Prepare the melt data: APL
+
 ```{r}
@@
 ## 3.1.2 Prepare the melt data: CBS
+
 ```{r}
@@
 #### 3.2.1 Making the plot seperately for CBS & APL
+
 ```{r}
@@
 ```
+
 #### 3.2.2 Making APL & CBS in one plot
+
 ```{r}
@@
 ```
+
 #### 3.2.3 Plot for transformed curved
+
 ```{r}
@@
 ```
+
 ## 4 Misc Analysis
@@
 #### 4.1 Sensitivity Analysis
+
 ```{r}
@@
 
+**Notes:** In this part we also do the analysis for those got vaccinated \>=28 days, however it does not change the shape of the lines so much. So we just group by doses (0/\>=1).
+
 ```{r}
-# re-do the melt data
-tb_abbsv<-APLs %>%
-  .[.$abb_s_AU==1 & !(.$vax %in% c("Not Immunized", "Pre-Immunization")) &
-    .$DAYS_SINCE_IMM>=28 & !is.na(.$DAYS_SINCE_IMM),] %>%
+# re-do the melt data: no vax
+tb_abbsv0<-APLs %>%
+  .[.$abb_s_AU==1 & (.$vax %in% c("Not Immunized", "Pre-Immunization", "1 Dose")),] %>%
   group_by(samplemonth) %>%
@@
             )
-tb_abbsv$prop.abbs1<-tb_abbsv$pos_abbott_s/tb_abbsv$n
-tb_abbsv$prop.abbs2<-tb_abbsv$pos_abbott_s2/tb_abbsv$n
+tb_abbsv0$prop.abbs1<-tb_abbsv0$pos_abbott_s/tb_abbsv0$n
+tb_abbsv0$prop.abbs2<-tb_abbsv0$pos_abbott_s2/tb_abbsv0$n
 
 # clean NAs
-tb_abbsv<-tb_abbsv[!is.na(tb_abbsv$samplemonth),]
+tb_abbsv0<-tb_abbsv0[!is.na(tb_abbsv0$samplemonth),]
 
-tb_abbsv_melt<-reshape2::melt(select(tb_abbsv, -c(2:4)), 
+tb_abbsv0_melt<-reshape2::melt(select(tb_abbsv0, -c(2:4)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "COI",
+                        value.name = "prop")
+
+# re-do the melt data: at least 1 dose
+tb_abbsv1<-APLs %>%
+  .[.$abb_s_AU==1 & .$vax %in% c("1-2 Doses","2 Doses","3 Doses",
+                             "4 Doses") ,] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
+            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
+            )
+tb_abbsv1$prop.abbs1<-tb_abbsv1$pos_abbott_s/tb_abbsv1$n
+tb_abbsv1$prop.abbs2<-tb_abbsv1$pos_abbott_s2/tb_abbsv1$n
+
+# clean NAs
+tb_abbsv1<-tb_abbsv1[!is.na(tb_abbsv1$samplemonth),]
+
+tb_abbsv1_melt<-reshape2::melt(select(tb_abbsv1, -c(2:4)), 
                         id.vars = c("samplemonth"),
@@
 # re-do the plot
-# APL vax only
-p09v0<-tb_abbsv_melt %>%
+# APL: not vaccinated
+p09v0<-tb_abbsv0_melt %>%
   ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
-  geom_point(aes(colour=COI)) +
-  geom_line(aes(colour=COI))  +
+  geom_point(colour="red") +
+  geom_line(aes(linetype=COI), colour="red")  +
   theme_bw() + 
-  ylim(c(0,1))
+  ylim(c(0,1)) +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_linetype_manual("", values=c(1,3), labels = c("APL-Abbott (COI=50)", 
+                                                      "APL-Abbott (COI=402)"))+
+  theme(legend.position='top') + 
+  ggtitle("Stratum: not vaccinated or 1 dose only")
 p09v0
 ```
+
+```{r}
+# Plot: at least 1 dose
+p09v1<-tb_abbsv1_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
+  geom_point(colour="red") +
+  geom_line(aes(linetype=COI), colour="red")  +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_linetype_manual("", values=c(1,3), labels = c("APL-Abbott (COI=50)", 
+                                                      "APL-Abbott (COI=402)"))+
+  theme(legend.position='top')+
+  ggtitle("Stratum: at least 2 doses")
+p09v1
+```
+##### Is 1-doser the cause of the "peak" and "bucket"?
+```{r}
+tb_abbsv11<-APLs %>%
+  .[.$abb_s_AU==1 & .$vax %in% c("1 Dose") ,] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
+            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
+            )
+tb_abbsv11$prop.abbs1<-tb_abbsv11$pos_abbott_s/tb_abbsv11$n
+tb_abbsv11$prop.abbs2<-tb_abbsv11$pos_abbott_s2/tb_abbsv11$n
+
+# clean NAs
+tb_abbsv11<-tb_abbsv11[!is.na(tb_abbsv11$samplemonth),]
+
+tb_abbsv11_melt<-reshape2::melt(select(tb_abbsv11, -c(2:4)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "COI",
+                        value.name = "prop")
+```
+
+```{r}
+# Plot: 1 dose only
+p09v11<-tb_abbsv11_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
+  geom_point(colour="red") +
+  geom_line(aes(linetype=COI), colour="red")  +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_linetype_manual("", values=c(1,3), labels = c("APL-Abbott (COI=50)", 
+                                                      "APL-Abbott (COI=402)"))+
+  theme(legend.position='top')+
+  ggtitle("Stratum: 1 dose only")
+p09v11
+```
+#### Add time indicator to the figure
+
+```{r}
+# time when vax is available to Alberta health care workers and first responders
+# Jan: medical workers
+# Feb: Phase-1: age 75+
+# Mar: Phase-2: group A
+# Apr: Phase-2: group B, C
+# May: Phase-3: age 12+
+
+p09v1 + 
+  geom_vline(xintercept = as.yearmon(c("2021-01", "2021-02", "2021-03",
+                                       "2021-04", "2021-05"))
+             , linetype="dotted", size = 1.0, color="grey") +
+  geom_text(aes(x=as.yearmon("Jan 2021"), label="\nHealth Care Workers", y=0.25), 
+            colour="navy", angle=90)+
+  geom_text(aes(x=as.yearmon("Feb 2021"), label="\nPhase1: aged 75+", y=0.25), 
+            colour="navy", angle=90)+
+  geom_text(aes(x=as.yearmon("Mar 2021"), label="\nPhase2A: aged 65+", y=0.25), 
+            colour="navy", angle=90)+
+  geom_text(aes(x=as.yearmon("Apr 2021"), label="\nPhase2B, 2C: aged 18+ with risks", y=0.40), 
+            colour="navy", angle=90)+
+  geom_text(aes(x=as.yearmon("May 2021"), label="\nPhase3: aged 12+", y=0.25), 
+            colour="navy", angle=90)
+```
+
+#### Figure panel
+
+```{r}
+# use cowplot package
+p09all.upper<-cowplot::plot_grid(p09c + ylab("Proportion") +
+                           theme(legend.position='top') +
+                           guides(color = guide_legend(nrow = 2)), 
+                           p09v0 + xlab ("") + ylab(""), nrow=1,
+                           labels = c("A", "B"),  align = "h")
+p09all.upper
+
+p09all.lower<-cowplot::plot_grid(
+                           NULL,  p09v1 + theme(legend.position="none"), nrow=1,
+                           labels = c("", "C"))
+p09all.lower
+
+png(file="../4_output/figs/Fig.4 Immunity.png", 
+    width=2400, height=2000, res=300)
+cowplot::plot_grid(p09all.upper, p09all.lower, nrow=2, rel_heights = c(1, 0.8))
+dev.off()
+```
+
 ## 4.2 The vaccination rate among APL
+
 The APL data Carmen shared with us has vaccination info, the 2024 Jan ver of it even has the vax brand.
@@
 However, a paper written by Sheila has the vax% of those in CBS data, which we can use for comparison on vax%.
+
+#### 4.2.1 Vax rate by APL vax data
 
@@
 ```
+
 **Alton's comments:** "It might be worth comparing to reported vax rate for Alberta similar to what Sheila and Steve did in that figure." As they used (S+ N-)% to estimate the vax%, it will inevitably results in underestimation, as those who are vaccinated and infected was not counted.
+
+#### 4.2.2 Vax by anti-S and anti-N
 
@@
 ```
+
 ```{r}
 tb_vax_est<-APLs %>%
-  .[.$abb_s_AU==1 & !is.na(.$pos_abbott_s) & !is.na(.$pos_abbott),] %>%
+#  .[!is.na(.$pos_abbott_s) & !is.na(.$pos_abbott),] %>%
   group_by(samplemonth) %>%
@@
 ```
+
+**Notes:** Estimating the vax% using the anti-S (+) and anti-N (-) assay results.
+For our analysis, this way of estimation might be problematic.
+We don't have pre-Omicron wave anti-N data in APL and the prevalence of ever infected (S+ & N+) raised sharply in 2022.
+This will result underestimated vax% in 2022, as presented in the fig above.
+
+#### 4.2.3 The Vax rate in AB general population
+
+```{r}
+# load the data from Canada Statistics
+vax_AB<-read.csv("https://health-infobase.canada.ca/src/data/covidLive/vaccination-coverage-map.csv")
+
+# create AB subset and clean the data
+vax_AB<-vax_AB[vax_AB$prename=="Alberta",]
+vax_ABs<-vax_AB %>%
+  select(week_end, numtotal_atleast1dose, proptotal_atleast1dose)
+colnames(vax_ABs)<-c("weekend", "num.1dose", "prop.1dose")
+vax_ABs$weekend<-as.yearmon(vax_ABs$weekend)
+
+# calculate monthly mean prop for vax rate
+tb_vax_AB<-vax_ABs %>%
+  group_by(weekend) %>%
+  summarise(ave.1dose = mean(prop.1dose))
+
+tb_vax_AB<-tb_vax_AB[!is.na(tb_vax_AB$ave.1dose),]
+tb_vax_AB$data<-rep("AB_general", length(tb_vax_AB$weekend))
+
+d01<-tb_vax_AB
+colnames(d01)<-c("samplemonth", "prop.vax", "data")
+```
+
+##### creating a merged data
+
+```{r}
+# Estimated vaccination rate: by anti-S + and anti-N neg
+View(tb_vax_est)
+d02<-select(tb_vax_est, c(1,4))
+d02$data<-rep("S_pos_N_Neg", length(d02$samplemonth))
+
+# Estimated vaccination rate: by anti-S only
+View(tb_abbs)
+d03<-select(tb_abbs, samplemonth, prop.abbs1, study)
+colnames(d03)<-c("samplemonth", "prop.vax", "data")
+d03$data<-rep("S_pos", length(d03$samplemonth))
+
+# Reported vax rate in APL
+View(tb_vax)
+d04<-select(tb_vax, c(1,4))
+d04$data<-rep("APL_vax", length(d04$samplemonth))
+
+# merge all
+dta.vax<-rbind.data.frame(d01, d02, d03, d04)
+rm(d01, d02, d03, d04)
+
+dta.vax$prop.vax[dta.vax$data=="AB_general"]<-dta.vax$prop.vax[dta.vax$data=="AB_general"]/100
+```
+
+##### Making the plot
+
+```{r}
+p12<-dta.vax %>%
+  ggplot(aes(x=samplemonth, y=prop.vax, group=data, colour=data)) +
+  geom_point(aes(colour=data)) +
+  geom_line(aes(colour=data))  +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  ylab("Vaccinated%")
+p12
+```
+
+## 4.3 Iris's Imputation on Vax history
+
+```{r}
+load("/home/jchen/../iganser/CBS_data/temp/2024-04-16_df4_schange_new_corrected2.RData")
+save(df4_schange_new_corrected2, file = "../1_data/private/CBS_vax_imputed.RData")
+
+# beware iris is using an updated ver of CBS, there's few thousands more observations available in this ver of CBS.
+CBS_vax_imputed<-df4_schange_new_corrected2
+rm(df4_schange_new_corrected2)
+```
+
+#### 4.3.1 Data cleaning and merging
+```{r}
+# The imputed vaccination variable
+table(CBS_vax_imputed$series_s, useNA = 'ifany')
+
+# add imputed variable to CBSs data
+CBSs_AB<-merge(CBSs_AB, 
+      select(CBS_vax_imputed, pid, sampledate, series_s), 
+             by = c("pid", "sampledate"),
+      all.x = T, all.y = F)
+
+table(CBSs_AB$series_s, useNA = 'ifany')
+# N(missing) = 24393. The combination of pid and sampledate is unique in both CBS and Iris's ver of CBS. So some of the samples we are working on are not matching.
+
+CBSs_AB$series_bin[CBSs_AB$series_s=="not vaccinated/1st only"]<-0
+CBSs_AB$series_bin[CBSs_AB$series_s!="not vaccinated/1st only" & !is.na(CBSs_AB$series_s)]<-1
+
+table(CBSs_AB$series_s, CBSs_AB$series_bin, useNA = 'ifany')
+```
+Iris algorithm cannot take unvaccinated and 1 dose only apart. In our case we can create a binary var <2 doses and >=2 doses, so we can do the same stratification with APL data.
+```{r}
+# re-do the melt data: no vax or 1 dose only
+tb_rosv0<-CBSs_AB %>%
+  .[.$series_bin == 0,] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_ros= sum(pos_roche_s, na.rm = T), 
+            pos_ros2= sum(pos_roche_s2, na.rm = T)
+            )
+tb_rosv0$prop.ros1<-tb_rosv0$pos_ros/tb_rosv0$n
+tb_rosv0$prop.ros2<-tb_rosv0$pos_ros2/tb_rosv0$n
+
+# clean NAs
+tb_rosv0<-tb_rosv0[!is.na(tb_rosv0$samplemonth),]
+
+tb_rosv0_melt<-reshape2::melt(select(tb_rosv0, -c(2:4)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "COI",
+                        value.name = "prop")
+
+# re-do the melt data: at least 1 dose
+tb_rosv1<-CBSs_AB %>%
+  .[.$series_bin == 1,] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_ros= sum(pos_roche_s, na.rm = T), 
+            pos_ros2= sum(pos_roche_s2, na.rm = T)
+            )
+tb_rosv1$prop.ros1<-tb_rosv1$pos_ros/tb_rosv1$n
+tb_rosv1$prop.ros2<-tb_rosv1$pos_ros2/tb_rosv1$n
+
+# clean NAs
+tb_rosv1<-tb_rosv1[!is.na(tb_rosv1$samplemonth),]
+
+tb_rosv1_melt<-reshape2::melt(select(tb_rosv1, -c(2:4)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "COI",
+                        value.name = "prop")
+```
+
+```{r}
+# re-do the plot
+# APL: not vaccinated
+p13v0<-tb_rosv0_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
+  geom_point(colour="blue") +
+  geom_line(aes(linetype=COI), colour="blue")  +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_linetype_manual("", values=c(1,3), labels = c("CBS-Roche (COI=0.8)", 
+                                                      "CBS-Roche (COI=179)"))+
+  theme(legend.position='top') + 
+  ggtitle("Stratum: not vaccinated or 1 dose only")
+p13v0
+```
+
+```{r}
+# Plot: at least 1 dose
+p13v1<-tb_rosv1_melt %>%
+  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
+  geom_point(colour="blue") +
+  geom_line(aes(linetype=COI), colour="blue")  +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_linetype_manual("", values=c(1,3), labels = c("CBS-Roche (COI=0.8)", 
+                                                      "CBS-Roche (COI=179)"))+
+  theme(legend.position='top')+
+  ggtitle("Stratum: at least 2 doses")
+p13v1
+```

### 2024-06-10 2_scripts/2.1 APL Data Quality Check.qmd
emergency backups

@@
 ```{r}
+# APL first version: provided by Carmen in 2023
 load(file = "../1_data/private/RFD4682e.RData")
+
+# Second version of APL: provided by Carmen in Jan 2024
+
 ```
@@
 #### As required by Matthew, add quintmat & qunitsoc variable
+
 ```{r}
@@
 library(readxl)
-RFD4682_v2 <- read_excel("1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
+RFD4682_v2 <- read_excel("../1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
     sheet = "RFD4682_Full_Data", col_types = c("text", 
@@
 ```{r}
+
 # only keep the month, remove the data

### 2024-06-10 5_manuscript/manuscript.qmd
emergency backups

@@
 ---
-title: "Assessing methods to adjust for assay differences across COVID-19 serological surveillance studies in Canada"
+title: "HOW ASSAY DIFFERENCES AND STATISTICAL ADJUSTMENT IMPACT INFERENCES FROM SARS-COV-2 SEROLOGICAL SURVEILLANCE STUDIES"
 execute: 
@@
 ```
-Jiacheng Chen^1,2^, Yuan Yu^1^, Sheila O'Brien^3^, Yu Nakagama^4^ ..., W. Alton Russell^1,2^
+Jiacheng Chen^1,2^, Yuan Yu^1^, Sheila O'Brien^3,4^, Carmen L. Charlton^5,6,7^, Steven Drews^5,6^, Yu Nakagama^8^, Yasutoshi Kido^8^, David L. Buckeridge^1,2^, W. Alton Russell^1,2^
 
@@
 
-^2^COVID-19 Immunity Task Force
+^2^COVID-19 Immunity Task Force, Montreal, Canada
 
-^3^Canadian Blood Services
+^3^Epidemiology and Surveillance, Canadian Blood Services, Ottawa, Ontario, Canada
 
-^4^Graduate School of Medicine, Osaka City University
+^4^School of Epidemiology and Public Health, University of Ottawa, Ottawa, Canada
+
+^5^Microbiology, Canadian Blood Services, Edmonton, Alberta, Canada
+
+^6^Laboratory Medicine and Pathology, University of Alberta, Edmonton, Alberta, Canada
+
+^7^Alberta Precision Laboratories (ProvLab), University of Alberta Hospital, Edmonton, Alberta, Canada
+
+^8^Graduate School of Medicine, Osaka Metropolitan University, Osaka, Japan
 
@@
 
-**Corresponding author:**
+**Corresponding author:** W. Alton Russell, McGill University, alton.russell\@mcgill.ca
 
-**Key words:**
+**Key words:** population health surveillance, serology, assay adjustment
 
-**Running title:**
+**Running title:** Serologic assay adjustment
 
@@
 
-**Background:** A
-
-**Methods:** A
-
-**Results:** A
-
-**Conclusions:** A
+Previous research shows substantial variability in SARS-CoV-2 seroprevalence estimates using different serological assays.
+The Roche Elecsys® anti-nucleocapsid (anti-N) assay demonstrates less waning sensitivity over time compared to the Abbott anti-N assay.
+We analyzed two Alberta serosurveillance datasets: 124,008 blood donor samples by Canadian Blood Services (CBS) and 214,780 outpatient samples by Alberta Precision Labs (APL).
+For anti-N analysis, we estimated seroprevalence using Rogen-Gladen (RG) adjustments and Bayesian intervals, with and without regression models.
+We also estimated rolling incidence using multinomial regression models.
+Early pandemic anti-N seropositivity was similar, but diverged after May 2022 due to greater waning sensitivity in the Abbott assay.
+RG adjustments had minimal impact, while regression improved concordance.
+For anti-S assays, 95% of blood donors and 90% of APL outpatients had seroconverted by the end of 2021, with 90% and 80% respectively achieving high antibody levels.
+Serosurveillance estimates are sensitive to the choice of anti-N assay, with common adjustments having limited impact.
+Regression-based approaches enhance comparability, and rolling incidence is a more population health-relevant measurement.
+Anti-S assays performed similarly between two datasets, with minor differences likely due to vaccination rates.
+These methods help to integrate and compare serosurveillance data across populations.
+(194/200 words)
 
@@
 # Introduction
-
-Test of citation: [@Langham2018a]. See @fig-sample, @cor-sample, @thm-sample, @thm-data, and @lem-sample.
 
@@
 NOTES:
-- The methods section of this project is partly unfinished, and the details of it needs constantly updating overtime, as the project goes. The potential obstacles were listed below:
+- The methods section of this project is finished, however, there's few more tasks to finish before submitting the manuscript to the journal:
 
-1. APL assay data before the Omicron wave was mostly missing, only Omicron wave data is available. Also we do not know what assay they used for anti-N (probably Abbott anti-N).
+1. APL anti-N assay data before the Omicron wave was recently found misplaced in the anti-S column. Data cleaning is on-going and we will update our analysis with updated anti-N data later.
 
@@
 In AB Outpatient Lab Data, they changed the assay (Diasorin S -- Abbott S) in the midway, this would affect the validity of analysis based on quantitative test results, as the results from different assay are not directly comparable.
-Hopefully this can be clarified by the APL data team.
 Our currently analysis assume the Anti-N/S assays used in omicron wave are all Abbott's, based on the paper published by Carmen in the post-vax stage: https://doi.org/10.1080/23744235.2022.2080250
-
-3.
-Canpath (removed): At least two types of anti-N assays are utilized: Mount Sinai Hosp anti-N (a in-lab developed assay) and Abbott anti-N (a commercial assay).
-Need to figure out which prov at what time used what assay, so we can apply sens for adjustment accordingly.
-As the assay difference project is mainly comparing CBS to APL, and Canpath has very few data available at later phases of the pandemic when sero-prevalence diverge between APL & CBS, we decided to have it excluded from analysis.
 
@@
 (last checked on 30-Apr-2024, CBS 2023-11-30 is the version we are working on, lastest CBS has 2023 data updated) As the assay difference project is mainly comparing CBS to APL, we don't have to include 2023 CBS data as APL has no comparable data.
+
+5. We've added Iris's algorithm for vaccination imputation in CBS. This algothrim has some minor limitations and those issues are discussed in the discussion section.
 --->
@@
 
-To estimate the SARS-CoV-2 seroprevalence and rolling incidence in Alberta, we used data from the Alberta Outpatient Lab (APL) and Canadian Blood Services (CBS).
-Alberta Health Services (AHS) tested \>160,000 APL blood draws for SARS-CoV-2 anti-N and anti-S antibodies from June 2020 until October 2022.
-CBS has tested \>450,000 blood donations for both antibodies levels since March 2020.
-The CBS Alberta subset has tested 124,008 samples in the same data collection period as CBS.
-Besides of assay results, both datasets include demographic variables such as donor's sex, self-reported race/ethnicity, age, and Forward Sortation Area (FSA) codes.
-From their FSA codes, neighborhood material/social deprivation and urban/rural indicator have been derived.
+To estimate the SARS-CoV-2 seroprevalence and rolling incidence, we analyzed two residual blood serosurveillance datasets in Alberta, Canada: 124,008 blood donor plasma samples tested with Roche Elecsys® Anti-SARS-CoV-2 (anti-N) and Anti-SARS-CoV-2 S (anti-S) by Canadian Blood Services (CBS) and 214,780 outpatient lab plasma or sera samples tested with the Abbott SARS-CoV-2 IgG Assay (anti-N) and SARS-CoV-2 IgG QUANT (anti-S) by Alberta Precision Labs (APL).
 
@@
 However, the anti-N assay results are only available starting from December 2021 in the APL data.
-@thm-data summarized the differences among two datasets.
-
-@thm-data CBS and APL
-
-[insert @thm-data here if necessary]
-
-Regarding the data we used to train the regression models, we identified several studies conducted in the early phases of the Covid-19 pandemic.
-Blood samples are collected from France, Jamaica, Japan and UK \[**add citation**\], with known sample positivity at the time of collection (more details are available at the supplemental methods section).
-The samples in Jamaica, Japan and UK were tested both for Abbott anti-N and Roche anti-N assay, which makes them suitable for transformation model training.
 
@@
 
-To evaluate the severity of Covid-19 pandemic overtime in Alberta province, first we calculated the anti-N assay test positivity.
-Then, to bridge the gap between test positivity and population-level seroprevalence, we employed the adjustment using Rogen-Gladen (RG) equation and Bayesian models.
-We conducted analysis separately for CBS and APL data then presented the results in figure X (@fig-seroprevalence).
+To evaluate the severity of Covid-19 pandemic overtime in each dataset, we estimated crude seroprevalence, and seroprevalence with the Rogen-Gladen (RG) adjustment for previously reported sensitivity and specificity using Bayesian credible intervals.
 
@@
 
-Where as "Prevalence" indicate RG adjusted seroprevalence estimates.
+Where "Prevalence" indicate RG adjusted seroprevalence estimates.
 "Test Positivity" indicates the proportion of test samples with positive results.
-"Spec" and "Sens" indicate the specificity and sensitivity of the assay, respectively.
+"Spec" and "Sens" indicate the specificity and sensitivity of the anti-N assay, respectively.
 
@@
 
-## **Seroprevalence based on raw/transformed data**
+## **Regression-based assay transformation**
 
-Following seroprevalence estimation, we further estimated monthly rolling incidence proportion of SARS-CoV-2 infection.
-Using matched data from samples tested with both assays early in the pandemic, we developed log-linear regression models to transform CBS Roche qualitative results to Abbott and applied the Abbott cutoff index and made the opposite transformation to convert APL to Roche.
-The seroprevalence estimates based on raw and transformed data are presented in figure X (@fig-transformed).
+Using data from samples tested with both assays early in the pandemic (in 2020), we developed linear regression models to transform CBS Roche qualitative results to Abbott and applied the Abbott cutoff index and made the opposite transformation to convert APL to Roche.
+Then we estimated seroprevalence in each dataset based on raw and transformed data.
+
+\[Add our transformation equation here?\]
 
@@
 
-Next, we developed multinomial logistic regression models using data collected in the early pandemic to estimate the probability a tested sample was tested 0-3, 4-6, 7-9, or \>9 months from infection based on the qualitative anti-N assay results.
+Following seroprevalence estimation, we further estimated monthly rolling incidence proportion of SARS-CoV-2 infection.
+We developed multinomial logistic regression models using data collected in the early pandemic to estimate the probability a tested sample was tested 0-3, 4-6, 7-9, or \>9 months from infection based on the qualitative anti-N assay results.
 Then we applied this model to the CBS and APL datasets to estimate rolling incidence from 2021 to 2023.
 
-We presented the rolling incidence proportion estimates in a paneled figure (@fig-rolling-incidence), displaying regression-based estimation of proportion of population infected in the last ever or in the past 3, 6, or 9 months.
+## **PRNT50 and corresponding antibody level**
 
-## Attack rates - TBD
-
-The cumulative attack rates is no longer of interest, as the majority of the Canadian population has been infected at least once since the Omicron wave.
-As the comparison rolling incidence proportion is a measurement with more meaningful public health implications.
-
-## Mixture Models - TBD
-
-In mixture models, we assume the dist of certain variables (in our case it is S/Co of Anti-N/S assay results) has two (or maybe three) distinct (though partially overlapping) distributions.
-Mixture models were utilized by previous studies to estimate cumulative incidence.
-
-<https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1008728>
-
-<https://www.sciencedirect.com/science/article/pii/S175543652200024X>
-
-## **NT50 and corresponding antibody level:**
-
-In evaluating whether participants possess humoral immunity levels sufficient to shield them from future infections, from either vaccination or infection, we applied a threshold of protection to the quantitative Anti-S assay results.
-The threshold is cited from the package insert of each assays \[Roche and Abbott\], indicating sufficient antibody level to neutralize over 50% virus, with certain times of dilution (1:20).
-We also presented the fluctuation of the proportion sufficiently protected over time in another figure (@fig-immunity), and we conducted sensitivity analyses in which we compared alternative threshold levels.
-The details of the sensitivity analysis is presented in the **Fig**.
-Y (haven't scratched this one yet).
+In evaluating whether participants possess anti-S antibody levels potentially providing protection against SARS-CoV-2 infection, first we estimated the positivity of anti-S assays among CBS blood donors and APL outpatients by applying manufacturers' cutoffs related to the presence of antibody.
+Then we applied the cutoffs used by previous researchers to estimate the proportion of population with relatively high antibody titers.
+These cutoffs are associated with a 50% reduction at plague reduction neutralization test (PRNT) at 1:120 dilution.
+Additionally, we conducted stratified analysis by immunization status.
 
@@
 
-\[Describe fig: seroprevalence\] A paragraph about test sample positivity, Rogen-Gladen adjusted seroprevalence and Bayesian model adjusted seroprevalence.
-Describe the trend of changing overtime, difference and consistency among different adjusting methos, regions and datasets.
-(Cite figures accordingly if necessary.)The datasets had similar seropositivity earlier in the pandemic but diverged after May 2022 due to greater waning sensitivity in the Abbott assay used by APL (Figure X).RG and Bayesian adjustments had minimal impact, increasing sero-prevalence estimates by 2.04% for the CBS Alberta subset and 7.76% for APL (Figure X).
+The datasets had similar anti-N seropositivity earlier in the pandemic but diverged after May 2022 due to greater waning sensitivity in the Abbott anti-N assay used by APL (Figure 1).
+RG and Bayesian adjustments had minimal impact, increasing seroprevalence estimates by 2.04% for the CBS Alberta subset and 7.76% for APL (Panel A), according to the estimates in October 2022.
+Our regression-based approaches increased concordance between estimated seropositivity (Panel B) and led to more similar estimates of rolling incidence (Panel C).
+The gap between CBS and APL seropositivity estimates for October 2022 was 48% using raw seropositivity, 15% after transforming CBS to Abbott, and 14% after transforming APL to Roche.
 
-\[Describe fig: seroprevalence based on raw/transformed data\] Our regression-based approaches increased concordance between estimated seropositivity (Figure X).
-However, we still observed divergence after May 2022.The gap between CBS and APL seropositivity estimates for October 2022 was 48% using raw seropositivity, 15% after transforming CBS to Abbott, and 14% after transforming APL to Roche.
+\[Should we use paneled fig our separate figures?\]
 
-\[Describe fig: rolling inc prop\] In the rolling incidence estimates, we obderved more similar estimates of rolling incidence in each panels.
-The difference on rolling incidence estimates ar October 2022 are X%, X%, X% and X%, for ever infected %, 3-month, 6-month, and 9-month rolling incidence, respectively.
+Regarding the results of the anti-S assay, among CBS blood donors 95% had seroconversion by the end of 2021, with over 90% achieving high antibody levels that potentially protect them from the infection of SARS-CoV-2 (Figure 2).
+In contrast, among APL outpatients, 90% had seroconversion at the same time, yet only 80% reached high antibody titers, leaving a wider gap (10%) between assay positivity and the proportion of population gained strong immunity.
+In the stratified analysis, 75% of the unvaccinated APL outpatients gained anti-S seropositivity by the end of 2021, with over 60% having high antibody titers.
+Among outpatients who received at least one dose of vaccine, 98% of them are anti-S positive at the same time, and over 95% achieved high antibody titers.
 
-\[Describe fig: NT50 and corresponding Ab level\] A paragraph about estimated proportion with sufficient humoral immunity to protect them from future possible SARS-CoV-2 infection.
-This proportion is also adjusted by Rogen-Gladen equation and Bayesian models.
+\[we need to adjust this part to match Iris' imputation on vax status\]
 
@@
 
-The gap between Rogen-Gladen adjusted sero-prevalence and unadjused sero-positivity: In our case, the sensitivity is where most variance come from.
-As the proof of excellent assay performance, assay manufacturer published the sensitivity and specificity of their assay.
-However, their test performance figures derived from hospitalized patients with more severe symptoms, who were samples at the time when sero antibody level reached its highest level.
-Due to the representativeness issue, it is not recommended only take the test performance figures published by the assays manufacturer into consideration.
+## **RG and Bayesian models**
 
-Since sens is imperfect (0.85--1.0) and specificity is mostly 100% trustworthy (\~1.0), the R-G equation can be simplified as:
+After applying RG and Bayesian models seroprevalence estimating, the RG adjusted seroprevalence increased 2.04% among CBS blood donors and 7.76% among APL outpatients.
+The wider gap in APL can be explained by lower assay sensitivity we applied.
+The test performance figures, including sensitivity and specificity, were estimated by previous researchers \[citation: Public Health England\] in England.
+Where sensitivities are 0.927 for Abbott anti-N assay and 0.972 for Roche anti-N assay, and specificities are mostly 1.0 (0.999 and 0.998 for Abbott and Roche anti-N assay respectively).
+Therefore the RG equation can be simplified as:
 
@@
 
-According to the equation above, the worse the sensitivity (smaller in value), the larger the gap between the adjusted and unadjusted.
+Where "Prevalence" indicate simplified RG adjusted seroprevalence estimates.
+"Test Positivity" indicates the proportion of test samples with positive results.
+"Sens" indicate the sensitivity of the anti-N assay.
+
+According to **Eq**.3, the worse the sensitivity (smaller in value), the larger the gap between the adjusted and unadjusted seroprevalence curves in the figure.
 We found larger gaps among the Abbott assay results, as the sensitivity we used for Abbott assays are smaller, when compared to sensitivity we used for Roche assays.
@@
 ```
-\[Difference/Consistence among adjusting methods, regions in Canada\] Is our finding consistent with previous studies?\]
+## Comparison among studies
+
+\[Difference/Consistence among adjusting methods, regions in Canada\] Is our finding consistent with previous studies?
+
+## **Strength and Limitation**
 
@@
 
-(to-be-deleted: we do not utilize most of the demographic variabes in our analysis) Besides of serology test results and demographic data, CBS has linked the administrative data for the cohort, which provided additional information on their hospitalizations, clinic visits, diagnoses, COVID vaccinations, and long-term disability claims.
-Similarly, AHS and CANPATH have completed linkages and included additional vaccination history.
-
 ```{=html}
 <!---
-Notes: Current ver of APL has completed very limited data linkage, which only includes PCR test results/time and vaccination time.
-The paragraph briefly describe the data linkage for three datasets, however for now we do not need those variables from linked administrative data.
+Notes: Sub-sections to be added for the supplemental section:
 
-Delete this part later
+Yuan's comments: add a short paragraph explaining the definition of sensitivity and specificity, difference in those figures btw Roche and Abbott, and their potential effect on RG-adjusted sero-prevalence estimates.
+This para can be added into somewhere in the discussion section.
+
+Definition on sens and specs are too basic to our potential readers in AJE. Let's focus on their potential effect then.
 --->
 ```
-
-A short paragraph about weighting: Our analysis is based on individual level.
-As we are calculating measurements at a monthly scale, samples sampled at different months were unweighted as they contributed independently to each month's point prevalence estimates.
+\[A short paragraph about weighting\] Our analysis is based on individual level data.
+As we are calculated measurements at a monthly scale, samples sampled at different months were unweighted as they contributed independently to each month's point prevalence estimates.
 While in all of those three dataset, it's possible to have one individual being sampled multiple times within the same month, in this case we applied weights to each individual according to the number of samples they have in the same month.
@@
 
-A paragraph about the training data: Cite the table summarizing the features of training data (Table S1). Clarify which data has matched assay results.
+\[A paragraph about the training data\] Cite the table summarizing the features of training data (Table S1).
+Clarify which data has matched assay results, and what data we used for each part of analysis.
+
+Regarding the data we used to train the regression models, we identified several studies conducted in the early phases of the Covid-19 pandemic (in 2020).
+Blood samples are collected from France, Jamaica, Japan and UK \[**add citation**\], with known sample positivity at the time of collection.
+The data from France and UK includes anti-N assay data for convalescent patients, the data provided by Osaka Metropolitan University in Japan included anti-N assay results for both convalescent patients and negative controls, and the data from Jamaica are used for negative controls only.
+Among all four datasets, the samples in Jamaica, Japan and UK were tested both for Abbott anti-N and Roche anti-N assay, which makes them suitable for transformation model training (figure).
+
+\[A short paragraph about anti-N sensitivity analysis\]
+
+\[A short paragraph about anti-S sensitivity analysis\]
 

### 2024-06-11 2_scripts/2.1 APL Data Quality Check_new.qmd
Finished APL ver2 data cleaning (Jan 2024)

--- 2024-06-10 2_scripts/2.1 APL Data Quality Check.qmd
@@
 # Second version of APL: provided by Carmen in Jan 2024
+# We load all columns as char, to avoid potential data corruption issues
+library(readxl)
+RFD4682_v2 <- read_excel("../1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
+    sheet = "RFD4682_Full_Data", col_types = c("text", 
+        "text", "text", "text", "text", 
+        "date", "text", "text", "date", "date", 
+        "text", "text", "numeric", "text", 
+        "text", "text", "text", "text", "text", 
+        "text", "text", "text", "text", "text", 
+        "text"))
+View(RFD4682_v2)
 
+RFD4682_v2_vax <- read_excel("../1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
+    sheet = "Vax Data", col_types = c("text", "numeric", "date", "text"))
+
+RFD4682_v2_NAAT <- read_excel("../1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
+    sheet = "NAAT Results", col_types = c("text", "date", "text"))
+
+# There's some minor updates they made to the v2 (Jan 2024), so it's recommended to rerun the data cleaning code on v2, instead of adding new variables to v1 (June 2023)
 ```
@@
 library(dplyr)         # Data management
-library(lubridate)     # Manage variables in Date formats
+# library(lubridate)     # Manage variables in Date formats
 library(zoo)           # for yearmonth Date format
-library(reshape2)      # for melt() function
+# library(reshape2)      # for melt() function
 ```
@@
 
-## The distribution of time from NAAT to Sero-assay
+## 1. APL v1 (June 2023) - Data Cleaning
+
+#### 1.1 Some Desc Analysis
+
+##### The distribution of time from NAAT to Sero-assay
 
@@
 
-## Records are not patients, how many individual has been included?
+##### Records are not patients, how many individual has been included?
 
@@
 
-## Why unique PERSON_ID and ULI_ID are not matching with each other?
+##### Why unique PERSON_ID and ULI_ID are not matching with each other?
 
@@
 
-## Scenario Examples Showcase
+#### 1.2 Scenario Examples Showcase
 
@@
 
-## Add clean_IDe and Scenario back to dataset (encrypted)
+#### 1.3 Data cleaning
+
+##### Add clean_IDe and Scenario back to dataset (encrypted)
 
@@
 
-#### ULI ID and PERSON_ID, in NAAT+ subset
+##### ULI ID and PERSON_ID, in NAAT+ subset
 
@@
 
-## Check N=2865 cases with missing in both ULI and Person_ID
+##### Check N=2865 cases with missing in both ULI and Person_ID
 
@@
 
-## The Whole APL dataset
+##### The Whole APL dataset
 
@@
 
-## Categorize the time since last NAAT test
+##### Categorize the time since last NAAT test
 
@@
 
-## Vaccination Status
+##### Vaccination Status
 
@@
 Estimation on number of individuals, using cleaned version of IDe
-
 ```{r}
@@
 Some additional cleaning/editing, making anti-S quantitative
-
 ```{r}
@@
 
-#### Yuan found this: anti-N is only available for the Omicron stage
+#### 1.4 Other data cleaning
 
+##### Yuan found this: anti-N is only available for the Omicron stage
+**Notes:** This is an error in two ver of APL. The anti-N assay results are available and misplaced in the anti-S-RBD column.
 ```{r}
@@
 # load updated version of APL
-library(readxl)
-RFD4682_v2 <- read_excel("../1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
-    sheet = "RFD4682_Full_Data", col_types = c("text", 
-        "numeric", "text", "numeric", "numeric", 
-        "date", "text", "text", "date", "date", 
-        "text", "numeric", "numeric", "text", 
-        "text", "text", "text", "text", "numeric", 
-        "numeric", "numeric", "numeric", 
-        "numeric", "numeric", "numeric"))
+
 View(RFD4682_v2)
@@
 
-#### Save a cleaned ver of RFD4682e
+#### 1.5 Save a cleaned ver of RFD4682e
 
@@
 
-save(RFD4682_e1, file = "../1_data/private/RFD4682e1.RData")
+save(RFD4682_e1, file = "../1_data/private/RFD4682e1.RData") # ver for matt only
 ```
+
+## 2. APL v2 (Jan 2024) - Data Cleaning
+
+#### ADD order ID and other ids -- doesn't work
+```{r}
+RFD4682_v2$order_ID<-RFD4682_e$order_ID
+
+cbind.data.frame(RFD4682_e$order_ID, RFD4682_e$SERO_ID, 
+                 RFD4682_v2$order_ID, RFD4682_v2$SERO_ID)
+```
+
+#### 2.1 De-duplicate
+```{r}
+# Does duplicated records exist?
+table(duplicated(RFD4682_v2))  # no-all records are unique
+
+# use the saved duplicated SERO_ID variables to de_duplicate:
+RFD4682_v2dup<-RFD4682_v2 %>%
+ .[duplicated(RFD4682_v2$SERO_ID)==T,]
+
+# Duplicated records to clean: 
+RFD4682_v2dup2<-RFD4682_v2 %>%
+  .[.$SERO_ID %in% RFD4682_v2dup$SERO_ID,]
+RFD4682_v2dup2$SERO_ID[9:16]<-paste0(RFD4682_v2dup2$SERO_ID[9:16],
+                                     rep(c("_1", "_2"),4))
+
+# Clean, now SERO_ID is unique
+RFD4682_v2<-RFD4682_v2 %>%
+  .[!(.$SERO_ID %in% RFD4682_v2dup$SERO_ID),]
+
+RFD4682_v2<-rbind.data.frame(RFD4682_v2, RFD4682_v2dup2[c(2,4,6,8,9:16),])
+rm(RFD4682_v2dup, RFD4682_v2dup2)
+```
+##### De-duplicate SERO_ID for APL v1
+```{r}
+# use the saved duplicated SERO_ID variables to de_duplicate:
+RFD4682_edup<-RFD4682_e %>%
+ .[duplicated(RFD4682_e$SERO_ID)==T,]
+
+# Duplicated records to clean: 
+RFD4682_edup2<-RFD4682_e %>%
+  .[.$SERO_ID %in% RFD4682_edup$SERO_ID,]
+RFD4682_edup2<-RFD4682_edup2 %>%
+  arrange(order_ID)
+RFD4682_edup2$SERO_ID[9:16]<-paste0(RFD4682_edup2$SERO_ID[9:16],
+                                     rep(c("_1", "_2"),4))
+
+# Clean, now SERO_ID is unique
+RFD4682_e<-RFD4682_e %>%
+  .[!(.$SERO_ID %in% RFD4682_edup$SERO_ID),]
+
+RFD4682_e<-rbind.data.frame(RFD4682_e, RFD4682_edup2[c(2,4,6,8,9:16),])
+rm(RFD4682_edup, RFD4682_edup2)
+```
+
+#### 2.2 Add ULI_IDe from v1
+```{r}
+library(duawranglr)    # use this to encrypt the ids
+
+## get crosswalk and admin data files
+dua_cw_file <- system.file('extdata', 'dua_cw.csv', package = 'duawranglr')
+admin_file <- system.file('extdata', 'admin_data.csv', package = 'duawranglr')
+
+## set the DUA crosswalk
+set_dua_cw(dua_cw_file)
+```
+
+```{r}
+# Use SERO_ID and 'RBD-IgGII_RESULT' to identify each records
+RFD4682_v2<-merge(RFD4682_v2, select(RFD4682_e, SERO_ID, order_ID, PERSON_IDe, ULI_IDe,
+                         clean_IDe, QUINTMAT, QUINTSOC), 
+      by ="SERO_ID", all = T
+      )
+# ULI - ULI_IDe crosswalk
+ULI_crosswalk<-select(RFD4682_v2, ULI_ID, ULI_IDe)
+ULI_crosswalk<-distinct(ULI_crosswalk)
+
+table(is.na(ULI_crosswalk$ULI_IDe))
+ULI_crosswalk %>%
+  .[is.na(.$ULI_IDe),]
+
+# APL ver2 seems have solved missing value issue in the ULI ID, needs to add those
+table(is.na(RFD4682_v2$ULI_ID), is.na(RFD4682_v2$ULI_IDe))
+
+table(is.na(RFD4682_v2$PERSON_ID), is.na(RFD4682_v2$PERSON_IDe))
+
+# encrypt new ULI_ID
+set.seed(240611)
+ULI_crosswalk_e <- deid_dua(ULI_crosswalk, 
+                      id_col = 'ULI_ID',
+                      new_id_name = 'ULI_IDe2',
+                      write_crosswalk = F, id_length = 15)
+
+ULI_crosswalk$ULI_IDe2<-ULI_crosswalk_e$ULI_IDe2
+
+# clean the ULI_ID crosswalk file
+table(is.na(ULI_crosswalk$ULI_ID))
+ULI_crosswalk<-ULI_crosswalk[!is.na(ULI_crosswalk$ULI_ID),]
+ULI_crosswalk<-ULI_crosswalk[-64591,]
+ULI_crosswalk$ULI_IDe[is.na(ULI_crosswalk$ULI_IDe)]<-ULI_crosswalk$ULI_IDe2[is.na(ULI_crosswalk$ULI_IDe)]
+
+# apply the cross-walk to the v2 data
+df.temp1<-RFD4682_v2[!is.na(RFD4682_v2$ULI_ID) & is.na(RFD4682_v2$clean_IDe),]
+df.temp2<-RFD4682_v2[!(!is.na(RFD4682_v2$ULI_ID) & is.na(RFD4682_v2$clean_IDe)),]
+
+df.temp1<-merge(select(df.temp1, -c("ULI_IDe")), 
+             ULI_crosswalk[,1:2],
+      by="ULI_ID", all.x = T, all.y = F)
+df.temp1$clean_IDe<-df.temp1$ULI_IDe
+
+RFD4682_v2<-bind_rows(df.temp1, df.temp2)
+rm(df.temp1, df.temp2, ULI_crosswalk_e)
+```
+
+```{r}
+# Still, there's some missing value in clean_IDe
+RFD4682_v2 %>%
+  .[is.na(.$clean_IDe),]
+```
+#### 2.3 Clean anti-N and anti-S
+
+##### 2.3.1 Anti-N cleaning (remove index)
+```{r}
+# Create an indicator for misplaced anti-N
+RFD4682_v2$abb_s_index<-grepl("index", RFD4682_v2$`RBD-IgGII_RESULT`, ignore.case = T)
+RFD4682_v2$abb_s_AU<-grepl("ml", RFD4682_v2$`RBD-IgGII_RESULT`, ignore.case = T)
+table(RFD4682_v2$abb_s_index, RFD4682_v2$abb_s_AU, useNA = 'ifany')
+# 36 samples missing abbott_s info
+
+RFD4682_v2$misplace<-rep("anti_S missing", 214776)
+RFD4682_v2$misplace[RFD4682_v2$abb_s_index==T]<-"index"
+RFD4682_v2$misplace[RFD4682_v2$abb_s_AU==T]<-"AU"
+table(RFD4682_v2$misplace, RFD4682_v2$abb_s_index, useNA = 'ifany')
+
+RFD4682_v2$`N-IgG_RESULT`[RFD4682_v2$misplace=="index"]<-
+  RFD4682_v2$`RBD-IgGII_RESULT`[RFD4682_v2$misplace=="index"]
+
+RFD4682_v2$`RBD-IgGII_RESULT`[RFD4682_v2$misplace=="index"]<-NA
+
+RFD4682_v2<-select(RFD4682_v2, -c("abb_s_index", "abb_s_AU"))
+```
+
+```{r}
+# Remove "index" for anti-N
+RFD4682_v2$`N-IgG_RESULT`<-gsub('Index', '', RFD4682_v2$`N-IgG_RESULT`)
+RFD4682_v2$`N-IgG_RESULT`<-as.numeric(RFD4682_v2$`N-IgG_RESULT`)
+
+# Anti-N interp:
+summary(RFD4682_v2$`N-IgG_RESULT`[RFD4682_v2$`N-IgG_INTERP`=="Positive"]) # good
+summary(RFD4682_v2$`N-IgG_RESULT`[RFD4682_v2$`N-IgG_INTERP`=="Negative"]) # good
+summary(RFD4682_v2$`N-IgG_RESULT`[is.na(RFD4682_v2$`N-IgG_INTERP`)]) # needs correction
+
+RFD4682_v2$`N-IgG_INTERP`[RFD4682_v2$`N-IgG_RESULT` < 1.4 & 
+                          !is.na(RFD4682_v2$`N-IgG_RESULT`)]<-"Negative"
+RFD4682_v2$`N-IgG_INTERP`[RFD4682_v2$`N-IgG_RESULT` >= 1.4 & 
+                          !is.na(RFD4682_v2$`N-IgG_RESULT`)]<-"Positive"
+
+summary(RFD4682_v2$COLLECTION_DATE[!is.na(RFD4682_v2$`N-IgG_INTERP`)])
+```
+
+##### 2.3.2 Anti-S cleaning (remove AU/ml)
+```{r}
+RFD4682_v2$`RBD-IgGII_RESULT`<-gsub('[a-z]', '', RFD4682_v2$`RBD-IgGII_RESULT`)
+RFD4682_v2$`RBD-IgGII_RESULT`<-gsub('[A-Z]', '', RFD4682_v2$`RBD-IgGII_RESULT`)
+RFD4682_v2$`RBD-IgGII_RESULT`<-gsub('[/<>]', '', RFD4682_v2$`RBD-IgGII_RESULT`)
+RFD4682_v2$`RBD-IgGII_RESULT`<-as.numeric(RFD4682_v2$`RBD-IgGII_RESULT`)
+summary(RFD4682_v2$`RBD-IgGII_RESULT`)
+
+# >800000 is recoded as 800000
+RFD4682_v2 %>%
+  .[.$`RBD-IgGII_RESULT`<800000 & 100000<.$`RBD-IgGII_RESULT` &
+    !is.na(.$`RBD-IgGII_RESULT`),]
+
+# Double check in anti_S interpretation
+summary(RFD4682_v2$`RBD-IgGII_RESULT`[RFD4682_v2$`RBD-IgGII_INTERP`=="Positive"]) # -->NA
+summary(RFD4682_v2$`RBD-IgGII_RESULT`[RFD4682_v2$`RBD-IgGII_INTERP`=="Negative"]) # -->NA
+summary(RFD4682_v2$`RBD-IgGII_RESULT`[is.na(RFD4682_v2$`RBD-IgGII_INTERP`)]) # good
+
+# correct anti-S interpretations
+RFD4682_v2$`RBD-IgGII_INTERP`[is.na(RFD4682_v2$`RBD-IgGII_RESULT`)]<-NA
+
+table(RFD4682_v2$`RBD-IgGII_INTERP`, useNA = 'ifany')
+
+```
+#### 2.4 Remove unencryoted individual inditifier
+```{r}
+RFD4682_ev2<-select(RFD4682_v2, -c(ULI_ID, PERSON_ID, misplace))
+```
+
+#### 2.5 Save the encrypted ver of APL data
+```{r}
+save(RFD4682_v2, file = "../1_data/private/RFD4682_v2.RData")
+save(RFD4682_ev2, file = "../1_data/private/RFD4682_ev2.RData")
+
+save(RFD4682_ev2, file = "~/../../jmy/APL/RFD4682_ev2.RData")
+```
+
+## 3. Clean other two APL sub datasets (Vax, NAAT)
+
+```{r}
+# Add encrypted IDe to vax and NAAT data
+RFD4682_v2_vaxe<-merge(RFD4682_v2_vax, select(ULI_crosswalk, ULI_ID, ULI_IDe),
+      by.x="ULI", by.y="ULI_ID", all.x = T, all.y = F)
+# any missing?
+table(is.na(RFD4682_v2_vaxe$ULI_IDe))
+
+# also encrypt the ULI_ID in the RFD4682_vax
+
+list_temp<-RFD4682_v2_vaxe %>%
+  .[is.na(.$ULI_IDe),]
+```

### 2024-06-11 2_scripts/2.4 PRNT50.qmd
Finished APL ver2 data cleaning (Jan 2024)

@@
 ```
+
 ##### Is 1-doser the cause of the "peak" and "bucket"?
+
 ```{r}
@@
 ```
+
 #### Add time indicator to the figure
@@
 #### 4.3.1 Data cleaning and merging
+
 ```{r}
@@
 ```
-Iris algorithm cannot take unvaccinated and 1 dose only apart. In our case we can create a binary var <2 doses and >=2 doses, so we can do the same stratification with APL data.
+
+Iris algorithm cannot take unvaccinated and 1 dose only apart. In our case we can create a binary var \<2 doses and \>=2 doses, so we can do the same stratification with APL data.
+
 ```{r}

### 2024-06-13 2_scripts/2.1 APL Data Quality Check_new.qmd
Update codes using APL v2 (unfinished) Further investigation on vaccination time gap

@@
 
+table(RFD4682_v2$`N-IgG_INTERP`, useNA = 'ifany')
 summary(RFD4682_v2$COLLECTION_DATE[!is.na(RFD4682_v2$`N-IgG_INTERP`)])
@@
 
+## Unfinished ##
 ```{r}

### 2024-06-13 2_scripts/2.3 Prob Model for PCR pos.qmd
Update codes using APL v2 (unfinished) Further investigation on vaccination time gap

--- 2024-05-06 2_scripts/2.3 Prob Model for PCR pos.qmd
@@
 
+# How many obs we have in each provinces?
+table(CBSs$province, useNA = 'ifany') # n(AB)=124008
+
 # The APL data
-load(file = "../1_data/private/RFD4682e.RData")
-
-# The cleaned ver of Canpath data
-
+load(file = "../1_data/private/RFD4682e.RData")   # ver.1
+load(file = "../1_data/private/RFD4682ev2.RData") # ver.2 with additional anti-N data
 ```
@@
 #### 1.2 Model mnlm04 same model but for abbott
+
 ```{r}
@@
 ```
+
 #### 2.2.3 Make plots by region
@@
 #### 2.3.1 Load & Clean
+
 ```{r}
@@
 # create a subset of APL
-APLs<-RFD4682_e %>%
-  select(order_ID, AGE_AT_COLLECTION, GENDER, N, `N-IgG_INTERP`, Col_Month, PAT3)
+APLs<-RFD4682_ev2 %>%
+  select(order_ID, SERO_ID, clean_IDe,
+         AGE_AT_COLLECTION, GENDER, `N-IgG_RESULT`,`N-IgG_INTERP`, COLLECTION_DATE)
 
 # unify colnames
-colnames(APLs)<-c("order_ID","age","gender","abbott_n", "pos_abbott","samplemonth","fsa")
-APLs$samplemonth<-as.yearmon(APLs$samplemonth)
+colnames(APLs)<-c("order_ID", "SERO_ID", "clean_IDe",
+                  "age","gender","abbott_n", "pos_abbott","sampledate")
+
+APLs$samplemonth<-as.yearmon(APLs$sampledate)
 
 # add log_abbott_n
-## find 0 values bf transformation
+## find 0 values bf transformation, correct them as 0.001
 APLs$abbott_n[APLs$abbott_n==0 & 
@@
 APLs$log_abbott_n <- log(APLs$abbott_n)
+
+# Recode abbott_n test results pos/neg as 1/0
+APLs$pos_abbott[APLs$pos_abbott=="Negative"]<-0
+APLs$pos_abbott[APLs$pos_abbott=="Positive"]<-1
+APLs$pos_abbott<-as.numeric(APLs$pos_abbott)
+
 ```
@@
 #### 2.3.2 Apply model (Abbott)
+
 ```{r}
@@
 #### 2.3.3 making Plots
+
 ```{r}
@@
 ```
+
 **Additional plotting**
+
 ```{r}
@@
 
-##===================== Part III. Exp: Reg Model Abbott N~Roche N =============##
-
-This idea came from Alton, we can use the data we have for prob model training to generate another model: Abbott-N~Roche-N.
+##===================== Part III. Exp: Reg Model Abbott N\~Roche N =============##
+
+This idea came from Alton, we can use the data we have for prob model training to generate another model: Abbott-N\~Roche-N.
 Then apply the model to the CBS data and generate the sero-prevalence curve, see if it gets closer to the APL curve
 
-#### 3.1.1 The linear (log) model: Abbott ~ Roche
+#### 3.1.1 The linear (log) model: Abbott \~ Roche
+
 ```{r}
@@
 
-#### 3.1.2  The linear (log) model: Roche ~ Abbott
+#### 3.1.2 The linear (log) model: Roche \~ Abbott
+
 ```{r}
@@
 #### 3.2.1 Apply the model lm01 to the CBS data
+
 ```{r}
@@
 #### 3.2.2 Apply model lm01r to APL data
+
 ```{r}
@@
 #### 3.3.1 plot: sero-prevalence, using predicted Abbott positivity
+
 ```{r}
@@
 #### 3.3.2 plot: sero-prevalence, using predicted Roche pos
+
 ```{r}
@@
 ```
+
 #### 3.3.3 Plot both roche and abbott, raw/transformed in on figure
+
 ```{r}
@@
 
-Alton's comments:
-- solid/dash to indicate raw/transformed
-- red/blue to indicate Abbott/Roche
-- Merge the color & linetype indicator in the legend
+Alton's comments: - solid/dash to indicate raw/transformed - red/blue to indicate Abbott/Roche - Merge the color & linetype indicator in the legend
 
@@
 ```
+
 The gap between CBS and APL seropositivity estimates for October 2023 was X% using raw seropositivity, Y% after transforming CBS to Abbott, and Z% after transforming APL to Roche.
@@
 #### 3.4 Plot the correlation btw Roche-N and Abbott-N:
+
 ```{r}
@@
 ```
+
 #### Add alternative COI
+
 ```{r}
@@
 ```
+
 #### Alton suggests using marginal dist with {ggExtra}
+
 ```{r}
@@
 ```
+
 #### 3.5 Create figure panel for AABB conference abstract
@@
 #### 3.5.1 Panel A: the serosurveillance figure and RG/Bayesian adjustment
+
 Why: they only alow on figure.
+
 ```{r}
@@
 #### 3.5.2 Panel B: transformed assay result fig
+
 ```{r}
@@
 #### 3.5.3 Panel C: rolling incidence fig
+
 ```{r}
@@
 #### 3.5.4 Make all three figures into one
+
 ```{r}
@@
 Generally speaking, blood donors (CBS, FR-Leon data) is relatively older, while Covid-19 convalescent patients are younger.
+
 ```{r}
@@
 #### Create a sample based on the sampling prob:
+
 ```{r}
@@
 ## Regression on sampled data:
+
 ```{r}
@@
 #### 4.2.1 Using Jamaica-Osaka-UK data:
+
 ```{r}
@@
 ```
+
 #### 4.2.2 Refit using CBS matched data
+
 ```{r}
@@
 ```
+
 #### Adjusting the weights of those with negative test results
+
 ```{r}
@@
 ```
+
+#### 4.2.3 Add regression lines to the figure
+
+```{r}
+# Modify p04 for sensitivity analysis
+p04<-dta_all %>%
+  .[!is.na(.$log_abbott_n) & !is.na(.$log_roche_n),] %>%
+  ggplot(aes(x=log_roche_n, y=log_abbott_n, 
+             study, colour=study)) +
+  geom_point(aes(colour=as.factor(study)), alpha=0.5) +
+  geom_vline(xintercept = 0, linewidth=1.0, linetype="dotted", col="darkgrey") +
+  geom_hline(yintercept = log(1.4), linewidth=1.0, linetype="dotted", col="darkgrey") +
+  xlab("Roche Anti-N S/Co ratio (log scale)") + 
+  ylab("Abbott Anti-N S/Co ratio (log scale)") +
+  labs(color='') +
+  scale_colour_manual(labels = c("Jamaica (-)", "Osaka, JP (-)",
+                                 "Osaka, JP (+)","UK (+)"), 
+                      values = c("red", "blue", "darkgreen", "purple")) +
+  theme_bw() + 
+  theme(legend.position = "bottom")
+
+p04.2a<-p04 + geom_abline(aes(intercept = -1.60, slope = 0.68, color="Roche->Abbott",),
+                  size=0.8) +
+      geom_abline(aes(intercept = -1.68, slope = 0.78, color="Abbott->Roche",), 
+                  size=0.8) +
+scale_colour_manual(values = c("red", "blue", "darkgreen", "purple","blue","red"),
+                    guide = guide_legend(nrow = 3,
+                            override.aes = list(
+                            linetype = c(rep("blank", 4), "solid", "solid"),
+                            shape = c(rep(16, 4), NA, NA)))) +
+     ggtitle("Jamaica-Osaka-UK matched data")
+```
+
+```{r}
+# Calculating Beta estimates for model 01r-reverted
+(-1)*lm01r$coefficients[1]/lm01r$coefficients[2] #beta_0
+1/lm01r$coefficients[2]                          #beta_1
+
+# Calculating Beta estimates for model 05r-reverted
+(-1)*lm05r$coefficients[1]/lm05r$coefficients[2] #beta_0
+1/lm05r$coefficients[2]                          #beta_1
+
+# Calculating Beta estimates for model 05rw-reverted
+(-1)*lm05r.w$coefficients[1]/lm05r.w$coefficients[2] #beta_0
+1/lm05r.w$coefficients[2]                            #beta_1
+```
+
+```{r}
+# Refit models for CBS data, modify p05
+p04.2b<-dta_CBS %>%
+  .[!is.na(.$log_abbott_n) & !is.na(.$log_roche_n) &
+    .$study=="CBS matched",] %>%
+  ggplot(aes(x=log_roche_n, y=log_abbott_n, 
+             study, colour=study)) +
+  geom_point(aes(colour=as.factor(study)), alpha=0.40) +
+  geom_vline(xintercept = 0, linewidth=1.0, linetype="dotted", col="darkgrey") +
+  geom_hline(yintercept = log(1.4), linewidth=1.0, linetype="dotted", col="darkgrey") +
+  xlab("Roche Anti-N S/Co ratio (log scale)") + 
+  ylab("Abbott Anti-N S/Co ratio (log scale)") +
+  labs(color='') +
+  scale_colour_manual(labels = c("CBS matched"), 
+                      values = c("lightblue")) +
+  theme_bw() + 
+  theme(legend.position = "bottom")
+
+p04.2b<-p04.2b + geom_abline(aes(intercept = -1.87, slope = 0.64, color="Roche->Abbott",),
+                  size=0.8) +
+         geom_abline(aes(intercept = 2.02, slope = 2.26, color="Abbott->Roche",), 
+                  size=0.8) +
+         geom_abline(aes(intercept = -2.14, slope = 0.85, 
+                         color="Abbott->Roche (down-weighted)",), 
+                  size=0.8, linetype=2) +
+         scale_colour_manual(values = c("lightblue","blue","red","red"),
+                             guide = guide_legend(nrow=2,
+                             override.aes = list(
+                             linetype = c(rep("blank", 1), "solid", "solid","dashed"),
+                             shape = c(rep(16, 1), NA, NA, NA))
+                             )) +
+         ggtitle("CBS matched data")
+p04.2b
+```
+
+```{r}
+## make two into one paneled fig
+png(file="../4_output/figs/Fig.S1 Sensitivity Analysis.png", 
+    width=2400, height=1440, res=300)
+cowplot::plot_grid(p04.2a, 
+                   p04.2b  + ylab(""), nrow=1,
+                   align = 'h',
+                   labels = c("A", "B"))
+dev.off()
+```
+
+#### 4.3 RG Adjusted Sero-prevalence using Bayesian model's priors
+
+##### 4.3.1 Create monthly anti-N positivity data for CBS
+```{r}
+# CBSs AB subset, N=124008
+CBSs_AB<-CBSs[CBSs$province == "AB",]
+
+# Monthly Positivity in CBS
+tb_431<-CBSs_AB %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_roche= sum(pos_roche, na.rm = T)
+            )
+tb_431$prop.raw<-tb_431$pos_roche/tb_431$n
+
+# Add RG adjustment
+tb_431$prop.adj.p1<-(tb_431$prop.raw + 0.998 -1)/(0.998 + 0.979 -1)
+tb_431$prop.adj.p2<-(tb_431$prop.raw + 0.993 -1)/(0.993 + 0.862 -1)
+tb_431$study <- rep("CBS", length(tb_431$samplemonth))
+```
+
+##### 4.3.2 Create monthly anti-N positivity data for APL
+```{r}
+# APL data, N=214776
+
+# Monthly Positivity in APL
+tb_432<-APLs %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_abbott= sum(pos_abbott, na.rm = T)
+            )
+tb_432$prop.raw<-tb_432$pos_abbott/tb_432$n
+
+# remove missing data
+tb_432<-tb_432[!is.na(tb_432$samplemonth),]
+tb_432<-tb_432[tb_432$pos_abbott!=0,]
+
+# Add RG adjustment
+tb_432$prop.adj.p1<-(tb_432$prop.raw + 0.995 -1)/(0.995 + 0.960 -1)
+tb_432$prop.adj.p2<-(tb_432$prop.raw + 0.995 -1)/(0.995 + 0.838 -1)
+tb_432$study <- rep("APL", length(tb_432$samplemonth))
+
+# merge CBS and APL table
+tb_43<-bind_rows(tb_431[, c(1, 4:7)], tb_432[-c(1), c(1, 4:7)])
+# April 2020 row is removed
+
+tb_43m<-reshape2::melt(tb_43, 
+                        id.vars = c("samplemonth", "study"),
+                        variable.name = "adjustment",
+                        value.name = "prop")
+rm(tb_431, tb_432)
+```
+
+##### 4.3.3 Make plots
+```{r}
+p4.3.3<-tb_43m %>%
+  ggplot(aes(x=samplemonth, y=prop, colour=study)) +
+  geom_point(aes(colour=study)) +
+  geom_line(aes(colour=study, linetype=adjustment))  +
+  xlab("Month of Sample Collection") +
+  ylab("Seroprevalence") +
+  coord_cartesian(xlim = c(as.yearmon("May 2020"), as.yearmon("Jan 2022")),
+                  ylim = c(0, 0.15)) +
+#  ylim(c(as.yearmon("Jan 2022"), as.yearmon("Feb 2023"))) +
+  scale_colour_manual("", values = c("red","blue")) +
+  scale_linetype_manual("", values=c(1,2,3),
+                        labels= c("Unadjusted", "RG adjusted (informed priors)",
+                                  "RG adjusted (nautral priors)")) +
+  theme_bw()
+
+p4.3.3
+```

### 2024-06-13 2_scripts/2.4 PRNT50.qmd
Update codes using APL v2 (unfinished) Further investigation on vaccination time gap

@@
 # The CBS Data, extracted using "CITF_CBS_desc_stats.Rmd"
+
+load(file = "../1_data/private/CBS.RData") # original CBS
 load(file = "../1_data/private/CBSs.RData") # a clean subset of CBS
@@
 # APL Data
-load()
 load(file = "../1_data/private/APLs.RData") # a clean subset
@@
 library(dplyr)         # Data management
-#library(reshape2)      # for melt() function
+library(reshape2)      # for melt() function
 library(ggplot2)       # ggplot
@@
 
-#### 1.1 Use the APLs created
+#### 1.1 Use the APLs created, add anti-S assay results and interp
 
 ```{r}
-APLs<-merge(APLs, select(RFD4682_e, order_ID, 
-                   'RBD-IgGII_RESULT', `RBD-IgGII_INTERP`,`RBD-IgGII_RESULT_DATE`
+APLs<-merge(APLs, select(RFD4682_ev2, order_ID, 
+                   `RBD-IgGII_RESULT`, `RBD-IgGII_INTERP`,`RBD-IgGII_RESULT_DATE`
                    ), by="order_ID",all=T)
-colnames(APLs)<-c(names(APLs)[1:15], "abbott_s", "pos_abbott_s", "month_abbott_s")
-APLs$abb_s_index<-grepl("index", APLs$abbott_s, ignore.case = T)
-APLs$abb_s_AU<-grepl("ml", APLs$abbott_s, ignore.case = T)
-table(APLs$abb_s_index, APLs$abb_s_AU)
-# 36 samples missing abbott_s info
-APLs$abb_s_index[APLs$abb_s_index==0 & APLs$abb_s_AU==0]<-NA
-APLs$abb_s_AU[is.na(APLs$abb_s_index)]<-NA
-
-APLs$abb_s_AU<-as.numeric(APLs$abb_s_AU)
-APLs$abb_s_index<-as.numeric(APLs$abb_s_index)
-table(APLs$abb_s_index, APLs$abb_s_AU, useNA = 'ifany')
-
-# remove units
-APLs$abbott_s<-gsub('[a-z]', '', APLs$abbott_s)
-APLs$abbott_s<-gsub('[A-Z]', '', APLs$abbott_s)
-APLs$abbott_s<-gsub('[/<>]', '', APLs$abbott_s)
-APLs$abbott_s<-as.numeric(APLs$abbott_s)
+colnames(APLs)<-c(names(APLs)[1:10], "abbott_s", "pos_abbott_s", "month_abbott_s")
 
 # check the COI
-summary(APLs[APLs$abb_s_index==1 & APLs$pos_abbott_s=="Positive",]$abbott_s)
-# 1.4 - 9.6 the COI is 1.4 not 1?
-summary(APLs[APLs$abb_s_index==1 & APLs$pos_abbott_s=="Negative",]$abbott_s)
-# 8.1 index but negative?
-
-summary(APLs[APLs$abb_s_AU==1 & APLs$pos_abbott_s=="Positive",]$abbott_s)
+summary(APLs[APLs$pos_abbott_s=="Positive",]$abbott_s)
 # 50 - 800000 the COI = 50 good
-summary(APLs[APLs$abb_s_AU==1 & APLs$pos_abbott_s=="Negative",]$abbott_s)
+summary(APLs[APLs$pos_abbott_s=="Negative",]$abbott_s)
 # 0-49, COI = 50 good
-
-# COI for anti_n
-summary(APLs[APLs$pos_abbott==1 & !is.na(APLs$abbott_n),]$abbott_n)
-# 1.4 - 14.19 the COI = 1.4
-summary(APLs[APLs$pos_abbott==0 & !is.na(APLs$abbott_n),]$abbott_n)
-# 0.001 - 1.39
-
-table(APLs$pos_abbott)
-```
-
-```{r}
-# abbott_n misclassified as abbott_s with index?
-APLs %>%
-  .[is.na(.$abbott_n) & .$abb_s_index==1 & !is.na(.$abb_s_index),] # N=104946
-
-APLs %>%
-  .[is.na(.$abbott_n) & .$abb_s_AU==1 & !is.na(.$abb_s_AU),] # N=69108
-
-APLs %>%
-  .[!is.na(.$abbott_n) & .$abb_s_index==TRUE & !is.na(.$abb_s_index),] # N=0
-
-APLs %>%
-  .[!is.na(.$abbott_n) & .$abb_s_AU==1 & !is.na(.$abb_s_AU),]  # N=40690
-
-APLs %>%
-  .[!is.na(.$abbott_n) & is.na(.$abb_s_AU),]                   # N=36
-
-table(is.na(APLs[APLs$abb_s_AU==1,]$abbott_n))
-
-# month of data collection for those five groups:
-# Group 1: "Abbott-S" (Index) only
-fivenum(APLs[APLs$abb_s_index==1,]$samplemonth, na.rm = T)
-
-# Group 2: True Abbott-S (AU) only
-fivenum(APLs[is.na(APLs$abbott_n) & APLs$abb_s_AU==1,]$samplemonth, na.rm = T)
-
-# Group 3: True Abbott-S (AU) and Abbott-N?
-fivenum(APLs[!is.na(APLs$abbott_n) & APLs$abb_s_AU==1,]$samplemonth, na.rm = T)
-
-# Group 4: no available anti-S data, only Abbott-N
-fivenum(APLs[!is.na(APLs$abbott_n) & is.na(APLs$abb_s_AU),]$samplemonth, na.rm = T)
 ```
@@
 Updates on May 28th: Carmen replied and confirmed they are misplaced anti-N assay results.
-
-#### 1.2 Move the misplaced anti-N back to where they are
-
-```{r}
-APLs$abbott_n[APLs$abb_s_index==1 &
-              !is.na(APLs$abbott_s)]<-APLs$abbott_s[APLs$abb_s_index==1 & 
-                                                   !is.na(APLs$abbott_s)]
-APLs$abbott_s[APLs$abb_s_index==1 & !is.na(APLs$abbott_s)]<-NA
-
-# add anti-N interpretation
-APLs$pos_abbott[APLs$abbott_n<1.4 & !is.na(APLs$abbott_n)]<-0
-APLs$pos_abbott[APLs$abbott_n>=1.4 & !is.na(APLs$abbott_n)]<-1
-
-# redo the anti-S interpretation
-APLs$pos_abbott_s<-rep(NA,214780)
-APLs$pos_abbott_s[APLs$abbott_s<50 & !is.na(APLs$abbott_s)]<-0
-APLs$pos_abbott_s[APLs$abbott_s>=50 & !is.na(APLs$abbott_s)]<-1
-table(APLs$pos_abbott_s, useNA = 'ifany')
-```
+APL data has been cleaned and uploaded to the server to share with Yuan and Matthew.
 
@@
 ```{r}
-APLs$pos_abbott_s2<-rep(NA, 214780)
-APLs$pos_abbott_s2[APLs$abb_s_AU==1 & !is.na(APLs$abb_s_AU) &
-                   APLs$abbott_s>=402.36 & !is.na(APLs$abbott_s)]<-"Positive"
-APLs$pos_abbott_s2[APLs$abb_s_AU==1 & !is.na(APLs$abb_s_AU) &
-                     APLs$abbott_s<402.36 & !is.na(APLs$abbott_s)]<-"Negative"
+APLs$pos_abbott_s2<-rep(NA, 214776)
+APLs$pos_abbott_s2[APLs$abbott_s>=402.36 & !is.na(APLs$abbott_s)]<-"Positive"
+APLs$pos_abbott_s2[APLs$abbott_s<402.36 & !is.na(APLs$abbott_s)]<-"Negative"
 table(APLs$pos_abbott_s2, useNA = 'ifany')
@@
 tb_abbs<-APLs %>%
-  .[.$abb_s_AU==1,] %>%
   group_by(samplemonth) %>%
@@
 tb_APLCBS2<-rbind.data.frame(tb_abbs_melt, tb_ros_melt)
+rm(tb_abbs, tb_abbs_melt, tb_ros, tb_ros_melt)
 
@@
 # Add vax variable
-APLs<-merge(APLs, select(RFD4682_e, order_ID, VACCINATION_STATUS),
+APLs<-merge(APLs, select(RFD4682_ev2, order_ID, VACCINATION_STATUS),
       by="order_ID", all=T)
@@
 # Add days since imm (vax)
-APLs<-merge(APLs, select(RFD4682_e, order_ID, DAYS_SINCE_IMM),
+APLs<-merge(APLs, select(RFD4682_ev2, order_ID, DAYS_SINCE_IMM),
       by="order_ID", all=T)
@@
 tb_abbsv0<-APLs %>%
-  .[.$abb_s_AU==1 & (.$vax %in% c("Not Immunized", "Pre-Immunization", "1 Dose")),] %>%
+  .[.$vax %in% c("Not Immunized", "Pre-Immunization"),] %>%
   group_by(samplemonth) %>%
@@
 tb_abbsv0<-tb_abbsv0[!is.na(tb_abbsv0$samplemonth),]
+tb_abbsv0<-tb_abbsv0[!tb_abbsv0$pos_abbott_s==0,]
 
@@
 
-# re-do the melt data: at least 1 dose
+# re-do the melt data: 1 dose only
 tb_abbsv1<-APLs %>%
-  .[.$abb_s_AU==1 & .$vax %in% c("1-2 Doses","2 Doses","3 Doses",
-                             "4 Doses") ,] %>%
+  .[.$vax %in% c("1 Dose") ,] %>%
   group_by(samplemonth) %>%
@@
 tb_abbsv1_melt<-reshape2::melt(select(tb_abbsv1, -c(2:4)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "COI",
+                        value.name = "prop")
+
+# re-do the melt data: 2 doses +
+tb_abbsv2<-APLs %>%
+  .[.$vax %in% c("1-2 Doses", "2 Doses", "3 Doses", "4 Doses") ,] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
+            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
+            )
+tb_abbsv2$prop.abbs1<-tb_abbsv2$pos_abbott_s/tb_abbsv2$n
+tb_abbsv2$prop.abbs2<-tb_abbsv2$pos_abbott_s2/tb_abbsv2$n
+
+# clean NAs
+tb_abbsv2<-tb_abbsv2[!is.na(tb_abbsv2$samplemonth),]
+
+tb_abbsv2_melt<-reshape2::melt(select(tb_abbsv2, -c(2:4)), 
                         id.vars = c("samplemonth"),
@@
 ```{r}
-# Plot: at least 1 dose
+# Plot: 1 dose only
 p09v1<-tb_abbsv1_melt %>%
@@
   theme(legend.position='top')+
-  ggtitle("Stratum: at least 2 doses")
+  ggtitle("Stratum: 1 dose only")
 p09v1
@@
 
-##### Is 1-doser the cause of the "peak" and "bucket"?
-
 ```{r}
-tb_abbsv11<-APLs %>%
-  .[.$abb_s_AU==1 & .$vax %in% c("1 Dose") ,] %>%
-  group_by(samplemonth) %>%
-  summarise(n = n() , 
-            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
-            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
-            )
-tb_abbsv11$prop.abbs1<-tb_abbsv11$pos_abbott_s/tb_abbsv11$n
-tb_abbsv11$prop.abbs2<-tb_abbsv11$pos_abbott_s2/tb_abbsv11$n
-
-# clean NAs
-tb_abbsv11<-tb_abbsv11[!is.na(tb_abbsv11$samplemonth),]
-
-tb_abbsv11_melt<-reshape2::melt(select(tb_abbsv11, -c(2:4)), 
-                        id.vars = c("samplemonth"),
-                        variable.name = "COI",
-                        value.name = "prop")
-```
-
-```{r}
-# Plot: 1 dose only
-p09v11<-tb_abbsv11_melt %>%
+# Plot: 2 doses +
+p09v2<-tb_abbsv2_melt %>%
   ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
@@
   theme(legend.position='top')+
-  ggtitle("Stratum: 1 dose only")
-p09v11
+  ggtitle("Stratum: 2 doses +")
+p09v2
 ```
+
+##### Is 1-doser the cause of the "peak" and "bucket"?
 
@@
 ```
+
+## 4.4 The vaccination policy and those "1-Doser"
+
+#### 4.4.1 add prop (1-dose only)
+```{r}
+# number of personals in each month
+tb_abbsvall<-APLs %>%
+  .[!(.$vax %in% c("No_ULI")),] %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
+            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
+            )
+tb_abbsvall$n.all<-tb_abbsvall$n
+
+tb_abbsv1<-merge(tb_abbsv1, select(tb_abbsvall, samplemonth, n.all),
+      by="samplemonth", all.x = T, all.y = F)
+tb_abbsv1$prop.d1<-tb_abbsv1$n/tb_abbsv1$n.all
+```
+
+#### 4.4.2 Making the plot:
+```{r}
+p09v1a<-tb_abbsv1 %>%
+  ggplot(aes(x=samplemonth)) +
+  geom_point(aes(y=prop.abbs1), colour="red") +
+  geom_point(aes(y=prop.abbs2), colour="red") +
+  geom_line(aes(y=prop.abbs1), colour="red")  +
+  geom_line(aes(y=prop.abbs2), linetype=2, colour="red")  +
+  geom_line(aes(y=prop.d1), colour="darkgrey") +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_linetype_manual("", values=c(1,3), labels = c("APL-Abbott (COI=50)", 
+                                                      "APL-Abbott (COI=402)"))+
+  theme(legend.position='top')+
+  ggtitle("Stratum: 1 dose only")
+p09v1a
+```
+
+## 4.4.3 Check the time gap btw 1st dose and 2nd dose. 
+```{r}
+APL_vax<-RFD4682_v2_vaxe %>%
+  select(ULI, ULI_IDe, DOSE_COUNT, IMMUNIZATION_DATE) %>%
+  .[.$DOSE_COUNT==1 | .$DOSE_COUNT==2, ]
+
+APL_vax<-reshape(APL_vax[,c(2:4)], idvar="ULI_IDe", timevar = "DOSE_COUNT",
+        direction = "wide")
+
+# Calculate time gap btw 1st and 2nd dose
+APL_vax$time_gap<-as.Date(APL_vax$IMMUNIZATION_DATE.2)-
+                  as.Date(APL_vax$IMMUNIZATION_DATE.1)
+
+# check distribution
+hist(as.numeric(APL_vax$time_gap), breaks=25)
+
+summary(as.numeric(APL_vax$time_gap))
+```

### 2024-06-18 2_scripts/2.4 PRNT50.qmd
Update figures for the manuscript.

@@
 
-#### Figure panel
-
-```{r}
-# use cowplot package
-p09all.upper<-cowplot::plot_grid(p09c + ylab("Proportion") +
-                           theme(legend.position='top') +
-                           guides(color = guide_legend(nrow = 2)), 
-                           p09v0 + xlab ("") + ylab(""), nrow=1,
-                           labels = c("A", "B"),  align = "h")
-p09all.upper
-
-p09all.lower<-cowplot::plot_grid(
-                           NULL,  p09v1 + theme(legend.position="none"), nrow=1,
-                           labels = c("", "C"))
-p09all.lower
-
-png(file="../4_output/figs/Fig.4 Immunity.png", 
-    width=2400, height=2000, res=300)
-cowplot::plot_grid(p09all.upper, p09all.lower, nrow=2, rel_heights = c(1, 0.8))
-dev.off()
-```
-
 ## 4.2 The vaccination rate among APL
@@
 
-## 4.4.3 Check the time gap btw 1st dose and 2nd dose. 
+#### 4.4.3 Check the time gap btw 1st dose and 2nd dose. 
 ```{r}
@@
 ```
+##  4.5 Adding the proportion of each stratum
+
+### Make the plot for three strata
+
+#### The unvaccinated stratum
+```{r}
+# add proportion of the stratum into the data table
+tb_abbsv0<-merge(tb_abbsv0, select(tb_abbsv1, samplemonth, n.all),
+      by="samplemonth", all.x=T, all.y=F)
+tb_abbsv0<-tb_abbsv0 %>%
+  mutate(
+    prop.stra = n/n.all
+  )
+p09v0.lab<-c("APL-Abbott (COI=50)", "APL-Abbott (COI=402)",
+             "Proportion of the stratum")
+
+tb_abbsv0_melt<-reshape2::melt(select(tb_abbsv0, -c(2:4,7)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "COI",
+                        value.name = "prop")
+
+# making the plot
+p09v0<-tb_abbsv0_melt %>%
+  ggplot(aes(x=samplemonth, y=prop)) +
+  geom_point(aes(colour=COI)) +
+  geom_line(aes(linetype=COI, colour=COI))  +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_colour_manual("", values = c("red", "red", "darkgrey"),
+                      labels = p09v0.lab)+
+  scale_linetype_manual("", values=c(1,3,3), labels = p09v0.lab) +
+  theme(legend.position='top') +
+  guides(color = guide_legend(nrow = 2)) +
+  ggtitle("Stratum: unvaccinated")
+p09v0
+```
+
+#### The 1-dose only stratum
+```{r}
+tb_abbsv1_melt<-reshape2::melt(select(tb_abbsv1, -c(2:4,7)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "COI",
+                        value.name = "prop")
+
+p09v1<-tb_abbsv1_melt %>%
+  ggplot(aes(x=samplemonth, y=prop)) +
+  geom_point(aes(colour=COI)) +
+  geom_line(aes(linetype=COI, colour=COI))  +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_colour_manual("", values = c("red", "red", "darkgrey"),
+                      labels = p09v0.lab)+
+  scale_linetype_manual("", values=c(1,3,3), labels = p09v0.lab) +
+  theme(legend.position='none') +
+  guides(color = guide_legend(nrow = 2)) +
+  ggtitle("Stratum: 1 dose only")
+p09v1
+```
+
+#### The 2-dose+ stratum
+```{r}
+tb_abbsv2<-tb_abbsv2[-1,]
+
+# add proportion of the stratum into the data table
+tb_abbsv2<-merge(tb_abbsv2, select(tb_abbsv1, samplemonth, n.all),
+      by="samplemonth", all.x=T, all.y=F)
+tb_abbsv2<-tb_abbsv2 %>%
+  mutate(
+    prop.stra = n/n.all
+  )
+```
+
+```{r}
+# melt the data tab
+tb_abbsv2_melt<-reshape2::melt(select(tb_abbsv2, -c(2:4,7)), 
+                        id.vars = c("samplemonth"),
+                        variable.name = "COI",
+                        value.name = "prop")
+
+p09v2<-tb_abbsv2_melt %>%
+  ggplot(aes(x=samplemonth, y=prop)) +
+  geom_point(aes(colour=COI)) +
+  geom_line(aes(linetype=COI, colour=COI))  +
+  theme_bw() + 
+  ylim(c(0,1)) +
+  xlab("Time of Sample Collection") + 
+  ylab("Proportion")+ 
+  scale_colour_manual("", values = c("red", "red", "darkgrey"),
+                      labels = p09v0.lab)+
+  scale_linetype_manual("", values=c(1,3,3), labels = p09v0.lab) +
+  theme(legend.position='none') +
+  guides(color = guide_legend(nrow = 2)) +
+  ggtitle("Stratum: 2 doses +")
+p09v2
+```
+
+### merge all four into one figue panel
+```{r}
+rm(p09v0a, p09v1a, p09v2a)
+```
+
+#### Figure panel
+```{r}
+# use cowplot package
+p09.upper<-cowplot::plot_grid(p09c + ylab("Proportion") + xlab ("") +
+                           theme(legend.position='top') +
+                           guides(color = guide_legend(nrow = 2)), 
+                           p09v0 + xlab ("") + ylab(""),
+                           labels = c("A", "B"),
+                           nrow=1,  align = "h")
+
+p09.lower<-cowplot::plot_grid(p09v1, p09v2 + ylab(""),
+                              labels = c("C", "D"),
+                              nrow=1,  align = "h")
+p09all<-cowplot::plot_grid(p09.upper, p09.lower,
+                              labels = c("", ""),
+                              nrow=2,  align = "v", rel_heights = c(1,0.8))
+p09all
+
+png(file="../4_output/figs/Fig.4 Immunity.png", 
+    width=2400, height=2000, res=300)
+p09all
+dev.off()
+```

### 2024-08-08 2_scripts/2.4 PRNT50.qmd
Codes for vax%

@@
 
+APLs$pos_abbott_s_na<-is.na(APLs$pos_abbott_s)
 tb_abbs<-APLs %>%
@@
   summarise(n = n() , 
+            n_na_s = sum(pos_abbott_s_na, na.rm = T),
             pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
@@
             )
-tb_abbs$prop.abbs1<-tb_abbs$pos_abbott_s/tb_abbs$n
-tb_abbs$prop.abbs2<-tb_abbs$pos_abbott_s2/tb_abbs$n
+tb_abbs$prop.abbs1<-tb_abbs$pos_abbott_s/(tb_abbs$n-tb_abbs$n_na_s)
+tb_abbs$prop.abbs2<-tb_abbs$pos_abbott_s2/(tb_abbs$n-tb_abbs$n_na_s)
+tb_abbs$prop.abbs1[1:9]<-rep("NA",9)
+tb_abbs$prop.abbs2[1:9]<-rep("NA",9)
 
@@
 # create vax% cat var
-APLs$vax_cat<-rep(NA, 214780)
+APLs$vax_cat<-rep(NA, length(APLs$clean_IDe))
 APLs$vax_cat[APLs$vax %in% c("1 Dose","1-2 Doses","2 Doses","3 Doses",
@@
 tb_vax<-APLs %>%
-  .[.$abb_s_AU==1,] %>%
   group_by(samplemonth) %>%
@@
 **Alton's comments:** "It might be worth comparing to reported vax rate for Alberta similar to what Sheila and Steve did in that figure." As they used (S+ N-)% to estimate the vax%, it will inevitably results in underestimation, as those who are vaccinated and infected was not counted.
+An alternative solution is to estimate vax% using (S+ N-) by (N-), assuming vaccination is not related to anti N positivity (which is probably not true and we can cross-verify using our own data).
 
@@
 
-APLs$vax_est<-rep(0, 214780)
+APLs$vax_est<-rep(0, length(APLs$clean_IDe))
 APLs$vax_est[APLs$pos_abbott==0 & APLs$pos_abbott_s==1]<-1
@@
 ```{r}
+# Modify APLs data, add na indicator, N=69108
+table(APLs$pos_abbott, useNA = "ifany")
+APLs$pos_abbott_na <- is.na(APLs$pos_abbott)
+table(APLs$pos_abbott, APLs$pos_abbott_na, useNA = 'ifany')
+
 tb_vax_est<-APLs %>%
-#  .[!is.na(.$pos_abbott_s) & !is.na(.$pos_abbott),] %>%
-  group_by(samplemonth) %>%
-  summarise(n = n() , 
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            n_pos = sum(pos_abbott, na.rm = T),
+            n_na = sum(pos_abbott_na, na.rm=T),
             n_vax= sum(vax_est, na.rm = T)
             )
-tb_vax_est$prop.vax<-tb_vax_est$n_vax/tb_vax_est$n
+tb_vax_est$n_neg = tb_vax_est$n - tb_vax_est$n_pos - tb_vax_est$n_na
+tb_vax_est$prop.vax<-tb_vax_est$n_vax/tb_vax_est$n_neg
 
 # clean NAs
+tb_vax_est$prop.vax[1:17]<-rep(NA, 17)
 tb_vax_est<-tb_vax_est[!is.na(tb_vax_est$samplemonth),]
@@
 colnames(d01)<-c("samplemonth", "prop.vax", "data")
+d01$prop.vax = d01$prop.vax/100
 ```
@@
 View(tb_vax_est)
-d02<-select(tb_vax_est, c(1,4))
+d02<-select(tb_vax_est, samplemonth, prop.vax)
 d02$data<-rep("S_pos_N_Neg", length(d02$samplemonth))
@@
 View(tb_abbs)
-d03<-select(tb_abbs, samplemonth, prop.abbs1, study)
-colnames(d03)<-c("samplemonth", "prop.vax", "data")
+d03<-select(tb_abbs, samplemonth, prop.abbs1)
+colnames(d03)<-c("samplemonth", "prop.vax")
 d03$data<-rep("S_pos", length(d03$samplemonth))
@@
 rm(d01, d02, d03, d04)
-
-dta.vax$prop.vax[dta.vax$data=="AB_general"]<-dta.vax$prop.vax[dta.vax$data=="AB_general"]/100
+dta.vax$prop.vax<-as.numeric(dta.vax$prop.vax)
 ```
@@
   theme_bw() + 
-  ylim(c(0,1)) +
-  ylab("Vaccinated%")
+  ylim(0,1)+
+  ylab("Vaccinated%") +
+  scale_colour_manual("", values = c("orange", "red", "red", "purple"),
+                      labels = c("AB General Pop.", "APL Vax%", 
+                                 "Anti-S(+)", "Anti-S(+) N(-)")
+                      )
+#  scale_linetype_manual("", values=c(3,3,1,1), labels = p09.lab) 
 p12
+```
+
+#### 4.2.4 Apply (S+ N-)/N- for vax% estimation
+
+First we calculate (S+ N-)/N- among CBS samples
+```{r}
+table(CBSs_AB$pos_roche, CBSs_AB$pos_roche_s, useNA = 'ifany')
+# good data quality, only two NAs in roche_n and roche_s
+
+# remove two NAs
+CBSs_AB<-CBSs_AB[!(is.na(CBSs_AB$pos_roche) | is.na(CBSs_AB$pos_roche_s)), ]
+
+# add vax_est
+CBSs_AB$vax_est<-rep(0, length(CBSs_AB$pid))
+CBSs_AB$vax_est[CBSs_AB$pos_roche==0 & CBSs_AB$pos_roche_s==1]<-1
+table(CBSs_AB$vax_est)
+```
+
+Then we calculate the vax% using S(+) and (S+ & N-)/(N-)
+```{r}
+tbCBS_vax_est<-CBSs_AB %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            n_pos = sum(pos_roche, na.rm = T),
+            s_pos = sum(pos_roche_s, na.rm = T),    
+            n_vax= sum(vax_est, na.rm = T)
+            )
+tbCBS_vax_est$n_neg = tbCBS_vax_est$n - tbCBS_vax_est$n_pos
+tbCBS_vax_est$prop.vax<-tbCBS_vax_est$n_vax/tbCBS_vax_est$n_neg
+
+## add CBS data to merged data
+d05<-select(tbCBS_vax_est, c(1,7))
+colnames(d05)<-c("samplemonth", "prop.vax")
+d05$data<-rep("CBS: S_pos_N_Neg", length(d05$samplemonth))
+
+dta.vax<-rbind.data.frame(dta.vax, d05)
+```
+
+#### Now redo the plot
+```{r}
+# Add study and adjustment
+dta.vax$study[dta.vax$data %in% c("AB_general")]<-"AB_general"
+dta.vax$study[dta.vax$data %in% c("CBS: S_pos_N_Neg")]<-"CBS"
+dta.vax$study[dta.vax$data %in% c("APL_vax", "S_pos", "S_pos_N_Neg")]<-"APL"
+
+dta.vax$adj[dta.vax$data %in% c("AB_general", "APL_vax")]<-"raw"
+dta.vax$adj[dta.vax$data %in% c("CBS: S_pos_N_Neg", "S_pos_N_Neg")]<-"S_pos_N_Neg"
+dta.vax$adj[dta.vax$data %in% c("S_pos")]<-"S_pos"
+
+p12.lab <- c("AB General Pop.", "APL Vax%", 
+             "APL: Anti-S(+)", "APL: Anti-S(+) N(-)", "CBS: Anti-S(+) N(-)")
+p12<-dta.vax %>%
+  ggplot(aes(x=samplemonth, y=prop.vax, colour=interaction(study, adj),
+                                        linetype=interaction(study, adj))) +
+  geom_point() +
+  geom_line()  +
+  theme_bw() + 
+  ylim(0,1)+
+  ylab("Vaccinated%") +
+  scale_colour_manual("", values = c("orange", "red", "red","purple", "blue"),
+                      labels = p12.lab)+
+  scale_linetype_manual("", values=c(1,1,3,3,3), labels = p12.lab) 
+p12
+```
+#### Print p12
+```{r}
+png(file="../4_output/figs/Fig.S2 Vax.png", 
+    width=2400, height=1440, res=300)
+p12
+dev.off()
 ```

### 2025-04-16 2_scripts/2.4 PRNT50.qmd
Update figures and misc analysis

@@
 # ggplot
-p09.lab<-c("APL-Abbott (COI=402)", "CBS-Roche (COI=179)", 
-           "APL-Abbott (COI=50)", "CBS-Roche (COI=0.8)")
+p09.lab<-c("APL (PRNT-50 cutoff)", "CBS (PRNT-50 cutoff)", 
+           "APL (manufacturer cutoff)", "CBS (manufacturer cutoff)")
 p09c<-tb_APLCBS2 %>%
@@
                       labels = p09.lab)+
-  scale_linetype_manual("", values=c(3,3,1,1), labels = p09.lab) 
+  scale_linetype_manual("", values=c(3,3,1,1), labels = p09.lab)+
+  ggtitle("Unstratified")
 #  theme(legend.position="top")+
@@
   )
-p09v0.lab<-c("APL-Abbott (COI=50)", "APL-Abbott (COI=402)",
+p09v0.lab<-c("APL (manufacturer cutoff)", "APL (PRNT-50 cutoff)",
              "Proportion of the stratum")
@@
 p09.upper<-cowplot::plot_grid(p09c + ylab("Proportion") + xlab ("") +
+                           coord_cartesian(xlim=as.yearmon(c("Jan 2021", "Oct 2022"))) +
                            theme(legend.position='top') +
@@
 ```
+## 5. Additional analysis for manuscript
+```{r}
+# median age in CBS Alberta and APL
+summary(CBSs_AB$age) 
+summary(APLs$age) # age=0?
+
+APLs %>%
+  .[.$age %in% c(0:17),]
+
+# age categories
+CBSs_AB$age_cat[CBSs_AB$age %in% c(16:17)]<-"<18"
+CBSs_AB$age_cat[CBSs_AB$age %in% c(18:35)]<-"18-35"
+CBSs_AB$age_cat[CBSs_AB$age %in% c(36:64)]<-"36-64"
+CBSs_AB$age_cat[CBSs_AB$age %in% c(65:199)]<-"65+"
+table(CBSs_AB$age_cat, useNA = "ifany")
+prop.table(table(CBSs_AB$age_cat, useNA = "ifany"))
+
+APLs$age_cat[APLs$age %in% c(0:17)]<-"<18"
+APLs$age_cat[APLs$age %in% c(18:35)]<-"18-35"
+APLs$age_cat[APLs$age %in% c(36:64)]<-"36-64"
+APLs$age_cat[APLs$age %in% c(65:199)]<-"65+"
+table(APLs$age_cat)
+prop.table(table(APLs$age_cat))
+
+# Sex
+table(CBSs_AB$sex)
+prop.table(table(CBSs_AB$sex))
+
+table(APLs$gender)
+prop.table(table(APLs$gender))
+```
+
+Among samples from our auxiliary datasets tested with both Abbott and Roche Anti-N assays, the COI was lower with Abbott than with Roche in X% of samples
+```{r}
+# This involves directly comparison btw Roche anti-N and Abbott-anti-N
+
+# How many samples each?
+table(dta_all$study)
+
+# Non missing in Abbott & Roche anti-N
+table(dta_all[(!is.na(dta_all$roche_n) & !is.na(dta_all$abbott_n)),]$study)
+dta_nmiss<-dta_all[(!is.na(dta_all$roche_n) & !is.na(dta_all$abbott_n)),]
+
+table(dta_nmiss$abbott_n<dta_nmiss$roche_n)
+778/(778+43)
+```
+
+```{r}
+#Numbers at 2022 July
+tb_melt_2 %>%
+  .[.$samplemonth == as.yearmon("July 2022"),]
+```
+
+#### PCR-verified test positivity:
+```{r}
+# How many test results are previously PCR-verified?
+
+# CBS data (AB subset) - ver Dec 2023, this should be the same ver of CBS data we received by late Nov of 2023, the same ver of data we used for the project. No PCR test results were found.
+
+# How many records in AB?
+CBSs %>%
+  filter(province=="AB") %>%  # N=124,008
+  select(pid, sampledate) %>%
+  unique()                    # No duplicated record, good
+
+# APL data: check the APL data ev2, this one is uploaded to the jmy folder on the server
+load("/home/jchen/../../jmy/APL/RFD4682_ev2.RData")
+
+# N=214776, de-duplicated, good.
+
+# Is sero-id unique (this is the sample ID)
+length(unique(RFD4682_ev2$SERO_ID)) # N=214,776, good
+length(unique(RFD4682_ev2$SID)) # N=213,139, duplicates, use SERO_ID
+
+# how many has valid PCR test results?
+t1<-table(RFD4682_ev2$NAAT_STATUS, useNA = 'ifany')
+t1
+prop.table(t1)
+```
+
+```{r}
+# Unstratified test positivity 
+# Abbott Anti-N
+RFD4682_ev2 %>%
+  filter(NAAT_STATUS == "Positive") %>%
+  count(`N-IgG_INTERP`)
+
+# Abbott Anti-S-RBD
+RFD4682_ev2 %>%
+  filter(NAAT_STATUS == "Positive") %>%
+  count(`RBD-IgGII_INTERP`)
+```
+```{r}
+# Stratified analysis
+RFD4682_ev2$DAYS_SINCE_NAAT<-as.numeric(RFD4682_ev2$DAYS_SINCE_NAAT)
+hist(RFD4682_ev2$DAYS_SINCE_NAAT)
+
+RFD4682_ev2$NAAT_month_cat<-rep(NA, length(RFD4682_ev2$SERO_ID))
+RFD4682_ev2[!is.na(RFD4682_ev2$DAYS_SINCE_NAAT) &
+            0<=RFD4682_ev2$DAYS_SINCE_NAAT &
+            RFD4682_ev2$DAYS_SINCE_NAAT<90,]$NAAT_month_cat<-"<3 months"
+RFD4682_ev2[!is.na(RFD4682_ev2$DAYS_SINCE_NAAT) &
+            90<=RFD4682_ev2$DAYS_SINCE_NAAT &
+            RFD4682_ev2$DAYS_SINCE_NAAT<180,]$NAAT_month_cat<-"3-6 months"
+RFD4682_ev2[!is.na(RFD4682_ev2$DAYS_SINCE_NAAT) &
+            180<=RFD4682_ev2$DAYS_SINCE_NAAT &
+            RFD4682_ev2$DAYS_SINCE_NAAT<360,]$NAAT_month_cat<-"6-12 months"
+RFD4682_ev2[!is.na(RFD4682_ev2$DAYS_SINCE_NAAT) &
+            360<=RFD4682_ev2$DAYS_SINCE_NAAT,]$NAAT_month_cat<-">12 months"
+table(RFD4682_ev2$NAAT_month_cat, useNA = 'ifany')
+
+RFD4682_ev2 %>%
+  filter(NAAT_STATUS == "Positive") %>%
+  group_by(NAAT_month_cat) %>%
+  count(`N-IgG_INTERP`)
+```

### 2025-04-17 2_scripts/2.3 Prob Model for PCR pos.qmd
Update figures for the manuscript.
Updated Fig 1 for rev 1

@@
   xlab("Time of Sample Collection") + 
-  ylab("Sero-prevalence")+ 
+  ylab("Seroprevalence")+ 
   scale_colour_manual("",   values = c("red", "red", "blue", "blue"))+
@@
 # use this chunk to skip the paste()
-p03.lab<-c("APL (raw)", "CBS (raw)", 
-           "APL (transformed)", "CBS (transformed)")
+p03.lab<-c("APL Abbott (actual)", "CBS Roche (actual)", 
+           "APL Roche (transformed)", "CBS Abbott (transformed)")
 p03.3<-tb_alberta %>%
@@
   xlab("Time of Sample Collection") + 
-  ylab("Sero-prevalence")+ 
+  ylab("Seroprevalence")+ 
   scale_colour_manual("", values = c("red", "blue", "red", "blue"),
@@
 # rebuild the figure
-p03.5lab<-c("APL (raw)", "CBS (raw)", 
-            "APL (RG-Bayesian)", "CBS (RG-Bayesian)")
+p03.5lab<-c("APL Abbott (actual)", "CBS Roche (actual)", 
+            "APL Abbott (RG-Bayesian)", "CBS Roche (RG-Bayesian)")
 
@@
   xlab("Time of Sample Collection") + 
-  ylab("Sero-prevalence") +
+  ylab("Seroprevalence") +
   scale_colour_manual("", values = c("red", "blue", "red", "blue"),
@@
                    labels = c("","C"), ncol=1, rel_heights = c(1, 1))
+dev.off()
+```
+
+#### 3.6 Create fig panel for the manuscript
+```{r}
+# use cowplot package
+p08.upper<-cowplot::plot_grid(p03.5  + ylab("Seroprevalence") +
+                              ylim(0,0.8), 
+                              p03.3  + ylab("") +
+                              ylim(0,0.8)+
+                              scale_y_continuous(breaks = c(0,0.2,0.4,0.6,0.8)),
+                              nrow=1, align = "h",
+                              labels = c("A", "B"))
+p08.upper
+
+png(file="../4_output/figs/Fig.1m Seroprevalence.png", 
+    width=2400, height=1000, res=300)
+p08.upper
 dev.off()

### 2025-04-17 2_scripts/2.4 PRNT50.qmd
Updated Fig 1 for rev 1

@@
 ## 5. Additional analysis for manuscript
+
+####  5.1 Some desc analysis
 ```{r}
@@
 
-#### PCR-verified test positivity:
+#### 5.2 PCR-verified test positivity:
+We did this analysis as suggested by the reviewer-1
 ```{r}

### 2025-05-12 2_scripts/2.4 PRNT50.qmd
Add Panel B for Figure S.2

@@
 load(file = "../1_data/private/APLs.RData") # a clean subset
+# Updated APL v2 data (ID cleaned and encrypted)
+load(file = "../1_data/private/RFD4682_ev2.RData")
 ```
@@
 APLs<-merge(APLs, select(RFD4682_ev2, order_ID, 
-                   `RBD-IgGII_RESULT`, `RBD-IgGII_INTERP`,`RBD-IgGII_RESULT_DATE`
-                   ), by="order_ID",all=T)
-colnames(APLs)<-c(names(APLs)[1:10], "abbott_s", "pos_abbott_s", "month_abbott_s")
+                   `RBD-IgGII_RESULT`, `RBD-IgGII_INTERP`,
+                   `RBD-IgGII_RESULT_DATE`,
+                   VACCINATION_STATUS,
+                   ), by="order_ID",all.x=T, all.y=F)
+
+# Remove duplicated records (N=4)
+APLs<-APLs %>%
+  filter(order_ID %in% RFD4682_ev2$order_ID) %>%
+  rename(abbott_s = `RBD-IgGII_RESULT`,
+         pos_abbott_s = `RBD-IgGII_INTERP`,
+         date_abbott_s = `RBD-IgGII_RESULT_DATE`,
+         vax = VACCINATION_STATUS)
+APLs$month_abbott_s<-as.yearmon(APLs$date_abbott_s)
 
@@
 ```{r}
+CBS<-arrange(CBS, pid)
+CBSs<-arrange(CBSs, pid)
 CBSs$pos_roche_s<-CBS$interp_roche_s
@@
 
+```{r}
+# Save the modified CBSs, APLs, and CBSs_AB data
+save(APLs, file = "../1_data/private/APLs.RData")
+save(CBSs, file = "../1_data/private/CBSs.RData")
+save(CBSs_AB, file = "../1_data/private/CBSs_AB.RData")
+```
+
 **Notes:** After increasing the COI from 0.8 to 178.81, 6732 of 108,514 (6.20%) samples were re-categorized as negative.
@@
 
-#### 2.2 Roche-S \<--\> Abbott-S transformation
+#### 2.2 Roche-S --> Abbott-S transformation
+This part is skipped and not mentioned in the manuscript.
 
@@
 The APL data Carmen shared with us has vaccination info, the 2024 Jan ver of it even has the vax brand.
-While the CBS data we have doesnot include vax info.
+While the CBS data we have no vax info.
 However, a paper written by Sheila has the vax% of those in CBS data, which we can use for comparison on vax%.
@@
 # create vax% cat var
-APLs$vax_cat<-rep(NA, length(APLs$clean_IDe))
+APLs$vax_cat<-rep(NA, length(APLs$order_ID))
 APLs$vax_cat[APLs$vax %in% c("1 Dose","1-2 Doses","2 Doses","3 Doses",
@@
 
-APLs$vax_est<-rep(0, length(APLs$clean_IDe))
+APLs$vax_est<-rep(0, length(APLs$order_ID))
 APLs$vax_est[APLs$pos_abbott==0 & APLs$pos_abbott_s==1]<-1
@@
 ```
-
 #### 4.2.4 Apply (S+ N-)/N- for vax% estimation
@@
                       labels = p12.lab)+
-  scale_linetype_manual("", values=c(1,1,3,3,3), labels = p12.lab) 
+  scale_linetype_manual("", values=c(1,1,3,3,3), labels = p12.lab) +
+  theme(legend.position = c(0.8,0.25))
 p12
@@
 p12
+dev.off()
+```
+#### 4.2.5 Figure S2 Panel B using transformed data
+```{r}
+# APL S+/N- N is transformed (Abbott-->Roche)
+
+table(APLs$pos_abbott, APLs$pos_roche, useNA = 'ifany')
+# if transformed into Roche, N=11501 participants will be identified as positive
+
+table(APLs$pos_abbott_s, APLs$pos_roche, useNA = 'ifany')
+
+# Add vax indicator using S+ and N- (transformed N Abb->Ro)
+APLs$vax_est_t<-rep(0, length(APLs$order_ID))
+APLs$vax_est_t[APLs$pos_roche==0 & APLs$pos_abbott_s==1]<-1
+APLs$vax_est_t[is.na(APLs$pos_roche) | is.na(APLs$pos_abbott_s)]<-NA
+table(APLs$vax_est_t, useNA = 'ifany')
+
+tbNT_vax_est<-APLs %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            n_pos = sum(pos_roche, na.rm = T),
+            s_pos = sum(pos_abbott_s, na.rm = T),    
+            n_vax= sum(vax_est_t, na.rm = T)
+            )
+tbNT_vax_est$n_neg = tbNT_vax_est$n - tbNT_vax_est$n_pos
+tbNT_vax_est$prop.vax<-tbNT_vax_est$n_vax/tbNT_vax_est$n_neg
+
+# clean the table
+tbNT_vax_est<-tbNT_vax_est[!is.na(tbNT_vax_est$samplemonth),]
+tbNT_vax_est$prop.vax[tbNT_vax_est$prop.vax==0]<-NA
+
+## add CBS data to merged data
+d06<-select(tbNT_vax_est, samplemonth, prop.vax)
+d06$data<-rep("S_pos_N_Neg (Anti-N Transformed)", length(d06$samplemonth))
+
+#dta.vax<-rbind.data.frame(dta.vax, d05)
+```
+
+```{r}
+# CBS S+/N- N is transformed (Roche-->Abbott)
+
+# Add vax indicator using S+ and N- (transformed N Abb->Ro)
+CBSs_AB$vax_est_t<-rep(0, length(CBSs_AB$pid))
+CBSs_AB$vax_est_t[CBSs_AB$pos_abbott==0 & CBSs_AB$pos_roche_s==1]<-1
+CBSs_AB$vax_est_t[is.na(CBSs_AB$pos_abbott) | is.na(CBSs_AB$pos_roche_s)]<-NA
+table(CBSs_AB$vax_est_t, useNA = 'ifany')
+
+tbCBST_vax_est<-CBSs_AB %>%
+  group_by(samplemonth) %>%
+  summarise(n = n() , 
+            n_pos = sum(pos_abbott, na.rm = T),
+            s_pos = sum(pos_roche_s, na.rm = T),    
+            n_vax= sum(vax_est_t, na.rm = T)
+            )
+tbCBST_vax_est$n_neg = tbCBST_vax_est$n - tbCBST_vax_est$n_pos
+tbCBST_vax_est$prop.vax<-tbCBST_vax_est$n_vax/tbCBST_vax_est$n_neg
+
+## add CBS data to merged data
+d07<-select(tbCBST_vax_est, samplemonth, prop.vax)
+d07$data<-rep("CBS: S_pos_N_Neg (trasformed)", length(d07$samplemonth))
+
+dta.vax.2<-rbind.data.frame(d06, d07, d02, d05)
+
+```
+
+```{r}
+# Use figure to present data
+p13.lab <- c("CBS Raw: Anti-S(+) N(-)", "CBS Transformed: Anti-S(+) N(-)", 
+             "APL Raw: Anti-S(+) N(-)", "APL Transformed: Anti-S(+) N(-)")
+p13<-dta.vax.2 %>%
+  ggplot(aes(x=samplemonth, y=prop.vax, colour=data, linetype=data)) +
+  geom_point() +
+  geom_line()  +
+  theme_bw() + 
+  ylim(0,1)+
+  ylab("Vaccinated%") +
+  scale_colour_manual("", values = c("blue", "brown", "purple","green"),
+                      labels = p13.lab)+
+  scale_linetype_manual("", values=c(3,3,3,3), labels = p13.lab)+
+  theme(legend.position = c(0.8,0.2))
+p13
+```
+##### Print p12 and p13 together
+```{r}
+p1213<-plot_grid(p12 + xlab ("Time of Sample Collection") +
+                 coord_cartesian(xlim=as.yearmon(c("Oct 2020", "Feb 2023")))+
+                 theme(legend.position = c(0.67,0.25)), 
+                 p13 + xlab ("Time of Sample Collection") + ylab("") +
+                 coord_cartesian(xlim=as.yearmon(c("Oct 2020", "Feb 2023")))+
+                 theme(legend.position = c(0.65,0.2)),
+                 labels = c("A", "B"),
+                 nrow=1,  align = "h")
+p1213
+
+png(file="../4_output/figs/Fig.S2 Vax.png", 
+    width=2400, height=1200, res=300)
+p1213
 dev.off()
@@
 ```
+#### 5.3 Modify the legends and y-axis labels of figure 1
+```{r}
+# Check the codes in qmd script 2.3
+# use key words "Fig.1" "Fig.2" to search
+```

### 2025-07-03 2_scripts/2.4 PRNT50.qmd
Redo figures in pdf

@@
 
-tb_abbs_melt<-reshape2::melt(select(tb_abbs, -c(2:4)), 
+tb_abbs_melt<-reshape2::melt(select(tb_abbs, 
+                                    samplemonth, prop.abbs1, prop.abbs2), 
                         id.vars = c("samplemonth"),
@@
 tb_APLCBS2<-rbind.data.frame(tb_abbs_melt, tb_ros_melt)
-rm(tb_abbs, tb_abbs_melt, tb_ros, tb_ros_melt)
+#rm(tb_abbs, tb_abbs_melt, tb_ros, tb_ros_melt)
 
@@
 tb_APLCBS2$COI_ty[tb_APLCBS2$COI %in% c("402.36", "178.81")]<-"Immunity"
+
+tb_APLCBS2<-tb_APLCBS2 %>%
+  mutate(prop=as.numeric(prop))
 ```
@@
 p09c<-tb_APLCBS2 %>%
-  ggplot(aes(x=samplemonth, y=prop, colour=interaction(study, COI_ty),
-                                    linetype=interaction(study, COI_ty))) +
+  filter(!is.na(COI_ty)) %>%
+  ggplot(aes(x=samplemonth, y=prop, colour=interaction(study, COI_ty))) +
   geom_point() +
-  geom_line()  +
+  geom_line(aes(linetype = interaction(study, COI_ty)))  +
   theme_bw() +
@@
 ```
+
 ##### Print p12 and p13 together
@@
 dev.off()
+```
+##### Print p12 and p13 in pdf formats, as requested by AJE
+```{r}
+ggsave(file="../4_output/figs/Fig.S2 Vax.pdf", 
+       width=2400, height=1200, dpi=300, units = "px", 
+       plot = p1213)
 ```
@@
 
-png(file="../4_output/figs/Fig.4 Immunity.png", 
-    width=2400, height=2000, res=300)
-p09all
-dev.off()
+ggsave(file="../4_output/figs/Fig.3 Immunity.pdf",
+       plot = p09all,
+        width=2400, height=2000, 
+        dpi=300, units ="px")
 ```

### 2025-07-07 2_scripts/2.1 APLv1 Data Cleaning.qmd
Removed unnecessary scripts

--- 2024-06-10 2_scripts/2.1 APL Data Quality Check.qmd
@@
 # Second version of APL: provided by Carmen in Jan 2024
-
+library(readxl)
+RFD4682_v2 <- read_excel("../1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
+    sheet = "RFD4682_Full_Data", col_types = c("text", 
+        "text", "text", "text", "text", 
+        "date", "text", "text", "date", "date", 
+        "text", "text", "numeric", "text", 
+        "text", "text", "text", "text", "numeric", 
+        "numeric", "text", "text", 
+        "numeric", "numeric", "numeric"))
+View(RFD4682_v2)
 ```

### 2025-07-07 2_scripts/2.1.1 APLv2 Data Cleaning.qmd
Removed unnecessary scripts

--- 2024-06-13 2_scripts/2.1 APL Data Quality Check_new.qmd
@@
 Estimation on number of individuals, using cleaned version of IDe
+
 ```{r}
@@
 Some additional cleaning/editing, making anti-S quantitative
+
 ```{r}
@@
 ##### Yuan found this: anti-N is only available for the Omicron stage
+
 **Notes:** This is an error in two ver of APL. The anti-N assay results are available and misplaced in the anti-S-RBD column.
+
 ```{r}
@@
 #### ADD order ID and other ids -- doesn't work
+
 ```{r}
@@
 #### 2.1 De-duplicate
+
 ```{r}
@@
 ```
+
 ##### De-duplicate SERO_ID for APL v1
+
 ```{r}
@@
 #### 2.2 Add ULI_IDe from v1
+
 ```{r}
@@
       )
-# ULI - ULI_IDe crosswalk
-ULI_crosswalk<-select(RFD4682_v2, ULI_ID, ULI_IDe)
-ULI_crosswalk<-distinct(ULI_crosswalk)
-
-table(is.na(ULI_crosswalk$ULI_IDe))
-ULI_crosswalk %>%
-  .[is.na(.$ULI_IDe),]
-
-# APL ver2 seems have solved missing value issue in the ULI ID, needs to add those
-table(is.na(RFD4682_v2$ULI_ID), is.na(RFD4682_v2$ULI_IDe))
-
-table(is.na(RFD4682_v2$PERSON_ID), is.na(RFD4682_v2$PERSON_IDe))
-
-# encrypt new ULI_ID
-set.seed(240611)
-ULI_crosswalk_e <- deid_dua(ULI_crosswalk, 
-                      id_col = 'ULI_ID',
-                      new_id_name = 'ULI_IDe2',
-                      write_crosswalk = F, id_length = 15)
-
-ULI_crosswalk$ULI_IDe2<-ULI_crosswalk_e$ULI_IDe2
-
-# clean the ULI_ID crosswalk file
-table(is.na(ULI_crosswalk$ULI_ID))
-ULI_crosswalk<-ULI_crosswalk[!is.na(ULI_crosswalk$ULI_ID),]
-ULI_crosswalk<-ULI_crosswalk[-64591,]
-ULI_crosswalk$ULI_IDe[is.na(ULI_crosswalk$ULI_IDe)]<-ULI_crosswalk$ULI_IDe2[is.na(ULI_crosswalk$ULI_IDe)]
-
-# apply the cross-walk to the v2 data
-df.temp1<-RFD4682_v2[!is.na(RFD4682_v2$ULI_ID) & is.na(RFD4682_v2$clean_IDe),]
-df.temp2<-RFD4682_v2[!(!is.na(RFD4682_v2$ULI_ID) & is.na(RFD4682_v2$clean_IDe)),]
-
-df.temp1<-merge(select(df.temp1, -c("ULI_IDe")), 
-             ULI_crosswalk[,1:2],
-      by="ULI_ID", all.x = T, all.y = F)
-df.temp1$clean_IDe<-df.temp1$ULI_IDe
-
-RFD4682_v2<-bind_rows(df.temp1, df.temp2)
-rm(df.temp1, df.temp2, ULI_crosswalk_e)
 ```
@@
 ```
+
 #### 2.3 Clean anti-N and anti-S
@@
 ##### 2.3.1 Anti-N cleaning (remove index)
+
 ```{r}
@@
 ##### 2.3.2 Anti-S cleaning (remove AU/ml)
+
 ```{r}
@@
 ```
+
 #### 2.4 Remove unencryoted individual inditifier
+
 ```{r}
@@
 #### 2.5 Save the encrypted ver of APL data
+
 ```{r}
@@
 ```
-
-## 3. Clean other two APL sub datasets (Vax, NAAT)
-
-## Unfinished ##
-```{r}
-# Add encrypted IDe to vax and NAAT data
-RFD4682_v2_vaxe<-merge(RFD4682_v2_vax, select(ULI_crosswalk, ULI_ID, ULI_IDe),
-      by.x="ULI", by.y="ULI_ID", all.x = T, all.y = F)
-# any missing?
-table(is.na(RFD4682_v2_vaxe$ULI_IDe))
-
-# also encrypt the ULI_ID in the RFD4682_vax
-
-list_temp<-RFD4682_v2_vaxe %>%
-  .[is.na(.$ULI_IDe),]
-```

### 2025-07-07 2_scripts/2.2.1 CBS Preprocessing.qmd
Removed unnecessary scripts

--- 2024-04-05 2_scripts/2.2.a CBS Preprocessing and Fig.qmd
@@
 
-#### 2.2 APL data
-
-**Notes:** Seroprevalence estimation requires individual analysis and a little bit weighting.
-Those who has been sampled multiple times should be down-weighted properly.
-Those who are sampled multiple times in different months will not be down-weighted.
-
-```{r}
-
-```
-
-#### 2.3 Canpath data
-
-**Notes:** In Canpath data we don't have to worry about weighting, as one people has been sampled only once, then tested by multiple asays (either Gingras x3 or Abbott x2)
-
 ##### 2.3.1 Pre-process Canpath data
@@
 ```
-
-#### 3.2 Canpath: make plots by region
-
-```{r}
-# Reshape meltted data
-Canpaths_melt<-reshape2::melt(select(Canpaths_pr, samplemonth, region,
-                                       prop.n, prop.n_RG), 
-                        id.vars = c("samplemonth", "region"),
-                        variable.name = "Adj",
-                        value.name = "prop")
-
-p02<-Canpaths_melt %>%
-  ggplot(aes(x=samplemonth, y=prop, group=Adj, colour=Adj)) +
-  geom_point(aes(colour=Adj)) +
-#  geom_smooth(method = "loess", aes(colour=Adj), se=FALSE) +
-  geom_line(aes(colour=Adj))  + 
-  theme_bw()+
-  scale_y_continuous(limits = c(0,1))
-
-p02 + facet_wrap(~region)
-
-```

### 2025-08-12 2_scripts/2.3 Prob Model for PCR pos.qmd
Minor updates

@@
 ```{r}
+# What is the numbers in 2022 Oct?
+tb_alberta %>%
+  filter(samplemonth == as.yearmon("2022-10-01"))
+
 # How gaps were calculated: 
@@
   select(data, samplemonth, prop, raw) %>%
-  .[.$samplemonth == as.yearmon("Oct 2022"),] %>%
-  mutate(X = .$prop[4] - .$prop[1]) %>%
-  mutate(Y = .$prop[3] - .$prop[1]) %>%
-  mutate(Z = .$prop[4] - .$prop[2])
+  filter(samplemonth == as.yearmon("2022-10-01")) %>%
+  summarise(
+    A = prop[1], B = prop[4],
+    X = .$prop[4] - .$prop[1],
+    Y = .$prop[3] - .$prop[1],
+    Z = .$prop[4] - .$prop[2]
+  )
+
 ```
@@
 #### 3.6 Create fig panel for the manuscript
+
 ```{r}
@@
 ##### 4.3.1 Create monthly anti-N positivity data for CBS
+
 ```{r}
@@
 ##### 4.3.2 Create monthly anti-N positivity data for APL
+
 ```{r}
@@
 ##### 4.3.3 Make plots
+
 ```{r}

### 2025-08-12 2_scripts/2.4 PRNT50.qmd
All figs match w the ver in the rev_1
Minor updates

@@
 
-#### 2.2 Roche-S --> Abbott-S transformation
+#### 2.2 Roche-S --\> Abbott-S transformation
+
 This part is skipped and not mentioned in the manuscript.
@@
 ```
+
 #### 4.2.4 Apply (S+ N-)/N- for vax% estimation
@@
 First we calculate (S+ N-)/N- among CBS samples
+
 ```{r}
@@
 Then we calculate the vax% using S(+) and (S+ & N-)/(N-)
+
 ```{r}
@@
 #### Now redo the plot
+
 ```{r}
@@
 ```
+
 #### Print p12
+
 ```{r}
@@
 ```
+
 #### 4.2.5 Figure S2 Panel B using transformed data
+
 ```{r}
@@
 ##### Print p12 and p13 together
+
 ```{r}
@@
 ```
+
 ##### Print p12 and p13 in pdf formats, as requested by AJE
+
 ```{r}
@@
 #### 4.4.1 add prop (1-dose only)
+
 ```{r}
@@
 #### 4.4.2 Making the plot:
+
 ```{r}
@@
 #### 4.4.3 Check the time gap btw 1st dose and 2nd dose.
+
 ```{r}
@@
 ```
+
 ## 4.5 Adding the proportion of each stratum
@@
 #### The unvaccinated stratum
+
 ```{r}
@@
 #### The 1-dose only stratum
+
 ```{r}
@@
 #### The 2-dose+ stratum
+
 ```{r}
@@
 ### merge all four into one figue panel
+
 ```{r}
@@
 #### Figure panel
+
 ```{r}
@@
        plot = p09all,
-        width=2400, height=2000, 
+        width=2550, height=2000, 
         dpi=300, units ="px")
 ```
+
 ## 5. Additional analysis for manuscript
@@
 #### 5.1 Some desc analysis
+
 ```{r}
@@
 Among samples from our auxiliary datasets tested with both Abbott and Roche Anti-N assays, the COI was lower with Abbott than with Roche in X% of samples
+
 ```{r}
@@
 #### 5.2 PCR-verified test positivity:
+
 We did this analysis as suggested by the reviewer-1
+
 ```{r}
@@
 ```
+
 #### 5.3 Modify the legends and y-axis labels of figure 1
+
 ```{r}
